### 2019-06-05 Latex/followup.tex
Eerste implementatie space-only hierarchical basis FEM

\documentclass{amsart}
\usepackage{amssymb,epsfig,mathrsfs,mathpazo}
%\usepackage{showkeys}
\usepackage{color}
\usepackage{epsfig,bm}
\usepackage[multiple]{footmisc}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\numberwithin{equation}{section}

\newcommand{\eps}{\varepsilon}
\newcommand{\cA}{\mathcal A}
\newcommand{\R}{\mathbb R}
\newcommand{\C}{\mathbb C}
\newcommand{\K}{\mathbb K}
\newcommand{\N}{\mathbb N}
\newcommand{\Z}{\mathbb Z}

\newcommand{\cF}{\mathcal F}
\newcommand{\cY}{\mathcal Y}
\newcommand{\cX}{\mathcal X}
\newcommand{\cE}{\mathcal E}
\newcommand{\cZ}{\mathcal Z}
\newcommand{\cL}{\mathcal L}
\newcommand{\Lis}{\cL\mathrm{is}}

\newcommand{\identity}{\mathrm{Id}}
\newcommand{\cB}{\mathcal{B}}

\DeclareMathOperator{\ran}{ran}
\DeclareMathOperator{\supp}{supp}
\DeclareMathOperator{\clos}{clos}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\blockdiag}{blockdiag}
\DeclareMathOperator{\diam}{diam}
\DeclareMathOperator{\dist}{dist}

\newcommand{\ds}{\displaystyle}

\DeclareMathOperator{\divv}{div}
\DeclareMathOperator{\moddiv}{\mbox{$\widetilde{\rm div}$}}
\DeclareMathOperator{\Span}{span}
\newcommand{\veecirc}{\raisebox{.21ex}[1.7ex]{$\stackrel{\raisebox{-.6ex}{$\scriptstyle \circ$}}{\vee}$}}

\newcommand{\nrm}{| \! | \! |}

\renewcommand{\mod}{\mathrm{mod}}

\newcommand{\rem}[1]{{\color{blue}{[#1]}}}

\newcommand{\rs}[1]{{\color{red}{RS: #1}}}
\newcommand{\jw}[1]{{\color{red}{JW: #1}}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}

\newcommand{\1}{\mathbb 1}

\newcommand{\bbT}{\mathbb{T}}
\newcommand{\bbP}{\mathbb{P}}
\newcommand{\tria}{{\mathcal T}}
\newcommand{\cP}{{\mathcal P}}

\newcommand{\aumlaut}{{\"a}}
\newcommand{\uumlaut}{{\"u}}

\newcommand{\parent}{{\tt parent}} 
\newcommand{\gen}{{\tt gen}}

\title{BLA parabolic}

\date{\today}

\author{RS}

\address{
Korteweg-de Vries (KdV) Institute for Mathematics, University of Amsterdam, P.O. Box 94248, 1090 GE Amsterdam, The Netherlands.
}
%\email{r.p.stevenson@uva.nl, j.h.westerdiep@uva.nl}

%\thanks{The first author has been supported by NSF Grant DMS 172029. The second author has been supported by the Netherlands Organization for Scientific Research (NWO) under contract.~no.~613.001.652}

\subjclass[2010]{
}

%\keywords{Parabolic PDEs, space-time variational formulations, quasi-best approximations, stability}

\begin{document}

\begin{abstract}  
\end{abstract}

\maketitle

\section{Inf-sup stability} \label{Sinfsup}
$V, H$ be separable Hilbert spaces. Gelfand triple $V \hookrightarrow H \simeq H' \hookrightarrow V'$.

$X:=L_2(I;V) \cap H^1(I;V)$, $Y:=L_2(I;V)$.
Central  issue in this section is the construction of families of pairs of closed subspaces $X^\delta \subset X$, $Y^\delta \subset Y$, s.t.
\be \label{infinfsup}
\inf_{\delta \in \Delta} \inf_{0 \neq x^\delta \in X^\delta} \sup_{0 \neq y^\delta \in Y^\delta} \frac{\langle \partial_t x^\delta,y^\delta \rangle_{L_2(I)\otimes H}}{\|\partial_t x^\delta\|_{Y'}\|y^\delta\|_{Y}} \gtrsim1.
\ee

The following result generalizes \cite[Lemma 6.2]{11} and \cite[Proposition 2.5]{258.4} in the sense that $W \neq \tilde W$ is allowed.

\begin{proposition} \label{prop1} Let $W$ and $\tilde W$ be closed subspaces of $H$ such that there exists a (biorthogonal) projector $Q \in \cL(H,H)$  exists with $\ran Q=W$ and $\ran (\identity -Q)={\tilde W}^\perp$. 
Then, when $W \subset V$, it holds that $\gamma:=\inf_{0 \neq \tilde w \in \tilde W}\sup_{0 \neq w \in W}\frac{\langle \tilde w,w\rangle}{\|\tilde w\|_{V'}\|w\|_V}>0$ if and only if $Q \in \cL(V,V)$, in which case $\gamma=\|Q\|_{\cL(V,V)}^{-1}$.
\end{proposition}

\begin{proof} If $Q\in \cL(V,V)$, then for $\tilde w \in \tilde W$,
$$
\sup_{0 \neq w \in W}\frac{\langle \tilde w,w\rangle}{\|w\|_V}=\sup_{0 \neq v \in V}\frac{\langle \tilde w, v\rangle}{\|Qv\|_V}
\geq \|Q\|_{\cL(V,V)}^{-1} \|\tilde w\|_{V'},
$$
or $\gamma \geq \|Q\|_{\cL(V,V)}^{-1}$. If on the other hand $\gamma>0$, then for any $u \in H$,
$$
\gamma \|Q' u\|_{V'} \leq \sup_{0 \neq w \in W}\frac{\langle Q'u,w\rangle}{\|w\|_V} =
 \sup_{0 \neq w \in W}\frac{\langle u,w\rangle}{\|w\|_V} \leq \|u\|_{V'},
 $$
so that from $H$ being dense in $V'$, $\|Q\|_{\cL(V,V)}=\|Q'\|_{\cL(V',V')}\leq \gamma^{-1}$.
\end{proof}

\rem{Voor $H=L_2(\Omega)$, $V=H^1_0(\Omega)$, $W=\tilde W$ continuous fem space w.r.t. partition created by NVB in 2D, condities vervuld.
Wellicht ook interessant $W$ van hogere orde te nemen.

Vb. $W\neq \tilde W$: Als $W$ cont. piecewise quadratics w.r.t. partition, en $\tilde W$ cont. piecewise linears w.r.t. verfijnde partitie, dan biorthogonale $Q \in \cL(H,H)$ bestaat (zie \cite{56}). Uit $Q=\Pi+Q(I-\Pi)$ voor zeg $\Pi$ is Scott-Zhang, volgt in ieder geval voor quasi-uniforme meshes dat $Q \in \cL(V,V)$ uniform in de mesh-size.

(Ik vermoed dat $W \neq \tilde W$ in dit verhaal uiteindelijk geen nuttige toepassingen zal hebben.)}

\newpage
{\LARGE Unnecessarily complicated:}

{\tiny \begin{theorem}[{\cite[Theorem 4.1]{249.95}}] \label{thm1} Let $\Psi=\{\psi_\lambda\colon \lambda \in \vee\}$, $\tilde{\Psi}=\{\tilde{\psi}_\lambda\colon \lambda \in \vee\}$ be biorthogonal Riesz bases for $L_2(I)$ such that $\{2^{-|\lambda|}\psi_\lambda\colon \lambda \in \vee\}$ is a Riesz basis for $H^1(I)$, and
for some $\hat{\lambda} \in \nabla$, $\psi_{\hat{\lambda}}\in \Span\{\1\}$. Then with $\underline{\vee}:=\vee \setminus \{\hat{\lambda}\}$, it holds that
$\Psi^-=\{\psi^-_\lambda\colon \lambda \in \underline{\vee}\}$ defined by $\psi^-_\lambda:=2^{-|\lambda|}\psi_\lambda'$ is a Riesz basis for $L_2(I)$ with dual basis
$\tilde{\Psi}^+=\{\tilde{\psi}^+_\lambda\colon \lambda \in \underline{\vee}\}$ defined by $\tilde{\psi}^+_\lambda:=x \mapsto \int_0^x 2^{|\lambda|} \tilde{\psi}_\lambda \,dx$.
Moreover $\supp \psi^-_\lambda \subset \supp \psi_\lambda$ and ${\rm conv hull}(\supp \tilde{\psi}^+_\lambda) \subset \supp \tilde{\psi}_\lambda$.
\end{theorem}

\rem{Corresponding results about scaling functions in \cite[Proposition 5.1]{249.95}.}

\rem{Vb. neem $\Psi=\tilde{\Psi}$ piecewise polynomial orthogonal (multi-) wavelet basis uit \cite{65.5} aangepast aan interval zonder randvoorwaarden (kan maar is denk ik niet gedaan).}

\rem{Vb. Neem $\Psi \neq \tilde{\Psi}$ de biorthogonal wavelets uit \cite{45} (primals splines, duals implicitly defined) aangepast aan interval zonder randvoorwaarden. Zie bijv. \cite[Ch. 2]{64.65} of \cite{245.1}.}

\rem{Vb. Neem $\Psi \neq \tilde{\Psi}$ de biorthogonal (multi-) wavelets uit \cite{75.6} (primals and duals splines) aangepast aan interval zonder randvoorwaarden (kan maar is denk ik niet gedaan).}

\begin{proposition} \label{prop2}
Let $\Psi^-=\{\psi^-_\lambda\colon \lambda \in \underline{\vee}\}$,
$\tilde{\Psi}^+=\{\tilde{\psi}^+_\lambda\colon \lambda \in \underline{\vee}\}$ be biorthogonal Riesz bases for $L_2(I)$, and let $\kappa$ denote the spectral condition number of $\langle \Psi^-,\Psi^-\rangle$.

Let ${\mathcal O}$ be a collection of pairs of closed subspaces of $H$ as in Proposition~\ref{prop1} such that for some $\mu>0$, for each $(W,\tilde W) \in {\mathcal O}$, 
$$
\inf_{0 \neq \tilde w \in \tilde W}\sup_{0 \neq w \in W}\frac{\langle \tilde w,w\rangle}{\|\tilde w\|_{V'}\|w\|_V} \geq \mu.
$$
For $\delta \in \Delta$ and $\lambda \in \underline{\vee}$, select $(W^\delta_\lambda,{\tilde W}^\delta_\lambda) \in {\mathcal O}$ or $(W^\delta_\lambda,{\tilde W}^\delta_\lambda):=\{(0,0)\}$. Then for $G^\delta:=\clos_{V'} \Span \{\psi^-_\lambda \otimes \tilde{W}^\delta_\lambda\}$ and $Y^\delta:=\clos_{V} \Span \{\tilde{\psi}^+_\lambda \otimes W^\delta_\lambda\}$, it holds that
$$
\inf_{0 \neq g \in G^\delta}\sup_{0 \neq y \in Y^\delta} \frac{\langle g, y\rangle_{L_2(I;H)}}{\|g\|_{Y'} \|y\|_Y} \geq \mu \kappa^{-\frac12}.
$$
\end{proposition}

\begin{proof} For any finite $\Lambda \subset \underline{\vee}$, let $g:=\sum_{\lambda \in \Lambda} \psi^-_\lambda \otimes \tilde{w}_\lambda$ for some ${\tilde w}_\lambda \in {\tilde W}^\delta_\lambda$. For any fixed $\eps \in (0,\mu)$, there exists a $w_\lambda \in W^\delta_\lambda$ with $\|w_\lambda\|_V=\|\tilde{w}_\lambda\|_{V'}$ and $\langle \tilde{w}_\lambda,w_\lambda\rangle \geq (\mu-\eps) \|\tilde{w}_\lambda\|_{V'}^2$.
For $y:=\sum_{\lambda \in \Lambda} \tilde{\psi}^+_\lambda \otimes w_\lambda$, we find that
\begin{align*}
\langle g, y\rangle_{L_2(I;H)} &= \sum_{\lambda \in \Lambda} \langle \tilde{w}_\lambda,w_\lambda\rangle
\geq (\mu-\eps) \sum_{\lambda \in \Lambda}\|\tilde{w}_\lambda\|_{V'}^2\\
&=(\mu-\eps) \sqrt{\sum_{\lambda \in \Lambda} \|\tilde{w}_\lambda\|_{V'}^2}
\sqrt{\sum_{\lambda \in \Lambda} \|w_\lambda\|_{V}^2}\\
&\geq  (\mu-\eps)\rho\big(\langle \Psi^-,\Psi^-\rangle_{L_2(I)}\big)^{-\frac12}\big\| g\big\|_{Y'}
\rho\big(\langle \Psi^-,\Psi^-\rangle_{L_2(I)}^{-1}\big)^{-\frac12} \big\|y\|_{Y},
\end{align*}
where we used that $\langle \tilde{\Psi}^+,\tilde{\Psi}^+\rangle_{L_2(I)}=\langle \Psi^-,\Psi^-\rangle_{L_2(I)}^{-1}$. Since $\eps \in (0,\mu)$ was arbitrary, the proof is completed.
\end{proof}

\rem{In ``full'' tensor product setting is alles simpeler: Zij $G^\delta=R \otimes \tilde{W}$ en $Y^\delta=\tilde{R} \otimes W$ z.d.d. $(R,\tilde{R})$ is $(L_2,L_2)$ inf-sup stable (bijv $R=\tilde R$), en $(\tilde W,W)$ is $(V',V)$ inf-sup stable, dan $(G^\delta,Y^\delta$) is $(Y',Y)$ inf-sup stable.}

\rem{Neem je in Proposition~\ref{prop2} $\Psi^-,\tilde{\Psi}^+$ als geconstrueerd in Thm.~\ref{thm1} dan is probleem dat alle $\tilde{\psi}^+_\lambda$ verdwijnen op rand.
Dit geeft beroerde approximatie eigenschappen voor $Y^\delta$.
Plan: Maak $L_2$-Riesz basis $\hat{\Psi}$ z.d.d. het  approximatie properties heeft als $\tilde{\Psi}^+$ maar dan ook voor functies welke niet verdwijnen aan de rand, en z.d.d. de interior wavelets gelijk zijn, en $\Span\{\hat{\psi}_\lambda \colon |\lambda|\leq L\} \supset \Span\{\tilde{\psi}^+_\lambda \colon |\lambda|\leq L\}$.
Gegeven $G^\delta:=\clos_{V'} \Span \{\psi^-_\lambda \otimes \tilde{W}^\delta_\lambda\}$ neem  dan $Y^\delta \supset \clos_{V} \Span \{\tilde{\psi}^+_\lambda \otimes W^\delta_\lambda\}$, dan 
inf-sup als in Proposition~\ref{prop2}}

\rem{We denken aan $X^\delta:=\Span\{\psi_\lambda \otimes \tilde{W}^\delta_\lambda\colon \lambda \in \Lambda\}$ met ook $\tilde{W}^\delta_\lambda \subset V$, en
$\Span\{\tilde{\psi}^+_\lambda  \otimes W^\delta_\lambda\colon \lambda \in \Lambda \setminus \{\hat{\lambda}\} \}\subset Y^\delta $.}
} % end tiny

\newpage
{\LARGE Better approach:}

Let ${\mathcal O}$ be a collection of pairs of closed subspaces $(W,\tilde{W})$ as in Proposition~\ref{prop1} with a uniform inf-sup constant $\gamma>0$.

Let $\Sigma=\{\sigma_\lambda \colon \lambda \in \vee_\Sigma\}$ be, properly scaled, a Riesz basis for $L_2(I)$ and $H^1(I)$ (important later for preconditioning).

Let $\Psi=\{\psi_\lambda \colon \lambda \in \vee_\Psi\}$ be a Riesz basis for $L_2(I)$, with dual basis denoted as $\tilde{\Psi}$.

Given
$$
\framebox{$X^\delta=\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes \tilde{W}_\lambda^\delta,$}
$$
where \framebox{$(W_\lambda^\delta,\tilde{W}_\lambda^\delta) \in {\mathcal O}$ or $W_\lambda^\delta=\tilde{W}_\lambda^\delta=\{0\}$}, let
$$
\framebox{$Y^\delta=\sum_{\mu \in \vee_\Psi} \psi_\mu \otimes V_\mu^\delta$},
$$
 $\tilde{Y}^\delta=\sum_{\mu \in \vee_\Psi} \tilde{\psi}_\mu \otimes \tilde{V}_\mu^\delta$ 
where \framebox{$(V^\delta_\mu,\tilde{V}^\delta_\mu) \in {\mathcal O}$ or $V_\lambda^\delta=\tilde{V}_\lambda^\delta=\{0\}$} is such that 
\be \label{19}
\langle \sigma'_\lambda,\psi_\mu\rangle_{L_2(I)} \neq 0 \Longrightarrow \tilde{V}^\delta_\mu \supset \tilde{W}^\delta_\lambda.
\ee
Thanks to $\sigma'_\lambda=\sum_{\mu \in \vee_\Psi} \langle \sigma'_\lambda,\psi_\mu\rangle_{L_2(I)}\tilde{\psi}_\mu$ this ensures that
\be \label{20}
\partial_t X^\delta \subset \tilde{Y}^\delta.
\ee

Let $\tilde v=\sum_{\mu \in \vee_\Psi} \tilde{\psi}_\mu \otimes \tilde{v}_\mu \in \tilde{Y}^\delta$.
Then, for any constant $\eps\in (0,\gamma)$, there exists a $v=\sum_{\mu \in \vee_\Psi} \psi_\mu \otimes v_\mu \in Y^\delta$  with $\langle \tilde v_\mu,v_\mu\rangle \geq (\gamma-\eps) \|\tilde{v}_\mu\|_{V'} \|v_\mu\|_V$ and $\|\tilde{v}_\mu\|_{V'}= \|v_\mu\|_V$, and so
\be \label{21}
\langle\tilde v, v\rangle_{L_2(I)\otimes H}
= \sum_{\mu \in \vee_\Psi} \langle \tilde{v}_\mu,v_\mu\rangle \geq (\gamma-\eps) 
\sum_{\mu \in \vee_\Psi} \|\tilde{v}_\lambda\|_{V'}^2
\eqsim \|\tilde v\|_{Y'} \|v\|_Y.
\ee
From \eqref{20} and \eqref{21}, we conclude that $(X^\delta,Y^\delta)_{\delta \in \Delta}$ satisfies \eqref{infinfsup} with uniform inf-sup constant $\gamma$.

To ensure \eqref{19}, we will assume that 
that for  $\ell \in \N_0$,
$$
\Span\{\sigma'_\lambda\colon |\lambda|\leq \ell\} \subset \Span\{\tilde{\psi}_\lambda\colon |\lambda|\leq \ell\}.
$$
 so that
\be \label{22}
|\mu| >|\lambda| \Longrightarrow \langle \sigma'_\lambda,\psi_\mu\rangle_{L_2(I)}=0.
\ee
Now \eqref{19} is ensured under the condition that
\be \label{23}
|\mu| \leq |\lambda| \wedge|\supp \psi_\mu \cap \supp \sigma_\lambda|>0 \Longrightarrow \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda.
\ee

We will assume that $\Sigma$ and $\Psi$ are locally supported, which in particular means that
$\diam \supp \sigma_\lambda \lesssim 2^{-|\lambda|}$ and $\diam \supp \psi_\lambda \lesssim 2^{-|\lambda|}$.
Furthermore, to ensure an efficient stiffness matrix evaluation, in the next section we will impose a multi-level constraint on $X^\delta$ which will have the consequence
that for any $\lambda \in \vee_\Sigma$ with $\tilde{W}_\lambda^\delta \neq \{0\}$, and any $\ell <|\lambda|$, there exists a $\mu \in \vee_\Sigma$ with $|\mu|=\ell$ ($\mu$ being the `ancestor' of $\lambda$)
with $\dist(\supp \sigma_\mu,\supp \sigma_\lambda)\lesssim 2^{-|\mu|}$ and $\tilde{W}_\mu^\delta \supset \tilde{W}_\lambda^\delta$.
We infer that \eqref{23} and thus \eqref{20} can be ensured for $Y^\delta$ with $\dim Y^\delta \lesssim \dim X^\delta$.

In particular, for $\lambda \in \vee_\Psi$, let $S(\lambda)$ be the neighborhood of $\supp \psi_\lambda$ with $\diam S(\lambda) \lesssim 2^{-|\lambda|}$ s.t. $S(\lambda) \subset S(\mu)$ whenever $\lambda$ is a child of $\mu \in \vee_\Psi$.
Then assuming the multilevel constraint on $X^\delta$, \eqref{23} is ensured under the condition that
\be \label{24}
|\mu| = |\lambda| \wedge|\supp \psi_\mu \cap \supp S(\lambda)|>0 \Longrightarrow \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda.
\ee

\begin{example} Let $W=\tilde{W}$ for any pair $(W,\tilde{W}) \in {\mathcal O}$. Let any $\sigma_\lambda$ be a continuous piecewise polynomial of degree $d-1$ w.r.t. a uniform partition of $I$ into $2^{|\lambda|}$ subintervals, and let 
$\Psi=\tilde{\Psi}$ be such that 
$\Span\{\tilde{\psi}_\lambda\colon |\mu| \leq |\lambda|\}$ includes all piecewise polynomials of degree $d-2$ w.r.t. a uniform partition of $I$ into $2^{|\lambda|}$ subintervals (Alpert basis) so that \eqref{22} is satisfied.
If $\Span\{\tilde{\psi}_\lambda\colon |\mu| \leq |\lambda|\}$ even includes all piecewise polynomials of degree $d-1$ w.r.t. a uniform partition of $I$ into $2^{|\lambda|}$ subintervals, then 
$Y^\delta$ as constructed above additionally satisfies
$$
\framebox{$X^\delta \subset Y^\delta$}.
$$
\rem{We zullen $X^\delta \subset Y^\delta$ nodig hebben alsook $\partial X^\delta \subset \tilde{Y}^\delta$. Met het oog hierop zie ik momenteel geen toepassingen voor $W \neq \tilde{W}$ en $\Psi \neq \tilde{\Psi}$}

Although later we might consider quadratic temporal wavelets and quadratic spatial finite element spaces, perhaps wise to start with considering 
linear temporal wavelets and linear spatial finite element spaces.
Then for $\Sigma$ we can take the 3-point hierarchical basis. For this basis $S(\lambda)=\supp \sigma_\lambda$.
Then we take for $\Psi$ the orthonormal (discontinuous) piecewise linear wavelet basis. See figures.
\begin{figure}
\begin{center}
\input{threepoint.pdf_t}
\end{center}
\caption{3-point hierarchical basis.}
\end{figure}
\begin{figure}
\begin{center}
\input{discont_linears.pdf_t}
\end{center}
\caption{Orthonormal (w.r.t. $L_2(0,1)$) (discontinuous) piecewise linear wavelet basis $\Psi$.}
\end{figure}
\end{example}

\section{Preconditioning}
For each $\delta \in \Delta$, let $X^\delta$,  $Y^\delta$ be finite dimensional.

Thanks to $\Psi$ Riesz for $L_2(I)$,  any $y \in Y$ is of the form $\sum_{\lambda \in \vee_\Psi} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$ (with $\sum_{\lambda \in \vee_\Psi} \|w_\lambda\|_V^2<\infty$), on $Y \times Y$ we define the bounded, symmetric, and coercive bilinear form
$$
(D_Y \sum_{\lambda \in \vee_\Psi} \psi_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_\Psi} \psi_\mu \otimes v_\lambda):=\sum_{\lambda \in \vee_\Psi} \langle w_\lambda, v_\lambda\rangle_V.
$$
Denoting with $E_Y^\delta$ the trivial embedding of $Y^\delta=\sum_{\lambda\in \vee_\Psi} \psi_\lambda \otimes V_\lambda^\delta$ into $Y$, we set $D_Y^\delta:={E_Y^\delta}' D_Y E_Y^\delta\in \Lis(Y^\delta,{Y^\delta}')$, whose norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$.
Equipping $Y^\delta$ with basis of type $\cup_{\lambda} \psi_\lambda \otimes \Sigma_\lambda^\delta$, the matrix representation of $D_Y^\delta$ reads as
$$
{\bf D}_Y^\delta=\blockdiag[{\bf A}_\lambda^\delta]_{\lambda}.
$$
where ${\bf A}_\lambda^\delta:=\langle \Sigma_\lambda^\delta, \Sigma_\lambda^\delta\rangle_V$.
For ${\bf B}_\lambda^\delta \eqsim ({\bf A}_\lambda^\delta)^{-1}$, we have
$$
({\bf D}_Y^\delta)^{-1}\eqsim \blockdiag[{\bf B}_\lambda^\delta]_{\lambda},
$$
so that the right-hand side is a uniform preconditioner.
For ${\mathcal O}$ being a family of finite element spaces w.r.t. possibly locally refined partitions, such ${\bf B}_\lambda^\delta$, that moreover can be applied in linear complexity, are provided by \emph{multi-grid methods}.

\rem{Om te beginnen ${\bf B}_\lambda^\delta = ({\bf A}_\lambda^\delta)^{-1}$ nemen, d.w.z. directe solver in spatial direction.}

Let $\Sigma=\{\sigma_\lambda \colon \lambda \in \vee_\Sigma\}$ a Riesz basis for $L_2(I)$ s.t.
$\{2^{-|\lambda|} \sigma_\lambda \colon \lambda \in \vee_\Sigma\}$ is a Riesz basis for $H^1(I)$.
Using that any $u \in X$ is of the form $\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$
with $\|x\|_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} \|w_\lambda\|_V^2+4^{-|\lambda|}\|w_\lambda\|_{V'}^2<\infty$, on $X \times X$ we define the bounded, symmetric, and coercive bilinear form
$$
(D_X \sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_\Sigma} \sigma_\mu \otimes v_\mu):=\sum_{\lambda \in \vee_\Sigma} \langle w_\lambda, v_\lambda\rangle_V+4^{-|\lambda|}\langle w_\lambda, v_\lambda\rangle_{V'}.
$$
Denoting with $E_X^\delta$ the trivial embedding of $X^\delta=\sum_{\lambda\in \vee_\Sigma} \sigma_\lambda \otimes \tilde{W}_\lambda^\delta$ into $X$, we set $D_X^\delta:={E_X^\delta}' D_X E_X^\delta\in \Lis(X^\delta,{X^\delta}')$, whose norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$.
Equipping $X^\delta$ with basis of type $\cup_{\lambda} \psi_\lambda \otimes \tilde{\Sigma}_\lambda^\delta$, the matrix representation of $D_X^\delta$ reads as
$$
{\bf D}_X^\delta=\blockdiag[{\bf A}_\lambda^\delta+4^{-|\lambda|}\langle \tilde{\Sigma}_\lambda^\delta, \tilde{\Sigma}_\lambda^\delta\rangle_{V'}]_{\lambda}.
$$
where ${\bf A}_\lambda^\delta:=\langle \tilde{\Sigma}_\lambda^\delta, \tilde{\Sigma}_\lambda^\delta\rangle_V$.
For $u \in \tilde{W}_\lambda^\delta$, it holds that $\|u\|_{V'} \geq \sup_{0 \neq w \in \tilde{W}_\lambda^\delta} \frac{\langle u,w\rangle}{\|w\|_V} \geq \gamma \|u\|_{V'}$ \rem{aannemende dat $L_2$-orthogonal projector onto $\tilde{W}_\lambda^\delta$ uniform bounded is in $V$-norm, hetgeen ok als voor de collectie van paren $(W,\tilde{W}) \in {\mathcal O}$ geldt dat $\tilde{W}=W$}.
With ${\bf u}$  denoting the representation of $u$ w.r.t. $\tilde{\Sigma}_\lambda^\delta$, it holds that
$$
\sup_{0 \neq w \in \tilde{W}_\lambda^\delta} \frac{\langle u,w\rangle}{\|w\|_V}=
\|({\bf A}_\lambda^\delta)^{-\frac12} {\bf M}_\lambda^\delta {\bf u}\|,
$$
where ${\bf M}_\lambda^\delta=\langle \tilde{\Sigma}_\lambda^\delta, \tilde{\Sigma}_\lambda^\delta\rangle$, so that
$$
\gamma^2 \langle \tilde{\Sigma}_\lambda^\delta, \tilde{\Sigma}_\lambda^\delta\rangle_{V'} \leq  
{\bf M}_\lambda^\delta ({\bf A}_\lambda^\delta)^{-1} {\bf M}_\lambda^\delta
 \leq \langle \tilde{\Sigma}_\lambda^\delta, \tilde{\Sigma}_\lambda^\delta\rangle_{V'}.
$$

As follows from \cite[Thm. 4]{242.817}, it holds that
\begin{align*}
{\textstyle \frac{1}{2}}\big( {\bf A}_\lambda^\delta\!+\!4^{-|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda\big)
\!\leq \! ({\bf A}_\lambda^\delta\!+\!4^{-|\lambda|}{\bf M}^\delta_\lambda) ({\bf A}_\lambda^\delta)^{-1} ({\bf A}_\lambda^\delta\!+\!4^{-|\lambda|}{\bf M}^\delta_\lambda)
\!\leq\!
{\bf A}_\lambda^\delta\!+\!4^{-|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda.
\end{align*}
Now assuming that
\be \label{robustmg}
{\bf C}^\delta_{\lambda} \eqsim ({\bf A}_\lambda^\delta+4^{-|\lambda|}{\bf M}^\delta_\lambda)^{-1}
\ee
(uniformly in $\delta$ and $|\lambda| \in \N_0$), we infer that
$$
({\bf D}_X^\delta)^{-1} \eqsim
\blockdiag\Big[{\bf C}^\delta_{\lambda} {\bf A}_\lambda^\delta {\bf C}^\delta_{\lambda}\Big]_{\lambda}
.
$$
In \cite{241.1,173.4} it was shown that a V-cycle multi-grid method yields a ${\bf C}^\delta_{\lambda}$ that satisfies \eqref{robustmg}, and that moreover  can be applied in linear complexity.
These results were shown for uniformly refined meshes, but similarly as for $4^{-|\lambda|} = 0$, we expect them to hold for locally refined meshes as well.
\rem{Dit is waar door grofste rooster in space decomposition sufficiently fine te nemen (Joachim) zodat tegelijkertijd op dat level $A+4^{-|\lambda|} M \eqsim 4^{-|\lambda|} M$ dus well-conditioned.}

\rem{Om te beginnen ${\bf C}^\delta_{\lambda} = ({\bf A}_\lambda^\delta+4^{-|\lambda|}{\bf M}^\delta_\lambda)^{-1}$ nemen, d.w.z. directe solver in spatial direction.}

\section{Evaluation stiffness matrix in linear complexity}
For $* \in \{\ldots\}$, let $\Psi^*=\{\psi^*_\lambda\colon \lambda \in \vee^*\}$ be a (multilevel) collection of functions on some domain $\Omega$.
We assume that the $\psi_\lambda^*$ are \emph{locally supported} in the sense that with $|\lambda| \in \N_0$ denoting the \emph{level} of $\lambda$, 
\begin{align} \label{c1}
& \sup_{\lambda \in \vee^*} 2^{|\lambda|} \diam \psi^*_\lambda <\infty,\\
\label{c2}
& \sup_{\ell \in \N_0} \sup_{x \in \Omega} \#\{ \lambda \in \vee^*\colon |\lambda|=\ell \wedge
 \supp \psi^*_\lambda \cap B(x;2^{-\ell}) \neq\emptyset\} <\infty.
\end{align}
\rem{Factor $2$ in \eqref{c1}-\eqref{c2} en verder, kan gelezen worden als e.o.a. constante $>1$, evt. verschillend voor temporal en spatial functions.
Bijv. voor NVB in $n$ space dimensies, ligt het afhankelijk van de definitie van de level van een functie misschien voor de hand om $2^{1/n}$ te nemen.}

We will refer to the functions $\psi^*_\lambda$ as being {\em wavelets}, although not necessarily they have vanishing moments or other specific wavelet properties.

For $\ell \in \N_0$, and any $\Lambda \subset \vee^*$, we set $\Lambda_\ell:=\{\lambda \in \Lambda:|\lambda|=\ell\}$ and 
$\Lambda_{\ell\uparrow}:=\{\lambda \in \Lambda:|\lambda| \geq \ell\}$, and write $\Psi^*_\ell:=\Psi^*|_{\vee^*_\ell}$.
%We write $\Psi^*_\ell:=\Psi^*|_{\vee^*_\ell}$, and add the harmless assumption that for any $\ell \in \N_0$,
%\begin{equation} \label{c3}
%\Omega = \cup_{\psi^*_\lambda \in \Psi^*_\ell} \supp \psi^*_\lambda.
%\end{equation}

For $\ell \in \N_0$, we assume a collection
$
\Phi^*_\ell=\{\phi^*_\lambda\colon \lambda \in \Delta^*_\ell\},
$
whose members will be referred to as being {\em scaling functions},
with
\begin{align} \label{c4}
& \Span \Phi^*_{\ell+1} \supseteq  \Span \Phi^*_{\ell} \cup \Psi^*_{\ell+1},\quad \Phi^*_0=\Psi^*_0 \quad(\Delta_0^*:=\vee_0^*), \\
\label{c5}
& \sup_{\ell \in \N_0}\sup_{\lambda \in \Delta^*_\ell} 2^{\ell} \diam \supp \phi^*_\lambda <\infty, \\
\label{c6}
& \sup_{\ell \in \N_0} \sup_{x \in \Omega} \#\{\lambda \in \Delta^*_\ell: \supp \phi^*_\lambda \cap B(x;2^{-\ell}) \neq \emptyset\} <\infty,\\
\label{c7}
& \{\phi^*_\lambda|_\Sigma: \lambda \in \Delta^*_\ell,\,\phi^*_\lambda|_\Sigma \not\equiv 0\} \text{ is independent } (\text{for all open } \Sigma \subset \Omega,\,\ell \in \N_0).
\end{align}
W.l.o.g. we assume that the index sets $\Delta^*_\ell$ for different $\ell$ are mutually disjoint, and set $\Phi^*:=\cup_{\ell \in \N_0} \Phi^*_\ell$ with index set $\Delta^*:=\cup_{\ell \in \N_0} \Delta^*_\ell$. For $\lambda \in \Delta^*$, we set $|\lambda|:=\ell$ when $\lambda \in \Delta^*_\ell$.

Viewing $\Psi^*_\ell$, $\Phi^*_\ell$ as column vectors, the assumptions we made so far guarantee the existence of matrices $p^*_{\ell}$, $q^*_{\ell}$ such that
$$
\left[\begin{array}{@{}cc@{}} (\Phi^*_{\ell-1})^\top & (\Psi^*_\ell)^\top \end{array} \right] = (\Phi^*_{\ell})^\top
\left[\begin{array}{@{}cc@{}} p^*_{\ell} & q^*_{\ell} \end{array} \right],
$$
where the number of non-zeros per row and column of $p^*_{\ell}$ and $q^*_{\ell}$ is finite, uniformly in the rows and columns and in $\ell \in \N$ (here also \eqref{c7} has been used).

To each $\lambda \in \vee^*$ with $|\lambda|>0$, we associate one or more $\mu \in \vee^*$ with $|\mu|=|\lambda|-1$ and $|\supp \psi^*_\lambda \cap \supp \psi^*_\mu|>0$.
We call $\mu$ the (a) {\em parent} of $\lambda$, and so $\lambda$ a {\em child} of $\mu$.

To each $\lambda \in \vee^*$, we associate some neighbourhood  $S^*(\lambda)$ of $\supp \psi^*_\lambda$, with diameter $\lesssim2^{-|\lambda|}$, such that for $|\lambda|>0$,
$S^*(\lambda) \subset \cup_{\mu \in \parent(\lambda)} S^*(\mu)$.

\begin{remark}
Such a neighborhood always exists even when a child has only one parent. Indeed with $C:=\sup_{\lambda \in \vee^*} 2^{|\lambda|} \diam \psi^*_\lambda$ and
$S^*(\lambda):=\{x \in \Omega \colon \dist(x,\supp \psi^*_\lambda) < C 2^{-|\lambda|}\}$,
for $\mu$ being a parent of $\lambda$ and  $x \in S^*(\lambda)$, $\dist(x,\supp \psi^*_\mu)\leq \dist(x,\supp \psi^*_\lambda)+\diam(\supp \psi^*_\lambda)< 2 C 2^{-|\lambda|}=C 2^{-|\mu|}$, i.e.,
$x \in S^*(\mu)$.
\end{remark}

%We call a finite $\Lambda \subset \vee^\ast$ a {\em tree}, if it contains all $\lambda \in \vee^*$ with $|\lambda|=0$, as well as the parent of any $\lambda \in \Lambda$ with $|\lambda|>0$.
%A finite subset $\Lambda \subset \vee_{\ell \uparrow}^\ast$ is called an $\ell$-tree, or simply a tree when the value of $\ell$ is clear from the context, if for any $\lambda \in \Lambda$ its parent in $\vee_{\ell \uparrow}^\ast$ is in $\Lambda$.

A finite $\Lambda \subset \vee_{\ell \uparrow}^\ast$ is called an \emph{$\ell$-tree}, or simply a \emph{tree} when $\ell=0$,  when for any $\lambda \in \Lambda$ its parent(s) in $\vee_{\ell \uparrow}^\ast$ is(are) in $\Lambda$.

\bigskip
\rem{Bekijk conforming NVB in 2 dimensies. Neem aan initial triangulation ${\mathcal T}_\bot$ satisfies matching condition. Bekijk figuur:
\begin{figure}[h]
\begin{center}
\input{nvb.pdf_t}
\end{center}
\end{figure}

In deze situatie geldt altijd $\gen(T_1)=\gen(T_2)$. Call nodal basis function associated to new node the hierarchical basis associated to this node (it is independent of the rest of the triangulation).
It parents are the hierarchical basis functions associated to the nodes near the arrows. Def. $|\lambda|=\gen(T_i)+1$. Dan $\diam \supp \psi_\lambda \eqsim (\sqrt{2})^{-|\lambda|}$.
With this parent-child relation, there is an 1-1 correspondency between conforming $P_1$ finite element spaces w.r.t. conforming NVB triangulations and the span of hierarchical basis functions whose indices form trees (with the additional condition that all roots should be part of a tree). See \cite{64.59}.
}

%%%%%%
\subsection{A routine \texttt{eval}} \label{Seval}
On several places the restriction of a vector (of scalars or of functions) to its indices in some subset of the index set should be read as the vector of full length where the entries with indices outside this subset are replaced by zeros.

Let $a\colon \Span \breve{\Phi} \times \Span \Phi \rightarrow \R$ be linear in its first argument, and local in the sense that $a(v,u)=a(v,u|_{\supp v})$ for all $(v,u) \in \Span \breve{\Phi} \times \Span \Phi$.
(Maybe better later swap the arguments of $a$).\medskip

{\em \begin{tabbing}
$[\vec{e},\,\vec{f}]:=\mathtt{eval}(a)(\ell,  \breve{\Pi},\breve{\Lambda},\Pi,\Lambda,\vec{d},\vec{c})$\\
%
{\em \%} Input: \= $\ell \in \N$,
$\breve{\Pi} \subset \breve{\Delta}_{\ell-1}, \Pi \subset \Delta_{\ell-1}$, 
$\ell$-trees $\breve{\Lambda} \subset \breve{\vee}_{\ell\uparrow}$ and $\Lambda \subset {\vee}_{\ell\uparrow}$,\\
{\em \%} \> $\vec{d}\in \R^{\# \Pi},\, \vec{c}\in  \R^{\#  \Lambda}$. \\
%
{\em \%} Output: With $u:=\vec{d}^\top \Phi|_{\Pi}+\vec{c}^\top \Psi|_{\Lambda}$,\\
{\em \%} \> $\vec{e}=a(\breve{\Phi}|_{\breve{\Pi}},u)$,
$\vec{f}=a(\breve{\Psi}|_{\breve{\Lambda}},u)$.\\ \\
%
%
{\em \texttt{if}} \= $\breve{\Pi}\cup \breve{\Lambda}\neq \emptyset$ {\em \texttt{then}} \\
%
%
\>$\breve{\Pi}_B:=\{\lambda \in \breve{\Pi}\colon \big|\supp \breve{\phi}_\lambda \cap \cup_{\mu \in \Lambda_\ell}S(\mu)\big|>0$, $\breve{\Pi}_A:=\breve{\Pi} \setminus \breve{\Pi}_B$\\
\>$\Pi_B:=\{\lambda \in \Pi\colon \big|\supp \phi_\lambda \cap \big(\cup_{\mu \in \breve{\Lambda}_\ell} \breve{S}(\mu) \cup_{\gamma \in \breve{\Pi}_B} \supp \breve{\phi}_\gamma \big)\big|>0\}$, $\Pi_A:=\Pi \setminus \Pi_B$\\
\>$\breve{\underline{\Pi}}:=\{\lambda \in \breve{\Delta}_{\ell}\colon \supp \breve{\phi}_\lambda \subset \big(\cup_{\mu \in \breve{\Pi}_B} \supp \breve{\phi}_\mu \cup_{\gamma \in \breve{\Lambda}_\ell} \supp\breve{\psi}_\gamma\big)\}$\\
\>$\underline{\Pi}:=\{\lambda \in \Delta_{\ell}\colon \supp \phi_\lambda \subset \big(\cup_{\mu \in \Pi_B} \supp \phi_\mu \cup_{\gamma \in \Lambda_\ell} \supp\psi_\gamma\big)\}$\\ 
%
\>$\underline{\vec{d}}:=\big(\mathfrak{p}_{\ell} \vec{d}|_{\Pi_B}+\mathfrak{q}_\ell \vec{c}|_{\Lambda_{\ell}}\big)|_{\underline{\Pi}}$\\
\>$[\underline{\vec{e}},\,\underline{\vec{f}}]:=\mathtt{eval}(a)(\ell+1, \breve{\underline{\Pi}}, \breve{\Lambda}_{\ell+1\uparrow}, \underline{\Pi}, \Lambda_{\ell+1\uparrow},\underline{\vec{d}}, \vec{c}|_{\Lambda_{\ell+1\uparrow}})$\\
\>$\vec{e}=
\left[\begin{array}{@{}l@{}} \vec{e}|_{\breve{\Pi}_A}\\ \vec{e}|_{\breve{\Pi}_B}\end{array}\right]
:=
\left[\begin{array}{@{}l@{}} a(\breve{\Phi}|_{\breve{\Pi}_A},\vec{d}^\top\Phi|_{\Pi}) \\ 
(\breve{\mathfrak{p}}_{\ell} ^\top \underline{\vec{e}})|_{\breve{\Pi}_B}
\end{array}\right]$\\
\>$\vec{f}=
\left[\begin{array}{@{}l@{}} \vec{f}|_{\breve{\Lambda}_\ell} \\ \vec{f}|_{\breve{\Lambda}_{\ell+1\uparrow}}\end{array}\right]
:=
\left[\begin{array}{@{}l@{}} (\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}})|_{\breve{\Lambda}_{\ell}} \\ \underline{\vec{f}} \end{array}\right]$\\
{\em \texttt{endif}}
 \end{tabbing}}

   \begin{remark}  Let $\breve{\Lambda} \subset \breve{\vee}$, $\Lambda \subset \vee$ be  trees, and $\vec{c} \in \ell_2(\Lambda)$, then
 $$
  [\vec{e},\,\vec{f}]:=\mathtt{eval}(a)(1,\breve{\Lambda}_{0},\breve{\Lambda}_{1 \uparrow},\Lambda_{0},\Lambda_{1 \uparrow},\vec{c}|_{\Lambda_0},\vec{c}|_{\Lambda_{1 \uparrow}}),
 $$
satisfies
$$
a(\breve{\Psi}|_{\breve{\Lambda}},\vec{c}^\top \Psi|_{\Lambda})
=
\left[\begin{array}{@{}c@{}} \vec{e} \\ \vec{f}
 \end{array}\right].
$$
\end{remark}

\newpage

 \begin{theorem} A call of \texttt{eval} yields the output as specified, at the cost of
 ${\mathcal O}(\# \breve{\Pi}+\# \breve{\Lambda}+\# \Pi+\# \Lambda)$ operations.

 \end{theorem}

 \begin{proof}
By its definition, $\# \breve{\underline{\Pi}} \lesssim \#\breve{\Pi}_B +\# \breve{\Lambda}_\ell \lesssim \# \Lambda_\ell+\# \breve{\Lambda}_\ell$. So after sufficiently many recursive calls, the current set  $\breve{\Pi}  \cup \breve{\Lambda}$ will be empty.
For use later, we note that 
 $\# \underline{\Pi}  \lesssim \#\Pi_B +\#\Lambda_\ell \lesssim \#\breve{\Lambda}_\ell+\#\breve{\Pi}_B +\#\Lambda_\ell \lesssim \#\Lambda_\ell+\#\breve{\Lambda}_\ell$.

 For $\breve{\Pi}  \cup \breve{\Lambda}= \emptyset$, the call produces nothing, which is correct.

Now let  $\breve{\Pi}  \cup \breve{\Lambda} \neq \emptyset$.
From $\Lambda$ being an $\ell$-tree, the definitions of $S(\cdot)$ and $\breve{\Pi}_A$, and the locality of $a$, one has
 $$
 \vec{e}|_{\breve{\Pi}_A}=a(\breve{\Phi}|_{\breve{\Pi}_A},u)=a(\breve{\Phi}|_{\breve{\Pi}_A},\vec{d}^\top\Phi|_{\Pi}).
 $$

 Thanks to \eqref{c7}, the definition of $\underline{\Pi}$ shows that $\supp \big( \mathfrak{p}_{\ell} \vec{d}|_{\Pi_B}+\mathfrak{q}_\ell \vec{c}|_{\Lambda_{\ell}}\big) \subset \underline{\Pi}$, and so
  $$
 \underline{u}:=\underline{\vec{d}}^\top\Phi|_{\underline{\Pi}}+\vec{c}|_{\Lambda_{\ell+1\uparrow}}^\top \Psi|_{\Lambda_{\ell+1\uparrow}} =(\vec{d}|_{\Pi_B})^\top \Phi|_{\Pi_B} +\vec{c}^\top \Psi|_{\Lambda}=u-(\vec{d}|_{\Pi_A})^\top \Phi|_{\Pi_A}.
 $$

By induction the recursive call  yields
 $\underline{\vec{e}}=a(\breve{\Phi}|_{\breve{\underline{\Pi}}}, \underline{u})$, and
 $\underline{\vec{f}}=a(\breve{\Psi}|_{\breve{\Lambda}_{\ell+1 \uparrow}}, \underline{u})$.
 From  $\breve{\Lambda}$ being an $\ell$-tree, the definitions of $\breve{S}(\cdot)$ and $\Pi_A$, and the locality of $a$, we have
 $$
 a(\breve{\Psi}|_{\breve{\Lambda}_{\ell \uparrow}}, u)=a(\breve{\Psi}|_{\breve{\Lambda}_{\ell \uparrow}}, \underline{u}),
 $$
and so in particular $\vec{f}|_{\Lambda_{\ell+1,\uparrow}}=\underline{\vec{f}}$.

Thanks to \eqref{c7}, the definition of $\breve{\underline{\Pi}}$ shows that
$$
 \breve{\Phi}|_{\breve{\Pi}_B}=(\breve{\mathfrak{p}}_{\ell} ^\top \breve{\Phi}|_{\underline{\breve{\Pi}}})|_{\breve{\Pi}_B},\quad \breve{\Psi}|_{\breve{\Lambda}_\ell}=(\mathfrak{\breve{q}}_\ell^\top \breve{\Phi}|_{\underline{\breve{\Pi}}})|_{\breve{\Lambda}_\ell}.
 $$
We conclude that 
 $$
  \vec{f}|_{\breve{\Lambda}_\ell}= a(\breve{\Psi}|_{\breve{\Lambda}_\ell}, u) =a(\breve{\Psi}|_{\breve{\Lambda}_\ell}, \underline{u})=
   \big(\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}}\big)|_{\breve{\Lambda}_\ell},
 $$ 
 and from $|\supp \phi_\lambda \cap \supp \breve{\phi}_\mu|=0$ for $(\lambda,\mu) \in \Pi_A \times\breve{\Pi}_B$, that
$$
 \vec{e}|_{\breve{\Pi}_B}=  a(\breve{\Phi}|_{\breve{\Pi}_B}, u)= a(\breve{\Phi}|_{\breve{\Pi}_B}, \underline{u})=   \big(\breve{\mathfrak{p}}_{\ell} ^\top \underline{\vec{e}}\big)|_{\breve{\Pi}_B}.
$$

 From the assumptions on the collections $\Phi$, $\breve{\Phi}$, $\breve{\Psi}$, and $\Psi$, and their consequences on the sparsity of the matrices $\mathfrak{p}_{\ell} $, $\breve{\mathfrak{p}}_{\ell}$, $\mathfrak{q}_\ell$, and $\mathfrak{\breve{q}}_\ell$, one infers that the total cost of the evaluations of the statements in \texttt{eval} is ${\mathcal O}(\# \breve{\Pi}+\# \breve{\Lambda}_\ell+\# \Pi+\# \Lambda_\ell)$ plus the cost of the recursive call. Using 
 $ \# \breve{\underline{\Pi}} + \# \underline{\Pi} \lesssim \# \breve{\Lambda}_\ell+\# \Lambda_\ell$
 and induction, we conclude the second statement of the theorem.
 \end{proof}

 \newpage
 \subsection{A routine \texttt{evalupp} } \mbox{}

Let $a\colon \Span \breve{\Phi} \times \Span \Phi \rightarrow \R$ be local and {\em bilinear}.
Set $U=[a(\breve{\psi}_\lambda,\psi_\mu)]_{|\lambda| \leq |\mu|}$, $L=[a(\breve{\psi}_\lambda,\psi_\mu)]_{|\lambda| > |\mu|}$, so that $A=[a(\breve{\psi}_\lambda,\psi_\mu)]_{(\lambda,\mu) \in \breve{\vee}\times \vee}$ satisfies $A=L+U$. This splitting is going to be useful in the tensor product setting. \bigskip

{\em \begin{tabbing}
$[\vec{e},\,\vec{f}]:=\mathtt{evalupp}(a)(\ell,  \breve{\Pi},\breve{\Lambda},\Pi,\Lambda,\vec{d},\vec{c})$\\
%
{\em \%} Input: \= $\ell \in \N$,
$\breve{\Pi} \subset \breve{\Delta}_{\ell-1}, \Pi \subset \Delta_{\ell-1}$, 
$\ell$-trees $\breve{\Lambda} \subset \breve{\vee}_{\ell\uparrow}$ and $\Lambda \subset {\vee}_{\ell\uparrow}$,\\
{\em \%} \> $\vec{d}\in \R^{\# \Pi},\, \vec{c}\in  \R^{\#  \Lambda}$. \\
%
{\em \%} Output: With $u:=\vec{d}^\top \Phi|_{\Pi}+\vec{c}^\top \Psi|_{\Lambda}$,\\
{\em \%} \> $\vec{e}=a(\breve{\Phi}|_{\breve{\Pi}},u)$,
$\vec{f}=U|_{\breve{\Lambda} \times \Lambda} \vec{c}$.\\ \\
%
%
%
%
{\em \texttt{if}} \= $\breve{\Pi}\cup \breve{\Lambda}\neq \emptyset$ {\em \texttt{then}} \\
\>$\breve{\Pi}_B:=\{\lambda \in \breve{\Pi}\colon \big|\supp \breve{\phi}_\lambda \cap \cup_{\mu \in \Lambda_\ell}S(\mu)\big|>0$, $\breve{\Pi}_A:=\breve{\Pi} \setminus \breve{\Pi}_B$\\
\>$\breve{\underline{\Pi}}:=\{\lambda \in \breve{\Delta}_{\ell}\colon \supp \breve{\phi}_\lambda \subset \big(\cup_{\mu \in \breve{\Pi}_B} \supp \breve{\phi}_\mu \cup_{\gamma \in \breve{\Lambda}_\ell} \supp\breve{\psi}_\gamma\big)\}$\\
\>$\underline{\Pi}:=\{\lambda \in \Delta_{\ell}\colon \supp \phi_\lambda \subset \ \cup_{\gamma \in \Lambda_\ell} \supp\psi_\gamma\}$\\ 
%
\>$\underline{\vec{d}}:=\big(\mathfrak{q}_\ell \vec{c}|_{\Lambda_{\ell}}\big)|_{\underline{\Pi}}$\\
\>$[\underline{\vec{e}},\,\underline{\vec{f}}]:=\mathtt{evalupp}(a)(\ell+1, \breve{\underline{\Pi}}, \breve{\Lambda}_{\ell+1\uparrow}, \underline{\Pi}, \Lambda_{\ell+1\uparrow},\underline{\vec{d}}, \vec{c}|_{\Lambda_{\ell+1\uparrow}})$\\
\>$\vec{e}=
\left[\begin{array}{@{}l@{}} \vec{e}|_{\breve{\Pi}_A}\\ \vec{e}|_{\breve{\Pi}_B}\end{array}\right]
:=
\left[\begin{array}{@{}l@{}} a(\breve{\Phi}|_{\breve{\Pi}_A},\vec{d}^\top\Phi|_{\Pi}) \\ 
a(\breve{\Phi}|_{\breve{\Pi}_B},\vec{d}^\top\Phi|_{\Pi})+(\breve{\mathfrak{p}}_{\ell} ^\top \underline{\vec{e}})|_{\breve{\Pi}_B}
\end{array}\right]$\\
\>$\vec{f}=
\left[\begin{array}{@{}l@{}} \vec{f}|_{\breve{\Lambda}_\ell} \\ \vec{f}|_{\breve{\Lambda}_{\ell+1\uparrow}}\end{array}\right]
:=
\left[\begin{array}{@{}l@{}} (\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}})|_{\breve{\Lambda}_{\ell}} \\ \underline{\vec{f}} \end{array}\right]$\\
{\em \texttt{endif}}
 \end{tabbing}}

  \begin{remark} Let $\breve{\Lambda} \subset \breve{\vee}$, $\Lambda \subset \vee$ be  trees, and $\vec{c} \in \ell_2(\Lambda)$, then
   $$
 [\vec{e},\,\vec{f}]:=\mathtt{evalupp}(a)(1,\breve{\Lambda}_{0},\breve{\Lambda}_{1 \uparrow},\Lambda_0,\Lambda_{1 \uparrow},\vec{c}|_{\Lambda_0},\vec{c}|_{\Lambda_{1 \uparrow}}),
 $$
satisfies
$$
U|_{\breve{\Lambda} \times \Lambda} \vec{c}
=
\left[\begin{array}{@{}c@{}} \vec{e} \\ \vec{f}
 \end{array}\right].
$$
\end{remark}

\newpage
 \begin{theorem} A call of \texttt{evalupp} yields the output as specified, at the cost of
 ${\mathcal O}(\# \breve{\Pi}+\# \breve{\Lambda}+\# \Pi+\# \Lambda)$ operations.
 \end{theorem}

 \begin{proof}
By its definition, $\# \breve{\underline{\Pi}} \lesssim \#\breve{\Pi}_B +\# \breve{\Lambda}_\ell \lesssim \# \Lambda_\ell+\# \breve{\Lambda}_\ell$. So after sufficiently many recursive calls, the current set  $\breve{\Pi}  \cup \breve{\Lambda}$ will be empty.
Notice that 
 $\# \underline{\Pi}  \lesssim \#\Lambda_\ell$.

 For $\breve{\Pi}  \cup \breve{\Lambda}= \emptyset$, the call produces nothing, which is correct.

Now let  $\breve{\Pi}  \cup \breve{\Lambda} \neq \emptyset$.
From $\Lambda$ being an $\ell$-tree, the definitions of $S(\cdot)$ and $\breve{\Pi}_A$, and the locality of $a$, one has
 $$
 \vec{e}|_{\breve{\Pi}_A}=a(\breve{\Phi}|_{\breve{\Pi}_A},u)=a(\breve{\Phi}|_{\breve{\Pi}_A},\vec{d}^\top\Phi|_{\Pi}).
 $$

 Thanks to \eqref{c7}, the definition of $\underline{\Pi}$ shows that $\supp \big(\mathfrak{q}_\ell \vec{c}|_{\Lambda_{\ell}}\big) \subset \underline{\Pi}$, and so
  $$
 \underline{u}:=\underline{\vec{d}}^\top\Phi|_{\underline{\Pi}}+\vec{c}|_{\Lambda_{\ell+1\uparrow}}^\top \Psi|_{\Lambda_{\ell+1\uparrow}} =\vec{c}^\top \Psi|_{\Lambda}=u-\vec{d}^\top \Phi|_{\Pi}.
 $$

By induction the recursive call  yields
 $\underline{\vec{e}}=a(\breve{\Phi}|_{\breve{\underline{\Pi}}}, \underline{u})$, and
 $\underline{\vec{f}}=U_{\breve{\Lambda}_{\ell+1 \uparrow} \times \Lambda_{\ell+1 \uparrow}} c|_{\Lambda_{\ell+1 \uparrow}}=\vec{f}|_{\breve{\Lambda}_{\ell+1 \uparrow}}$.

Thanks to \eqref{c7}, the definition of $\breve{\underline{\Pi}}$ shows that
$$
 \breve{\Phi}|_{\breve{\Pi}_B}=(\breve{\mathfrak{p}}_{\ell} ^\top \breve{\Phi}|_{\underline{\breve{\Pi}}})|_{\breve{\Pi}_B},\quad \breve{\Psi}|_{\breve{\Lambda}_\ell}=(\mathfrak{\breve{q}}_\ell^\top \breve{\Phi}|_{\underline{\breve{\Pi}}})|_{\breve{\Lambda}_\ell}.
 $$
We conclude that 
 $$
  \vec{f}|_{\breve{\Lambda}_\ell}= a(\breve{\Psi}|_{\breve{\Lambda}_\ell}, \vec{c}^\top \Psi|_{\Lambda}) =a(\breve{\Psi}|_{\breve{\Lambda}_\ell}, \underline{u})=
   \big(\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}}\big)|_{\breve{\Lambda}_\ell},
 $$ 
and
$$
 \vec{e}|_{\breve{\Pi}_B}=  a(\breve{\Phi}|_{\breve{\Pi}_B}, u)= a(\breve{\Phi}|_{\breve{\Pi}_B}, \underline{u})+a(\breve{\Phi}|_{\breve{\Pi}_B},\vec{d}^\top\Phi|_{\Pi})=   \big(\mathfrak{p}_{\ell} ^\top \underline{\vec{e}}\big)|_{\breve{\Pi}_B}+a(\breve{\Phi}|_{\breve{\Pi}_B},\vec{d}^\top\Phi|_{\Pi}).
$$

 From the assumptions on the collections $\Phi$, $\breve{\Phi}$, $\breve{\Psi}$, and $\Psi$, and their consequences on the sparsity of the matrices $\mathfrak{p}_{\ell} $, $\breve{\mathfrak{p}}_{\ell}$, $\mathfrak{q}_\ell$, and $\mathfrak{\breve{q}}_\ell$, one infers that the total cost of the evaluations of the statements in \texttt{eval} is ${\mathcal O}(\# \breve{\Pi}+\# \breve{\Lambda}_\ell+\# \Pi+\# \Lambda_\ell)$ plus the cost of the recursive call. Using 
 $ \# \breve{\underline{\Pi}} + \# \underline{\Pi} \lesssim \# \breve{\Lambda}_\ell+\# \Lambda_\ell$
 and induction, we conclude the second statement of the theorem.
 \end{proof}

 \newpage
  \subsection{A routine \texttt{evallow}} \label{SevalL}
  Let $a\colon \Span \breve{\Phi} \times \Span \Phi \rightarrow \R$ be local and {\em bilinear}.
Recall $L=[a(\breve{\psi}_\lambda,\psi_\mu)]_{|\lambda| > |\mu|}$.

 {\em \begin{tabbing}
$\vec{f}:=\mathtt{evallow}(a)(\ell,\breve{\Lambda},\Pi,\Lambda,\vec{d},\vec{c})$\\
%
{\em \%} Input: \= $\ell \in \N$,
$\Pi\subset \Delta_{\ell-1}$, 
$\ell$-trees $\breve{\Lambda} \subset \breve{\vee}_{\ell\uparrow}$ and $\Lambda \subset {\vee}_{\ell\uparrow}$,\\
{\em \%} \> $\vec{d} \in \R^{\# \Pi},\, \vec{c} \in \R^{\# \Lambda}$. \\
%
{\em \%} Output: 
$\vec{f}=a(\breve{\Psi}|_{\breve{\Lambda}},\Phi|_{\Pi})\vec{d}+L|_{\breve{\Lambda} \times \Lambda} \vec{c}$.\\ \\
%
%
{\em \texttt{if}} \= $\breve{\Pi} \cup \breve{\Lambda}\neq \emptyset$ {\em \texttt{then}}\\
%
%
\>$\Pi_B:=\{\lambda \in \Pi\colon \big|\supp \phi_\lambda \cap \cup_{\mu \in \breve{\Lambda}_\ell} \breve{S}(\mu) \big|>0\}$,\\
\>$\underline{\Pi}:=\{\lambda \in \Delta_{\ell}\colon \supp \phi_\lambda \subset \big(\cup_{\mu \in \Pi_B} \supp \phi_\mu \cup_{\gamma \in \Lambda_\ell} \supp\psi_\gamma\big)\}$\\

\>$\underline{\Pi}_B:=\{\lambda \in \Delta_{\ell}\colon \supp \phi_\lambda \subset \cup_{\mu \in \Pi_B} \supp \phi_\mu\}$\\
\>$\underline{\breve{\Pi}}_B:=\{\lambda \in \breve{\Delta}_{\ell}\colon \supp \breve{\phi}_\lambda \subset \cup_{\lambda  \in \breve{\Lambda}_\ell} \supp \psi_\lambda\}$\\
%
\>$\underline{\vec{e}}:=a(\breve{\Phi}|_{\underline{\breve{\Pi}}_B},\Phi|_{\underline{\Pi}_B}) \mathfrak{p}_{\ell} \vec{d}|_{\Pi_B}$\\
\>$\underline{\vec{d}}:=(\mathfrak{p}_{\ell} \vec{d}|_{\Pi_B}+\mathfrak{q}_\ell \vec{c}|_{\Lambda_{\ell}})|_{\underline{\Pi}}$\\
\>$\vec{f}=
\left[\begin{array}{@{}l@{}} \vec{f}|_{\breve{\Lambda}_\ell} \\ \vec{f}|_{\breve{\Lambda}_{\ell+1\uparrow}}\end{array}\right]
:=
\left[\begin{array}{@{}l@{}} (\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}})|_{\breve{\Lambda}_{\ell}} \\ 
\mathtt{evallow}(a)(\ell+1, \breve{\Lambda}_{\ell+1\uparrow}, \underline{\Pi}, \Lambda_{\ell+1\uparrow},\underline{\vec{d}}, \vec{c}|_{\Lambda_{\ell+1\uparrow}}) \end{array}\right]$\\
{\em \texttt{endif}}
 \end{tabbing}}

 \begin{remark} \label{rem3}
 Let $\breve{\Lambda} \subset \breve{\vee}$, $\Lambda \subset \vee$ be trees, and $\vec{c} \in \ell_2(\Lambda)$, then 
 $$
 L|_{\breve{\Lambda} \times \Lambda} \vec{c}=
\mathtt{evallow}(a)(1,\breve{\Lambda}_{1 \uparrow},\Lambda_0,\Lambda_{1 \uparrow},\vec{c}|_{\Lambda_0},\vec{c}|_{\Lambda_{1 \uparrow}}).
 $$
\end{remark}
 \newpage

\begin{theorem} A call of \texttt{evallow} yields the output as specified, at the cost of
 ${\mathcal O}(\# \breve{\Lambda}+\# \Pi+\# \Lambda)$ operations.

 \end{theorem}

 \begin{proof}
Notice that
 $\# \underline{\Pi}  \lesssim \#\Lambda_\ell+\#\Pi_B \lesssim  \#\Lambda_\ell+\#\breve{\Lambda}_\ell$.

 For $\breve{\Pi} \cup \breve{\Lambda}= \emptyset$, the call produces nothing, which is correct.

Now let  $\breve{\Pi} \cup \breve{\Lambda}\neq \emptyset$. The definitions of $\underline{\breve{\Pi}}_B$ and $\underline{\Pi}_B$ together with \eqref{c7} show that \begin{align*}
\vec{f}|_{\breve{\Lambda}_\ell} = a(\breve{\Psi}_{\breve{\Lambda}_\ell},\Phi|_{\Pi})\vec{d}= a(\breve{\Psi}_{\breve{\Lambda}_\ell},\Phi|_{\Pi})\vec{d}|_{\Pi_B}&=(\mathfrak{\breve{q}}_\ell^\top a(\breve{\Phi}|_{\underline{\breve{\Pi}}_B},\Phi|_{\underline{\Pi}_B})\mathfrak{p}_{\ell} \vec{d}|_{\Pi_B})|_{\breve{\Lambda}_{\ell}}\\
&=(\mathfrak{\breve{q}}_\ell^\top \underline{\vec{e}})|_{\breve{\Lambda}_{\ell}}
\end{align*}

From  $\breve{\Lambda}$ being an $\ell$-tree, the definitions of $\breve{S}(\cdot)$ and $\Pi_B$, and the locality of $a$, and for the third equality, the definition of $\underline{\Pi}$ and \eqref{c7}, one has
\begin{align*}
f|_{\breve{\Lambda}_{\ell+1 \uparrow}}&=a(\breve{\Psi}|_{\breve{\Lambda}_{\ell+1 \uparrow}},\Phi|_\Pi)\vec{d}
+
L|_{\breve{\Lambda}_{\ell+1 \uparrow} \times \Lambda_\ell}\vec{c}|_{\Lambda_\ell}+
L|_{\breve{\Lambda}_{\ell+1 \uparrow} \times \Lambda_{\ell+1 \uparrow }}\vec{c}|_{\Lambda_{\ell+1\uparrow}}\\
& = a(\breve{\Psi}|_{\breve{\Lambda}_{\ell+1 \uparrow}},\Phi|_\Pi)\vec{d}|_{\Pi_B}+
a(\breve{\Psi}|_{\breve{\Lambda}_{\ell+1 \uparrow}},\Psi|_{\Lambda_\ell})
\vec{c}|_{\Lambda_\ell}+L|_{\breve{\Lambda}_{\ell+1 \uparrow} \times \Lambda_{\ell+1 \uparrow }}\vec{c}|_{\Lambda_{\ell+1\uparrow}}
\\
& = a(\breve{\Psi}|_{\breve{\Lambda}_{\ell+1 \uparrow}},\Phi|_{\underline{\Pi}})\underline{\vec{d}}+L|_{\breve{\Lambda}_{\ell+1 \uparrow} \times \Lambda_{\ell+1 \uparrow }}\vec{c}|_{\Lambda_{\ell+1\uparrow}}\\
& =\mathtt{evallow}(a)(\ell+1, \breve{\Lambda}_{\ell+1\uparrow}, \underline{\Pi}, \Lambda_{\ell+1\uparrow},\underline{\vec{d}}, \vec{c}|_{\Lambda_{\ell+1\uparrow}}) \end{align*}
 by induction.

 From the assumptions on the collections $\Phi$, $\breve{\Psi}$, and $\Psi$, and their consequences on the sparsity of the matrices $\mathfrak{p}_{\ell} $, $\mathfrak{q}_\ell$, and $\mathfrak{\breve{q}}_\ell$, one easily infers that the total cost of the evaluations of the statements in \texttt{evallow} is ${\mathcal O}(\# \breve{\Lambda}_\ell+\# \Pi+\# \Lambda_\ell)$ plus the cost of the recursive call. Using 
 $ \# \underline{\Pi} \lesssim \# \breve{\Lambda}_\ell+\# \Lambda_\ell$
 and induction, we conclude the second statement of the theorem.
 \end{proof}

 \newpage
%%%%%%%%%%% 
\section{The application of tensor product operators} \label{Sappl-of-tensors}
For $i \in \{1,2\}$, let $a_i\colon \Span \breve{\Phi}_i \times \Span \Phi_i \rightarrow \R$ be local and {\em bilinear}, and 
let $A_i:=a_i(\breve{\Psi}_i,\Psi_i)$ be splitted into $L_i+U_i$, where $U_i:=[(A_i)_{\lambda,\mu}]_{|\lambda| \leq |\mu|}$, $L_i:=[(A_i)_{\lambda,\mu}]_{|\lambda| > |\mu|}$.

For $i \in \{1,2\}$, let $\neg i$ be the value in $\{1,2\}$ unequal to $i$.

Let the coordinate projector $P_i (b_1,b_2):=b_i$. \rem{In our setting, it seems best when first coordinate corresponds to the temporal spaces, and the second coordinate corresponds to the spatial spaces.} For $\bm{\Lambda} \subset \{\breve{\vee}^{1} \times \breve{\vee}^2,\vee^1 \times \breve{\vee}^2, \breve{\vee}^1 \times \vee^2, \vee^1 \times \vee^2\}$, we call $\bm{\Lambda}$ an {\em double-tree} when for $i \in \{1,2\}$ and any $\mu\in P_{\lnot i}\bm{\Lambda}$, the fiber
$$
\bm{\Lambda}_{i,\mu}:=P_i(P_{\lnot i}|_{\bm{\Lambda}})^{-1} \{\mu\}
$$
is a tree (in $\breve{\vee}^i $ or $\vee^i$). That is, $\bm{\Lambda}$ is a double-tree when `frozen' in each of its coordinates, at any value of that coordinate, it is a tree in the remaining coordinate.

\rem{$X^\delta=\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes \tilde{W}_\lambda^\delta$ is a multi-tree iff $\tilde{W}_\lambda^\delta \subset \tilde{W}_\mu^\delta$ whenever $\mu$ is parent of $\lambda$, and each $\tilde{W}_\lambda^\delta$ is either $\{0\}$ or a fem space w.r.t. conforming NVB partition (with fixed $\tria_\bot$).}

\begin{figure}
\begin{center}
\input{multitrees.pdf_t}
\end{center}
\caption{Projection of a subset $\bm{\Lambda}$ of the Cartesian product of two index sets onto a coordinate, and fibers.}
\end{figure}

From $\bm{\Lambda}=\cup_{\mu\in P_{\lnot i}\bm{\Lambda}} (P_{\lnot i}|_{\bm{\Lambda}})^{-1} \{\mu\}$, we have 
$P_i \bm{\Lambda}=\cup_{\mu\in P_{\lnot i}\bm{\Lambda}} \bm{\Lambda}_{i,\mu}$, which, being a union of trees, is a tree itself.

For a subset $\lhd$ of a (double) index set $\Diamond$, let $I_\lhd^\Diamond$ denote the extension operator with zeros of a vector supported on $\lhd$ to one on $\Diamond$, and let $R_\lhd^\Diamond$ denotes its (formal) adjoint, being the restriction operator of a vector supported on $\Diamond$ to one on $\lhd$. Since the set $\Diamond$ will always be clear from the context, we will denote these operators simply by $I_\lhd$ and $R_\lhd$.%

A generalization of the following theorem to double-trees, i.e., multi-index sets that frozen in all but any coordinate direction are trees, was given in \cite[Thm.~3.1]{171.7}, with a different definition of a (single) tree though.
Therefore, for completeness we include a proof.

\begin{theorem} \label{thm1} Let $\bm{\breve{\Lambda}} \subset \breve{\vee}^1 \times \breve{\vee}^2$, $\bm{\Lambda} \subset \vee^1 \times \vee^2$ be finite double-trees. Then
\begin{align*}
\bm{\Sigma}&:=\bigcup_{\lambda \in P_1 \bm{\Lambda}} \{\lambda\} \times \big\{
\bm{\breve{\Lambda}}_{2,\mu}
:\mu \in P_1 \bm{\breve{\Lambda}},\,|\mu|=|\lambda|+1,\,|\breve{S}^1(\mu) \cap S^1(\lambda)|>0\big\},\\
\bm{\Theta}&:=\bigcup_{\lambda \in P_{2} \bm{\Lambda}} 
\{\mu \in  P_{1} \bm{\breve{\Lambda}} : \exists \gamma \in \bm{\Lambda}_{1,\lambda} \text{ s.t. } |\gamma|= |\mu|,\,|\breve{S}^1(\mu) \cap S^1(\gamma)|>0\}
 \times \{\lambda\},
\end{align*}
are double-trees with $\# \bm{\Sigma} \lesssim \# \bm{\breve{\Lambda}}$ and $\# \bm{\Theta} \lesssim \# \bm{\Lambda}$, and
\begin{equation} \label{splitting}
\begin{split}
R_{\bm{\breve{\Lambda}}} (A_1 \otimes A_2) I_{\bm{\Lambda}}=
 & R_{\bm{\breve{\Lambda}}} (L_1 \otimes \mathrm{Id}) I_{\bm{\Sigma}}  R_{\bm{\Sigma}} (\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}+\\
& R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2) I_{\bm{\Theta}}  R_{\bm{\Theta}} (U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}.
\end{split}
\end{equation}
\end{theorem}

\begin{proof}We write
\begin{align} \nonumber
R_{\bm{\breve{\Lambda}}} (A_1 \otimes A_2) I_{\bm{\Lambda}}=&R_{\bm{\breve{\Lambda}}} ((L_1+U_1) \otimes A_2) I_{\bm{\Lambda}}\\ 
\label{first}
=&R_{\bm{\breve{\Lambda}}} (L_1\otimes \mathrm{Id})(\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}+
\\ 
\label{second}
&R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2) (U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}.
\end{align}

Considering \eqref{first}, the range of $(\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}$ consists of vectors whose entries with first index outside $P_1\bm{\Lambda}$ are zero.
In view of the subsequent application of $L_1 \otimes I$, furthermore only those indices $(\lambda,\gamma) \in P_1\bm{\Lambda} \times \breve{\vee}^2$ of these vectors might be relevant for which $\exists (\mu,\gamma) \in \bm{\breve{\Lambda}}$, i.e. $\gamma \in \bm{\Lambda}_{2,\mu}$, with $|\mu|>|\lambda|$ and $|\breve{S}^1(\mu) \cap S^1(\lambda)|>0$.
Indeed $|\breve{S}^1(\mu) \cap S^1(\lambda)|=0$ implies
$|\supp \breve{\psi}^1_\mu \cap \supp \psi^1_\lambda|=0$, and so $a_1(\breve{\psi}^1_\mu,\psi^1_\lambda)=0$. 
If for given $(\lambda,\gamma)$ such a pair $(\mu,\gamma)$ exists for  $|\mu|>|\lambda|$, then such a pair exists for $|\mu|=|\lambda|+1$ as well, because $\bm{\breve{\Lambda}}_{1,\gamma}$ is a tree, and $\breve{S}^1(\mu') \supset \breve{S}^1(\mu)$ for any ancestor $\mu'$ of $\mu$.
In order words, the condition $|\mu|>|\lambda|$ can be read as $|\mu|=|\lambda|+1$. The set of $(\lambda,\gamma)$ that we just described is given by the set $\bm{\Sigma}$, and so we infer that
$$
R_{\bm{\breve{\Lambda}}} (L_1\otimes \mathrm{Id})(\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}=R_{\bm{\breve{\Lambda}}} (L_1\otimes \mathrm{Id})I_{\bm{\Sigma}}  R_{\bm{\Sigma}}(\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}.
$$

Now let $(\lambda,\gamma) \in \bm{\Sigma}$. Using that $P_1\bm{\Lambda}$ is a tree, and $S^1(\lambda) \subset S^1(\lambda')$ for any ancestor $\lambda'$ of $\lambda$, we infer that $(\lambda',\gamma) \in \bm{\Sigma}$. Using that for any $\mu \in P_1 \bm{\breve{\Lambda}}$, $\bm{\breve{\Lambda}}_{2,\mu}$ is a tree, we infer that for any ancestor $\gamma'$ of $\gamma$, $(\lambda,\gamma') \in \bm{\Sigma}$, so that $\bm{\Sigma}$ is a double-tree.

For any $\mu \in \breve{\vee}^1$, the number of $\lambda \in \vee^1$ with $|\mu|=|\lambda|+1$ and $|\breve{S}^1(\mu) \cap S^1(\lambda)|>0$ is uniformly bounded, 
from which we infer that $\# \bm{\Sigma} \lesssim \sum_{\mu \in P_1 \bm{\breve{\Lambda}}} \#\bm{\breve{\Lambda}}_{2,\mu}=\#\bm{\breve{\Lambda}}$.

Considering \eqref{second}, the range of $(U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}$ consists of vectors that can only have non-zero entries for indices $(\mu,\lambda) \in \breve{\vee}^1 \times P_2\bm{\Lambda}$ for which there exists a $\gamma \in \bm{\Lambda}_{1,\lambda}$ with $|\gamma|\geq |\mu|$ and $|\breve{S}^1(\mu) \cap S^1(\gamma)|>0$.
Since $\bm{\Lambda}_{1,\lambda}$ is a tree, and $S^1(\gamma') \supset S^1(\gamma)$ for any ancestor $\gamma'$ of $\gamma$, 
equivalently $|\gamma|\geq |\mu|$ can be read as $|\gamma|= |\mu|$. Furthermore, in view of the subsequent application of $R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2)$, it suffices to consider 
those indices $(\mu,\lambda)$ with $\mu \in P_1\bm{\breve{\Lambda}}$. 
The set of $(\mu,\lambda)$ that we just described is given by the set $\bm{\Theta}$, and so we infer that
$$
R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2) (U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}=R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2) I_{\bm{\Theta}} R_{\bm{\Theta}} (U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}.
$$

Now let $(\mu,\lambda) \in \bm{\Theta}$. If $\lambda'$ is an ancestor of $\lambda$, then 
from $P_1 \bm{\Lambda}$ being a tree, and $\bm{\Lambda}_{1,\lambda} \subset \bm{\Lambda}_{1,\lambda'}$, we have $(\mu,\lambda') \in \bm{\Theta}$. If $\mu'$ is an ancestor of $\mu$, then from $P_1 \bm{\breve{\Lambda}}$ being a tree, and $\breve{S}^1(\mu') \supset \breve{S}^1(\mu)$, we infer that $(\mu',\lambda) \in \bm{\Theta}$, and thus that $\bm{\Theta}$ is a double-tree.

For any $\gamma \in \vee^1$, the number of $\mu \in \breve{\vee}^1$ with $|\mu|=|\gamma|$ and $|\breve{S}^1(\mu) \cap S^1(\gamma)|>0$ is uniformly bounded, 
from which we infer that $\# \bm{\Theta} \lesssim \sum_{\lambda \in P_2 \bm{\Lambda}} \#\bm{\Lambda}_{1,\lambda}=\#\bm{\Lambda}$.
\end{proof}

The application of $R_{\bm{\breve{\Lambda}}} (L_1 \otimes \mathrm{Id}) I_{\bm{\Sigma}}$ boils down to the application of 
$R_{\bm{\breve{\Lambda}}_{1,\mu}} L_1 I_{\bm{\Sigma}_{1,\mu}}$ for any $\mu \in P_2 \bm{\Sigma}\cap P_2\bm{\breve{\Lambda}}$.
Such an application can be performed in ${\mathcal O}(\#\bm{\breve{\Lambda}}_{1,\mu}+\#\bm{\Sigma}_{1,\mu})$ operations by means of a call of $\mathtt{evallow}(a_1)$.
Since
$\sum_{\mu \in \breve{\vee}_2}\#\bm{\breve{\Lambda}}_{1,\mu} + \#\bm{\Sigma}_{1,\mu} = \# \bm{\breve{\Lambda}}+ \# \bm{\Sigma}$,
we conclude that the application of $R_{\bm{\breve{\Lambda}}} (L_1 \otimes \mathrm{Id}) I_{\bm{\Sigma}}$ can be performed in ${\mathcal O}(\#\bm{\breve{\Lambda}}+\#\bm{\Sigma})$ operations.

Similarly, the applications of 
$R_{\bm{\Sigma}} (\mathrm{Id} \otimes A_2) I_{\bm{\Lambda}}$,
$R_{\bm{\breve{\Lambda}}} (\mathrm{Id} \otimes A_2) I_{\bm{\Theta}}$, and
$R_{\bm{\Theta}} (U_1 \otimes \mathrm{Id}) I_{\bm{\Lambda}}$
by means of calls of $\mathtt{eval}(a_2)$, $\mathtt{eval}(a_2)$, and $\mathtt{evalupp}(a_1)$, respectively,
can be performed in ${\mathcal O}(\# \bm{\Sigma}+ \# \bm{\Lambda})$, ${\mathcal O}(\#\bm{\breve{\Lambda}}+ \#\bm{\Theta})$, and ${\mathcal O}(\#\bm{\Theta}+ \#\bm{\Lambda})$ operations. From the bounds $\# \bm{\Sigma} \lesssim \# \bm{\breve{\Lambda}}$ and $\# \bm{\Theta} \lesssim \# \bm{\Lambda}$ we conclude the following:

\begin{corollary} Let $\bm{\breve{\Lambda}} \subset \breve{\vee}^1 \times \breve{\vee}^2$, $\bm{\Lambda} \subset \vee^1 \times \vee^2$ be finite double-trees, then $R_{\bm{\breve{\Lambda}}} (A_1 \otimes A_2) I_{\bm{\Lambda}}$ can be applied in ${\mathcal O}(\# \bm{\breve{\Lambda}}+\#\bm{\Lambda})$ operations. 
\end{corollary}

\rem{Voor efficiente evaluatie van tensor product operatoren op $X^\delta=\sum_\lambda \sigma_\lambda \otimes \tilde{W}_\lambda^\delta$ ($\tilde{W}_\lambda^\delta=W_\lambda^\delta$) en/of $Y^\delta=\sum_\lambda \psi_\lambda \otimes V_\lambda^\delta$ met $W_\lambda^\delta$, $V_\lambda^\delta$ continuous piecewise linears w.r.t conforme NVB partitie moeten we deze fem spaces uitrusten met hierarchische bases. Noteer zo'n basis als $\Psi^{hb}_\Lambda$ met $\Lambda$ een familie-tree (child heeft gewoonlijk 2 parents).
Binnen de evaluatie van die tensor product operatoren moeten we voor verschillende familie-trees $\Lambda$, $a_2(\Psi^{hb}_\Lambda,\Psi^{hb}_\Lambda)$ evalueren (misschien ook $a_2(\Psi^{hb}_{\Lambda_1},\Psi^{hb}_{\Lambda_2})$ voor $\Lambda_1\neq\Lambda_2$).

We kunnen dit doen met de routine eval($a_2$). Hiertoe moeten we dan ook $\Phi_\ell$ definieren. Bijv. als de nodale basis van de uniform verfijnde partitie op level $\ell$. Wellicht is er een betere optie:

$\Span \Psi_\Lambda^{hb}$ is een standaard fem space w.r.t. een conforme NVB triangulatie ${\mathcal T}$. Zo'n space heeft een natural single scale basis $\Phi_{\mathcal T}$, (welke voor ${\mathcal T}$ een niet-uniforme verfijning van de initial triangulatie $\tria_\bot$ geen subset is van $\cup_\ell \Phi_\ell$).
De evaluatie van $a_2(\Phi_{\mathcal T},\Phi_{\mathcal T})$ is echter standaard (bereken de lokale stiffness/mass per triangle and assemble), en daar willen we dus gebruik van maken (i.p.v. van de algemene routine eval($a_2$)).
 Met $T$ z.d.d. $(\Psi^{hb}_{\Lambda})^\top=\Phi_{\mathcal T}^\top T^\top$, geldt
 $a_2(\Psi^{hb}_\Lambda,\Psi^{hb}_\Lambda)= T^\top a_2(\Phi_{\mathcal T},\Phi_{\mathcal T}) T$.
 De toepassing van $T$ kan bottom up in linear complexity (assuming a suitable data structure) (en dus transpose top-down) geconstrueerd als compositie van deze elementaire transformaties:
\begin{figure}[h]
\begin{center}
\input{nvb2.pdf_t}
$
\left[
\begin{array}{@{}ccccc@{}}
1 & 0 & 0 & 0 & 0\\
0 & 1 & 0 & 0 & 0\\
0 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 1 & 0\\
0 & 0 & \frac12 & \frac12 & 1\\
\end{array}
\right]
$
\end{center}
\end{figure}
Merk op de parents 1, 2 van 5 zijn hier niet van belang, maar de godfathers 3 en 4.
}

\section{Error reduction} \label{Sreduction}
Consider abstract elliptic problem to find $u \in X$ (here some Hilbert space) s.t.
\be \label{n1}
a(u,v)=f(v)\quad (v \in X).
\ee
On $X$, define energy-norm $\|\cdot\|_E=a(\cdot,\cdot)^{\frac12} \eqsim \|\cdot\|_X$.
We aim to improve a current approximation $\tilde{u}^\delta$ for $u$.
Typically $\tilde{u}^\delta$ is an approximate Galerkin solution from some closed subspace $X^\delta \subset X$.

We assume that we can determine a closed subspace $\bar{X}^\delta \subset X$ with $X^\delta \subset \bar{X}^\delta$ s.t. for some constant $\varsigma<1$, and with $\bar{u}^\delta$ denoting the Galerkin solution from $\bar{X}^\delta$, it holds that
\be \label{n2}
\|u-\bar{u}^\delta\|_E \leq \varsigma \|u-\tilde{u}^\delta\|_E \quad \text{(saturation assumption)}.
\ee
(Verified in \cite{35.937} for Poisson problem under a smallness condition on the data-oscillation)

With the aim to achieve the best possible rate, we do not simply take $\bar{u}^\delta$ as the next approximation.
Instead, for some constants $\eps$ and $\rho<1$, let $v \in \bar{X}^\delta$ 
a Galerkin approximation on some space intermediate of $X^\delta$ and $\bar{X}^\delta$ constructed by bulk chasing such that
\be \label{n3}
\|\hat{u}^\delta-v\|_E \leq \rho \|\hat{u}^\delta-\tilde{u}^\delta\|_E,
\ee
where $\hat{u}^\delta \in \bar{X^\delta}$ is close to $\hat{u}^\delta$ in the sense that
\be \label{n4}
\|u-\hat{u}^\delta\|_E \leq \sqrt{1+\eps^2} \|u-\bar{u}^\delta\|_E.
\ee
Think of $\hat{u}^\delta$ being the Galerkin solution on $\bar{X}^\delta$ of a slightly perturbed elliptic problem.

Then one can show that
$$
\|u-v\|_E \leq \sqrt{\varsigma^2+\tilde{\rho}^2(1-\varsigma^2)}\|u-\tilde{u}^\delta\|_E
$$
where $\tilde{\rho}=(1-\rho) \eps \sqrt{\frac{\varsigma^2}{1-\varsigma^2}}+\rho$.
We will have to ensure that $\eps \varsigma$ is small enough such that $\tilde{\rho}<1$ (we don't want to assume that $\rho$ is small enough).

To realize bulk chasing: Consider first the case that $\tilde{u}^\delta$ and $\hat{u}^\delta$ are the Galerkin solutions of the unperturbed elliptic problem from $X^\delta$ and $\bar{X}^\delta$, respectively.
Let $\bar{X}^\delta=X^\delta \oplus V^\delta$ be a stable decomposition w.r.t. $a(\,,\,)$, and let $\bar{X}^\delta$ be equipped with a basis that consists of some basis for $X^\delta$ plus a basis for $V^\delta$, where the latter basis is $a(\,,\,)$-stable.
Then compute the residual of $\tilde{u}^\delta$ restricted to those entries corresponding to the basis functions for $V^\delta$ (the other entries are zero), and collect the absolute largest entries that make up a $\theta$-fraction of the whole residual.
Add the span of those basis functions to  $X^\delta$ (and extend to multi-tree)  to form the new Galerkin space, cf. \cite[Lemmas 5 and 6, and Proposition 5]{249.92}.
Then convergence and optimal rates are realized.
Perturbutions as a consequence of $\varsigma>0$, and $\eps>0$ (application of a perturbed elliptic problem in the bulk chasing process)
 yet have to be dealt with.

\subsection{Application to parabolic problem}
With $X=L_2(I;V) \cap H^1(I;V')$, $Y:=L_2(I,V)$, consider $\left[\begin{array}{@{}c@{}} B \\ \gamma_0\end{array}\right] u=\left[\begin{array}{@{}c@{}} g \\ u_0\end{array} \right]$. Recall that $\left[\begin{array}{@{}c@{}} B \\ \gamma_0\end{array}\right] \in \Lis(X,Y' \times H)$.
With $0<\bar{A}=\bar{A}' \in \Lis(Y,Y')$ (not necessarily $\bar{A}=A_s$), on $Y \times H \times X$ consider
\be \label{m1}
\left[\begin{array}{@{}ccc@{}} \bar{A} & 0 & B\\ 0 & \identity & \gamma_0\\ B' & \gamma_0' & 0\end{array}\right]
\left[\begin{array}{@{}c@{}} \mu \\ \sigma \\ u \end{array}\right]=
\left[\begin{array}{@{}c@{}} g \\ u_0 \\ 0 \end{array}\right],
\ee
or
\be \label{m2}
(B' \bar{A}^{-1} B+\gamma_0'\gamma_0)u=B' \bar{A}^{-1}g+\gamma_0' u_0.
\ee
The elliptic system \eqref{m2} will play the role of \eqref{n1}.

We will ``ensure'' saturation by enlarging the current $X^\delta$ in the spatial and temporal direction (for the moment using heuristics) to a space $\bar{X}^\delta$.
\rem{Voorstel:
Zij $X^\delta=\sum_{\lambda} \sigma_\lambda \otimes W_{\lambda}^\delta$ waarbij ik alleen sommeer over die $\lambda$ waarvoor $W_{\lambda}^\delta \neq \{0\}$ (de set of these $\lambda$ form a tree $\Lambda$).
Construeer $\bar{X}^\delta$ door, voor iedere $\lambda \in \Lambda$, de triangulatie $\tria_\lambda$ welke correspondeert met de fem space $W_{\lambda}^\delta$ te vervangen door $\tria^{++}_\lambda$, en verder,
voor alle $\lambda \not\in \Lambda$ met een parent in $\Lambda$, de ruimte $\sigma_\lambda \otimes W_0$ aan $X^\delta$ toe te voegen, waarbij $W_0$ de fem-space is w.r.t. the initial mesh $\tria_\bot$.

Merk op: Resulterende $\bar{X}^\delta$ is een multi-tree.

Voor het realiseren van de bulk chasing procedure zullen we gebruiken maken van het feit 
dat de splitting van $\bar{X}^\delta$ in $X^\delta$ en de ruimte die we toegevoegd hebben $X$-stabiel is, en dat we de ruimte die we toegevoegd hebben kunnen uitrusten met en $X$-stabiele basis.
}

Given any close subspace $X^\delta \subset X$, we will approximate the Galerkin solution of \eqref{m2} by the third component of the Galerkin solution of \eqref{m1} with subspace $Y^\delta \times H^\delta \times X^\delta$.
This is the `slightly perturbed' elliptic problem mentioned in Sect.~\ref{Sreduction}. 
Assuming $Y^\delta \supset X^\delta$ and $H^\delta \supset \ran \gamma_0|_{X^\delta}$, we have \eqref{n4} with
\be \label{m3}
\sqrt{1+\eps^2}\leq\Big(\inf_{u \in X^\delta} \sup_{0 \neq w \in Y^\delta} \frac{(\partial_t u)(w)}{\|\partial_t u\|_{Y'}\|w\|_Y} 
\Big)^{-1}
\ee
\rem{This estimate isn't possible with the new variational formulation developed in \cite{249.99}. Indeed, with that formulation the upper bound for $\|u -\hat{u}^\delta\|_X$ in \eqref{n4} will also depend on $\inf_{w \in Y^\delta}\|u-w\|_Y$ which, unless $X^\delta \subset Y^\delta$ (in which case the new formulation doesn't provide advantages), cannot be controlled by a multiple of $\|u-u^\delta\|_X$}
In Sect.~\ref{Sinfsup} (``Better approach'') we have seen how to choose $Y^\delta \times H^\delta$, dependent on $X^\delta$ to ensure 
 $Y^\delta \supset \bar{X}^\delta$, $H^\delta \supset \ran \gamma_0|_{X^\delta}$, and \eqref{m3}, with in particular the right-hand side of \eqref{m3} bounded uniformly in $X^\delta$.
For the analysis in Sect.~\ref{Sreduction} it is needed that $\eps$ is sufficiently small, which we expect can be guaranteed by making $Y^\delta$ sufficiently large.
Likely this isn't needed in practice, so for the moment let us not go into that.

\bibliographystyle{alpha}
\bibliography{../ref}
\end{document}

### 2019-06-05 Python/space-impl.py

@@ -1,0 +1,412 @@
+import itertools
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.tri as mpltri
+from mpl_toolkits.mplot3d import Axes3D
+from numpy.linalg import norm
+import time
+from scipy.sparse.linalg import LinearOperator, cg
+
+
+class Vertex(object):
+    """ A vertex in a locally refined triangulation. """
+
+    def __init__(self, x, y, on_domain_boundary):
+        self.x = x
+        self.y = y
+        self.on_domain_boundary = on_domain_boundary
+
+    def as_array(self):
+        return np.array([self.x, self.y], dtype=float)
+
+
+class Triangle(object):
+    """ A triangle as part of a locally refined triangulation. """
+
+    def __init__(self, idx, vertex_ids, parent=None):
+        """ Instantiates the triangle object.
+
+        Arguments:
+            idx: our index in the `triangulation.tris` array.
+            vertex_ids: vertex id's in the `triangulation.verts` array.
+            parent: reference to the parent of this triangle.
+            """
+        self.idx = idx
+        self.vertex_ids = vertex_ids
+        self.parent = parent
+        self.children = []  # References to the children of this triangle.
+        # Indices of my neighbours, ordered by the edge opposite vertex i.
+        self.neighbours = [None, None, None]
+
+        if self.parent:
+            self.area = self.parent.area / 2
+
+    def newest_vertex_id(self):
+        """ Returns the newest vertex, i.e., vertex 0. """
+        return self.vertex_ids[0]
+
+    def edge(self, i):
+        assert 0 <= i <= 2
+        return [self.vertex_ids[(i + 1) % 3], self.vertex_ids[(i + 2) % 3]]
+
+    def reversed_edge(self, i):
+        assert 0 <= i <= 2
+        return [self.vertex_ids[(i + 2) % 3], self.vertex_ids[(i + 1) % 3]]
+
+    def is_leaf(self):
+        return not len(self.children)
+
+
+class Triangulation(object):
+    def __init__(self, verts, tris):
+        """ Instantiates the triangulation given the (vertices, triangles).
+
+        Arguments:
+            verts: Vx2 matrix of floats: the coordinates of the vertices.
+            tris: Tx3 matrix of integers: the indices inside the verts array.
+        """
+        self.verts = [
+            Vertex(*vert, on_domain_boundary=False) for vert in verts
+        ]
[…]

### 2019-06-05 Latex/space-fem.tex
rename space-fem files

\documentclass[11pt,a4paper]{amsart}

\usepackage{amsmath, amssymb}
\usepackage{dsfont}
\usepackage[dutch]{babel}
\usepackage{geometry}
\usepackage{commath}
\usepackage{url}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{todonotes}
%\setlength{\parindent}{0pt}
\geometry{%
  includeheadfoot,
  margin=2cm
}

\newtheorem*{problem}{Problem}
\newtheorem*{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem*{answer}{Answer}

\newcommand{\bbone}{\mathds{1}}
\DeclareMathOperator*{\diam}{diam}
\DeclareMathOperator*{\spann}{span}
\DeclareMathOperator*{\init}{init}
\DeclareMathOperator*{\ext}{ext}
\newcommand{\udlrarrow}{\mathrlap{\,{\scriptscriptstyle \updownarrow}}{\scriptscriptstyle \leftrightarrow}}

\input{/Users/janwesterdiep/Dropbox/Commands.tex}

\begin{document}

\title{Eerste implementatie blauwe tekstje, \today}
\author{Jan Westerdiep}
\maketitle

Gegeven is een initial matching conforming triangulatie $\T_0$. We gaan hier
vertices in toevoegen door driehoeken te verfijnen volgens nvb. We willen, gegeven
een lokaal verfijnde triangulatie $\T \geq \T_0$, \emph{snel} een matvec
$a_2(\Psi_{\text{hb}}, \Psi_{\text{hb}}) \vec v$ uit kunnen rekenen voor
$\Psi_{\text{hb}}$ de hierarchische basis wrt vertices van $\T$. Dit kan in
$\mathcal O(\# \T)$ operaties, waarbij $\# \T$ het aantal vertices in $\T$ is.
Hiertoe zullen we gebruik maken van de single-scale basis $\Phi_{\text{ss}}$ op
$\T$.

\section*{De \texttt{Vertex} klasse}
We hebben een klasse \texttt{Vertex}, met attributen
\begin{description}
  \item[x, y] de $x$- en $y$-coordinaten van de vertex;
  \item[on\_bdr] of de vertex op de domain boundary ligt.
\end{description}

\section*{De \texttt{Triangle} klasse}
We hebben een klasse \texttt{Triangle}, met attributen
\begin{description}
  \item[idx] de index in de \texttt{Triangulation.tris}-array;
  \item[vertex\_ids] de indices van de drie hoekpunten in de \texttt{Triangulation.verts}-array;
  \item[parent] een reference/pointer naar de parent van deze driehoek (mogelijk \texttt{None});
  \item[children] twee pointers naar de kinderen van deze driehoek (mogelijk \texttt{None});
  \item[nbrs] een 3-tuple van pointers naar onze buren. Nbr $0$ zit \emph{tegenover} vertex 0, etc;
  \item[area] de area van deze driehoek.
\end{description}
Deze klasse heeft een paar convenience-methodes:
\begin{description}
  \item[edge(i)] returned \texttt{[self.vertex\_ids[(i + 1) \% 3], self.vertex\_ids[(i + 2) \% 3]]};
  \item[is\_leaf()] returned \texttt{not len(self.children)}.
\end{description}

\section*{De \texttt{Triangulation} klasse}
Als laatste de klasse \texttt{Triangulation}, met attributen
\begin{description}
  \item[verts] een array van \texttt{Vertex}-objecten;
  \item[tris] een array van \texttt{Triangle}-objecten;
  \item[root\_ids] een \texttt{set} met de indices van driehoeken met generatie 0;
  \item[history] array 2-tupels \texttt{(new\_vertex\_id, bisected\_triangle)} welke we zo omschrijven.
\end{description}
Deze klasse heeft twee methodes voor het initialiseren en refinen:
\begin{description}
  \item[init(verts, tris)] initialiseert $\T$ en de \texttt{Triangle}- en
    \texttt{Vertex}-attributen (\texttt{nbrs} resp.~ \texttt{on\_bdr});
  \item[refine(tri)] maakt een conforming refinement waarin \texttt{tri} gerefined is. Psuedocode:
\begin{verbatim}
  als tri geen leaf is: return # deze driehoek is al ge-refined.
  nbr = tri.nbrs[0]
  if nbr == None:  # we zitten aan de rand van het domein.
    self.bisect(tri)
  else if nbr.edge(0) tegenover tri.edge(0):  # shared refinement edge.
    self.bisect_pair(tri, nbr)
  else:
    self.refine(nbr)
    self.bisect_pair(tri, het kind van nbr grenzend aan tri.edge(0))
\end{verbatim}
\end{description}
De \texttt{refine}-methode maakt gebruik van de volgende \emph{private} methodes.
Wij zorgen dat deze slechts aangeroepen worden wanneer het mag (om conformiteit van de triangulatie te bewaren).
\begin{description}
  \item[bisect(tri, new\_vertex\_id=None)] doet NVB van de gegeven driehoek. Pseudocode:
\begin{verbatim}
  als new_vertex_id == None:
    bouw object new_vertex en plaats in de self.verts array
    plaats (new_vertex_id, tri.idx) achteraan in de self.history array
  bouw de kinderen C1, C2 van tri en plaats in de self.tris array
  C1.nbrs = [tri.nbrs[2], None, C2]; C2.nbrs = [tri.nbrs[1], C1, None]
  vervang tri door C1 in tri.nbrs[2], door C2 in tri.nbrs[1]
  return C1, C2
\end{verbatim}
  \item[bisect\_pair(T1, T2)] bisect \texttt{T1} en \texttt{T2}, en zet alle neighbour-info goed.
\end{description}

Met deze basis kunnen we gemakkelijk twee methodes bouwen:
\begin{description}
  \item[applyT(v)] pakt een vector $\vec v \in \R^{\# \T}$ coefficienten en geeft de vector $\vec w \in \R^{\# \T}$ zdd $\Psi_{\text{hb}}^\top \vec v = \Phi_{\text{ss}}^\top \vec w$,
    i.e.,~\texttt{applyT(v)} applied de hierarchic-to-singlescale operator.
    Pseudocode:
\begin{verbatim}
  w = v
  for (vi, Ti) in self.history:
    godfathers = self.tris[Ti].edge(0)
    w[vi] += w[godfathers[0]]/2 + w[godfathers[1]]/2
  return w
\end{verbatim}
  \item[applyT\_transpose(v)] applies de getransponeerde, met pseudocode
\begin{verbatim}
  w = v
  for (vi, Ti) in reversed(self.history):  # NB: in reversed order
    godfathers = self.tris[Ti].edge(0)
    w[godfathers[0]] += w[vi]/2
    w[godfathers[1]] += w[vi]/2
  return w
\end{verbatim}
  \item[apply\_HB\_*(v)] hebben voor $* \in \set{\mathtt{mass}, \mathtt{stiffness}}$ dan een simpele
    implementatie:
\begin{verbatim}
  return self.applyT_transpose(self.apply_SS_*(self.applyT(v)))
\end{verbatim}
\end{description}
Daarvoor is het wel zaak om de mass- en stiffness matrices snel te applyen. Dit kan door
\begin{description}
  \item[apply\_SS\_mass(v)] gaat in $\mathcal O(\# \mathtt{self.tris}) \eqsim \mathcal O(\# \T)$ operaties volgens
\begin{verbatim}
  elt_mass = 1/12 * [[2, 1, 1], [1, 2, 1], [1, 1, 2]]
  w = [0, ..., 0]
  for tri in self.tris:
    if not tri.is_leaf(): continue
    N = tri.vertex_ids
    for i in [0,1,2] en j in [0,1,2]:
      w[N[j]] += elt_mass[i,j] * tri.area * v[N[i]]
  return w
\end{verbatim}
  \item[apply\_SS\_stiffness(v)] gaat in dezelfde complexiteit. Ik vond dit in een paper:
\begin{verbatim}
  w = [0, ..., 0]
  for tri in self.tris:
    if not tri.is_leaf(): continue
    N = tri.vertex_ids
    V = [self.verts[ni] for ni in N]
    D = [[V[2].x - V[1].x, V[0].x - V[2].x, V[1].x - V[0].x],
         [V[2].y - V[1].y, V[0].y - V[2].y, V[1].y - V[0].y]]
    elt_stiff = matmat(D.T, D) / (4 * tri.area)
    for i in [0,1,2] en j in [0,1,2]:
      w[N[j]] += elt_stiff[i,j] * v[N[i]]
  return w
\end{verbatim}
\end{description}

\section*{Test hierarchic-to-singlescale implementatie}
Ik kan niet zo 1-2-3 echt netjes laten zien dat wat ik doe, het juiste is. Maar
het kwam er wel heel snel uitrollen toen ik op papier een beetje voorbeelden uit
ging werken. De methode \texttt{applyT()} is geen recursie, en stom genoeg kon ik
ook niet helemaal zien hoe je hem als recursie kon opschrijven.

\subsection*{Wat obvious tests}
Ik heb een paar obvious dingen getest, en die komen er allemaal positief uit:
\begin{itemize}
  \item is \texttt{applyT(v + a*w) == applyT(v) + a applyT(w)} voor een paar willekeurige $\vec v, a, \vec w$?
  \item is \texttt{< applyT(v), z > == <v, applyT\_transpose(z) >} voor een paar willekeurige $\vec v, \vec z$?
  \item is \texttt{vertex.on\_bdr} correct voor willekeurige vertices op een verfijnde square?
\end{itemize}

\subsection*{Een FE-oplossing}
De laatste test was om de vergelijking $- \triangle u = \bbone, u = 0$ op te lossen op $[-1,1]^2$.
In \url{http://people.inf.ethz.ch/arbenz/FEM17/pdfs/0-19-852868-X.pdf} staat
de oplossing uitgeschreven als oneindige som (welke we tot eoa precisie uit
kunnen rekenen en op een aantal punten kunnen evalueren). Hiervoor hebben we
inderdaad alle onderdelen van onze implementatie nodig.

\subsubsection*{De right-hand side} Als eerste: we weten dat ons rechterlid voldoet aan
\[
  \langle \Psi_{\text{hb}}, \bbone\rangle = T^\top \langle \Phi_{\text{ss}}, \bbone\rangle =
  \mathtt{applyT\_transpose(apply\_SS\_mass}(\vec 1))
\]
waar $\vec 1$ een vector met allemaal enen is.

\subsubsection*{Een kleine shortcut voor nu}
Merk op dat de methodes die we tot nu toe hebben gebouwd, allemaal werken
\emph{op elke vertex}, en niet enkel op vertices in de domain interior. De
stiffness matrix die we op deze manier hebben gebouwd, is singulier. Ik heb er
voor nu voor gekozen om de boundary condities op te leggen door na elke call naar
\texttt{apply\_HB\_stiffness()}, alle waarden wrt rand-vertices op nul te zetten.
Dit heb ik ook gedaan voor de right-hand side.

\subsubsection*{De solver}
Ik heb de ingebouwde CG-solver van \texttt{scipy.sparse.linalg} gebruikt, welke
geen matrix nodig heeft, slechts een functie die een matvec doet.

\subsubsection*{De triangulatie}
We starten met een standaard initial triangulatie met 2 driehoeken. Vervolgens
doen we 9 uniform refinements, dwz 8 keer een lijst leaves genereren en
\texttt{refine(leaf)} aanroepen voor elke leaf. Omdat we op elke iteratie een
hierarchische basis hebben, kunnen we na elke uniform refinement een \texttt{rhs}-variabele
genereren (zoals hierboven) en testen dat de eerste zoveel waardes overeen komen
met de \texttt{rhs} uit de vorige iteratie.

\subsubsection*{De oplossing}
We krijgen een oplossing in hierarchical basis-coefficienten. Om deze oplossing
te testen tegen de uitgerekende reference solution, zullen we \texttt{sol\_SS = applyT(sol)}
gebruiken. We testen:
\begin{itemize}
  \item $|\mathtt{sol\_SS[4]} - u(0,0)| < 10^{-3}$ waarbij volgens Mathematica, $u(0,0) = 0.2946854131260553$;
  \item $|\mathtt{sol\_SS[i]} - u(\tfrac{1}{2},\tfrac{1}{2})| < 10^{-3}$ voor $i \in \set{9, 10, 11, 12}$ waarbij $u(\tfrac{1}{2},\tfrac{1}{2}) = 0.181145$;
\end{itemize}
Op deze manier ben ik er vrij van verzekerd dat deze methode een aardige oplossing
vindt. Een plot ziet er ook goed uit (de assen worden wat verschillend geplot maar maxima komen overeen).
\begin{figure}[b]
  \centering
  \includegraphics[width=0.3\linewidth]{img/paper-sol}
  \includegraphics[width=0.3\linewidth]{img/fe-sol}
  \caption{Links: de oplossing uit het paper. Rechts: een plot van \texttt{sol\_SS}.}
\end{figure}
\clearpage

\subsection*{Een gekkigheidje}
We lopen niet level-voor-level door de vertices van de triangulatie. Zie de figuur
voor een illustratie. In de eerste figure zien we de progressie van de triangulaties
met de vertices genummerd. In de tweede figuur zie je de bijbehorende family-tree.
Een level-voor-level traversal zou gaan als
\[
  0 \to 1 \to 2 \to 3 \to 4 \to 5 \to 6 \to 8 \to 7
\]
whereas de \texttt{history}-array in ons geval een traversal geeft als
\[
  4 \to 5 \to 6 \to 7 \to 8.
\]
(Duidelijk worden in dit geval de roots van de family tree ook niet ge-traversed.)
Wel zijn beide traversals een \emph{topological ordering}, wat betekent dat een
kind pas aan bod komt wanneer beide parents al zijn geweest.

\begin{figure}[h!]
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below left] at (0,0){0};
    \node[below right] at (4,0){1};
    \node[above right] at (4,4){2};
    \node[above left] at (0,4){3};
    \node[below] at (2,-1){$\T_0$};
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below] at (2,-1){$\T_1$};

    \draw (0,0) -- (4,4); \node[right] at (2,2){4};
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below] at (2,-1){$\T_2$};

    \draw (0,0) -- (4,4);
    \draw (2,2) -- (2,4); \node[above] at (2,4){5};
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below] at (2,-1){$\T_3$};

    \draw (0,0) -- (4,4);
    \draw (2,2) -- (2,4);
    \draw (2,2) -- (0,2); \node[left] at (0,2){6};
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below] at (2,-1){$\T_4$};

    \draw (0,0) -- (4,4);
    \draw (2,2) -- (2,4);
    \draw (2,2) -- (0,2);
    \draw (0,2) -- (2,4); \node at (1,3){7};
  \end{tikzpicture}
  \hfill
  \begin{tikzpicture}[scale=0.4]
    \draw (0,0) -- (4,0) -- (4,4) -- (0,4) -- (0,0); \draw (4,0) -- (0,4);
    \node[below] at (2,-1){$\T_5$};

    \draw (0,0) -- (4,4);
    \draw (2,2) -- (2,4);
    \draw (2,2) -- (0,2);
    \draw (0,2) -- (2,4);
    \draw (2,2) -- (2,0); \node[below] at (2,0){8};
  \end{tikzpicture}
  \caption{Een voorbeeldprogressie. We schrijven enkel de nieuwe vertex op per iteratie.}
\end{figure}
\begin{figure}
\end{figure}

\subsection*{Visuele check}
Als we de methode \texttt{applyT()} toepassen op de eenheidsvector $\vec v = e_i$
horende bij een hierarchische basisfunctie $\psi_i$, dan krijgen we de coefficienten
van $\psi_i$ in termen van de nodale/singlescale hoedfuncties $\Phi_{\text{ss}}$. In de derde figuur
zien we het resultaat, namelijk dat de code hierarchische basisfuncties inderdaad
terugvindt.
\begin{figure}
  \begin{minipage}{0.4\linewidth}
    \includegraphics[width=\linewidth]{img/vertex-tree.png}
  \end{minipage}
  \begin{minipage}{0.55\linewidth}
    \includegraphics[width=\linewidth]{img/h0.png}\\
    \vspace{-1em}
    \includegraphics[width=\linewidth]{img/h2.png}\\
    \vspace{-2em}
  \end{minipage}
  \caption{Links: De bijbehorende vertex family-tree. Rechts: De functies $\psi_i$ en $\phi_i$ voor $i=0,2$---gebouwd dmv $\T_5$\texttt{.applyT(e\_i)}.}
\end{figure}

\end{document}

### 2019-07-30 Python/time/index_set.py
speed improvements

@@ -1,6 +1,8 @@
 import bisect
 import collections.abc
 
+from functools import lru_cache
+
 
 class IndexSet(collections.abc.Set):
     def __len__(self):
@@ -31,6 +33,7 @@ class SingleLevelIndexSet(IndexSet):
         for index in indices:
             assert index[0] == labda[0]
         self.indices = indices
+        self._neighbours = {}
 
         # TODO: if we assume that `indices` is given to us in lexicographical
         # ordering, we can save the list as `self.sorted` -- this also allows
@@ -47,13 +50,16 @@ class SingleLevelIndexSet(IndexSet):
     def neighbours(self, labda):
         """ Get the neighbours of this singlescale index.
         
-        Current complexity: once O(N log N) and later O(log N).
+        Current complexity: once O(N log N) per IndexSet, then
+                            once O(log N) per item, then O(1).
         Goal complexity: O(1).
-        TODO: this will become the bottleneck for N = 30K.
         """
-        i = bisect.bisect_left(self.asarray(), labda)
-        return (self.sorted[i - 1] if 0 < i < len(self.sorted) else None,
+        if not labda in self._neighbours:
+            i = bisect.bisect_left(self.asarray(), labda)
+            self._neighbours[labda] = (
+                self.sorted[i - 1] if 0 < i < len(self.sorted) else None,
                 self.sorted[i + 1] if i < len(self.sorted) - 1 else None)
+        return self._neighbours[labda]
 
     def __repr__(self):
         return r"SLIS(%s)" % self.indices

### 2019-07-30 Python/time/interval.py

@@ -1,18 +1,13 @@
-class Interval(object):
-    """ Represents an open or closed interval (a,b). """
+import collections.abc
 
-    def __init__(self, a, b=None):
-        if not b:
-            # Copy constructor.
-            assert isinstance(a, Interval)
-            self.a = a.a
-            self.b = a.b
-            self.mid = a.mid
-        else:
-            assert a < b
-            self.a = a
-            self.b = b
-            self.mid = (a + b) / 2
+
+class Interval(collections.namedtuple('Interval', 'a b')):
+    """ Represents an open or closed interval (a,b). """
+    __slots__ = ()
+
+    @property
+    def mid(self):
+        return (self.a + self.b) / 2
 
     def intersects(self, interval, closed=False):
         if not closed:
@@ -41,13 +36,13 @@ class IntervalSet(object):
         sorted_intervals = sorted(intervals, key=lambda interval: interval.a)
         stack = []
         if len(sorted_intervals):
-            stack.append(Interval(sorted_intervals[0]))
+            stack.append(Interval(*sorted_intervals[0]))
             for interval in sorted_intervals:
                 if interval.intersects(stack[-1], closed=True):
-                    stack[-1].b = max(stack[-1].b, interval.b)
+                    stack[-1] = Interval(stack[-1].a,
+                                         max(stack[-1].b, interval.b))
                 else:
-                    stack.append(Interval(interval))
-                    stack[-1].mid = (stack[-1].a + stack[-1].b) / 2
+                    stack.append(Interval(*interval))
             self.intervals = stack
         else:
             self.intervals = []

### 2019-07-30 Python/time/three_point_basis.py

@@ -30,7 +30,7 @@ def ms2ss(l, labda):
     return (l, 2**(l - labda[0]) * (2 * labda[1] + 1))
 
 
-@lru_cache(maxsize=4096)
+@lru_cache(maxsize=2**20)
 def ss2ms(labda):
     """ Multiscale index of a singlescale index labda.
     
@@ -79,6 +79,8 @@ class ThreePointBasis(Basis):
             }) for level in range(0, self.indices.maximum_level + 1)
         ]
         assert len(self.ss_indices) == self.indices.maximum_level + 1
+        self._scaling_support = {}
+        self._wavelet_support = {}
 
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
@@ -177,34 +179,41 @@ class ThreePointBasis(Basis):
         return LinearOperator(row)
 
     def scaling_support(self, labda):
-        l, n = labda
+        if labda not in self._scaling_support:
+            l, n = labda
 
-        left, right = self.scaling_indices_on_level(l).neighbours(labda)
+            left, right = self.scaling_indices_on_level(l).neighbours(labda)
 
-        if n == 0: return Interval(0, position_ss(right))
-        elif n == 2**l: return Interval(position_ss(left), 1)
+            if n == 0: return Interval(0, position_ss(right))
+            elif n == 2**l: return Interval(position_ss(left), 1)
 
-        return Interval(position_ss(left), position_ss(right))
+            self._scaling_support[labda] = Interval(position_ss(left),
+                                                    position_ss(right))
+        return self._scaling_support[labda]
 
     def wavelet_support(self, labda):
-        l, n = labda
-        if l == 0: return Interval(0, 1)
+        if labda not in self._wavelet_support:
+            l, n = labda
+            if l == 0: return Interval(0, 1)
 
-        left, right = self.scaling_indices_on_level(l).neighbours(
-            ms2ss(l, labda))
-        if n == 0:
-            left_side = 0
-        else:
-            left_left, _ = self.scaling_indices_on_level(l).neighbours(left)
-            left_side = position_ss(left_left)
+            left, right = self.scaling_indices_on_level(l).neighbours(
+                ms2ss(l, labda))
+            if n == 0:
+                left_side = 0
+            else:
+                left_left, _ = self.scaling_indices_on_level(l).neighbours(
+                    left)
+                left_side = position_ss(left_left)
 
-        if n == 2**(l - 1) - 1:
-            right_side = 1
-        else:
[…]

### 2019-07-30 Python/time/readme.tex
changed 1000 things lol

\documentclass[11pt,a4paper]{amsart}

\usepackage{amssymb}
\usepackage{geometry}
\usepackage{commath}
\usepackage{dsfont}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{todonotes}
%\setlength{\parindent}{0pt}

\newtheorem*{problem}{Problem}
\theoremstyle{definition}
\newtheorem*{answer}{Answer}
\DeclareMathOperator{\supp}{supp}
\DeclareMathOperator{\ran}{ran}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\spann}{span}
\DeclareMathOperator{\col}{col}
\DeclareMathOperator{\row}{row}
\DeclareMathOperator{\Id}{Id}
\renewcommand{\span}{\spann}
\newcommand{\bbone}{\mathds{1}}
\begin{document}

\title{A first implementation of Section 3 of ``Bla Parabolic''}
\author{Jan Westerdiep}
\maketitle

I've spent the past two weeks on a (not yet) linear time implementation of \S 3
from ``Bla Parabolic''. Currently it can only apply multiscale (mass) operator
in the time direction; I felt it was better to have a good understanding of one
aspect than to produce a half-baked implementation for the tensor-product case.

\subsection*{A little bit on notation} I am somewhat hesitant to call these
`mass matrices' \emph{matrices} because this implies that the rows and columns
are ordered, whereas we have mere \emph{sets} of indices (which are not necessarily
ordered). I will try to call everything \emph{linear operators} from one
\emph{vector space} to another, instead of matrices working on vectors.

The vector spaces that we will be working with are often sequence spaces
$\ell_2(\Lambda)$ for some finite index set $\Lambda$. If we were to order this
index set in, say, lexicographic ordering, then $\ell_2(\Lambda)$ corresponds
directly with $\mathbb{R}^{\# \Lambda}$.

\section{The Applicator}
The meat of the algorithm is in \texttt{applicator.py}. This \texttt{Applicator}
class has a constructor which takes, at minimum, a \texttt{Basis} object
$B = \Phi \cup \Psi$ and a singlescale \texttt{LinearOperator} object
$\mathcal A: \span \Phi \to (\span \Phi)'$. Both $\Phi$ and $\Psi$ are indexed by
\texttt{IndexSets}, with $\Phi$ indexed by a list of \texttt{SingleLevelIndexSet}s
$\Delta_\ell$ and $\Psi$ by a \texttt{MultiScaleIndexSet} (or \emph{tree})
which we call $\Lambda$. Remember that by assumption $\span \Phi = \span \Psi$.
It is also possible to construct the \texttt{Applicator} with different in- and
out-bases when the domain and range of $\mathcal A$ do not coincide.

It has three public methods: \texttt{apply()}, \texttt{apply\_upp()},
and \texttt{apply\_low()}. These methods in turn invoke recursive methods that
traverse $\Lambda$, starting from level $\ell=1$ until its maximum level, at
each level applying $\mathcal A$ to a subset of the singlescale indices
$\Delta_\ell$ on level $\ell$. In doing this, it needs the \texttt{SLIS}s
$\check \Pi_B^\ell := \check \Pi_B$ (named \texttt{Pi\_B\_out} in the code
corresponding with the out-basis), $\Pi_B^\ell = \Pi_B$ named
\texttt{Pi\_B\_in}, $\underline \Pi^\ell = \Pi^{\ell + 1}$ named \texttt{Pi\_bar\_in},
and $\underline{\check \Pi}^\ell = \check \Pi^{\ell+1}$. It is important to note 
that even when $\ran \mathcal A = \dom \mathcal A$, in general $\check \Pi_B \not= \Pi_B$
and hence $\underline{\check \Pi} \not= \underline \Pi$.

Computing these subsets $\check \Pi_B, \Pi_B, \underline{\check \Pi}, \underline \Pi$
in linear time is not an easy task. I was able to produce them in suboptimal complexity;
see the code. Maybe Rob will have an idea on how to tackle these.

\subsection{Constructing $\Pi_B$}
I am not \emph{at all} sure what $\Pi_B$ is supposed to do, but I feel like it's
the set of indices that are supposed to be lifted to level $\ell+1$. I computed 
\[
  \Pi_B := \set{\lambda \in \Pi: |\supp \phi_\lambda \cap
      (\cup_{\mu \in \check \Lambda_\ell} \supp \check \psi_\mu
       \cup_{\gamma \in \check \Pi_B} \supp \check \phi_\gamma)| > 0}
\]
by collecting all these supports $[a,b]$, sorting them\footnote{Blehhhh; see code
for a justification and a possible solution; still, this doesn't reduce the
operation to optimal complexity.} in order of increasing $a$, and merging them
pairwise to a minimum-size \texttt{IntervalSet} that covers the same intervals
as the original supports. We can loop over all $\lambda \in \Pi$ and check if
this \texttt{IntervalSet} \texttt{intersects()} its support. This is possibly
$\mathcal O[\# \Pi_B (\# \Lambda_\ell + \# \check \Pi_B) \log(\# \Lambda_\ell + \# \check \Pi_B)]$
operations where we would like $\mathcal O(\# \Pi_B$).  In practice though,
$\#\mathtt{IntervalSet} \ll \# \Lambda_\ell + \# \check \Pi_B$, so this helps
\emph{a lot} for faster runtimes.

On the other
hand $\Pi_A := \Pi \setminus \Pi_B$ is the set of indices we will want restrict
the result of applying $\mathcal A$ to, and these will not be lifted to the next
level. In Kestlers C++-code, (the equivalent of) $\Pi_A$ rather than $\Pi_B$ is
computed. This may be a way of doing it in linear time?

\subsection{Constructing $\underline \Pi$}
The construction of $\underline \Pi$ is very similar to the previous method, the
big difference being that we want to collect those indices $\lambda \in \Delta_\ell$
that are completely covered by the \texttt{IntervalSet}.

\subsection{Applying $P^\ell$ and $Q^\ell$, and restriction of vectors to subsets}
Throughout the algorithm, it is necessary to restrict a vector to a subset of
indices, e.g.
\[
  \underline{\vec d} := (P^\ell \vec d|_{\Pi_B} + Q^\ell \vec c|_{\Lambda_\ell})|_{\underline \Pi}.
\]
Instead of manually setting vector indices to zero, computing the result of the
linear operator on the full set $\Delta_{\ell + 1}$ and then again setting vector
indices to zero, instead we compute the result of the \texttt{LinearOperator}s
$P^\ell$ and $Q^\ell$ only on those indices in $\underline \Pi$, and in the inner
products of rows of the \texttt{LinearOperator} with the \texttt{IndexedVector},
we only consider those indices in $\Pi_B$ resp.~$\Lambda_\ell$, thus effectively
`zero-ing out' the other indices. We know that every
row of $P^\ell$ and $Q^\ell$ have $\lesssim 1$ number of nonzeros, so this inner
product can be computed in $O(1)$ operations.\footnote{One actually has to take
care in not looping over elements in $\Pi_B$ to see if the corresponding cell in
$P^\ell$ is nonzero, but rather loop over nonzeros in $P^\ell$ to see if they
are in $\Pi_B$. This mistake destroyed linearity at first, and took me quite some
time to fix.}

The linear operators $P^\ell$ and $Q^\ell$ are the unique ones such that
\[
  \Phi_{\ell-1}^\top = \Phi_\ell^\top P^\ell, \quad \Psi_\ell^\top = \Phi_\ell^\top Q^\ell.
\]
These exist because $\span \Phi_{\ell-1} \subset \span \Phi_\ell$ and
$\span \Psi_\ell \subset \span \Phi_\ell$ by assumption. For some $\phi_\lambda \in \Phi_{\ell-1}$,
we see that $\Phi_\ell^\top p_{\ell,\lambda}$: the linear operator $P^\ell$ gives us the
coefficients necessary to build a scaling function on the previous level using
scaling functions on this level: a little more on this later. Similar equalities
hold for the $Q^\ell$.

We also have to apply the transpose of $P^\ell$ and $Q^\ell$; building these
efficiently proved to be quite the obstacle for the more esoteric 3-point basis.

\subsection{Summation of vectors}
There is a number of places in the \texttt{Applicator} where vectors are built
by stacking multiple vectors indexed by disjoint sets (e.g.~$\vec e$ and $\vec f$).
We can be lazy and just write these as the sum of the two vectors.

\section{Bases}
In this proof-of-concept application, we will work with three bases: the Haar
basis, the orthonormal discontinuous piecewise linear basis, and the three-point
hierarchical basis. Every basis contains a \emph{singlescale} or \emph{scaling
basis} $\Phi$ indexed on every level by $\Delta_\ell$, and a \emph{multiscale}
or \emph{wavelet basis} $\Psi$ indexed by a tree $\Lambda$. Every index $lambda$
is a tuple $(\ell, n)$; its level and its \emph{offset}, even though in the
multiwavelet/multiscaling case, the offset also determines \emph{which} mother
scaling function $\phi_\lambda$ descends from.

Every \texttt{Basis} object is constructed by passing it the \texttt{MultiScaleIndexSet}
$\Lambda$. Upon construction, it exposes as properties the \texttt{LinearOperator}s
$P := (P^\ell)_{\ell}$ and $Q = (Q^\ell)_\ell$, as well as one or more singlescale
\texttt{LinearOperator}s like the mass and \emph{damping} operators
\[
  (M\Phi_\ell)(\Phi_\ell) := \left[\int_0^1 \phi_\lambda \phi_\mu \dif t \right]_{\lambda,\mu \in \Delta_\ell}, \quad
  (C\Phi_\ell)(\Phi_\ell) := \left[\int_0^1 \phi_\lambda' \phi_\mu \dif t \right]_{\lambda,\mu \in \Delta_\ell}.
\]

It has methods to compute the support of scaling- and wavelet functions (with
\texttt{Interval} return types), as well as the \emph{wavelet neighbourhood}
$S(\mu) \supset \supp \psi_\mu$ (which, for all our cases, just equals the support).

\subsection{Singlescale matrices}
I basically just computed the singlescale matrices using pen and paper or
Mathematica.

\subsection{$P^\ell$ and $Q^\ell$}
In general, the following equalities define $P^\ell$ and $Q^\ell$:
\[
\begin{cases}
  \phi_\lambda := \langle\Phi_\ell, \col_\lambda(P^\ell)\rangle_{\ell_2(\Delta_\ell)} = \sum_{\mu \in \Delta_\ell} P^\ell_{\mu,\lambda} \phi_\mu & \lambda \in \Delta_{\ell-1},\\
  \psi_\lambda := \langle \Phi_{\ell}, \col_\lambda(Q^\ell)\rangle_{\ell_2(\Delta_\ell)} = \sum_{\mu \in \Delta_\ell} Q^\ell_{\mu,\lambda} \phi_\mu & \lambda \in \Lambda_\ell.
\end{cases}
\]
This means that $P^\ell$ tells us exactly how to write a scaling function on the
previous level in terms of scaling functions on the current level, and that $Q^\ell$
tells us how to write wavelet functions on this level in terms of scaling functions
on this level.

The Haar basis has mother scaling resp.~wavelet functions
\[
  \phi(x) := \bbone_{[0,1)}(x), \quad \psi(x) := \bbone_{[0,\frac{1}{2})}(x) - \bbone_{[\frac{1}{2},1)}(x).
\]
At index $\lambda = (\ell, n)$, we have
\[
  \phi_\lambda(x) = \phi(2^\ell x - n), \quad \psi_\lambda(x) = \psi(2^{\ell-1}x - n).
\]
This $2^\ell$ versus $2^{\ell-1}$ is to balance the assumption that on level 0,
$\Psi_0 = \Phi_0$ (so that the mother wavelet function actually lives on level 1,
whereas the mother scaling function lives on level 0).

\subsubsection*{Haar basis}
The result is that for the Haar basis, both $P$ and $Q$ have very simple structure:
\[
  \phi_\lambda = \phi_{(\ell+1, 2n)} + \phi_{(\ell+1, 2n+1)}, \quad
  \psi_\lambda = \phi_{(\ell, 2n)} - \phi_{(\ell, 2n+1)}.
\]
In other words, every column of $P^\ell$ and $Q^\ell$ has two nonzeros. If we
were to order the indices in lexicographic ordering, then
\[
  P^{\ell} = \begin{bmatrix}1 \\ 1\end{bmatrix} \otimes \Id_{\ell}, \quad Q^{\ell} = \begin{bmatrix}1 \\ -1\end{bmatrix} \otimes \Id_{\ell}.
\]

\subsubsection*{Orthonormal basis}
The orthonormal discontinuous piecewise linear basis is a multiscaling/multiwavelet basis and has mother scaling functions
\[
  \phi_0(x) := \bbone_{[0,1)}(x), \quad \phi_1(x) := \sqrt{3} (2x-1) \bbone_{[0,1)}(x)
\]
with mother wavelet functions
\[
  \psi_0(x) := (1-6x) \bbone_{[0,\frac{1}{2})}(x) + (5-6x)\bbone_{[\frac{1}{2},1)}(x), \quad
  \psi_1(x) := \sqrt{3} (1-4x) \bbone_{[0,\frac{1}{2})}(x) + \sqrt{3}(4x-3)\bbone_{[\frac{1}{2},1)}(x)
\]
(see also the document by Rob for a more visual approach). We have
\[
  \phi_{(\ell, 2n)}(x) := \phi_0(2^\ell x - n), \quad \phi_{(\ell, 2n+1)}(x) := \phi_1(2^\ell x - n) \quad (\ell \geq 0, 0 \leq n < 2^{\ell+1})
\]
and
\[
  \psi_{(0, n)}(x) := \phi_n(x), \quad
  \psi_{(\ell, 2n)}(x) := 2^{(\ell-1)/2} \psi_0(2^{\ell-1} x - n), \quad
  \psi_{(\ell, 2n+1)}(x) := 2^{(\ell-1)/2} \psi_1(2^{\ell-1} x - n).
\]
This rescaling ensures that the multiscale mass operator is just the identity!
(Hence the name of the basis.) For this basis, the matrices $P^\ell$ and $Q^\ell$
are already a lot less obvious.\todo{somehow show in a figure how I derived them}

\subsubsection*{3-point basis}
The 3-point basis was introduced very tersely in the document, and I was not entirely
sure what Rob meant with the figures. I digged around and found a 1997 paper entitled
\emph{Experiments in 3D with a three-point hierarchical basis}, which seemed to
discuss the same basis.

First, let's consider the indices only. At level $\ell \geq 0$, a \emph{singlescale}
index can be any $\lambda = (\ell, n)$ with $0 \leq n \leq 2^\ell$, and
corresponds with a node at position $n/2^\ell$. A \emph{multiscale} index
$\lambda = (\ell, n)$ at level $\ell \geq 1$ has $0 \leq n \leq 2^{\ell-1}$ and
corresponds with a node at position $2^{-\ell} + 2^{1-\ell} n$. (Draw this out
for yourself on a piece of paper!) At level $\ell=0$, multiscale indices are
$(0,0)$ and $(0,1)$ at positions $0$ and $1$, respectively.

For $\ell \geq 2$, the \emph{parent} of a multiscale index $\lambda$ is the
unique multiscale index on level $\ell-1$ such that its position is adjacent to
$\lambda$ on the grid. The assumption that our multiscale index set $\Lambda$ is
a \emph{tree} in this setting coincides with the assumption that for any $\lambda$
with $\ell\geq 2$, its parent is also in $\Lambda$.

Let's take some $\Lambda$. Define $\Lambda_\ell := \set{\lambda \in \Lambda: |\lambda| = \ell}$ as the multiscale index sets on level $\ell$. By assumption,
$\Lambda_0 = \set{(0,0), (0,1)}$ (nodes at $0$ and $1$), and
$\Lambda_1 = \set{(1,0)}$ (single node at $\frac{1}{2}$). Then 
\[
  \begin{cases}
    \Lambda_2 \subset \set{(2,0), (2,1)} &\text{ nodes at } \tfrac{1}{4}, \tfrac{3}{4}, \\
    \Lambda_3 \subset \set{(3,0), (3,1), (3,2), (3,3)} &\text{ nodes at } \tfrac{1}{8}, \tfrac{3}{8}, \tfrac{5}{8}, \tfrac{7}{8}, \quad \text{etc.}
  \end{cases}
\]
Taking all multiscale indices until level $\ell$, we can define the set
$\Delta_\ell$ of singlescale indices that have the same node positions.
The idea is that the scaling functions on level $\ell$ are just hat functions
(rescaled with factor $2^{\ell/2}$) wrt the nodes of $\Delta_\ell$,\footnote{This
is different from what is sketched in figure 1 of the document.} and the
wavelet functions are linear combinations of (at most) three scaling functions.
\todo{show a couple of example $\Lambda$'s, $\Phi$'s and $\Psi$'s.}

\section{Tests}
I tried to test most essential parts of the code:
\begin{enumerate}
  \item Do the matrices $P^\ell$ and $Q^\ell$ work, i.e.~do we recover $\phi_\lambda$
    through $\langle \Phi_\ell, \col_\lambda(P^\ell) \rangle$?
  \item Are our multiscale mass matrices diagonal (Haar)/the identity (orthonormal)?
  \item Does the output of \texttt{apply()} correspond with values found through quadrature?
  \item Does the output of \texttt{apply()} equal \texttt{apply\_upp() + apply\_low()}?
\end{enumerate}
You can try running \texttt{pytest} in this folder, and it should pass all tests.

\section{Numerics}
\subsection{Bottlenecks}

\section{Towards tensor-product operators}

\end{document}

### 2019-07-30 Python/time/applicator.py

@@ -91,8 +91,7 @@ class Applicator(object):
         Necessary for computing Pi_B in apply/apply_upp/apply_low.
 
         Goal complexity: O(|Pi_B|).
-        Current complexity: O(|Pi_B| |Lambda_l|).
-        Current complexity: O(PiBin + LambdaL log[LabdaL]).
+        Current complexity: O(|PiBin| * |LambdaL| log[|LabdaL|]).
         TODO: make faster. it could be faster/easier to compute Pi_A?
 
         Arguments:
@@ -114,7 +113,7 @@ class Applicator(object):
         """ Similar to previous method, only with extra `Pi_B_out`.
 
         Goal complexity: O(|Pi_B_in|).
-        Current complexity: O(PiBin + (PiBout + LambdaL) log[PiBout + LabdaL]).
+        Current complexity: O(PiBin * (PiBout + LambdaL) log[PiBout + LabdaL]).
         """
         ivs = IntervalSet(
             [self.basis_out.wavelet_nbrhood(mu) for mu in Lambda_l_out] +

### 2019-07-30 Python/time/basis_test.py

@@ -277,22 +277,30 @@ def test_singlescale_mass_quadrature():
 def print_3point_functions():
     x = np.linspace(0, 1, 1025)
     for basis in [
-            ThreePointBasis.uniform_basis(max_level=4),
-            ThreePointBasis.origin_refined_basis(max_level=4)
+            HaarBasis.uniform_basis(max_level=2),
+            HaarBasis.origin_refined_basis(max_level=2),
+            OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=2),
+            OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
+                max_level=2),
+            ThreePointBasis.uniform_basis(max_level=2),
+            ThreePointBasis.origin_refined_basis(max_level=2)
     ]:
-        for l in range(1, basis.indices.maximum_level):
-            for labda in basis.scaling_indices_on_level(l):
+        for labda in basis.scaling_indices_on_level(2):
+            if basis.__class__ == ThreePointBasis:
                 plt.plot(
-                    [mu[1] / 2**l for mu in basis.scaling_indices_on_level(l)],
-                    [0] * len(basis.scaling_indices_on_level(l)), 'ko')
-                plt.plot(x, basis.eval_scaling(labda, x), label=labda)
-            plt.legend()
-            plt.show()
+                    [mu[1] / 2**2 for mu in basis.scaling_indices_on_level(2)],
+                    [0] * len(basis.scaling_indices_on_level(2)), 'ko')
+            plt.plot(x, basis.eval_scaling(labda, x), label=labda)
+        plt.title("Scaling functions for %s" % basis.__class__.__name__)
+        plt.legend()
+        plt.show()
 
         for labda in basis.indices:
-            plt.plot([position_ms(mu) for mu in basis.indices],
-                     [0] * len(basis.indices), 'ko')
+            if basis.__class__ == ThreePointBasis:
+                plt.plot([position_ms(mu) for mu in basis.indices],
+                         [0] * len(basis.indices), 'ko')
             plt.plot(x, basis.eval_wavelet(labda, x), label=labda)
+        plt.title("Wavelet functions for %s" % basis.__class__.__name__)
         plt.legend()
         plt.show()
 

### 2019-07-30 Python/time/haar_basis.py

@@ -8,6 +8,16 @@ import numpy as np
 
 
 class HaarBasis(Basis):
+    def __init__(self, indices):
+        super().__init__(indices)
+        self.ss_indices = []
+        for l in range(0, self.indices.maximum_level + 1):
+            Lambda_l = self.indices.on_level(l)
+            self.ss_indices.append(
+                SingleLevelIndexSet({(l, 2 * n + i)
+                                     for (_, n) in Lambda_l
+                                     for i in range(2)}))
+
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
         return MultiscaleIndexSet({(0, 0)} | {(l, n)
@@ -83,5 +93,6 @@ class HaarBasis(Basis):
         return Interval(2**-l * n, 2**-l * (n + 1))
 
     def scaling_indices_on_level(self, l):
-        # TODO: this can be a much smaller set when the grid is non-uniform.
-        return SingleLevelIndexSet({(l, n) for n in range(2**l)})
+        if l >= len(self.ss_indices):
+            return SingleLevelIndexSet({})
+        return self.ss_indices[l]

### 2019-07-30 Python/time/interval.py

@@ -29,10 +29,23 @@ class Interval(collections.namedtuple('Interval', 'a b')):
 
 
 class IntervalSet(object):
-    """ Represents a collection of intervals with fast merging of overlap. """
+    """ Represents a collection of intervals with fast merging of overlap.
+    
+    """
 
     def __init__(self, intervals):
-        """ This runs in O(#intervals) time. """
+        """ This runs in O(N log N) time.
+        
+        Current complexity: O(N log N) because of a sort step
+        Goal complexity: O(N)
+        TODO: if we assume to receive a list of intervals that is already sorted
+        in order of increasing left-hand side, then this sort step is
+        unnecessary. If we assume to get a number of these interval-lists, where
+        every list itself is already sorted, we can do a merge-sort-like thing
+        to sort the entire list in linear time. On the other hand: sorting in
+        Python is highly optimized and it may actually be faster to just use
+        the sort option.
+        """
         sorted_intervals = sorted(intervals, key=lambda interval: interval.a)
         stack = []
         if len(sorted_intervals):

### 2019-07-30 Python/time/orthonormal_basis.py

@@ -17,6 +17,16 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
     with the first, odd with the second.
     """
 
+    def __init__(self, indices):
+        super().__init__(indices)
+        self.ss_indices = []
+        for l in range(0, self.indices.maximum_level + 1):
+            Lambda_l = self.indices.on_level(l)
+            self.ss_indices.append(
+                SingleLevelIndexSet({(l, 2 * n + i)
+                                     for (_, n) in Lambda_l
+                                     for i in range(2)}))
+
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
         return MultiscaleIndexSet({(0, 0), (0, 1)}
@@ -104,8 +114,9 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
         return Interval(2**-l * (n // 2), 2**-l * (n // 2 + 1))
 
     def scaling_indices_on_level(self, l):
-        # TODO: this can be a much smaller set when the grid is non-uniform.
-        return SingleLevelIndexSet({(l, n) for n in range(2 * 2**l)})
+        if l >= len(self.ss_indices):
+            return SingleLevelIndexSet({})
+        return self.ss_indices[l]
 
     @property
     def singlescale_mass(self):

### 2019-07-30 Python/time/three_point_basis.py

@@ -79,6 +79,8 @@ class ThreePointBasis(Basis):
             }) for level in range(0, self.indices.maximum_level + 1)
         ]
         assert len(self.ss_indices) == self.indices.maximum_level + 1
+
+        # Some memoizations.
         self._scaling_support = {}
         self._wavelet_support = {}
 

### 2019-07-30 Python/time/basis_test.py
broke tests trying to fix a TODO, lol

@@ -101,10 +101,6 @@ def test_basis_PQ():
                 try:
                     assert np.allclose(inner, basis.eval_scaling(mu, x))
                 except AssertionError:
-                    plt.plot([
-                        mu[1] / 2**l
-                        for mu in basis.scaling_indices_on_level(l)
-                    ], [0] * len(basis.scaling_indices_on_level(l)), 'ko')
                     plt.plot(x, inner, label=r'$(\Phi_{l-1}^T P_l)_\mu$')
                     plt.plot(x, basis.eval_scaling(mu, x), label=r"$\phi_\mu$")
                     plt.legend()
@@ -277,23 +273,25 @@ def test_singlescale_mass_quadrature():
 def print_3point_functions():
     x = np.linspace(0, 1, 1025)
     for basis in [
-            HaarBasis.uniform_basis(max_level=2),
-            HaarBasis.origin_refined_basis(max_level=2),
-            OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=2),
+            HaarBasis.uniform_basis(max_level=4),
+            HaarBasis.origin_refined_basis(max_level=4),
+            OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=4),
             OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
-                max_level=2),
-            ThreePointBasis.uniform_basis(max_level=2),
-            ThreePointBasis.origin_refined_basis(max_level=2)
+                max_level=4),
+            ThreePointBasis.uniform_basis(max_level=4),
+            ThreePointBasis.origin_refined_basis(max_level=4)
     ]:
-        for labda in basis.scaling_indices_on_level(2):
-            if basis.__class__ == ThreePointBasis:
-                plt.plot(
-                    [mu[1] / 2**2 for mu in basis.scaling_indices_on_level(2)],
-                    [0] * len(basis.scaling_indices_on_level(2)), 'ko')
-            plt.plot(x, basis.eval_scaling(labda, x), label=labda)
-        plt.title("Scaling functions for %s" % basis.__class__.__name__)
-        plt.legend()
-        plt.show()
+        for l in range(basis.indices.maximum_level + 1):
+            for labda in basis.scaling_indices_on_level(l):
+                if basis.__class__ == ThreePointBasis:
+                    plt.plot([
+                        mu[1] / 2**l
+                        for mu in basis.scaling_indices_on_level(l)
[…]

### 2019-07-30 Python/time/haar_basis.py

@@ -10,13 +10,6 @@ import numpy as np
 class HaarBasis(Basis):
     def __init__(self, indices):
         super().__init__(indices)
-        self.ss_indices = []
-        for l in range(0, self.indices.maximum_level + 1):
-            Lambda_l = self.indices.on_level(l)
-            self.ss_indices.append(
-                SingleLevelIndexSet({(l, 2 * n + i)
-                                     for (_, n) in Lambda_l
-                                     for i in range(2)}))
 
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
@@ -93,6 +86,5 @@ class HaarBasis(Basis):
         return Interval(2**-l * n, 2**-l * (n + 1))
 
     def scaling_indices_on_level(self, l):
-        if l >= len(self.ss_indices):
-            return SingleLevelIndexSet({})
-        return self.ss_indices[l]
+        # TODO: this can be a much smaller set when the grid is non-uniform.
+        return SingleLevelIndexSet({(l, n) for n in range(2**l)})

### 2019-07-30 Python/time/index_set.py

@@ -32,7 +32,10 @@ class SingleLevelIndexSet(IndexSet):
             break
         for index in indices:
             assert index[0] == labda[0]
-        self.indices = indices
+        if isinstance(indices, SingleLevelIndexSet):
+            self.indices = indices.indices
+        else:
+            self.indices = indices
         self._neighbours = {}
 
         # TODO: if we assume that `indices` is given to us in lexicographical

### 2019-07-30 Python/time/orthonormal_basis.py

@@ -19,13 +19,6 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
 
     def __init__(self, indices):
         super().__init__(indices)
-        self.ss_indices = []
-        for l in range(0, self.indices.maximum_level + 1):
-            Lambda_l = self.indices.on_level(l)
-            self.ss_indices.append(
-                SingleLevelIndexSet({(l, 2 * n + i)
-                                     for (_, n) in Lambda_l
-                                     for i in range(2)}))
 
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
@@ -114,9 +107,8 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
         return Interval(2**-l * (n // 2), 2**-l * (n // 2 + 1))
 
     def scaling_indices_on_level(self, l):
-        if l >= len(self.ss_indices):
-            return SingleLevelIndexSet({})
-        return self.ss_indices[l]
+        # TODO: this can be a much smaller set when the grid is non-uniform.
+        return SingleLevelIndexSet({(l, n) for n in range(2 * 2**l)})
 
     @property
     def singlescale_mass(self):

### 2019-07-31 Python/time/readme.tex
writing document

@@
 \usepackage{todonotes}
+\geometry{margin=2.5cm}
 %\setlength{\parindent}{0pt}
@@
 
+\subsubsection*{Haar basis}
 The Haar basis has mother scaling resp.~wavelet functions
@@
 
-\subsubsection*{Haar basis}
 The result is that for the Haar basis, both $P$ and $Q$ have very simple structure:
@@
 In other words, every column of $P^\ell$ and $Q^\ell$ has two nonzeros. If we
-were to order the indices in lexicographic ordering, then
+were to take all indices $\Lambda_\ell = (\ell, n)$ with $0 \leq n < 2^l$ and
+order them in lexicographic ordering, then
 \[
@@
 \[
-  \psi_0(x) := (1-6x) \bbone_{[0,\frac{1}{2})}(x) + (5-6x)\bbone_{[\frac{1}{2},1)}(x), \quad
+  \begin{cases}
+    \psi_0(x) := (1-6x) \bbone_{[0,\frac{1}{2})}(x) + (5-6x)\bbone_{[\frac{1}{2},1)}(x), \\
     \psi_1(x) := \sqrt{3} (1-4x) \bbone_{[0,\frac{1}{2})}(x) + \sqrt{3}(4x-3)\bbone_{[\frac{1}{2},1)}(x)
+  \end{cases}
 \]
@@
 \[
-  \psi_{(0, n)}(x) := \phi_n(x), \quad
-  \psi_{(\ell, 2n)}(x) := 2^{(\ell-1)/2} \psi_0(2^{\ell-1} x - n), \quad
+  \begin{cases}
+  \psi_{(0, n)}(x) := \phi_n(x), \\
+  \psi_{(\ell, 2n)}(x) := 2^{(\ell-1)/2} \psi_0(2^{\ell-1} x - n), \\
   \psi_{(\ell, 2n+1)}(x) := 2^{(\ell-1)/2} \psi_1(2^{\ell-1} x - n).
+  \end{cases}
 \]
-This rescaling ensures that the multiscale mass operator is just the identity!
-(Hence the name of the basis.) For this basis, the matrices $P^\ell$ and $Q^\ell$
-are already a lot less obvious.\todo{somehow show in a figure how I derived them}
+This rescaling factors $2^{(\ell-1)/2}$ ensure that the multiscale mass operator
+is just the identity, hence the name of the basis. For this basis, the matrices
+$P^\ell$ and $Q^\ell$ are already a lot less obvious, but one can see that 
+$\phi_\lambda$ and $\psi_\lambda$ are both linear combinations of four scaling
+functions, and that the linear combination is different for even and odd offsets;
+we conclude that $P^\ell$ and $Q^\ell$ will be block-diagonal with blocks of size
+$4 \times 2$. See the code for the specific coefficients.
 
@@
 \emph{Experiments in 3D with a three-point hierarchical basis}, which seemed to
-discuss the same basis.
+discuss the same basis, and also discusses the non-uniform case.
 
@@
 index can be any $\lambda = (\ell, n)$ with $0 \leq n \leq 2^\ell$, and
-corresponds with a node at position $n/2^\ell$. A \emph{multiscale} index
+corresponds with a node inside $[0,1]$ at position $n/2^\ell$. A \emph{multiscale} index
 $\lambda = (\ell, n)$ at level $\ell \geq 1$ has $0 \leq n \leq 2^{\ell-1}$ and
@@
 $\lambda$ on the grid. The assumption that our multiscale index set $\Lambda$ is
-a \emph{tree} in this setting coincides with the assumption that for any $\lambda$
-with $\ell\geq 2$, its parent is also in $\Lambda$.
+a \emph{tree} means that $(0,0)$, $(0,1)$, and $(1,0)$ must be in $\Lambda$, and
+for any $\lambda \in \Lambda$ with $\ell\geq 2$, its parent is also in $\Lambda$.
 
@@
 is different from what is sketched in figure 1 of the document.} and the
-wavelet functions are linear combinations of (at most) three scaling functions.
-\todo{show a couple of example $\Lambda$'s, $\Phi$'s and $\Psi$'s.}
+wavelet functions are linear combinations of three scaling functions. On level 0,
+the wavelet functions coincide with the scaling functions. On level $\ell \geq 1$,
+the wavelet function $\psi_\lambda$ is a linear combination of $\phi_\lambda$,
+$\phi_{\mu_\lambda^l}$, and $\phi_{\mu_\lambda^r}$, where $\mu_\lambda^l$ resp.~$\mu_\lambda^r$
+are the left resp.~right neighbours of $\lambda$ on $\Delta_\ell$.
+
+\begin{figure}
+  \begin{tikzpicture}[scale=0.5]
+    \node[left] at (-1,0){$\Lambda_0$};
+    \draw (0,0) -- (4,0);
+    \node at (0,0){\textbullet};
+    \node at (4,0){\textbullet};
+
+    \node[left] at (-1,1){$\Lambda_1$};
+    \draw (0,1) -- (4,1);
+    \node at (2, 1){\textbullet};
+
+    \node[left] at (-1,2){$\Lambda_2$};
+    \draw (0,2) -- (4,2);
+    \node at (1,2){\textbullet};
+    \node at (3,2){\textbullet};
+
+    \node[left] at (-1,3){$\Lambda_3$};
+    \draw (0,3) -- (4,3);
+    \node at (1/2,3){\textbullet};
+    \node at (3/2,3){\textbullet};
+    \node at (7/2,3){\textbullet};
+
+    \draw[dotted] (-2, -1) -- (4, -1);
+    \node[left] at (-1, -2){$\Delta_3$};
+    \draw (0, -2) -- (4, -2);
+    \node at (0,-2){\textbullet};
+    \node at (1/2,-2){\textbullet};
+    \node at (1,-2){\textbullet};
+    \node at (3/2,-2){\textbullet};
+    \node at (2,-2){\textbullet};
+    \node at (3,-2){\textbullet};
+    \node at (7/2,-2){\textbullet};
+    \node at (4,-2){\textbullet};
+  \end{tikzpicture}
+  \begin{tikzpicture}[scale=0.5]
+    \draw (0,0) -- (0,4); \node[left] at(0,0){$0$}; \node[left] at(0,4){$1$};
+    \draw (0,0) -- (4,0); \node at (0,0){\textbullet}; \node at (4,0){\textbullet};
+    \draw[blue] (0,4) -- (4,0); \draw[brown] (0,0) -- (4,4);
+    \node[above,blue] at (0,4){$\psi_{(0,0)}$};
+    \node[above,brown] at (4,4){$\psi_{(0,1)}$};
+    \node[below] at (2,-0.5){$\ell=0$};
+  \end{tikzpicture}\hspace{-1em}
+  \begin{tikzpicture}[scale=0.5]
+    \draw (0,-2) -- (0,2); \node[left] at(0,-2){$-1$}; \node[left] at(0,2){$1$};
+    \draw (0,0) -- (4,0); \node at (0,0){$\circ$}; \node at (4,0){$\circ$};
+    \node at (2,0){\textbullet};
+    \draw[blue] (0,-2) -- (2,2) -- (4, -2);
+    \node[above,blue] at (2,2){$\psi_{(1,0)}$};
+    \node[below] at (2,-2.5){$\ell=1$};
+  \end{tikzpicture}\hspace{-0.7em}
+  \begin{tikzpicture}[scale=0.5]
+    \draw (0,-2) -- (0,2); \node[left] at(0,-2){$-\sqrt{2}$}; \node[left] at(0,2){$\sqrt{2}$};
+    \draw[dashed] (0,-1) -- (2,-1); \node[left] at (0,-1){$-\tfrac{\sqrt{2}}{2}$};
+    \draw (0,0) -- (4,0); \node at (0,0){$\circ$}; \node at (4,0){$\circ$};
+    \node at (2,0){$\circ$};
+    \node at (1,0){\textbullet};
+    \node at (3,0){\textbullet};
+    \draw[blue] (0,-2) -- (1,2) -- (2, -1) -- (3,0);
+    \node[above,blue] at (1,2){$\psi_{(2,0)}$};
+    \draw[brown] (1,0) -- (2, -1) -- (3, 2) -- (4, -2);
+    \node[above,brown] at (3,2){$\psi_{(2,1)}$};
+    \node[below] at (2,-2.5){$\ell=2$};
+  \end{tikzpicture}\hspace{-0.5em}
+  \begin{tikzpicture}[scale=0.5]
+    \draw (0,-2) -- (0,2); \node[left] at(0,-2){$-2$}; \node[left] at(0,2){$2$};
+    \draw[dashed] (0,-2/3) -- (2,-2/3); \node[left] at (0,-2/3){$-\tfrac{2}{3}$};
+    \draw (0,0) -- (4,0);
+    \node at (0,0){$\circ$}; \node at (4,0){$\circ$}; \node at (2,0){$\circ$};
+    \node at (1,0){$\circ$}; \node at (3,0){$\circ$};
+    \node at (1/2,0){\textbullet};
+    \node at (3/2,0){\textbullet};
+    \node at (7/2,0){\textbullet};
+
+    \draw[blue] (1/2,0) -- (1, -1) -- (3/2, 2) -- (2,-2/3) -- (3,0);
+    \node[above,blue] at (3/2,2){$\psi_{(3,1)}$};
+    \draw[brown] (2,0) -- (3,-2/3) -- (7/2,2) -- (4,-2);
+    \node[above,brown] at (7/2,2){$\psi_{(3,3)}$};
+    \node[below] at (2,-2.5){$\ell=3$};
+  \end{tikzpicture}
+  \caption{Leftmost: nodes in $\Lambda_\ell$ for $\ell=0,1,2,3$, and nodes in $\Delta_3$.
+Notice that $\Lambda_3$ is \emph{not} the full set as $\lambda=(3,2) \not\in \Lambda_3$.
+From left to right: selected wavelet functions on different levels.
+\textbullet\ is a node that is new on this level; $\circ$ is a node from previous levels.
+Note that $\psi_\lambda$ is linear combination of $\phi_\lambda$ and two $\phi_\mu$'s for $\mu$ its two neighbours on $\Delta_{\ell}$.
+Notice that the coefficient of $\phi_\lambda$ for $\psi_\lambda$ is always $+1$, but that of $\phi_\mu$ can be $-1$, $-1/2$, or $-1/3$.
+It is $-1$ when $|\mu| = 0$, $-1/2$ when the skip-neighbour $\nu$ of $\lambda$ through $\mu$ has $|\nu| = |\lambda|$, and $-1/3$ when $|\nu| < |\lambda|$.}
+\end{figure}
 
@@
 \begin{enumerate}
-  \item Do the matrices $P^\ell$ and $Q^\ell$ work, i.e.~do we recover $\phi_\lambda$
-    through $\langle \Phi_\ell, \col_\lambda(P^\ell) \rangle$?
+  \item Do the matrices $P^\ell$ and $Q^\ell$ work, i.e.~is $\phi_\lambda(t) = \langle \Phi_\ell(t), \col_\lambda(P^\ell) \rangle$ pointwise?
   \item Are our multiscale mass matrices diagonal (Haar)/the identity (orthonormal)?
-  \item Does the output of \texttt{apply()} correspond with values found through quadrature?
+  \item Does the output of \texttt{apply()} equal values found through quadrature?
   \item Does the output of \texttt{apply()} equal \texttt{apply\_upp() + apply\_low()}?
@@
 \section{Numerics}
+See Figure~\ref{fig:linear} for an impression. We see that it is pretty much linear-time.
+\begin{figure}
+  \includegraphics[width=0.5\linewidth]{../../Latex/img/time-scaling-31-jul.png}\hfill
+  \includegraphics[width=0.5\linewidth]{../../Latex/img/time-scaling-31-jul-loglog.png}
+  \caption{Wallclock time on my laptop for a single call to \texttt{apply()} on a uniform grid. Left: linear-linear plot; right: log-log plot.}
+  \label{fig:linear}
+\end{figure}
 \subsection{Bottlenecks}
+\begin{verbatim}
+      ncalls  tottime percall filename:lineno(function)
+(1)  8125836    4.824   0.000 indexed_vector.py:69(<genexpr>)
+(2)     1760    3.394   0.005 interval.py:36(__init__)
+(3)  8129028    3.009   0.000 {built-in method __new__ of type object at 0x10e3a2050}
+(4)  2752510    2.600   0.000 {built-in method builtins.sum}
+(5)     1368    2.293   0.002 indexed_vector.py:54(__add__)
+(6)  8129028    2.229   0.000 <string>:1(__new__)
+(7)  2752510    2.219   0.000 indexed_vector.py:67(dot)
+(8)      880    2.135   0.012 applicator.py:147(<setcomp>)
+(9)     1320    2.047   0.010 linear_operator.py:25(<dictcomp>)
+     1965762    1.909   0.000 orthonormal_basis.py:105(scaling_support)
+     2621416    1.889   0.000 interval.py:63(covers)
+     3538994    1.845   0.000 interval.py:12(intersects)
+     1965906    1.758   0.000 haar_basis.py:84(scaling_support)
+(10)    1960    1.661   0.001 {built-in method builtins.sorted}
+\end{verbatim}
+Total runtime to produce the above graph was about 1 minute (57 sec total).
+In the verbatim (creating using \texttt{cProfile}, see also \texttt{timing\_test.py},
+we see that there are no very obvious bottlenecks. Nonetheless, let's discuss some
+of the things we see (NB: there are numbers in the left-hand side of some lines of the verbatim; they correspond with numbers below).
+
+\begin{enumerate}
+  \item Most of the time is spent doing dot products, as can be seen from the code:
+    \begin{verbatim}
+69  return sum(self.vector[labda] * other[labda] for labda in other.keys()
+70             if labda in index_mask)\end{verbatim}
+    I am not sure we can do much about this. It makes sense to do it, I think.
+  \item We spend a lot of time merging these intervals, probably for constructing
+    $\Pi_B$ and $\underline \Pi$. If we think some more about optimizing these
+    calls, this will probably go away.
+  \item This is a little cryptic but is about creating \texttt{Interval} objects.
+    As we are not really doing much with them, possibly it would be better to
+    just use \texttt{tuple}s in a next iteration.
+  \item This is again to produce the dot-product; see snippet above.
+  \item We are lazy and append two disjoint vectors by just adding them; this
+    could maybe be improved.
+  \item This has to do with creating \texttt{Interval} objects, I think this line
+    is because \texttt{Interval} inherits from \texttt{collections.namedtuple}
+    which is supposed to be faster than a regular \texttt{object}.
+  \item More dot-product business.
+  \item This is where we compute $\underline \Pi$, through
+    \begin{verbatim}
+145  ivs = IntervalSet([basis.scaling_support(index) for index in Pi_B] +
+146                    [basis.wavelet_support(index) for index in Lambda_l])
+147  return SingleLevelIndexSet({
+148      index
+149      for index in basis.scaling_indices_on_level(l)
+150      if ivs.covers(basis.scaling_support(index))
+151  })\end{verbatim}
+    as you can see, every call to this is quite expensive. This is because the
+    size of $\Delta_\ell$ grows exponentially in $\ell$.\footnote{Actually, for
+    nonuniform grids, this is not true but \texttt{haar\_basis.py} and
+    \texttt{orthonormal\_basis.py} contain two TODO's in this direction.}
+  \item This is overhead of allocating new \texttt{IndexedVector}s at every
+    matvec. We could opt to do most matvecs in-place (by passing an
+    \texttt{out}-argument).
+  \item One can see that only 1.6 seconds is spent sorting, even though quite a
+    lot of our (nonoptimized) code uses sorting. This is nice.
+\end{enumerate}
+
+\subsection{Caveat on the 3-point basis}
+So I (more or less blindly) implemented the basis. But the multiscale mass matrix
+is not at all very sparse or something, which implies that these `wavelets' are
+not $L_2([0,1])$-orthogonal to one another. I am not completely sure if this was
+the goal of the basis, because if so, I have likely misunderstood something.
+\emph{Still, we can apply the matrix in linear time even though it has $\mathcal O(N^2)$
+elements :).}
 
 \section{Towards tensor-product operators}
+Before we can do things in space-time, a small number of things have to be fixed,
+or at least looked at, in the time-code. They are marked with TODO, so
+\texttt{grep -rI "TODO" .} should point you in the right directions.
 

### 2019-07-31 Python/time/applicator_test.py

@@ -60,11 +60,11 @@ def test_multiscale_mass_matrix_quadrature():
     """ Test that the mass matrix is the same as found with quadrature. """
     ml = 4
     for basis in [
-            HaarBasis.uniform_basis(max_level=ml),
-            HaarBasis.origin_refined_basis(max_level=ml),
-            OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml),
-            OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
-                max_level=ml),
+            #HaarBasis.uniform_basis(max_level=ml),
+            #HaarBasis.origin_refined_basis(max_level=ml),
+            #OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml),
+            #OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
+            #    max_level=ml),
             ThreePointBasis.uniform_basis(max_level=ml),
             ThreePointBasis.origin_refined_basis(max_level=ml)
     ]:
@@ -97,7 +97,8 @@ def test_multiscale_mass_matrix_quadrature():
                                         ])[0]
                         truemat[i, j] = true_val
             print(sorted(Lambda))
-            print(np.round(resmat - truemat, decimals=5))
+            print(np.round(truemat, decimals=3))
+            #print(np.round(resmat - truemat, decimals=5))
             assert np.allclose(resmat, truemat)
 
 

### 2019-07-31 Python/time/basis_test.py

@@ -170,7 +170,6 @@ def test_basis_PQ_matrix():
                 vec = IndexedVector(Pi_bar, bar_eye[i, :])
                 QT[i, :] = basis.Q.rmatvec(Pi_bar, Lambda_l, vec).asarray()
 
-            print(Q.T, QT)
             assert np.allclose(P.T, PT)
             assert np.allclose(Q.T, QT)
             assert np.allclose(PT @ P, (PT @ P).T)

### 2019-07-31 Python/time/timing_test.py

@@ -14,9 +14,11 @@ import pytest
 
 def plot_results(results):
     for basis_name in results:
-        plt.plot(results[basis_name]['N'],
-                 results[basis_name]['apply_time'],
-                 label=basis_name)
+        plt.loglog(results[basis_name]['N'],
+                   results[basis_name]['apply_time'],
+                   basex=2,
+                   basey=2,
+                   label=basis_name)
     #plt.loglog(results['ThreePointBasis']['N'], [
     #    results['ThreePointBasis']['apply_time'][-1] /
     #    results['ThreePointBasis']['N'][-1] * N
@@ -31,6 +33,10 @@ def plot_results(results):
     #],
     #           'k-.',
     #           label='quadratic')
+    plt.title("Complexity of applying mass matrices")
+    plt.xlabel("Number of degrees of freedom")
+    plt.ylabel("Seconds to apply 1 mass matrix")
+    plt.grid(which='both')
     plt.legend()
     plt.show()
     return results
@@ -40,7 +46,7 @@ def plot_results(results):
 def test_linear_complexity():
     results = {}
     try:
-        for level in range(1, 16):
+        for level in range(1, 17):
             for basis in [
                     HaarBasis.uniform_basis(max_level=level),
                     OrthonormalDiscontinuousLinearBasis.uniform_basis(

### 2019-07-31 Python/time/applicator_test.py
added tests for damping matrix and unequal in- and out basis

@@ -22,7 +22,7 @@ def test_haar_multiscale_mass():
     ]:
         for l in range(1, 6):
             Lambda = basis.indices.until_level(l)
-            applicator = Applicator(basis, basis.singlescale_mass, Lambda)
+            applicator = Applicator(basis, basis.singlescale_mass(), Lambda)
             for _ in range(10):
                 np_vec = np.random.rand(len(Lambda))
                 vec = IndexedVector(Lambda, np_vec)
@@ -41,7 +41,7 @@ def test_orthonormal_multiscale_mass():
     ]:
         for l in range(1, 6):
             Lambda = basis.indices.until_level(l)
-            applicator = Applicator(basis, basis.singlescale_mass, Lambda)
+            applicator = Applicator(basis, basis.singlescale_mass(), Lambda)
             eye = np.eye(len(Lambda))
             res_matrix = np.zeros([len(Lambda), len(Lambda)])
             for i in range(len(Lambda)):
@@ -56,36 +56,55 @@ def test_orthonormal_multiscale_mass():
                 assert np.allclose(vec, res.asarray())
 
 
-def test_multiscale_mass_matrix_quadrature():
-    """ Test that the mass matrix is the same as found with quadrature. """
+def test_multiscale_operator_quadrature():
+    """ Test that the multiscale matrix equals that found with quadrature. """
     ml = 4
-    for basis in [
-            #HaarBasis.uniform_basis(max_level=ml),
-            #HaarBasis.origin_refined_basis(max_level=ml),
-            #OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml),
-            #OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
-            #    max_level=ml),
-            ThreePointBasis.uniform_basis(max_level=ml),
-            ThreePointBasis.origin_refined_basis(max_level=ml)
+    hbu = HaarBasis.uniform_basis(max_level=ml)
+    hbo = HaarBasis.origin_refined_basis(max_level=ml)
+    oru = OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml)
+    oro = OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
+        max_level=ml)
+    tpu = ThreePointBasis.uniform_basis(max_level=ml)
+    tpo = ThreePointBasis.origin_refined_basis(max_level=ml)
+    for basis_in, basis_out, operator, deriv in [
[…]

### 2019-07-31 Python/time/basis.py

@@ -62,10 +62,10 @@ class Basis(object):
         """ The neighbourhood S(psi_labda). Defaults to its support. """
         return self.wavelet_support(labda)
 
-    def eval_scaling(self, labda, x):
+    def eval_scaling(self, labda, x, deriv=False):
         """ Debug method. """
         pass
 
-    def eval_wavelet(self, labda, x):
+    def eval_wavelet(self, labda, x, deriv=False):
         """ Debug method. """
         pass

### 2019-07-31 Python/time/basis_test.py

@@ -16,7 +16,7 @@ np.set_printoptions(linewidth=10000, precision=3)
 def test_haar_singlescale_mass():
     """ Test that the singlescale Haar mass matrix is indeed diagonal. """
     basis = HaarBasis.uniform_basis(max_level=5)
-    mass = basis.singlescale_mass
+    mass = basis.singlescale_mass()
     for l in range(1, 5):
         indices = basis.scaling_indices_on_level(l)
         assert len(indices.indices) == 2**l
@@ -29,7 +29,7 @@ def test_haar_singlescale_mass():
 
 def test_orthonormal_singlescale_mass():
     basis = OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=4)
-    mass = basis.singlescale_mass
+    mass = basis.singlescale_mass()
     for l in range(1, 5):
         indices = basis.scaling_indices_on_level(l)
         assert len(indices.indices) == 2 * 2**l
@@ -240,33 +240,59 @@ def test_3point_singlescale_indices():
 def test_singlescale_mass_quadrature():
     """ Use quadrature to test the singlescale mass matrices. """
     ml = 4
-    for basis in [
-            HaarBasis.uniform_basis(max_level=ml),
-            HaarBasis.origin_refined_basis(max_level=ml),
-            OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml),
-            OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
-                max_level=ml),
-            ThreePointBasis.uniform_basis(max_level=ml),
-            ThreePointBasis.origin_refined_basis(max_level=ml)
+    hbu = HaarBasis.uniform_basis(max_level=ml)
+    hbo = HaarBasis.origin_refined_basis(max_level=ml)
+    oru = OrthonormalDiscontinuousLinearBasis.uniform_basis(max_level=ml)
+    oro = OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
+        max_level=ml)
+    tpu = ThreePointBasis.uniform_basis(max_level=ml)
+    tpo = ThreePointBasis.origin_refined_basis(max_level=ml)
+    for basis_in, basis_out, operator, deriv in [
+        (hbu, hbu, hbu.singlescale_mass(hbu), (False, False)),
+        (hbo, hbo, hbo.singlescale_mass(hbo), (False, False)),
+        (hbu, hbo, hbu.singlescale_mass(hbo), (False, False)),
[…]

### 2019-07-31 Python/time/haar_basis.py

@@ -45,9 +45,10 @@ class HaarBasis(Basis):
 
         return LinearOperator(row, col)
 
-    @property
-    def singlescale_mass(self):
+    def singlescale_mass(self, basis_out=None):
         """ The singlescale Haar mass matrix is simply 2**-l * Id. """
+        if basis_out:
+            assert isinstance(basis_out, HaarBasis)
 
         def row(labda):
             l, n = labda
@@ -58,14 +59,16 @@ class HaarBasis(Basis):
     def eval_mother_scaling(self, x):
         return (0 <= x) & (x < 1)
 
-    def eval_scaling(self, labda, x):
+    def eval_scaling(self, labda, x, deriv=False):
+        assert deriv == False
         l, n = labda
         return 1.0 * self.eval_mother_scaling(2**l * x - n)
 
     def eval_mother_wavelet(self, x):
         return 1.0 * ((0 <= x) & (x < 0.5)) - 1.0 * ((0.5 <= x) & (x < 1.0))
 
-    def eval_wavelet(self, labda, x):
+    def eval_wavelet(self, labda, x, deriv=False):
+        assert deriv == False
         l, n = labda
         if l == 0:
             return self.eval_mother_scaling(x)

### 2019-07-31 Python/time/linear_operator.py

@@ -4,10 +4,19 @@ from indexed_vector import IndexedVector
 class LinearOperator(object):
     """ Linear operator. """
 
-    def __init__(self, row_fn, col_fn=None):
-        """ Initialize. """
-        self.row = row_fn
-        self.col = col_fn
+    def __init__(self, row, col=None):
+        """ Initialize the LinearOperator.
+        
+        Arguments:
+            row: a function taking an index and returning an IndexedVector,
+                being the nonzero entries of this operator at this row. Needed
+                for matrix-vector products.
+            col (optional): a similar function as `row` but for the column of
+                the operator. Needed for z = A^T x.
+                
+        """
+        self.row = row
+        self.col = col
 
     def matvec(self, indices_in, indices_out, vec, out=None):
         """ Computes matrix-vector product z = A x.

### 2019-07-31 Python/time/orthonormal_basis.py

@@ -65,33 +65,39 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
 
         return LinearOperator(row, col)
 
-    def eval_mother_scaling(self, odd, x):
-        if odd:
-            return sq3 * (2 * x - 1) * ((0 <= x) & (x < 1))
+    def eval_mother_scaling(self, odd, x, deriv):
+        if not deriv:
+            if odd: return sq3 * (2 * x - 1) * ((0 <= x) & (x < 1))
+            else: return (0 <= x) & (x < 1)
         else:
-            return (0 <= x) & (x < 1)
+            if odd: return sq3 * 2 * ((0 <= x) & (x < 1))
+            else: return 0 * ((0 <= x) & (x < 1))
 
-    def eval_scaling(self, labda, x):
+    def eval_scaling(self, labda, x, deriv=False):
         l, n = labda
-        return 1.0 * self.eval_mother_scaling(n % 2, 2**l * x - (n // 2))
+        chain_rule_constant = 2**l if deriv else 1.0
+        return chain_rule_constant * self.eval_mother_scaling(
+            n % 2, 2**l * x - (n // 2), deriv)
 
-    def eval_mother_wavelet(self, odd, x):
-        if odd:
-            return sq3 * (1 - 4 * x) * (
-                (0 <= x) & (x < 0.5)) + sq3 * (4 * x - 3) * ((0.5 <= x) &
-                                                             (x < 1))
+    def eval_mother_wavelet(self, odd, x, deriv):
+        mask1 = ((0 <= x) & (x < 0.5))
+        mask2 = ((0.5 <= x) & (x < 1))
+        if not deriv:
+            if odd: return sq3 * ((1 - 4 * x) * mask1 + (4 * x - 3) * mask2)
+            else: return (1 - 6 * x) * mask1 + (5 - 6 * x) * mask2
         else:
-            return (1 - 6 * x) * ((0 <= x) &
-                                  (x < 0.5)) + (5 - 6 * x) * ((0.5 <= x) &
-                                                              (x < 1))
+            if odd: return -4 * sq3 * mask1 + 4 * sq3 * mask2
+            else: return -6 * mask1 - 6 * mask2
 
-    def eval_wavelet(self, labda, x):
+    def eval_wavelet(self, labda, x, deriv=False):
         l, n = labda
         if l == 0:
-            return 1.0 * self.eval_mother_scaling(n % 2, x)
+            return 1.0 * self.eval_mother_scaling(n % 2, x, deriv)
         else:
[…]

### 2019-07-31 Python/time/three_point_basis.py

@@ -1,4 +1,5 @@
 from basis import Basis
+from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
 from index_set import MultiscaleIndexSet, SingleLevelIndexSet
 from indexed_vector import IndexedVector
 from interval import Interval
@@ -160,25 +161,51 @@ class ThreePointBasis(Basis):
 
         return LinearOperator(row, col)
 
-    @property
-    def singlescale_mass(self):
+    def singlescale_mass(self, basis_out=None):
+        if basis_out:
+            # I have currently only done the case where basis_in == basis_out.
+            assert isinstance(basis_out, ThreePointBasis) and self is basis_out
+
         def row(labda):
             l, n = labda
 
             left, right = self.scaling_indices_on_level(l).neighbours(labda)
 
-            res = {labda: 0.0}
+            res = {labda: 0.0, left: 0.0, right: 0.0}
             if n > 0:
                 dist_left = labda[1] - left[1]
                 res[left] = dist_left / 6
-                res[labda] += dist_left / 3
+                res[labda] = res[labda] + dist_left / 3
             if n < 2**l:
                 dist_right = right[1] - labda[1]
-                res[labda] += dist_right / 3
+                res[labda] = res[labda] + dist_right / 3
                 res[right] = dist_right / 6
             return res
 
         return LinearOperator(row)
+
+    def singlescale_damping(self, basis_out):
+        if isinstance(basis_out, ThreePointBasis):
+            # I have only built the case where the in- and out basis are equal.
+            assert self is basis_out
+
+            def row(labda):
+                l, n = labda
+                left, right = self.scaling_indices_on_level(l).neighbours(
+                    labda)
+
+                res = {}
+                if n == 0: res[labda] = -2**(l - 1)
+                else: res[left] = -2**(l - 1)
+
+                if n == 2**l: res[labda] = 2**(l - 1)
+                else: res[right] = 2**(l - 1)
+                return res
+
+            return LinearOperator(row)
+        elif isinstance(basis_out, OrthonormalDiscontinuousLinearBasis):
+            # TODO: this is probably necessary for the spacetime case.
+            assert False
 
     def scaling_support(self, labda):
         if labda not in self._scaling_support:
[…]

### 2019-08-01 Python/time/readme.tex
added tests for damping matrix and unequal in- and out basis
clearing up some things

@@
     size of $\Delta_\ell$ grows exponentially in $\ell$.\footnote{Actually, for
-    nonuniform grids, this is not true but \texttt{haar\_basis.py} and
-    \texttt{orthonormal\_basis.py} contain two TODO's in this direction.}
+    nonuniform grids, a much smaller set can be used, but \texttt{haar\_basis.py}
+    and \texttt{orthonormal\_basis.py} contain two TODO's in this direction.}
   \item This is overhead of allocating new \texttt{IndexedVector}s at every
@@
 
+In any case, right now,
+
 \end{document}

### 2019-08-01 Python/time/basis_test.py

@@ -237,8 +237,8 @@ def test_3point_singlescale_indices():
     }
 
 
-def test_singlescale_mass_quadrature():
-    """ Use quadrature to test the singlescale mass matrices. """
+def test_singlescale_quadrature():
+    """ Use quadrature to test the singlescale matrices. """
     ml = 4
     hbu = HaarBasis.uniform_basis(max_level=ml)
     hbo = HaarBasis.origin_refined_basis(max_level=ml)

### 2019-08-01 Python/time/haar_basis.py

@@ -89,5 +89,5 @@ class HaarBasis(Basis):
         return Interval(2**-l * n, 2**-l * (n + 1))
 
     def scaling_indices_on_level(self, l):
-        # TODO: this can be a much smaller set when the grid is non-uniform.
+        # TODO: this can be a much smaller set on non-uniform grids.
         return SingleLevelIndexSet({(l, n) for n in range(2**l)})

### 2019-08-01 Python/time/orthonormal_basis.py

@@ -65,6 +65,31 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
 
         return LinearOperator(row, col)
 
+    def singlescale_mass(self, basis_out=None):
+        """ The singlescale orthonormal mass matrix is simply 2**-l * Id. """
+        if basis_out:
+            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
+
+        def row(labda):
+            l, n = labda
+            return {(l, n): 2**-l}
+
+        return LinearOperator(row, None)
+
+    def singlescale_damping(self, basis_out=None):
+        """ The singlescale damping matrix int_0^1 phi_i' phi_j dt. """
+        if basis_out:
+            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
+
+        def row(labda):
+            l, n = labda
+            if n % 2 == 0:
+                return {(l, n + 1): 2 * sq3}
+            else:
+                return {}
+
+        return LinearOperator(row, None)
+
     def eval_mother_scaling(self, odd, x, deriv):
         if not deriv:
             if odd: return sq3 * (2 * x - 1) * ((0 <= x) & (x < 1))
@@ -113,30 +138,5 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
         return Interval(2**-l * (n // 2), 2**-l * (n // 2 + 1))
 
     def scaling_indices_on_level(self, l):
-        # TODO: this can be a much smaller set when the grid is non-uniform.
+        # TODO: this can be a much smaller set on non-uniform grids.
         return SingleLevelIndexSet({(l, n) for n in range(2 * 2**l)})
-
-    def singlescale_mass(self, basis_out=None):
-        """ The singlescale orthonormal mass matrix is simply 2**-l * Id. """
-        if basis_out:
-            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
-
-        def row(labda):
-            l, n = labda
-            return {(l, n): 2**-l}
-
-        return LinearOperator(row, None)
-
-    def singlescale_damping(self, basis_out=None):
-        """ The singlescale damping matrix int_0^1 phi_i' phi_j dt. """
-        if basis_out:
-            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
-
-        def row(labda):
-            l, n = labda
-            if n % 2 == 0:
-                return {(l, n + 1): 2 * sq3}
-            else:
-                return {}
-
-        return LinearOperator(row, None)

### 2019-08-01 Python/time/three_point_basis.py

@@ -205,7 +205,7 @@ class ThreePointBasis(Basis):
             return LinearOperator(row)
         elif isinstance(basis_out, OrthonormalDiscontinuousLinearBasis):
             # TODO: this is probably necessary for the spacetime case.
-            assert False
+            return LinearOperator(row=lambda x: {})
 
     def scaling_support(self, labda):
         if labda not in self._scaling_support:

### 2019-08-01 Python/time/readme.tex
added an ODE solver test

@@
 \usepackage{amssymb}
+\usepackage{bm}
 \usepackage{geometry}
@@
 
+\subsection{Some fun: solving an ODE}
+I used the three-point basis on a uniform grid to solve some ODEs, in order
+to test the (hopefully fast) application of the multiscale matrices inside an
+iterative linear solver. I didn't completely succed, however.
+
+My original goal was to solve $-u'' = f, u(0) = u(1) = 0$ using Galerkin on
+both the singlescale and multiscale basis. See \texttt{ode\_test.py}. In
+singlescale, it works perfectly and produces nice results (run the code). In
+multiscale however, I was unable to find the right subset of indices to impose
+a boundary condition on, in order to get results that were nice. I think I may
+have solved a different equation entirely, maybe Rob will have some more ideas
+(very likely, this idea was senseless). Still, I was able to solve the
+``zeroth-order'' ODE $u = f$ by using the multiscale mass matrix. This is nice
+enough.
+
 \subsection{Caveat on the 3-point basis}
@@
 
-In any case, right now,
+In any case, right now, we are (more or less) able to evaluate a time-operator
+in linear time. If we now use the \texttt{space-fem.py} code to convert a
+multiscale operator to a singlescale operator, we could in principe use a more
+optimized library for the singlescale space operations (or, in the mean time,
+use our own). With both of those elements in place, it is not a far leap to
+implementing Theorem 4.1 from the document, although I do not quite understand
+how to compute $\bm \Sigma$ and $\bm \Theta$.
+
+I am unsure if the current framework is suitable for spacetime computations
+as-is, because I was mostly concerned with building a working time-framework.
+It is very well possible that to efficiently compute the fibers and other
+spacetime index set stuff, we need different data structures altogether.
 

### 2019-08-01 Python/time/basis_test.py

@@ -267,6 +267,8 @@ def test_singlescale_quadrature():
             #(tpu, oru, tpu.singlescale_damping(oru), (True, False)),
             #(tpo, oro, tpo.singlescale_damping(oro), (True, False)),
             #(tpu, oro, tpu.singlescale_damping(oro), (True, False)),
+        (tpu, tpu, tpu.singlescale_stiffness(tpu), (True, True)),
+        (tpo, tpo, tpo.singlescale_stiffness(tpo), (True, True)),
     ]:
         for l in range(1, ml + 1):
             Delta_l_in = basis_in.scaling_indices_on_level(l)
@@ -290,8 +292,8 @@ def test_singlescale_quadrature():
                     except AssertionError:
                         print(basis_in.__class__.__name__,
                               basis_out.__class__.__name__, operator,
-                              Delta_l_in, labda, operator.row(labda), mu,
-                              true_val, out[mu])
+                              Delta_l_in, Delta_l_out, labda,
+                              operator.row(labda), mu, true_val, out[mu])
                         raise
 
 

### 2019-08-01 Python/time/indexed_vector.py

@@ -29,6 +29,9 @@ class IndexedVector(collections.abc.Mapping):
         """ Zero constructor. """
         return cls(index_set={})
 
+    def __setitem__(self, key, val):
+        self.vector[key] = val
+
     def __getitem__(self, key):
         if key in self.vector:
             return self.vector[key]

### 2019-08-01 Python/time/ode_test.py

@@ -1,0 +1,144 @@
+from applicator import Applicator
+
+from haar_basis import HaarBasis
+from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
+from three_point_basis import ThreePointBasis, position_ss, position_ms
+from index_set import MultiscaleIndexSet
+from indexed_vector import IndexedVector
+
+from scipy.integrate import quad
+from scipy.sparse import linalg
+import matplotlib.pyplot as plt
+import time
+import numpy as np
+import cProfile
+import pytest
+
+
+class ScipyLinearOperator(linalg.LinearOperator):
+    """ Helper class to interface with scipy.sparse.linalg.{gmres,cg}. """
+
+    def __init__(self, indices, boundary_condition, applicator):
+        self.indices = indices
+        self.boundary_condition = boundary_condition
+        self.applicator = applicator
+        self.shape = (len(self.indices), len(self.indices))
+        self.dtype = np.dtype('float')
+
+    def _matvec(self, array):
+        indexed_vector = IndexedVector(self.indices, array)
+        if isinstance(self.applicator, Applicator):
+            output = self.applicator.apply(indexed_vector)
+        else:
+            output = self.applicator.matvec(self.indices, self.indices,
+                                            indexed_vector)
+        for labda in self.boundary_condition:
+            output[labda] = 0.0
+        return output.asarray()
+
+
+def test_singlescale_ode_solve(plot=False):
+    """ Solve a Poisson problem -u'' = f using singlescale stiffnes matrix.
+    
+    This also works (upon making the correct changes) for the eqn u' = f
+    (using damping matrix). I suppose it also works for u = f using mass matrix.
+    """
+    prev_L2_error = 1.0
+    for ml in range(1, 6):
+        basis = ThreePointBasis.uniform_basis(max_level=ml)
+        indices = basis.scaling_indices_on_level(ml)
+        boundary_condition = {(ml, 0), (ml, 2**ml)}
+
+        u = lambda t: np.sin(2 * np.pi * t)
+        dtt_u = lambda t: -4 * np.pi**2 * np.sin(2 * np.pi * t)
+
+        # Build RHS, simply through quadrature.
+        rhs = {}
+        for labda in indices:
+            supp = basis.scaling_support(labda)
+            rhs[labda] = 0.0 if labda in boundary_condition else quad(
[…]

### 2019-08-01 Python/time/three_point_basis.py

@@ -171,15 +171,15 @@ class ThreePointBasis(Basis):
 
             left, right = self.scaling_indices_on_level(l).neighbours(labda)
 
-            res = {labda: 0.0, left: 0.0, right: 0.0}
+            res = {labda: 0.0}
             if n > 0:
-                dist_left = labda[1] - left[1]
-                res[left] = dist_left / 6
-                res[labda] = res[labda] + dist_left / 3
+                dist_left = (labda[1] - left[1]) / (2**l)
+                res[left] = (dist_left / 6) * 2**l
+                res[labda] = res[labda] + (dist_left / 3) * 2**l
             if n < 2**l:
-                dist_right = right[1] - labda[1]
-                res[labda] = res[labda] + dist_right / 3
-                res[right] = dist_right / 6
+                dist_right = (right[1] - labda[1]) / (2**l)
+                res[labda] = res[labda] + (dist_right / 3) * 2**l
+                res[right] = (dist_right / 6) * 2**l
             return res
 
         return LinearOperator(row)
@@ -197,15 +197,38 @@ class ThreePointBasis(Basis):
                 res = {}
                 if n == 0: res[labda] = -2**(l - 1)
                 else: res[left] = -2**(l - 1)
-
                 if n == 2**l: res[labda] = 2**(l - 1)
                 else: res[right] = 2**(l - 1)
+                assert None not in res
                 return res
 
             return LinearOperator(row)
         elif isinstance(basis_out, OrthonormalDiscontinuousLinearBasis):
             # TODO: this is probably necessary for the spacetime case.
             return LinearOperator(row=lambda x: {})
+
+    def singlescale_stiffness(self, basis_out=None):
+        if basis_out:
+            # I have currently only done the case where basis_in == basis_out.
+            assert isinstance(basis_out, ThreePointBasis) and self is basis_out
+
+        def row(labda):
+            l, n = labda
+
+            left, right = self.scaling_indices_on_level(l).neighbours(labda)
+
+            res = {labda: 0.0}
+            if n > 0:
+                dist_left = position_ss(labda) - position_ss(left)
+                res[left] = -1 / dist_left * 2**l
+                res[labda] = res[labda] + 1 / dist_left * 2**l
+            if n < 2**l:
+                dist_right = position_ss(right) - position_ss(labda)
+                res[labda] = res[labda] + 1 / dist_right * 2**l
+                res[right] = -1 / dist_right * 2**l
+            return res
+
[…]

### 2019-08-02 Python/time/readme.tex
ODE solve fixed!

@@
 
-I've spent the past two weeks on a (not yet) linear time implementation of \S 3
+I've spent the past two weeks on a (practically but not theoretically) linear time implementation of \S 3
 from ``Bla Parabolic''. Currently it can only apply multiscale (mass) operator
 in the time direction; I felt it was better to have a good understanding of one
-aspect than to produce a half-baked implementation for the tensor-product case.
+aspect than to produce a half-baked implementation for the tensor-product spacetime case.
 
@@
 to test the (hopefully fast) application of the multiscale matrices inside an
-iterative linear solver. I didn't completely succed, however.
-
-My original goal was to solve $-u'' = f, u(0) = u(1) = 0$ using Galerkin on
-both the singlescale and multiscale basis. See \texttt{ode\_test.py}. In
-singlescale, it works perfectly and produces nice results (run the code). In
-multiscale however, I was unable to find the right subset of indices to impose
-a boundary condition on, in order to get results that were nice. I think I may
-have solved a different equation entirely, maybe Rob will have some more ideas
-(very likely, this idea was senseless). Still, I was able to solve the
-``zeroth-order'' ODE $u = f$ by using the multiscale mass matrix. This is nice
-enough.
+iterative linear solver. I solve $-u'' = f, u(0) = u(1) = 0$ using Galerkin on
+both the singlescale and multiscale basis. See \texttt{ode\_test.py}. It works
+perfectly and produces nice results (run \texttt{python3 ode\_test.py}). With a
+small adaptation, this can be used to solve $u' = f$ and the like as well. Nice
+little proof-of-concept :-)
 

### 2019-08-02 Python/time/ode_test.py

@@ -2,7 +2,7 @@ from applicator import Applicator
 
 from haar_basis import HaarBasis
 from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
-from three_point_basis import ThreePointBasis, position_ss, position_ms
+from three_point_basis import ThreePointBasis, position_ss, position_ms, ms2ss
 from index_set import MultiscaleIndexSet
 from indexed_vector import IndexedVector
 
@@ -46,6 +46,7 @@ def test_singlescale_ode_solve(plot=False):
     prev_L2_error = 1.0
     for ml in range(1, 6):
         basis = ThreePointBasis.uniform_basis(max_level=ml)
+        basis.vanish_at_boundary = True
         indices = basis.scaling_indices_on_level(ml)
         boundary_condition = {(ml, 0), (ml, 2**ml)}
 
@@ -100,8 +101,9 @@ def test_multiscale_ode_solve(plot=False):
     prev_L2_error = 1.0
     for ml in range(1, 7):
         basis = ThreePointBasis.uniform_basis(max_level=ml)
+        basis.vanish_at_boundary = True
         indices = basis.indices
-        boundary_condition = {}
+        boundary_condition = {(0, 0), (0, 1)}
         u = lambda t: np.sin(2 * np.pi * t)
         dtt_u = lambda t: -4 * np.pi**2 * np.sin(2 * np.pi * t)
 
@@ -110,19 +112,19 @@ def test_multiscale_ode_solve(plot=False):
         for labda in indices:
             supp = basis.wavelet_support(labda)
             rhs_dict[labda] = 0.0 if labda in boundary_condition else quad(
-                lambda t: u(t) * basis.eval_wavelet(labda, t),
+                lambda t: -dtt_u(t) * basis.eval_wavelet(labda, t),
                 supp.a,
                 supp.b,
                 points=[supp.a, supp.mid, supp.b, 1.0])[0]
         rhs = IndexedVector(rhs_dict)
 
         # Build operator.
-        operator = basis.singlescale_mass(basis)
+        operator = basis.singlescale_stiffness(basis)
         applicator = Applicator(basis, operator, indices)
         scipy_op = ScipyLinearOperator(indices, boundary_condition, applicator)
         sol = IndexedVector(
             indices,
-            linalg.gmres(scipy_op, rhs.asarray(), atol='legacy')[0])
+            linalg.cg(scipy_op, rhs.asarray(), atol='legacy')[0])
 
[…]

### 2019-08-02 Python/time/three_point_basis.py

@@ -67,7 +67,7 @@ class ThreePointBasis(Basis):
     we build something that tracks neighbours of scaling indices.
     """
 
-    def __init__(self, indices):
+    def __init__(self, indices, vanish_at_boundary=False):
         super().__init__(indices)
 
         # TODO: This is a loop with complexity O(max_level * |indices|), oops!
@@ -84,6 +84,13 @@ class ThreePointBasis(Basis):
         # Some memoizations.
         self._scaling_support = {}
         self._wavelet_support = {}
+
+        # Incorporating these ``essential boundary conditions'' is mainly for
+        # testing in `ode_test.py` and will not be necessary for our parabolic
+        # situation, because the boundary conditions are imposed "weakly" in
+        # the sense that it is baked into the variational form rather than the
+        # spaces we work with.
+        self.vanish_at_boundary = vanish_at_boundary
 
     @classmethod
     def _uniform_multilevel_indices(cls, max_level):
@@ -128,9 +135,14 @@ class ThreePointBasis(Basis):
 
             # If we are the leftmost singlescale index, it can only interact
             # with multiscale index (l, 0).
-            if n == 0: return {(l, 0): -1.0}
+            if n == 0:
+                return {(l, 0): 0.0 if self.vanish_at_boundary else -1.0}
             # Same idea for the rightmost singlescale index.
-            if n == 2**l: return {(l, 2**(l - 1) - 1): -1.0}
+            if n == 2**l:
+                return {
+                    (l, 2**(l - 1) - 1):
+                    0.0 if self.vanish_at_boundary else -1.0
+                }
 
             # General case: we are between these two multiscale indices.
             return {
@@ -145,13 +157,17 @@ class ThreePointBasis(Basis):
             if (l, n) == (0, 0): return {(0, 0): 1.0}
             if (l, n) == (0, 1): return {(0, 1): 1.0}
 
-            if n == 0: left = -1
+            if n == 0: left = 0.0 if self.vanish_at_boundary else -1.0
+
             elif (l, n - 1) in self.indices: left = -1 / 2
             else: left = -1 / 3
 
-            if n == 2**(l - 1) - 1: right = -1
-            elif (l, n + 1) in self.indices: right = -1 / 2
-            else: right = -1 / 3
+            if n == 2**(l - 1) - 1:
+                right = 0.0 if self.vanish_at_boundary else -1.0
+            elif (l, n + 1) in self.indices:
[…]

### 2019-08-14 Python/time/three_point_basis.py
redid three_point_basis

@@ -1,96 +1,28 @@
+
 from basis import Basis
-from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
+from math import floor
 from index_set import MultiscaleIndexSet, SingleLevelIndexSet
 from indexed_vector import IndexedVector
-from interval import Interval
+from interval import Interval, IntervalSet
 from linear_operator import LinearOperator
 
-from functools import lru_cache
 import numpy as np
-
-sq2 = np.sqrt(2)
-sq3 = np.sqrt(3)
-
-
-def position_ms(labda):
-    """ Position in [0,1] of a multiscale index labda. """
-    if labda[0] == 0:
-        assert labda[1] in [0, 1]
-        return float(labda[1])
-    return float(2**(-labda[0]) + 2**(1 - labda[0]) * labda[1])
-
-
-def position_ss(labda):
-    """ Position in [0,1] of a singlescale index labda. """
-    return float(labda[1] / 2.0**labda[0])
-
-
-def ms2ss(l, labda):
-    """ Singlescale index on level l of a multiscale index labda. """
-    if labda[0] == 0: return (l, 2**l * labda[1])
-    return (l, 2**(l - labda[0]) * (2 * labda[1] + 1))
-
-
-@lru_cache(maxsize=2**20)
-def ss2ms(labda):
-    """ Multiscale index of a singlescale index labda.
-    
-    This recursive function is slow, could even be linear in time. Also magic.
-    We could fix this by tracking the multiscale index of every singlescale
-    index. A simpler option could be to cache the results.
-    """
-    l, n = labda
-    # Special case: we are one of the end points.
-    if n in [0, 2**l]:
-        return (0, n // (2**l))
-    # We are *right* on the middle point.
-    elif n == 2**(l - 1):
-        return (1, 0)
-
-    # Magic from here on.
-    # We are to the left of the middle point.
-    elif n < 2**(l - 1):
-        # Recurse, magic, and simply return.
-        ll, nn = ss2ms((l - 1, n))
-        return (ll + 1, nn)
-    # We are to the right of the middle point.
-    else:
-        # Recurse and *magic*.
-        ll, nn = ss2ms((l - 1, n - 2**(l - 1)))
-        return (ll + 1, nn + 2**(ll - 1))
 
 
 class ThreePointBasis(Basis):
-    """ Implements the three-point basis.
-    
-    Has a couple of expensive steps but we could fix that to be O(1) time if
[…]

### 2019-08-14 Python/time/three_point_basis_old.py

@@ -1,0 +1,404 @@
+from basis import Basis
+from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
+from index_set import MultiscaleIndexSet, SingleLevelIndexSet
+from indexed_vector import IndexedVector
+from interval import Interval
+from linear_operator import LinearOperator
+
+from functools import lru_cache
+import numpy as np
+
+sq2 = np.sqrt(2)
+sq3 = np.sqrt(3)
+
+
+def position_ms(labda):
+    """ Position in [0,1] of a multiscale index labda. """
+    if labda[0] == 0:
+        assert labda[1] in [0, 1]
+        return float(labda[1])
+    return float(2**(-labda[0]) + 2**(1 - labda[0]) * labda[1])
+
+
+def position_ss(labda):
+    """ Position in [0,1] of a singlescale index labda. """
+    return float(labda[1] / 2.0**labda[0])
+
+
+def ms2ss(l, labda):
+    """ Singlescale index on level l of a multiscale index labda. """
+    if labda[0] == 0: return (l, 2**l * labda[1])
+    return (l, 2**(l - labda[0]) * (2 * labda[1] + 1))
+
+
+@lru_cache(maxsize=2**20)
+def ss2ms(labda):
+    """ Multiscale index of a singlescale index labda.
+
+    This recursive function is slow, could even be linear in time. Also magic.
+    We could fix this by tracking the multiscale index of every singlescale
+    index. A simpler option could be to cache the results.
+    """
+    l, n = labda
+    # Special case: we are one of the end points.
+    if n in [0, 2**l]:
+        return (0, n // (2**l))
+    # We are *right* on the middle point.
+    elif n == 2**(l - 1):
+        return (1, 0)
+
+    # Magic from here on.
+    # We are to the left of the middle point.
+    elif n < 2**(l - 1):
+        # Recurse, magic, and simply return.
+        ll, nn = ss2ms((l - 1, n))
+        return (ll + 1, nn)
+    # We are to the right of the middle point.
+    else:
+        # Recurse and *magic*.
+        ll, nn = ss2ms((l - 1, n - 2**(l - 1)))
+        return (ll + 1, nn + 2**(ll - 1))
+
+
+class ThreePointBasis(Basis):
+    """ Implements the three-point basis.
+
+    Has a couple of expensive steps but we could fix that to be O(1) time if
+    we build something that tracks neighbours of scaling indices.
+    """
+
[…]

### 2019-08-14 Python/time/interval.py
let interval make use of rational numbers

@@ -1,9 +1,16 @@
-import collections.abc
+from fractions import Fraction
 
 
-class Interval(collections.namedtuple('Interval', 'a b')):
+class Interval(object):
     """ Represents an open or closed interval (a,b). """
-    __slots__ = ()
+    __slots__ = ('a','b')
+
+    def __init__(self, a, b):
+        print(type(a))
+        assert isinstance(a, int) or isinstance(a, Fraction)
+        assert isinstance(b, int) or isinstance(b, Fraction)
+        self.a = Fraction(a)
+        self.b = Fraction(b)
 
     @property
     def mid(self):
@@ -30,12 +37,12 @@ class Interval(collections.namedtuple('Interval', 'a b')):
 
 class IntervalSet(object):
     """ Represents a collection of intervals with fast merging of overlap.
-    
+
     """
 
     def __init__(self, intervals):
         """ This runs in O(N log N) time.
-        
+
         Current complexity: O(N log N) because of a sort step
         Goal complexity: O(N)
         TODO: if we assume to receive a list of intervals that is already sorted
@@ -49,7 +56,7 @@ class IntervalSet(object):
         sorted_intervals = sorted(intervals, key=lambda interval: interval.a)
         stack = []
         if len(sorted_intervals):
-            stack.append(Interval(*sorted_intervals[0]))
+            stack.append(Interval(sorted_intervals[0].a, sorted_intervals[0].b))
             for interval in sorted_intervals:
                 if interval.intersects(stack[-1], closed=True):
                     stack[-1] = Interval(stack[-1].a,
@@ -59,6 +66,9 @@ class IntervalSet(object):
             self.intervals = stack
         else:
             self.intervals = []
+
+    def __iter__(self):
+        return iter(self.intervals)
 
     def covers(self, interval):
         """ Test if the interval set contains the given interval completely."""

### 2019-08-14 Python/time/interval_test.py

@@ -1,18 +1,21 @@
+from fractions import Fraction
 from interval import Interval, IntervalSet
 
 
+def F(x,y): return Fraction(x,y)
+
 def test_interval_intersect():
-    assert not Interval(0, 0.5).intersects(Interval(0.75, 1.0))
-    assert not Interval(0.75, 1.0).intersects(Interval(0, 0.5))
-    assert not Interval(0.5, 1.0).intersects(Interval(0, 0.5), closed=False)
-    assert Interval(0.5, 1.0).intersects(Interval(0, 0.5), closed=True)
-    assert not Interval(0, 0.25).intersects(Interval(0.5, 1.0))
+    assert not Interval(0, F(1,2)).intersects(Interval(F(3,4), 1))
+    assert not Interval(F(3,4), 1).intersects(Interval(0, F(1,2)))
+    assert not Interval(F(1,2), 1).intersects(Interval(0, F(1,2)), closed=False)
+    assert Interval(F(1,2), 1).intersects(Interval(0, F(1,2)), closed=True)
+    assert not Interval(0, F(1,4)).intersects(Interval(F(1,2), 1))
 
 
 def test_interval_merge():
     ivs = IntervalSet(
         [Interval(0, 1),
-         Interval(0, 0.5),
+         Interval(0, F(1,2)),
          Interval(0, 1),
-         Interval(0, 0.5)])
+         Interval(0, F(1,2))])
     assert ivs.covers(Interval(0, 1))

### 2019-08-14 Python/time/basis.py
make use of fractions instead of floats for intervals

@@ -1,8 +1,12 @@
 from indexed_vector import IndexedVector
 from interval import Interval, IntervalSet
 import numpy as np
+from fractions import Fraction
 
 sq3 = np.sqrt(3)
+
+
+#TODO: Misschien moeten we de basis splitten in twee classes for schaling functies en waveletfuncties?
 
 
 class Basis(object):
@@ -30,14 +34,10 @@ class Basis(object):
         """ MultiscaleIndexSet of indices with refinement towards origin. """
         pass
 
-    def scaling_indices_on_level(self, l):
-        """ SingleLevelIndexSet of singlescale indices on level l; Delta_l. """
-        pass
-
     @property
     def P(self):
         """ The matrices {P_l}_l such that Phi_{l-1}^T = P_l Phi_l^T.
-        
+
         P is a LinearOperator object that implements `matvec()` and `rmatvec()`.
         """
         pass
@@ -45,6 +45,7 @@ class Basis(object):
     @property
     def Q(self):
         """ The matrices {Q_l}_l such that Psi_l^T = Q_l Phi_l^T.
+        TODO: Ik denk dat je bedoelt Psi_l = Q_l^T Phi_l.
 
         Q is a LinearOperator object that implements `matvec()` and `rmatvec()`.
         """
@@ -54,8 +55,24 @@ class Basis(object):
         """ The support of the scaling function phi_labda. """
         pass
 
+    def scaling_indices_on_level(self, l):
+        """ SingleLevelIndexSet of singlescale indices on level l; Delta_l. """
+        pass
+
+    def scaling_indices_nonzero_in_nbrhood(self, l, x):
+        """ Singlescale indices on level l that are nonzero in [x-eps,x+eps]. """
+        assert 0 <= x <= 1
+        assert isinstance(x, Fraction) or isinstance(x, int)
+
+    def scaling_mass(self):
+        """ Mass matrix applied to scaling functions. """
+
     def wavelet_support(self, labda):
         """ The support of the wavelet function psi_labda. """
+        pass
+
+    def wavelet_indices_on_level(self, l):
+        """ SingleLevelIndexSet of multiscale indices on level l; Lambda_l. """
         pass
 
     def wavelet_nbrhood(self, labda):
@@ -67,5 +84,7 @@ class Basis(object):
         pass
 
     def eval_wavelet(self, labda, x, deriv=False):
-        """ Debug method. """
+        """ Debug method.
+
+        TODO: shall we simply use Q_col and eval_scaling for this? """
         pass

### 2019-08-14 Python/time/haar_basis.py

@@ -1,13 +1,22 @@
 from basis import Basis
+from math import floor
 from index_set import MultiscaleIndexSet, SingleLevelIndexSet
 from indexed_vector import IndexedVector
 from interval import Interval, IntervalSet
 from linear_operator import LinearOperator
 
+from fractions import Fraction
+
 import numpy as np
 
 
 class HaarBasis(Basis):
+    """ Piecewise constant basis, with haar wavelets.
+
+    Scaling functions are piecewise constants, indexed by the index of the left
+    node. The haar wavelets of level l are indexed by the index of the
+    left node on level l - 1.
+    """
     def __init__(self, indices):
         super().__init__(indices)
 
@@ -45,17 +54,6 @@ class HaarBasis(Basis):
 
         return LinearOperator(row, col)
 
-    def singlescale_mass(self, basis_out=None):
-        """ The singlescale Haar mass matrix is simply 2**-l * Id. """
-        if basis_out:
-            assert isinstance(basis_out, HaarBasis)
-
-        def row(labda):
-            l, n = labda
-            return {(l, n): 2**-l}
-
-        return LinearOperator(row)
-
     def eval_mother_scaling(self, x):
         return (0 <= x) & (x < 1)
 
@@ -82,12 +80,41 @@ class HaarBasis(Basis):
             return Interval(0, 1)
         else:
             assert 0 <= n < 2**(l - 1)
-            return Interval(2**-(l - 1) * n, 2**-(l - 1) * (n + 1))
+            h = Fraction(1, 2**(l-1))
+            return Interval(h * n, h * (n + 1))
+
+    def wavelet_indices_on_level(self, l):
+        if l == 0:
+            return SingleLevelIndexSet({(0,0)})
+        else:
+            return SingleLevelIndexSet({(l, n) for n in range(2**(l-1))})
+
+    def scaling_mass(self):
+        """ The singlescale Haar mass matrix is simply 2**-l * Id. """
+        def row(labda):
+            l, n = labda
+            return {(l, n): 2**-l}
+        return LinearOperator(row)
 
     def scaling_support(self, labda):
         l, n = labda
-        return Interval(2**-l * n, 2**-l * (n + 1))
+        assert 0 <= n < 2**l
+        h = Fraction(1, 2**l)
+        return Interval(h * n, h * (n + 1))
+
+    def scaling_indices_nonzero_in_nbrhood(self, l, x):
+        super().scaling_indices_nonzero_in_nbrhood(l, x)
+
[…]

### 2019-08-14 Python/time/orthonormal_basis.py

@@ -4,6 +4,8 @@ from index_set import MultiscaleIndexSet, SingleLevelIndexSet
 from indexed_vector import IndexedVector
 from interval import Interval, IntervalSet
 from linear_operator import LinearOperator
+from math import floor
+from fractions import Fraction
 
 import numpy as np
 
@@ -11,7 +13,7 @@ sq3 = np.sqrt(3)
 
 
 class OrthonormalDiscontinuousLinearBasis(Basis):
-    """ We have a multiwavelet basis.
+    """ Discontinuous piecewise linear basis, with orthonormal wavelets.
 
     It has two wavelets and scaling functions. Even index-offsets correspond
     with the first, odd with the second.
@@ -65,31 +67,6 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
 
         return LinearOperator(row, col)
 
-    def singlescale_mass(self, basis_out=None):
-        """ The singlescale orthonormal mass matrix is simply 2**-l * Id. """
-        if basis_out:
-            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
-
-        def row(labda):
-            l, n = labda
-            return {(l, n): 2**-l}
-
-        return LinearOperator(row, None)
-
-    def singlescale_damping(self, basis_out=None):
-        """ The singlescale damping matrix int_0^1 phi_i' phi_j dt. """
-        if basis_out:
-            assert isinstance(basis_out, OrthonormalDiscontinuousLinearBasis)
-
-        def row(labda):
-            l, n = labda
-            if n % 2 == 0:
-                return {(l, n + 1): 2 * sq3}
-            else:
-                return {}
-
-        return LinearOperator(row, None)
-
     def eval_mother_scaling(self, odd, x, deriv):
         if not deriv:
             if odd: return sq3 * (2 * x - 1) * ((0 <= x) & (x < 1))
@@ -131,12 +108,54 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
             return Interval(0, 1)
         else:
             assert 0 <= n < 2 * 2**(l - 1)
-            return Interval(2**-(l - 1) * (n // 2), 2**-(l - 1) * (n // 2 + 1))
+            h = Fraction(1, 2**(l-1))
+            return Interval(h * (n // 2), h * (n // 2 + 1))
+
+    def wavelet_indices_on_level(self, l):
+        if l == 0:
+            return SingleLevelIndexSet({(0,0), (0,1)})
+        else:
[…]

### 2019-08-14 Python/time/three_point_basis.py

@@ -1,4 +1,3 @@
-
 from basis import Basis
 from math import floor
 from index_set import MultiscaleIndexSet, SingleLevelIndexSet
@@ -7,6 +6,7 @@ from interval import Interval, IntervalSet
 from linear_operator import LinearOperator
 
 import numpy as np
+from fractions import Fraction
 
 
 class ThreePointBasis(Basis):
@@ -165,14 +165,14 @@ class ThreePointBasis(Basis):
     def scaling_support(self, labda):
         l, n = labda
         assert 0 <= n <= 2**l
-
+        h = Fraction(1, 2**l)
         # The boundary scaling functions require special treatment.
-        left = max(2**-l * (n-1), 0)
-        right = min(2**-l * (n+1), 1)
+        left = max(h * (n-1), 0)
+        right = min(h * (n+1), 1)
         return Interval(left, right)
 
     def scaling_indices_nonzero_in_nbrhood(self, l, x):
-        assert 0 <= x <= 1
+        super().scaling_indices_nonzero_in_nbrhood(l, x)
         interval_index = floor(x * 2**l)
         return SingleLevelIndexSet({(l, interval_index, interval_index + 1)})
 
@@ -186,10 +186,11 @@ class ThreePointBasis(Basis):
             return Interval(0, 1)
         else:
             assert 0 <= n < 2**(l - 1)
+            h = Fraction(1, 2**l)
 
             # Boundary wavelets require special treatment.
-            left = max(2**-l + 2**-(l-1) * (n-1), 0)
-            right = min(2**-l + 2**-(l-1) * (n+1), 1)
+            left = max(h + 2 * h * (n-1), 0)
+            right = min(h + 2 * h * (n+1), 1)
             return Interval(left, right)
 
     def wavelet_indices_on_level(self, l):

### 2019-08-14 Python/time/applicator.py
redid the slow part of applicator

@@ -5,7 +5,7 @@ from interval import Interval, IntervalSet
 
 class Applicator(object):
     """ Class that can apply multiscale operators with minimal overhead.
-    
+
     See also github.com/skestler/lawa-phd-skestler/LAWA-lite/
         lawa/methods/adaptive/operators/localoperators/localoperator1d.{h,tcpp}
     for some (tough to read) C++ code that implements the same operators in
@@ -85,7 +85,7 @@ class Applicator(object):
     #  Private methods from here on out.
     def _construct_Pi_B_out(self, Pi_out, Lambda_l_in):
         """ Singlescale indices labda st supp(phi_labda) intersects Lambda_l.
-        
+
         Finds the subset Pi_B of Pi such that the support of phi_labda
         intersects some S(mu) for mu in Lambda_l.
         Necessary for computing Pi_B in apply/apply_upp/apply_low.
@@ -124,10 +124,10 @@ class Applicator(object):
             if ivs.intersects(self.basis_in.scaling_support(index))
         })
 
-    def _smallest_superset(self, l, basis, Pi_B, Lambda_l):
-        """ Find Pi_bar, the smallest index set covering Pi_B and Lambda_l.
+    def _largest_subset(self, l, basis, Pi_B, Lambda_l):
+        """ Find Pi_bar, the largest index set contained contained in Pi_B and Lambda_l.
 
-        span Phi_{Pi_B} cup span Psi_{Lambda_l} subset span Phi_{Pi_bar}.
+        span Phi_{Pi_bar} subset Phi_{Pi_B} cup span Psi_{Lambda_l}.
 
         Goal complexity: O(|Pi_bar|).
         Current compl: O(|Pi_bar|*((Pi_B + Lambda_l) log[Pi_B + Lambda_L])).
@@ -144,15 +144,22 @@ class Applicator(object):
         """
         ivs = IntervalSet([basis.scaling_support(index) for index in Pi_B] +
                           [basis.wavelet_support(index) for index in Lambda_l])
-        return SingleLevelIndexSet({
-            index
-            for index in basis.scaling_indices_on_level(l)
-            if ivs.covers(basis.scaling_support(index))
-        })
+
+        result = set()
+        for iv in ivs:
+            # Find the leftmost and rightmost index `active` in this interval
+            left_index = min(basis.scaling_indices_nonzero_in_nbrhood(l, iv.a))[1]
+            right_index = max(basis.scaling_indices_nonzero_in_nbrhood(l, iv.b))[1]
+
[…]

### 2019-08-14 Python/time/applicator_test.py

@@ -2,8 +2,9 @@ from applicator import Applicator
 
 from haar_basis import HaarBasis
 from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
-from three_point_basis import ThreePointBasis, ms2ss, ss2ms, position_ms
+from three_point_basis import ThreePointBasis
 from indexed_vector import IndexedVector
+from index_set import MultiscaleIndexSet, SingleLevelIndexSet
 
 import numpy as np
 from scipy.integrate import quad
@@ -14,6 +15,22 @@ np.random.seed(0)
 np.set_printoptions(linewidth=10000, precision=3)
 
 
+def test_haar_largest_subset():
+    basis = HaarBasis.uniform_basis(max_level=5)
+    applicator = Applicator(basis, basis.scaling_mass())
+
+    # Check the largest subset contained in mother scaling/wavelt
+    assert applicator._largest_subset(5, basis, Pi_B=[(0,0)], Lambda_l = []) == basis.scaling_indices_on_level(5)
+    assert applicator._largest_subset(5, basis, Pi_B=[], Lambda_l =[(0,0)]) == basis.scaling_indices_on_level(5)
+
+    # Check for a singlescale function on level 4
+    for index in range(2**4):
+        assert applicator._largest_subset(5, basis, Pi_B=[(4,index)], Lambda_l = []) == SingleLevelIndexSet([(5,index*2), (5,index*2+1)])
+
+    assert applicator._largest_subset(5, basis, Pi_B=[(4,0), (4,2), (4,3)], Lambda_l = []) == SingleLevelIndexSet([
+        (5,0), (5,1), (5,4), (5,5), (5,6), (5,7)])
+
+
 def test_haar_multiscale_mass():
     """ Test that the multiscale Haar mass matrix is indeed diagonal. """
     for basis in [
@@ -22,7 +39,7 @@ def test_haar_multiscale_mass():
     ]:
         for l in range(1, 6):
             Lambda = basis.indices.until_level(l)
-            applicator = Applicator(basis, basis.singlescale_mass(), Lambda)
+            applicator = Applicator(basis, basis.scaling_mass(), Lambda)
             for _ in range(10):
                 np_vec = np.random.rand(len(Lambda))
                 vec = IndexedVector(Lambda, np_vec)
@@ -36,12 +53,12 @@ def test_orthonormal_multiscale_mass():
     """ Test that the multiscale mass operator is the identity. """
     for basis in [
[…]

### 2019-08-14 Python/time/basis_test.py
updated tests and minor changes

@@ -1,22 +1,74 @@
 from haar_basis import HaarBasis
 from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
-from three_point_basis import ThreePointBasis, ms2ss, ss2ms, position_ms
+from three_point_basis import ThreePointBasis #, ms2ss, ss2ms, position_ms
 
 from index_set import MultiscaleIndexSet
 from indexed_vector import IndexedVector
+from linear_operator_test import check_linop_transpose
 
 import numpy as np
 from scipy.integrate import quad
+from fractions import Fraction
 import matplotlib.pyplot as plt
 import pytest
 
+
 np.set_printoptions(linewidth=10000, precision=3)
 
+def F(x,y): return Fraction(x,y)
 
-def test_haar_singlescale_mass():
+
+def test_haar_ss_nonzero_in_nbrhood():
+    """ Test that we always return the correct non-zero function. """
+    basis = HaarBasis([])
+
+    eps = F(1,1000)
+
+    # Check level 0
+    assert basis.scaling_indices_nonzero_in_nbrhood(0, 1-eps).indices == {(0,0)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(0, 0).indices == {(0,0)}
+
+    # Check level 2
+    assert basis.scaling_indices_nonzero_in_nbrhood(2, 0).indices == {(2,0)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(2, F(1,4)-eps).indices == {(2,0)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(2, F(1,4)).indices == {(2,0), (2,1)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(2, F(1,2)).indices == {(2,1), (2,2)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(2, F(3,4)).indices == {(2,2), (2,3)}
+
+    # Check level 40
+    assert basis.scaling_indices_nonzero_in_nbrhood(40, 0).indices == {(40,0)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(40, F(13,2**40)).indices == {(40,12), (40,13)}
+    assert basis.scaling_indices_nonzero_in_nbrhood(40, F(876,2**40)).indices == {(40,875), (40, 876)}
+
+    # Check 500 points for levels 0..123.
+    for l in range(123):
+        for pt in range(max(1, 2**l - 500), 2**l):
[…]

### 2019-08-14 Python/time/haar_basis.py

@@ -113,7 +113,7 @@ class HaarBasis(Basis):
         node = floor(x * 2 **l)
 
         # If this node coincides, we have to return the basis on left and right.
-        if x == node: return SingleLevelIndexSet({(l, node-1), (l, node)})
+        if x * 2 ** l == node: return SingleLevelIndexSet({(l, node-1), (l, node)})
         else: return SingleLevelIndexSet({(l, node)})
 
     def scaling_indices_on_level(self, l):

### 2019-08-14 Python/time/index_set.py

@@ -20,7 +20,7 @@ class IndexSet(collections.abc.Set):
 
 class SingleLevelIndexSet(IndexSet):
     """ Immutable set of indices on one level, either singlescale or multiscale.
-    
+
     Would be nice if the elements in a singlescale index set would be sorted by
     lexicographical order. Probably hard to do in linear time.
     """
@@ -35,7 +35,7 @@ class SingleLevelIndexSet(IndexSet):
         if isinstance(indices, SingleLevelIndexSet):
             self.indices = indices.indices
         else:
-            self.indices = indices
+            self.indices = set(indices)
         self._neighbours = {}
 
         # TODO: if we assume that `indices` is given to us in lexicographical
@@ -52,7 +52,7 @@ class SingleLevelIndexSet(IndexSet):
 
     def neighbours(self, labda):
         """ Get the neighbours of this singlescale index.
-        
+
         Current complexity: once O(N log N) per IndexSet, then
                             once O(log N) per item, then O(1).
         Goal complexity: O(1).

### 2019-08-14 Python/time/interval.py

@@ -6,7 +6,6 @@ class Interval(object):
     __slots__ = ('a','b')
 
     def __init__(self, a, b):
-        print(type(a))
         assert isinstance(a, int) or isinstance(a, Fraction)
         assert isinstance(b, int) or isinstance(b, Fraction)
         self.a = Fraction(a)
@@ -62,7 +61,7 @@ class IntervalSet(object):
                     stack[-1] = Interval(stack[-1].a,
                                          max(stack[-1].b, interval.b))
                 else:
-                    stack.append(Interval(*interval))
+                    stack.append(Interval(interval.a, interval.b))
             self.intervals = stack
         else:
             self.intervals = []

### 2019-08-14 Python/time/ode_test.py

@@ -2,7 +2,7 @@ from applicator import Applicator
 
 from haar_basis import HaarBasis
 from orthonormal_basis import OrthonormalDiscontinuousLinearBasis
-from three_point_basis import ThreePointBasis, position_ss, position_ms, ms2ss
+from three_point_basis import ThreePointBasis
 from index_set import MultiscaleIndexSet
 from indexed_vector import IndexedVector
 
@@ -39,7 +39,7 @@ class ScipyLinearOperator(linalg.LinearOperator):
 
 def test_singlescale_ode_solve(plot=False):
     """ Solve a Poisson problem -u'' = f using singlescale stiffnes matrix.
-    
+
     This also works (upon making the correct changes) for the eqn u' = f
     (using damping matrix). I suppose it also works for u = f using mass matrix.
     """
@@ -61,11 +61,11 @@ def test_singlescale_ode_solve(plot=False):
                 lambda t: -dtt_u(t) * basis.eval_scaling(labda, t),
                 supp.a,
                 supp.b,
-                points=[supp.a, supp.mid, supp.b, 1.0])[0]
+                points=[float(supp.a), float(supp.mid), float(supp.b), 1.0])[0]
         rhs = IndexedVector(rhs)
 
         # Build the operator. NB: stiffness matrix is pos-def so we can use CG.
-        operator = basis.singlescale_stiffness(basis)
+        operator = basis.scaling_stiffness()
         scipy_op = ScipyLinearOperator(indices, boundary_condition, operator)
         sol = IndexedVector(
             indices,
@@ -115,11 +115,11 @@ def test_multiscale_ode_solve(plot=False):
                 lambda t: -dtt_u(t) * basis.eval_wavelet(labda, t),
                 supp.a,
                 supp.b,
-                points=[supp.a, supp.mid, supp.b, 1.0])[0]
+                points=[float(supp.a), float(supp.mid), float(supp.b), 1.0])[0]
         rhs = IndexedVector(rhs_dict)
 
         # Build operator.
-        operator = basis.singlescale_stiffness(basis)
+        operator = basis.scaling_stiffness()
         applicator = Applicator(basis, operator, indices)
         scipy_op = ScipyLinearOperator(indices, boundary_condition, applicator)
         sol = IndexedVector(

### 2019-08-14 Python/time/orthonormal_basis.py

@@ -145,12 +145,12 @@ class OrthonormalDiscontinuousLinearBasis(Basis):
 
         # treat boundary seperately:
         if x == 0: return SingleLevelIndexSet({(l, 0), (l, 1)})
-        elif x == 1: return SingleLevelIndexSet({(l, 2**(l+1)-2, 2**(l+1)-1)})
+        elif x == 1: return SingleLevelIndexSet({(l, 2**(l+1)-2), (l, 2**(l+1)-1)})
 
         # Find the closest node on left of x
         node = floor(x * 2 **l)
 
-        if x == node:
+        if x * 2 ** l == node:
             # Return two basis functions on left and right of x
             return SingleLevelIndexSet({(l, n) for n in range(2*node-2, 2*node+2)})
         else:

### 2019-08-14 Python/time/three_point_basis.py

@@ -173,8 +173,17 @@ class ThreePointBasis(Basis):
 
     def scaling_indices_nonzero_in_nbrhood(self, l, x):
         super().scaling_indices_nonzero_in_nbrhood(l, x)
-        interval_index = floor(x * 2**l)
-        return SingleLevelIndexSet({(l, interval_index, interval_index + 1)})
+        # treat boundary seperately:
+        if x == 0: return SingleLevelIndexSet({(l, 0), (l, 1)})
+        elif x == 1: return SingleLevelIndexSet({(l, 2**l-1), (l, 2**l)})
+
+        # Find the closest node on left of x
+        node = floor(x * 2 **l)
+
+        if x * 2**l == node:
+            return SingleLevelIndexSet({(l, node-1), (l, node), (l, node + 1)})
+        else:
+            return SingleLevelIndexSet({(l, node), (l, node+1)})
 
     def scaling_indices_on_level(self, l):
         return SingleLevelIndexSet({(l, n) for n in range(2**l + 1)})
@@ -215,10 +224,10 @@ class ThreePointBasis(Basis):
         c = 2**l if deriv else 1.0
         return c * self.eval_mother_scaling(2**l * x - n, deriv)
 
-    def eval_wavelet(self, labda, x):
+    def eval_wavelet(self, labda, x, deriv=False):
         lin_comb_ss = self.Q.col(labda)
 
         result = 0
         for labda_ss, coeff_ss in lin_comb_ss.items():
-            result += coeff_ss * self.eval_scaling(labda_ss, x)
+            result += coeff_ss * self.eval_scaling(labda_ss, x, deriv)
         return result

### 2019-08-14 Python/time/timing_test.py

@@ -43,15 +43,16 @@ def plot_results(results):
 
 
 @pytest.mark.skip("timing test!")
+#@profile
 def test_linear_complexity():
     results = {}
     try:
-        for level in range(1, 17):
+        for level in range(1, 20):
             for basis in [
-                    HaarBasis.uniform_basis(max_level=level),
-                    OrthonormalDiscontinuousLinearBasis.uniform_basis(
-                        max_level=level - 1),
-                    ThreePointBasis.uniform_basis(max_level=level)
+                    HaarBasis.origin_refined_basis(max_level=level),
+#                    OrthonormalDiscontinuousLinearBasis.origin_refined_basis(
+#                        max_level=level - 1),
+#                    ThreePointBasis.uniform_basis(max_level=level)
             ]:
                 if not basis.__class__.__name__ in results:
                     results[basis.__class__.__name__] = {
@@ -59,7 +60,7 @@ def test_linear_complexity():
                         'apply_time': []
                     }
 
-                applicator = Applicator(basis, basis.singlescale_mass,
+                applicator = Applicator(basis, basis.singlescale_mass(),
                                         basis.indices)
                 N = len(basis.indices)
                 vec = IndexedVector(basis.indices, np.random.rand(N))
@@ -80,5 +81,6 @@ def test_linear_complexity():
 
 
 if __name__ == "__main__":
-    cProfile.run('results = test_linear_complexity()', sort='tottime')
-    plot_results(results)
+    results = test_linear_complexity()
+    #cProfile.run('results = test_linear_complexity()', sort='tottime')
+    #plot_results(results)

### 2019-10-24 C++/datastructures/tree.hpp
vertex tree

@@ -64,6 +64,7 @@ class Node : public NodeInterface<T> {
   int level_;
   bool marked_ = false;
   std::vector<std::shared_ptr<T>> parents_;
+  Node() : level_(-1) {}
 
  public:
   int level() const { return level_; }
@@ -73,8 +74,6 @@ class Node : public NodeInterface<T> {
   std::vector<std::shared_ptr<T>> children;
   const std::vector<std::shared_ptr<T>> parents() const { return parents_; }
 
-  Node() : level_(-1) {}
-
   explicit Node(const std::vector<std::shared_ptr<T>> &parents)
       : parents_(parents) {
     assert(parents.size());
@@ -83,4 +82,17 @@ class Node : public NodeInterface<T> {
       assert(parent->level() == level_ - 1);
     }
   }
+};
+
+template <typename T>
+class BinaryNode : public Node<T> {
+ protected:
+  using Node<T>::Node;
+
+ public:
+  bool is_full() const { return Node<T>::children.size() == 2; }
+  bool is_leaf() const { return Node<T>::children.size() == 0; }
+  std::shared_ptr<T> parent() const { return Node<T>::parents_[0]; }
+
+  explicit BinaryNode(std::shared_ptr<T> parent) : Node<T>({parent}) {}
 };

### 2019-10-24 C++/datastructures/triangulation.hpp

@@ -6,7 +6,7 @@
 #include <vector>
 #include "tree.hpp"
 
-struct Vertex;
+class Vertex;
 using VertexPtr = std::shared_ptr<Vertex>;
 using VectorVertexPtr = std::vector<VertexPtr>;
 template <size_t N>
@@ -19,18 +19,16 @@ template <size_t N>
 using ArrayElement2DPtr = std::array<Element2DPtr, N>;
 std::ostream &operator<<(std::ostream &os, const Element2D &elem);
 
-struct Vertex {
-  double x, y;
+class Vertex : public Node<Vertex> {
+ public:
+  const double x, y;
   bool on_domain_boundary;
-  VectorVertexPtr parents;
-  VectorVertexPtr children;
-  Vertex(double x, double y, bool on_domain_boundary, VectorVertexPtr parents,
-         VectorVertexPtr children)
-      : x(x),
-        y(y),
-        on_domain_boundary(on_domain_boundary),
-        parents(parents),
-        children(children) {}
+  VectorElement2DPtr patch;
+
+  Vertex() : Node(), x(-1), y(-1), on_domain_boundary(false) {}
+  Vertex(double x, double y, bool on_domain_boundary,
+         const VectorVertexPtr &parents)
+      : Node(parents), x(x), y(y), on_domain_boundary(on_domain_boundary) {}
 };
 
 std::ostream &operator<<(std::ostream &os, const Vertex &vertex) {
@@ -40,7 +38,7 @@ std::ostream &operator<<(std::ostream &os, const Vertex &vertex) {
 
 class InitialTriangulation;
 
-class Element2D : public Node<Element2D> {
+class Element2D : public BinaryNode<Element2D> {
  protected:
   double area_;
   VectorVertexPtr vertices_;
@@ -54,10 +52,10 @@ class Element2D : public Node<Element2D> {
     }
 
     ArrayVertexPtr<2> godparents{{vertices_[1], vertices_[2]}};
-    auto new_vertex = std::make_shared<Vertex>(
-        (godparents[0]->x + godparents[1]->x) / 2,
-        (godparents[0]->y + godparents[1]->y) / 2, nbr != nullptr,
-        vertex_parents, VectorVertexPtr{});
+    auto new_vertex =
+        std::make_shared<Vertex>((godparents[0]->x + godparents[1]->x) / 2,
+                                 (godparents[0]->y + godparents[1]->y) / 2,
+                                 nbr != nullptr, vertex_parents);
     for (auto vertex_parent : vertex_parents)
[…]

### 2019-10-24 C++/datastructures/hello_world_test.cpp
example bazel working, yaaay

@@ -1,0 +1,4 @@
+#include <iostream>
+#include "gtest/gtest.h"
+
+TEST(Hello, World) { std::cout << "Hello world" << std::endl; }

### 2019-10-25 C++/datastructures/hello_world_test.cpp
improved C++ version

@@ -1,4 +1,0 @@
-#include <iostream>
-#include "gtest/gtest.h"
-
-TEST(Hello, World) { std::cout << "Hello world" << std::endl; }

### 2019-10-25 C++/datastructures/tree.hpp

@@ -5,6 +5,7 @@
 #include <vector>
 #include "cassert"
 
+namespace datastructures {
 template <typename T>
 class NodeInterface : public std::enable_shared_from_this<T> {
   /**
@@ -30,7 +31,7 @@ virtual const std::vector<std::shared_ptr<T>> &parents() const = 0;
       nodes.emplace_back(node);
       node->set_marked(true);
       callback(node);
-      for (auto child : node->children) queue.emplace(child);
+      for (auto child : node->children()) queue.emplace(child);
     }
     for (auto node : nodes) {
       node->set_marked(false);
@@ -64,6 +65,7 @@ class Node : public NodeInterface<T> {
   int level_;
   bool marked_ = false;
   std::vector<std::shared_ptr<T>> parents_;
+  std::vector<std::shared_ptr<T>> children_;
   Node() : level_(-1) {}
 
  public:
@@ -71,8 +73,8 @@ class Node : public NodeInterface<T> {
   bool marked() const { return marked_; }
   void set_marked(bool value) { marked_ = value; }
   bool is_metaroot() const { return (level_ == -1); }
-  std::vector<std::shared_ptr<T>> children;
-  const std::vector<std::shared_ptr<T>> parents() const { return parents_; }
+  const std::vector<std::shared_ptr<T>> &parents() const { return parents_; }
+  std::vector<std::shared_ptr<T>> &children() { return children_; }
 
   explicit Node(const std::vector<std::shared_ptr<T>> &parents)
       : parents_(parents) {
@@ -90,9 +92,10 @@ class BinaryNode : public Node<T> {
   using Node<T>::Node;
 
  public:
-  bool is_full() const { return Node<T>::children.size() == 2; }
-  bool is_leaf() const { return Node<T>::children.size() == 0; }
+  bool is_full() const { return Node<T>::children_.size() == 2; }
+  bool is_leaf() const { return Node<T>::children_.size() == 0; }
   std::shared_ptr<T> parent() const { return Node<T>::parents_[0]; }
 
   explicit BinaryNode(std::shared_ptr<T> parent) : Node<T>({parent}) {}
 };
+}  // namespace datastructures

### 2019-10-25 C++/datastructures/triangulation.cpp

@@ -1,0 +1,162 @@
+#include "triangulation.hpp"
+
+namespace space {
+
+ArrayVertexPtr<2> Element2D::edge(int i) const {
+  assert(0 <= i && i <= 2);
+  return {{vertices_[(i + 1) % 3], vertices_[(i + 2) % 3]}};
+}
+
+ArrayVertexPtr<2> Element2D::reversed_edge(int i) const {
+  assert(0 <= i && i <= 2);
+  return {{vertices_[(i + 2) % 3], vertices_[(i + 1) % 3]}};
+}
+
+const VectorElement2DPtr &Element2D::refine() {
+  if (!is_full()) {
+    auto nbr = neighbours[0];
+    if (!nbr) {  // Refinement edge of `elem` is on domain boundary
+      bisect();
+    } else if (nbr->edge(0) != reversed_edge(0)) {
+      nbr->refine();
+      return refine();
+    } else {
+      bisect_with_nbr();
+    }
+  }
+  return children_;
+}
+
+VertexPtr Element2D::create_new_vertex(Element2DPtr nbr) {
+  assert(is_leaf());
+  VectorVertexPtr vertex_parents{newest_vertex()};
+  if (nbr) {
+    assert(nbr->edge(0) == reversed_edge(0));
+    vertex_parents.emplace_back(nbr->newest_vertex());
+  }
+
+  ArrayVertexPtr<2> godparents{{vertices_[1], vertices_[2]}};
+  auto new_vertex =
+      std::make_shared<Vertex>((godparents[0]->x + godparents[1]->x) / 2,
+                               (godparents[0]->y + godparents[1]->y) / 2,
+                               nbr == nullptr, vertex_parents);
+  for (auto vertex_parent : vertex_parents)
+    vertex_parent->children().push_back(new_vertex);
+  return new_vertex;
+}
+
+ArrayElement2DPtr<2> Element2D::bisect(VertexPtr new_vertex) {
+  assert(is_leaf());
+  if (!new_vertex) {
+    new_vertex = create_new_vertex();
+  }
+  auto child1 = std::make_shared<Element2D>(
+      shared_from_this(),
+      VectorVertexPtr{{new_vertex, vertices_[0], vertices_[1]}});
+  auto child2 = std::make_shared<Element2D>(
+      shared_from_this(),
+      VectorVertexPtr{{new_vertex, vertices_[2], vertices_[0]}});
+  children_ = {{child1, child2}};
[…]

### 2019-10-25 C++/datastructures/triangulation.hpp

@@ -6,6 +6,8 @@
 #include <vector>
 #include "tree.hpp"
 
+namespace space {
+
 class Vertex;
 using VertexPtr = std::shared_ptr<Vertex>;
 using VectorVertexPtr = std::vector<VertexPtr>;
@@ -17,116 +19,33 @@ using Element2DPtr = std::shared_ptr<Element2D>;
 using VectorElement2DPtr = std::vector<Element2DPtr>;
 template <size_t N>
 using ArrayElement2DPtr = std::array<Element2DPtr, N>;
-std::ostream &operator<<(std::ostream &os, const Element2D &elem);
 
-class Vertex : public Node<Vertex> {
+class Vertex : public datastructures::Node<Vertex> {
  public:
   const double x, y;
   bool on_domain_boundary;
   VectorElement2DPtr patch;
 
-  Vertex() : Node(), x(-1), y(-1), on_domain_boundary(false) {}
+  static VertexPtr CreateMetaroot() { return VertexPtr(new Vertex()); }
   Vertex(double x, double y, bool on_domain_boundary,
          const VectorVertexPtr &parents)
       : Node(parents), x(x), y(y), on_domain_boundary(on_domain_boundary) {}
+
+  friend std::ostream &operator<<(std::ostream &os, const Vertex &vertex) {
+    os << "V(" << vertex.x << ", " << vertex.y << ")";
+    return os;
+  }
+
+ protected:
+  Vertex() : Node(), x(-1), y(-1), on_domain_boundary(false) {}
 };
 
-std::ostream &operator<<(std::ostream &os, const Vertex &vertex) {
-  os << "V(" << vertex.x << ", " << vertex.y << ")";
-  return os;
-}
-
-class InitialTriangulation;
-
-class Element2D : public BinaryNode<Element2D> {
- protected:
-  double area_;
-  VectorVertexPtr vertices_;
-
-  VertexPtr create_new_vertex(Element2DPtr nbr = nullptr) {
-    assert(is_leaf());
-    VectorVertexPtr vertex_parents{newest_vertex()};
-    if (nbr) {
-      assert(nbr->edge(0) == reversed_edge(0));
-      vertex_parents.emplace_back(nbr->newest_vertex());
-    }
-
-    ArrayVertexPtr<2> godparents{{vertices_[1], vertices_[2]}};
-    auto new_vertex =
-        std::make_shared<Vertex>((godparents[0]->x + godparents[1]->x) / 2,
[…]

### 2019-10-25 C++/datastructures/triangulation_test.cpp

@@ -1,7 +1,126 @@
 #include "triangulation.hpp"
+#include <cmath>
+#include <map>
+#include <set>
+#include "gmock/gmock.h"
+#include "gtest/gtest.h"
 
+namespace space {
+using namespace testing;
+
+TEST(Triangulation, Refine) {
+  auto T = InitialTriangulation::UnitSquare();
+  T.elem_meta_root->UniformRefine(1);
+
+  auto elements = T.elem_meta_root->Bfs();
+  auto vertices = T.vertex_meta_root->Bfs();
+  EXPECT_THAT(vertices[4]->parents(), ElementsAre(vertices[0], vertices[1]));
+  EXPECT_THAT(vertices[0]->children(), ElementsAre(vertices[4]));
+  EXPECT_THAT(vertices[1]->children(), ElementsAre(vertices[4]));
+
+  elements[2]->refine();
+  elements = T.elem_meta_root->Bfs();
+  vertices = T.vertex_meta_root->Bfs();
+  ASSERT_TRUE(vertices[5]->on_domain_boundary);
+  EXPECT_THAT(vertices[5]->parents(), ElementsAre(vertices[4]));
+  EXPECT_THAT(vertices[4]->children(), ElementsAre(vertices[5]));
+
+  elements[4]->refine();
+  elements = T.elem_meta_root->Bfs();
+  vertices = T.vertex_meta_root->Bfs();
+  ASSERT_TRUE(vertices[6]->on_domain_boundary);
+  EXPECT_THAT(vertices[6]->parents(), ElementsAre(vertices[4]));
+  EXPECT_THAT(vertices[4]->children(), ElementsAre(vertices[5], vertices[6]));
+
+  elements[6]->refine();
+  elements = T.elem_meta_root->Bfs();
+  vertices = T.vertex_meta_root->Bfs();
+  ASSERT_FALSE(vertices[8]->on_domain_boundary);
+  EXPECT_THAT(vertices[8]->parents(), ElementsAre(vertices[5], vertices[7]));
+}
+
+TEST(Triangulation, OnDomainBoundary) {
+  auto init_triang = InitialTriangulation::UnitSquare();
+  for (auto vertex : init_triang.vertex_meta_root->Bfs()) {
+    ASSERT_TRUE(vertex->on_domain_boundary);
+  }
+
+  auto element_roots = init_triang.elem_meta_root->Bfs();
+  auto leaves =
+      std::set<Element2DPtr>(element_roots.begin(), element_roots.end());
[…]

### 2019-10-25 C++/datastructures/triangulation_test.cpp
removed some crap

@@ -6,7 +6,7 @@
 #include "gtest/gtest.h"
 
 namespace space {
-using namespace testing;
+using ::testing::ElementsAre;
 
 TEST(Triangulation, Refine) {
   auto T = InitialTriangulation::UnitSquare();
@@ -117,28 +117,4 @@ TEST(Triangulation, VertexPatch) {
       ASSERT_EQ(v->patch.size(), 4);
   }
 }
-
-/*
-int main() {
-  auto T = InitialTriangulation::UnitSquare();
-  auto elems = T.elem_meta_root->Bfs();
-
-  std::cout << "Elements in BFS before refine" << std::endl;
-  for (auto elem : elems) {
-    std::cout << *elem << ", ";
-  }
-  std::cout << std::endl;
-
-  T.elem_meta_root->UniformRefine(2);
-  elems = T.elem_meta_root->Bfs();
-  assert(elems.size() == 2 + 4 + 8);
-  std::cout << "Elements in BFS after refine" << std::endl;
-  for (auto elem : elems) {
-    std::cout << *elem << ", ";
-  }
-  std::cout << std::endl;
-
-  return 0;
-}
-*/
 }  // namespace space

### 2019-10-25 C++/datastructures/tree.hpp
improved some stuff

@@ -61,21 +61,7 @@ virtual const std::vector<std::shared_ptr<T>> &parents() const = 0;
 
 template <typename T>
 class Node : public NodeInterface<T> {
- protected:
-  int level_;
-  bool marked_ = false;
-  std::vector<std::shared_ptr<T>> parents_;
-  std::vector<std::shared_ptr<T>> children_;
-  Node() : level_(-1) {}
-
  public:
-  int level() const { return level_; }
-  bool marked() const { return marked_; }
-  void set_marked(bool value) { marked_ = value; }
-  bool is_metaroot() const { return (level_ == -1); }
-  const std::vector<std::shared_ptr<T>> &parents() const { return parents_; }
-  std::vector<std::shared_ptr<T>> &children() { return children_; }
-
   explicit Node(const std::vector<std::shared_ptr<T>> &parents)
       : parents_(parents) {
     assert(parents.size());
@@ -84,18 +70,32 @@ class Node : public NodeInterface<T> {
       assert(parent->level() == level_ - 1);
     }
   }
+
+  int level() const { return level_; }
+  bool marked() const { return marked_; }
+  void set_marked(bool value) { marked_ = value; }
+  bool is_metaroot() const { return (level_ == -1); }
+  const std::vector<std::shared_ptr<T>> &parents() const { return parents_; }
+  std::vector<std::shared_ptr<T>> &children() { return children_; }
+
+ protected:
+  int level_;
+  bool marked_ = false;
+  std::vector<std::shared_ptr<T>> parents_;
+  std::vector<std::shared_ptr<T>> children_;
+  Node() : level_(-1) {}
 };
 
 template <typename T>
 class BinaryNode : public Node<T> {
- protected:
-  using Node<T>::Node;
-
  public:
+  explicit BinaryNode(std::shared_ptr<T> parent) : Node<T>({parent}) {}
+
   bool is_full() const { return Node<T>::children_.size() == 2; }
   bool is_leaf() const { return Node<T>::children_.size() == 0; }
   std::shared_ptr<T> parent() const { return Node<T>::parents_[0]; }
 
-  explicit BinaryNode(std::shared_ptr<T> parent) : Node<T>({parent}) {}
+ protected:
+  using Node<T>::Node;
 };
 }  // namespace datastructures

### 2019-10-25 C++/datastructures/triangulation.cpp

@@ -153,6 +153,7 @@ InitialTriangulation::InitialTriangulation(
       }
   }
 }
+
 InitialTriangulation InitialTriangulation::UnitSquare() {
   std::vector<std::array<double, 2>> vertices = {
       {{0, 0}}, {{1, 1}}, {{1, 0}}, {{0, 1}}};

### 2019-10-25 C++/datastructures/tree.hpp
minor comment

@@ -91,11 +91,13 @@ class BinaryNode : public Node<T> {
  public:
   explicit BinaryNode(std::shared_ptr<T> parent) : Node<T>({parent}) {}
 
-  bool is_full() const { return Node<T>::children_.size() == 2; }
-  bool is_leaf() const { return Node<T>::children_.size() == 0; }
-  std::shared_ptr<T> parent() const { return Node<T>::parents_[0]; }
+  bool is_full() const { return children_.size() == 2; }
+  bool is_leaf() const { return children_.size() == 0; }
+  std::shared_ptr<T> parent() const { return parents_[0]; }
 
  protected:
   using Node<T>::Node;
+  using Node<T>::parents_;
+  using Node<T>::children_;
 };
 }  // namespace datastructures

### 2019-10-25 C++/datastructures/triangulation.hpp
use nan

@@ -37,7 +37,7 @@ class Vertex : public datastructures::Node<Vertex> {
   }
 
  protected:
-  Vertex() : Node(), x(-1), y(-1), on_domain_boundary(false) {}
+  Vertex() : Node(), x(NAN), y(NAN), on_domain_boundary(false) {}
 };
 
 class Element2D : public datastructures::BinaryNode<Element2D> {

### 2019-10-25 C++/datastructures/triangulation_test.cpp
added area test

@@ -39,6 +39,23 @@ TEST(Triangulation, Refine) {
   EXPECT_THAT(vertices[8]->parents(), ElementsAre(vertices[5], vertices[7]));
 }
 
+TEST(Triangulation, Area) {
+  auto T = InitialTriangulation::UnitSquare();
+  T.elem_meta_root->UniformRefine(2);
+
+  auto elements = T.elem_meta_root->Bfs();
+  auto vertices = T.vertex_meta_root->Bfs();
+  ASSERT_EQ(elements[0]->area(), 0.5);
+  ASSERT_EQ(elements[1]->area(), 0.5);
+
+  ASSERT_EQ(elements[2]->area(), 0.25);
+  ASSERT_EQ(elements[3]->area(), 0.25);
+  ASSERT_EQ(elements[4]->area(), 0.25);
+  ASSERT_EQ(elements[5]->area(), 0.25);
+
+  ASSERT_EQ(elements[6]->area(), 0.125);
+}
+
 TEST(Triangulation, OnDomainBoundary) {
   auto init_triang = InitialTriangulation::UnitSquare();
   for (auto vertex : init_triang.vertex_meta_root->Bfs()) {

### 2019-10-25 Python/space/triangulation_test.py
added silly test

@@ -5,6 +5,19 @@ import numpy as np
 from .triangulation import InitialTriangulation
 
 
+def test_area():
+    init_triang = InitialTriangulation.unit_square()
+    assert all(v.on_domain_boundary for v in init_triang.vertex_roots)
+    init_triang.element_roots[0].refine()
+
+    leaves = set(init_triang.element_roots)
+    for _ in range(100):
+        elem = leaves.pop()
+        leaves.update(elem.refine())
+    for elem in init_triang.elem_meta_root.bfs():
+        assert elem.area == 0.5**(elem.level + 1)
+
+
 def test_on_domain_bdr():
     init_triang = InitialTriangulation.unit_square()
     assert all(v.on_domain_boundary for v in init_triang.vertex_roots)

### 2019-10-25 C++/space/triangulation.hpp
build hell

@@ -4,7 +4,7 @@
 #include <iostream>
 #include <memory>
 #include <vector>
-#include "tree.hpp"
+#include "../datastructures/tree.hpp"
 
 namespace space {
 

### 2019-10-25 C++/space/triangulation.hpp
build hell fixed

@@ -4,7 +4,7 @@
 #include <iostream>
 #include <memory>
 #include <vector>
-#include "../datastructures/tree.hpp"
+#include "datastructures/tree.hpp"
 
 namespace space {
 

### 2019-10-25 C++/datastructures/tree.hpp
removed virtual methods comment

@@ -8,16 +8,6 @@
 namespace datastructures {
 template <typename T>
 class NodeInterface : public std::enable_shared_from_this<T> {
-  /**
-virtual int level() const = 0;
-virtual bool marked() = 0;
-virtual void set_marked(bool value) = 0;
-virtual bool is_full() const = 0;
-virtual bool is_metaroot() const = 0;
-virtual const std::vector<std::shared_ptr<T>> &children() const = 0;
-virtual const std::vector<std::shared_ptr<T>> &parents() const = 0;
-**/
-
  public:
   std::vector<std::shared_ptr<T>> Bfs(
       bool include_metaroot, std::function<void(std::shared_ptr<T>)> callback) {

### 2019-10-25 Output/error_reduction_analysis.py
some  small analyses

@@ -1,0 +1,96 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+from itertools import product
+import numpy as np
+import os
+
+
+def make_time_slices():
+    plt.figure()
+    plt.title(r"Time-slice error at various time steps")
+    plt.xlabel(r"$\dim X_\delta$")
+    plt.ylabel(r"$||\gamma_t(u - u_\delta)||_{L_2(\Omega)}$")
+    for i in range(0, 9, 2):
+        plt.loglog(data['dofs'], [err[i] for err in data['errors']],
+                   label='t=%g' % (i / 8))
+    plt.legend()
+    plt.grid()
+    plt.tight_layout()
+    plt.savefig("error_reduction_time_slices.png")
+
+
+def make_rates():
+    plt.figure()
+    plt.title(
+        "Average error reduction rates as function of start- and end-index")
+    plt.xlabel(r"$t$")
+    plt.ylabel(r"Average rate")
+
+    np.linspace(0, 1, 9)
+    len_dofs = len(data['dofs'])
+    for start, end in product([0, 1], [len_dofs, len_dofs - 1, len_dofs - 2]):
+        rates = [
+            -np.polyfit(np.log(data['dofs'])[start:end],
+                        np.log([err[i] for err in data['errors']])[start:end],
+                        deg=1)[0] for i in range(9)
+        ]
+        plt.plot(np.linspace(0, 1, 9),
+                 rates,
+                 label="data[%g:%g]" % (start, end))
+    plt.legend()
+    plt.tight_layout()
+    plt.savefig("error_reduction_rates.png")
+
+
+def make_linearity():
+    fig, (ax1, ax2) = plt.subplots(2, 1, sharex=True)
+    ax1.set_title("Time per apply")
+    ax1.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    ax1.set_ylabel("time (s)")
+    ax1.loglog(
+        data['dofs'],
+        [dofs * t for (dofs, t) in zip(data['dofs'], data['time_per_dof'])])
+    ax1.tick_params(axis='y')
+    ax1.grid()
+
+    ax2.set_title("Time per apply per dof")
+    ax1.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    ax2.set_ylabel("time (ms)")
+    ax2.loglog(data['dofs'], [1000 * t for t in data['time_per_dof']])
+    ax2.tick_params(axis='y')
[…]

### 2019-10-25 Python/applications/heat_equation_test.py

@@ -238,8 +238,12 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
     n_t = 9
     errors_quad = []
     ndofs = []
+    dims = []
     time_per_dof = []
     rates_quad = []
+    residual_norm_histories = []
+    residual_norms = []
+    minres_iters = []
     for level in range(2, max_level):
         # Create X^\delta as a sparse grid.
         X_delta = DoubleTree.from_metaroots(
@@ -248,25 +252,38 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
         print('X_delta: dofs time axis={}\tdofs space axis={}'.format(
             len(X_delta.project(0).bfs()), len(X_delta.project(1).bfs())))
 
+        # Create heat equation object.
+        heat_eq = HeatEquation(X_delta=X_delta)
+        rhs = example_rhs(heat_eq)
+
+        if level < 10:
+            residual_norm_history = []
+            callback = lambda vec: residual_norm_history.append(
+                np.linalg.norm(heat_eq.linop @ vec - rhs.to_array()))
+        else:
+            callback = None
+        # Now actually solve this beast!
+        sol, num_iters = heat_eq.solve(rhs, iter_callback=callback)
+
         # Count number of dofs (not on the boundary!)
         ndofs.append(
             len([
                 n for n in X_delta.bfs() if not n.nodes[1].on_domain_boundary
             ]))
 
-        # Create heat equation object.
-        heat_eq = HeatEquation(X_delta=X_delta)
-        rhs = example_rhs(heat_eq)
-
-        # Now actually solve this beast!
-        sol, num_iters = heat_eq.solve(rhs)
+        # Record some stuff for posterity.
+        dims.append(len(heat_eq.Y_delta.bfs()), len(X_delta.bfs()))
+        residual_norm_histories.append(residual_norm_history)
+        minres_iters.append(num_iters)
         residual_norm = np.linalg.norm(
             heat_eq.mat.apply(sol).to_array() - rhs.to_array())
+        residual_norms.append(residual_norm)
+        time_per_dof.append(heat_eq.time_per_dof())
+
         print('MINRES solved in {} iterations with a residual norm {}'.format(
             num_iters, residual_norm))
         print('Time per dof is approximately {}'.format(
             heat_eq.time_per_dof()))
-        time_per_dof.append(heat_eq.time_per_dof())
 
         u, u_order, u_slice_norm = example_solution_function()
 
[…]

### 2019-10-25 Python/applications/heat_equation_test.py
more

@@ -272,7 +272,7 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
             ]))
 
         # Record some stuff for posterity.
-        dims.append(len(heat_eq.Y_delta.bfs()), len(X_delta.bfs()))
+        dims.append([len(heat_eq.Y_delta.bfs()), len(X_delta.bfs())])
         residual_norm_histories.append(residual_norm_history)
         minres_iters.append(num_iters)
         residual_norm = np.linalg.norm(

### 2019-10-25 Output/error_reduction_analysis.tex
first pass at a quick report

\documentclass[11pt,a4paper]{amsart}

\usepackage{amsmath, amssymb}
\usepackage{dsfont}
\usepackage{geometry}
\usepackage{commath}
\usepackage{url}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{todonotes}
%\setlength{\parindent}{0pt}
\geometry{%
  includeheadfoot,
  margin=2cm
}

\newtheorem*{problem}{Problem}
\newtheorem*{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem*{answer}{Answer}

\newcommand{\bbone}{\mathds{1}}
\DeclareMathOperator*{\diam}{diam}
\DeclareMathOperator*{\spann}{span}
\DeclareMathOperator*{\init}{init}
\DeclareMathOperator*{\ext}{ext}
\newcommand{\udlrarrow}{\mathrlap{\,{\scriptscriptstyle \updownarrow}}{\scriptscriptstyle \leftrightarrow}}

%\input{/Users/janwesterdiep/Dropbox/Commands.tex}
\newcommand{\T}{{\mathcal T}}
\newcommand{\R}{\mathbb{R}}

\date{\today}
\begin{document}

\title{First look at Spacetime}
\maketitle

Let me keep it short. We did a first run of the spacetime project solving the
heat equation on the unit square $\Omega = [0,1]^2$ on the time interval
$I := [0,1]$. The true solution is 
\[
  u(t,x,y) := (t^2 + 1) (1-x)x(1-y)y.
\]

Given some level $L$, we build $X_\delta$ as the double tree consisting of all
double nodes $\lambda = (\lambda_{\text{time}}, \lambda_{\text{space}})$ for
which $2|\lambda_{\text{time}}| + |\lambda_{\text{space}}| \leq L$ (a \emph{sparse grid}).
We then build $Y_\delta$ through the formula halfway page 6 of \texttt{followup.pdf}.
We use the system on the bottom of page 6, and compute the data integrals in the
right-hand side using Gauss quadrature of sufficient order to be exact.

Denote the $2 \times 2$ block-matrix with $M$. We use MINRES with initial guess $x_0$
equal to the right-hand side $b$ of the system, and continue until the relative residual is `small':
\[
  \frac{\|Mx_k - b\|_2}{\|x_k\|_2} \leq 10^{-5}.
\]
This gives us a discrete solution $(\mu_\delta, u_\delta)$ for which we can measure
\[
  E(t) := t \mapsto \|\gamma_t(u_\delta - u)\|_{L_2(\Omega)}.
\]
We will use this as a proxy for the error $\|u - u_\delta\|_X$ which is difficult to measure.

Given $X_\delta$, we distinguish its dimension $\dim X_\delta$ and $\# X_\delta$,
being the total number of \emph{degrees of freedom}, i.e., the dimension of $X_\delta$
minus the number of Dirichlet double nodes (double nodes for which the vertex of
$\lambda_{\text{space}}$ is not on the domain boundary).

\section*{Convergence rate}
The code right now seems to solve the problem, as can be seen by Figure~\ref{fig:time-slice}:
Increasing $L$ (thereby increasing $\# X_\delta$, yields an error reduction.
We discern a pre-asymptotic regime: the error hardly decreases between levels $L=1$
and $2$. Moreover, errors seem to taper off near the final levels $L$, which may
have something to do with our solver tolerance.

In fact, the errors decay at an algebraic rate somewhere between $0.75$ and $0.85$,
depending on the time $t$ at which the errors are measured, as well as whether or
not we keep the full range of errors when calculating the rate (for instance, in
the brown line, we slice away level $1$ because it is in the pre-asymptotic regime,
and levels $11$ and $12$ because the convergence stagnates in this regime).
\begin{figure}[h!]
  \includegraphics[width=0.495\linewidth]{error_reduction_time_slices.pdf}
  \includegraphics[width=0.495\linewidth]{error_reduction_rates.pdf}
  \caption{Left: Time slice errors. Right: rates.}
  \label{fig:time-slice}
\end{figure}

\section*{Computational cost}
Looking at computation time, we measure the time it takes to perform a single
application of the system matrix, not taking into account the setup cost (like
computing $\Sigma$ and $\Theta$). Figure~\ref{fig:linearity} paints a slightly
less enthousiastic picture, but it is obvious that we need more data points
before concluding anything.\todo{BV: meten we nou per DoF, of per dimensie?}
\begin{figure}[h!]
  \includegraphics[width=0.8\linewidth]{error_reduction_linearity.pdf}
  \caption{Top: time (s) per apply; bottom: time (ms) per apply per DoF.}
  \label{fig:linearity}
\end{figure}

\end{document}

### 2019-10-28 Output/error_reduction_analysis.tex
wrote some stuff

@@
 \section*{Computational cost}
-Looking at computation time, we measure the time it takes to perform a single
-application of the system matrix, not taking into account the setup cost (like
+Let's measure the time it takes to perform a single application of the system
+matrix. We do not take into account the setup cost (like
 computing $\Sigma$ and $\Theta$). Figure~\ref{fig:linearity} paints a slightly
-less enthousiastic picture, but it is obvious that we need more data points
-before concluding anything.\todo{BV: meten we nou per DoF, of per dimensie?}
+demotivating picture, in that the bottom graph is not constant. We do have to
+place an asterisk here: for small systems, the produced numbers are averaged over
+a small number of iterations, so they change a lot from run to run.
 \begin{figure}[h!]
@@
 
+\section*{The solver}
+As mentioned before, we use MINRES with default parameters (tolerance $10^{-5}$,
+initial guess $x_0$ equal to the RHS $b$). In Figure~\ref{fig:minres}, we show
+the residual norm $\|M x_k - b\|_2$ as a function of $k$ for a selection of solves.
+\begin{figure}[h!]
+  \includegraphics[width=0.495\linewidth]{error_reduction_minres_history_semilogy.pdf}
+  \includegraphics[width=0.495\linewidth]{error_reduction_minres_history_loglog.pdf}
+  \caption{Left: semi-log-y MINRES convergence graph for some solves. Right: Same results in log-log format.}
+  \label{fig:minres}
+\end{figure}
+We see that the number of iterations increases strongly, suggesting that we could
+definitely use a good preconditioner.
+
 \end{document}

### 2019-10-28 Output/error_reduction_analysis.py

@@ -53,11 +53,12 @@ def make_linearity():
     ax1.tick_params(axis='y')
     ax1.grid()
 
-    ax2.set_title("Time per apply per dof")
-    ax1.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    ax2.set_title("Time per apply per dim")
+    ax2.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
     ax2.set_ylabel("time (ms)")
-    ax2.loglog(data['dofs'], [1000 * t for t in data['time_per_dof']])
+    ax2.semilogx(data['dofs'], [1000 * t for t in data['time_per_dof']])
     ax2.tick_params(axis='y')
+    ax2.set_ylim((0, 3.0))
     ax2.grid()
     plt.tight_layout()
     plt.savefig("error_reduction_linearity.pdf")
@@ -65,27 +66,37 @@ def make_linearity():
 
 def make_minres():
     histories = data['residual_norm_histories']
+    histories = histories[:8]
 
     plt.figure()
-    plt.plot([len(hist) for hist in histories])
+    plt.xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    plt.ylabel(r"MINRES iterations")
+    plt.loglog(data['dofs'], data['minres_iters'])
+    plt.grid()
     plt.savefig("error_reduction_minres_iters.pdf")
 
     plt.figure()
     plt.ylim(1e-5, 1e-0)
+    plt.ylabel(r"$||M x_k - b||_2$")
+    plt.xlabel(r"MINRES iteration")
     for i, hist in enumerate(histories):
         plt.semilogy(hist, label="solve %s" % (i + 1))
+    plt.grid()
     plt.legend()
     plt.savefig("error_reduction_minres_history_semilogy.pdf")
 
     plt.figure()
     plt.ylim(1e-5, 1e-0)
+    plt.ylabel(r"$||M x_k - b||_2$")
+    plt.xlabel(r"MINRES iteration")
     for i, hist in enumerate(histories):
         plt.loglog(hist, label="solve %s" % (i + 1))
+    plt.grid()
     plt.legend()
     plt.savefig("error_reduction_minres_history_loglog.pdf")
 
 
-data = pd.read_pickle("error_reduction.pickle")
+data = pd.read_pickle("error_reduction_v2.pickle")
 print(data)
 
 make_time_slices()

### 2019-10-28 Python/applications/heat_equation_test.py

@@ -273,7 +273,8 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
 
         # Record some stuff for posterity.
         dims.append([len(heat_eq.Y_delta.bfs()), len(X_delta.bfs())])
-        residual_norm_histories.append(residual_norm_history)
+        if level < 10:
+            residual_norm_histories.append(residual_norm_history)
         minres_iters.append(num_iters)
         residual_norm = np.linalg.norm(
             heat_eq.mat.apply(sol).to_array() - rhs.to_array())

### 2019-10-28 Output/error_reduction_analysis.tex
small changes in the document

@@
 
+\newcommand{\inp}[2]{\langle #1, #2 \rangle}
 \newcommand{\bbone}{\mathds{1}}
@@
 Denote the $2 \times 2$ block-matrix with $M$. We use MINRES with initial guess $x_0$
-equal to the right-hand side $b$ of the system, and continue until the relative residual is `small':
+equal to the right-hand side $b$ of the system, and continue until the relative residual norm is `small':
 \[
-  \frac{\|Mx_k - b\|_2}{\|x_k\|_2} \leq 10^{-5}.
+  \frac{\|Mx_k - b\|_2}{\|b\|_2} \leq 10^{-5}.
 \]
@@
 \[
-  E(t) := t \mapsto \|\gamma_t(u_\delta - u)\|_{L_2(\Omega)}.
+  E(t) := t \mapsto \|\gamma_t(u_\delta - u)\|_{L_2(\Omega)}
 \]
-We will use this as a proxy for the error $\|u - u_\delta\|_X$ which is difficult to measure.
+by computing
+\[
+  E(t)^2 = \inp{u(t)}{u(t)}_{L_2(\Omega)} + \inp{u_\delta(t)}{u_\delta(t)}_{L_2(\Omega)} - 2 \inp{u(t)}{u_\delta(t)}_{L_2(\Omega)}
+\]
+where we have the first term analytically, the second number can be computed by
+an applicator of the space-time mass matrix, and the final term is approximated
+using quadrature of sufficient degree. We will use this as a proxy for the error
+$\|u - u_\delta\|_X$, which is difficult to measure.
 
-Given $X_\delta$, we distinguish its dimension $\dim X_\delta$ and $\# X_\delta$,
-being the total number of \emph{degrees of freedom}, i.e., the dimension of $X_\delta$
-minus the number of Dirichlet double nodes (double nodes for which the vertex of
-$\lambda_{\text{space}}$ is not on the domain boundary).
+Given $X_\delta$, we distinguish the dimension $\dim X_\delta$ of the function space
+and $\# X_\delta$, being the total number of double nodes in the double tree.
 
@@
 The code right now seems to solve the problem, as can be seen by Figure~\ref{fig:time-slice}:
-Increasing $L$ (thereby increasing $\# X_\delta$, yields an error reduction.
+Increasing $L$ (thereby increasing $\dim X_\delta$, yields an error reduction.
 We discern a pre-asymptotic regime: the error hardly decreases between levels $L=1$
@@
 \begin{figure}[h!]
-  \includegraphics[width=0.8\linewidth]{error_reduction_linearity.pdf}
+  \includegraphics[width=0.6\linewidth]{error_reduction_linearity.pdf}
   \caption{Top: time (s) per apply; bottom: time (ms) per apply per DoF.}
@@
 initial guess $x_0$ equal to the RHS $b$). In Figure~\ref{fig:minres}, we show
-the residual norm $\|M x_k - b\|_2$ as a function of $k$ for a selection of solves.
+the \emph{absolute} residual norm $\|M x_k - b\|_2$ as a function of $k$ for a
+selection of solves.
 \begin{figure}[h!]

### 2019-10-28 Output/error_reduction_analysis.py

@@ -7,7 +7,7 @@ import numpy as np
 def make_time_slices():
     plt.figure()
     plt.title(r"Time-slice error at various time steps")
-    plt.xlabel(r"#$X_\delta$")
+    plt.xlabel(r"$\dim X_\delta$")
     plt.ylabel(r"$||\gamma_t(u - u_\delta)||_{L_2(\Omega)}$")
     for i in range(0, 9, 2):
         plt.loglog(data['dofs'], [err[i] for err in data['errors']],
@@ -45,7 +45,7 @@ def make_rates():
 def make_linearity():
     fig, (ax1, ax2) = plt.subplots(2, 1, sharex=True)
     ax1.set_title("Time per apply")
-    ax1.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    ax1.set_xlabel(r"$\# X_\delta + \# Y_\delta$")
     ax1.set_ylabel("time (s)")
     ax1.loglog(data['dofs'], [
         sum(dims) * t for (dims, t) in zip(data['dims'], data['time_per_dof'])
@@ -53,8 +53,8 @@ def make_linearity():
     ax1.tick_params(axis='y')
     ax1.grid()
 
-    ax2.set_title("Time per apply per dim")
-    ax2.set_xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    ax2.set_title("Time per apply per doublenode")
+    ax2.set_xlabel(r"$\# X_\delta + \# Y_\delta$")
     ax2.set_ylabel("time (ms)")
     ax2.semilogx(data['dofs'], [1000 * t for t in data['time_per_dof']])
     ax2.tick_params(axis='y')
@@ -69,7 +69,7 @@ def make_minres():
     histories = histories[:8]
 
     plt.figure()
-    plt.xlabel(r"$\dim X_\delta + \dim Y_\delta$")
+    plt.xlabel(r"$\# X_\delta + \# Y_\delta$")
     plt.ylabel(r"MINRES iterations")
     plt.loglog(data['dofs'], data['minres_iters'])
     plt.grid()

### 2019-10-28 Python/applications/heat_equation_test.py

@@ -220,7 +220,9 @@ def test_heat_eq_linear():
         assert np.allclose(heat_eq.linop.matvec(v_arr), heat_eq_mat.dot(v_arr))
 
 
-def test_heat_error_reduction(max_level=6, save_results_file=None):
+def test_heat_error_reduction(max_history_level=0,
+                              max_level=6,
+                              save_results_file=None):
     # Printing options.
     np.set_printoptions(precision=4)
     np.set_printoptions(linewidth=10000)
@@ -256,7 +258,7 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
         heat_eq = HeatEquation(X_delta=X_delta)
         rhs = example_rhs(heat_eq)
 
-        if level < 10:
+        if level <= max_history_level:
             residual_norm_history = []
             callback = lambda vec: residual_norm_history.append(
                 np.linalg.norm(heat_eq.linop @ vec - rhs.to_array()))
@@ -273,7 +275,7 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
 
         # Record some stuff for posterity.
         dims.append([len(heat_eq.Y_delta.bfs()), len(X_delta.bfs())])
-        if level < 10:
+        if level <= max_history_level:
             residual_norm_histories.append(residual_norm_history)
         minres_iters.append(num_iters)
         residual_norm = np.linalg.norm(
@@ -350,6 +352,7 @@ def test_heat_error_reduction(max_level=6, save_results_file=None):
 
 if __name__ == "__main__":
     test_heat_error_reduction(
+        max_history_level=9,
         max_level=16,
         save_results_file='error_reduction.pickle',
     )

### 2019-10-28 Output/error_reduction_analysis.tex
final change

@@
 where we have the first term analytically, the second number can be computed by
-an applicator of the space-time mass matrix, and the final term is approximated
+an applicator of the space-time mass matrix, and the final term is found
 using quadrature of sufficient degree. We will use this as a proxy for the error

### 2019-10-29 C++/datastructures/multi_tree_view.hpp
initial draft

@@ -1,0 +1,196 @@
+#pragma once
+#include <experimental/tuple>
+#include <memory>
+#include <numeric>
+#include <tuple>
+#include <utility>
+#include <vector>
+#include "tree.hpp"
+
+namespace datastructures {
+
+template <typename Func, std::size_t... Is>
+constexpr void static_for_impl(Func&& f, std::index_sequence<Is...>) {
+  (f(std::integral_constant<std::size_t, Is>{}), ...);
+}
+
+template <size_t N, typename Func>
+constexpr void static_for(Func&& f) {
+  static_for_impl(std::forward<Func>(f), std::make_index_sequence<N>{});
+}
+
+template <typename I, size_t dim>
+class MultiNodeViewInterface {
+ public:
+  int level() const {
+    const auto& nodes = static_cast<const I*>(this)->nodes();
+    // size_t result = 0;
+    // static_for<dim>([&result, &nodes](auto index) {
+    //  result += std::get<index>(nodes)->level();
+    //});
+    // return result;
+
+    return std::experimental::apply(
+        [](auto const&... e) -> decltype(auto) { return (e->level() + ...); },
+        nodes);
+  }
+
+  template <size_t i>
+  bool IsFull() const {
+    auto self = static_cast<const I*>(this);
+    return self->children(i).size() ==
+           std::get<i>(self->nodes())->children().size();
+  }
+
+  template <typename T>
+  std::shared_ptr<I> FindBrother(const T& nodes, size_t j, size_t i) {
+    auto self = static_cast<I*>(this);
+    if (nodes == self->nodes()) {
+      return self->shared_from_this();
+    }
+
+    for (const auto& parent_j : self->parents(j)) {
+      for (const auto& sibling_i : parent_j->children(i)) {
+        if (sibling_i->nodes() == nodes) {
+          return sibling_i;
+        }
+      }
+    }
+    assert(false);
+  }
+
+  template <size_t j, typename T>
+  void FindBrothers(const T& child_nodes, size_t i,
+                    std::array<std::vector<std::shared_ptr<I>>, dim>& result) {
+    for (auto child_parent_j : std::get<j>(child_nodes)->parents()) {
+      auto brother_nodes{child_nodes};  // Create a copy
+      std::get<j>(brother_nodes) = child_parent_j;
+      result[j].push_back(FindBrother(brother_nodes, j, i));
+    }
+
+    if constexpr (j > 0) {
+      FindBrothers<j - 1>(child_nodes, i, result);
+    }
+  }
+
+  template <typename T>
[…]

### 2019-10-29 C++/space/triangulation_view.cpp

@@ -1,0 +1,21 @@
+#include "datastructures/multi_tree_view.hpp"
+#include "triangulation.hpp"
+
+using namespace space;
+
+int main() {
+  auto T = InitialTriangulation::UnitSquare();
+  T.elem_meta_root->UniformRefine(1);
+
+  datastructures::NodeView<Vertex> root_view(T.vertex_meta_root);
+  std::shared_ptr<Vertex> child = T.vertex_meta_root->children()[0];
+  auto child_view = std::make_shared<datastructures::NodeView<Vertex>>(child);
+
+  root_view.children(0).push_back(child_view);
+
+  int x = child_view->level();
+
+  std::cout << x << std::endl;
+  root_view.Refine<0>();
+  return 0;
+}

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
first draft

@@ -1,17 +1,19 @@
 #pragma once
-#include <experimental/tuple>
 #include <memory>
 #include <numeric>
+#include <queue>
+#include <string>
 #include <tuple>
 #include <utility>
 #include <vector>
+
 #include "tree.hpp"
 
 namespace datastructures {
 
-template <typename Func, std::size_t... Is>
+template <typename Func, size_t... Is>
 constexpr void static_for_impl(Func&& f, std::index_sequence<Is...>) {
-  (f(std::integral_constant<std::size_t, Is>{}), ...);
+  (f(std::integral_constant<size_t, Is>{}), ...);
 }
 
 template <size_t N, typename Func>
@@ -19,159 +21,125 @@ constexpr void static_for(Func&& f) {
   static_for_impl(std::forward<Func>(f), std::make_index_sequence<N>{});
 }
 
-template <typename I, size_t dim>
-class MultiNodeViewInterface {
+template <typename I, typename Ts, size_t dim = std::tuple_size_v<Ts>>
+class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
  public:
+  static constexpr auto func_noop = [](const auto&... x) {};
+  static constexpr auto func_true = [](const auto&... x) { return true; };
+
+  const I& self() const { return static_cast<const I&>(*this); }
+  I& self() { return static_cast<I&>(*this); }
+
   int level() const {
-    const auto& nodes = static_cast<const I*>(this)->nodes();
-    // size_t result = 0;
-    // static_for<dim>([&result, &nodes](auto index) {
-    //  result += std::get<index>(nodes)->level();
-    //});
-    // return result;
-
-    return std::experimental::apply(
-        [](auto const&... e) -> decltype(auto) { return (e->level() + ...); },
-        nodes);
+    return std::apply([](const auto&... n) { return (n->level() + ...); },
+                      self().nodes());
   }
 
   template <size_t i>
-  bool IsFull() const {
-    auto self = static_cast<const I*>(this);
-    return self->children(i).size() ==
-           std::get<i>(self->nodes())->children().size();
+  bool is_full() const {
+    return self().children(i).size() ==
+           std::get<i>(self().nodes())->children().size();
   }
 
-  template <typename T>
-  std::shared_ptr<I> FindBrother(const T& nodes, size_t j, size_t i) {
-    auto self = static_cast<I*>(this);
-    if (nodes == self->nodes()) {
-      return self->shared_from_this();
-    }
+  bool is_leaf() const {
[…]

### 2019-10-30 C++/datastructures/tree.hpp

@@ -3,17 +3,18 @@
 #include <memory>
 #include <queue>
 #include <vector>
+
 #include "cassert"
 
 namespace datastructures {
-template <typename T>
-class NodeInterface : public std::enable_shared_from_this<T> {
+template <typename I>
+class NodeInterface : public std::enable_shared_from_this<I> {
  public:
-  std::vector<std::shared_ptr<T>> Bfs(
-      bool include_metaroot, std::function<void(std::shared_ptr<T>)> callback) {
-    std::vector<std::shared_ptr<T>> nodes;
-    std::queue<std::shared_ptr<T>> queue;
-    queue.emplace(static_cast<T *>(this)->shared_from_this());
+  template <typename Func>
+  std::vector<std::shared_ptr<I>> Bfs(bool include_metaroot, Func &&callback) {
+    std::vector<std::shared_ptr<I>> nodes;
+    std::queue<std::shared_ptr<I>> queue;
+    queue.emplace(static_cast<I *>(this)->shared_from_this());
     while (!queue.empty()) {
       auto node = queue.front();
       queue.pop();
@@ -21,14 +22,14 @@ class NodeInterface : public std::enable_shared_from_this<T> {
       nodes.emplace_back(node);
       node->set_marked(true);
       callback(node);
-      for (auto child : node->children()) queue.emplace(child);
+      for (const auto &child : node->children()) queue.emplace(child);
     }
-    for (auto node : nodes) {
+    for (const auto &node : nodes) {
       node->set_marked(false);
     }
     if (!include_metaroot) {
       nodes.erase(std::remove_if(nodes.begin(), nodes.end(),
-                                 [](const std::shared_ptr<T> &node) {
+                                 [](const std::shared_ptr<I> &node) {
                                    return node->is_metaroot();
                                  }),
                   nodes.end());
@@ -36,12 +37,12 @@ class NodeInterface : public std::enable_shared_from_this<T> {
     return nodes;
   }
 
-  std::vector<std::shared_ptr<T>> Bfs(bool include_metaroot = false) {
-    return Bfs(include_metaroot, [](std::shared_ptr<T> x) {});
+  std::vector<std::shared_ptr<I>> Bfs(bool include_metaroot = false) {
+    return Bfs(include_metaroot, [](std::shared_ptr<I> x) {});
   }
 
   void UniformRefine(int max_level) {
[…]

### 2019-10-30 C++/space/triangulation.cpp

@@ -40,7 +40,7 @@ VertexPtr Element2D::create_new_vertex(Element2DPtr nbr) {
       std::make_shared<Vertex>((godparents[0]->x + godparents[1]->x) / 2,
                                (godparents[0]->y + godparents[1]->y) / 2,
                                nbr == nullptr, vertex_parents);
-  for (auto vertex_parent : vertex_parents)
+  for (const auto &vertex_parent : vertex_parents)
     vertex_parent->children().push_back(new_vertex);
   return new_vertex;
 }
@@ -98,18 +98,17 @@ void Element2D::bisect_with_nbr() {
 InitialTriangulation::InitialTriangulation(
     const std::vector<std::array<double, 2>> &vertices,
     const std::vector<std::array<int, 3>> &elements)
-    : vertex_meta_root(Vertex::CreateMetaroot()),
-      elem_meta_root(Element2D::CreateMetaroot()) {
+    : vertex_meta_root(new Vertex()), elem_meta_root(new Element2D()) {
   // Convenient aliases
   auto &vertex_roots = vertex_meta_root->children();
   auto &element_roots = elem_meta_root->children();
 
-  for (auto vertex : vertices) {
+  for (const auto &vertex : vertices) {
     vertex_roots.push_back(std::make_shared<Vertex>(
         vertex[0], vertex[1], false, VectorVertexPtr{vertex_meta_root}));
   }
 
-  for (auto element : elements) {
+  for (const auto &element : elements) {
     double elem_area =
         0.5 * std::abs((vertices[element[0]][0] - vertices[element[2]][0]) *
                            (vertices[element[1]][1] - vertices[element[0]][1]) -
@@ -137,17 +136,17 @@ InitialTriangulation::InitialTriangulation(
   }
 
   // Determine patches for the vertices.
-  for (auto element : element_roots) {
-    for (auto vertex : element->vertices()) {
+  for (const auto &element : element_roots) {
+    for (const auto &vertex : element->vertices()) {
       vertex->patch.push_back(element);
     }
   }
 
   // Determine vertices that are on the boundary.
-  for (auto element : element_roots) {
+  for (const auto &element : element_roots) {
     for (int i = 0; i < 3; ++i)
       if (!element->neighbours[i]) {
-        for (auto v : element->edge(i)) {
+        for (const auto &v : element->edge(i)) {
           v->on_domain_boundary = true;
         }
       }

### 2019-10-30 C++/space/triangulation_view.cpp

@@ -5,17 +5,54 @@ using namespace space;
 
 int main() {
   auto T = InitialTriangulation::UnitSquare();
-  T.elem_meta_root->UniformRefine(1);
+  T.elem_meta_root->UniformRefine(3);
 
-  datastructures::NodeView<Vertex> root_view(T.vertex_meta_root);
-  std::shared_ptr<Vertex> child = T.vertex_meta_root->children()[0];
-  auto child_view = std::make_shared<datastructures::NodeView<Vertex>>(child);
+  auto root_view =
+      std::make_shared<datastructures::NodeView<Vertex>>(T.vertex_meta_root);
 
-  root_view.children(0).push_back(child_view);
+  std::cout << root_view->level() << std::endl;
+  std::cout << root_view->is_metaroot() << std::endl;
 
-  int x = child_view->level();
+  std::cout << root_view->children(0).size() << std::endl;
+  root_view->Refine<0>();
 
-  std::cout << x << std::endl;
-  root_view.Refine<0>();
+  std::cout << root_view->children(0).size() << std::endl;
+
+  auto bfs = root_view->Bfs();
+  for (auto nv : bfs) {
+    std::cout << *nv << std::endl;
+    nv->Refine<0>();
+  }
+
+  std::cout << " ---- after uniform refinemenet --- " << std::endl;
+  bfs = root_view->Bfs();
+  for (auto nv : bfs) {
+    std::cout << *nv << std::endl;
+    nv->Refine<0>();
+  }
+
+  std::cout << " ---- after uniform refinemenet --- " << std::endl;
+  bfs = root_view->Bfs();
+  for (auto nv : bfs) {
+    std::cout << *nv << std::endl;
+    nv->Refine<0>();
+  }
+
+  auto bla_view =
+      std::make_shared<datastructures::NodeView<Vertex>>(T.vertex_meta_root);
+  bla_view->DeepRefine();
+  root_view->Union(bla_view);
+
+  std::cout << " --- after union with a deep refine -- " << std::endl;
+  bfs = root_view->Bfs();
+  for (auto nv : bfs) {
+    std::cout << *nv << std::endl;
+    nv->Refine<0>();
+  }
+  //
+  //  int x = child_view->level();
+  //
+  //  std::cout << x << std::endl;
+  //  root_view.Refine<0>();
   return 0;
 }

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
removed unnecessary commentg

@@ -106,7 +106,7 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
  private:
   template <size_t i, size_t j>
   std::shared_ptr<I> FindBrother(const Ts& nodes, bool make_conforming);
-};  // namespace datastructures
+};
 
 template <typename I, typename... T>
 class MultiNodeView

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
added ability to union with other class

@@ -67,15 +67,15 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
                                       const Func& callback = func_noop,
                                       bool return_nodes = true);
 
-  // DeepRefine can be implemented using Bfs..
   template <typename FuncFilt = decltype(func_true),
             typename FuncPost = decltype(func_noop)>
   void DeepRefine(const FuncFilt& call_filter = func_true,
                   const FuncPost& call_postprocess = func_noop);
 
-  template <typename FuncFilt = decltype(func_true),
+  template <typename I_other = I, typename FuncFilt = decltype(func_true),
             typename FuncPost = decltype(func_noop)>
-  void Union(std::shared_ptr<I> other, const FuncFilt& call_filter = func_true,
+  void Union(std::shared_ptr<I_other> other,
+             const FuncFilt& call_filter = func_true,
              const FuncPost& call_postprocess = func_noop);
 
   template <size_t i, typename Func = decltype(func_true)>

### 2019-10-30 Output/error_reduction_analysis.tex
done

@@
 We discern a pre-asymptotic regime: the error hardly decreases between levels $L=1$
-and $2$. Moreover, errors seem to taper off near the final levels $L$, which may
+and $2$. Moreover, errors seem to stagnate/ near the final levels $L$, which may
 have something to do with our solver tolerance.
@@
 Let's measure the time it takes to perform a single application of the system
-matrix. We do not take into account the setup cost (like
-computing $\Sigma$ and $\Theta$). Figure~\ref{fig:linearity} paints a slightly
-demotivating picture, in that the bottom graph is not constant. We do have to
-place an asterisk here: for small systems, the produced numbers are averaged over
-a small number of iterations, so they change a lot from run to run.
+matrix. We do not take into account the setup cost (like computing $\Sigma$ and
+$\Theta$). Figure~\ref{fig:linearity} reveals a modest increase in the time spent
+per apply per doublenode, which hints at a very slight superlinearity in the
+computational cost of a single apply. However, we see only a factor-2 increase
+over the course of the simulation.
 \begin{figure}[h!]

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
added deepcopy

@@ -78,6 +78,10 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
              const FuncFilt& call_filter = func_true,
              const FuncPost& call_postprocess = func_noop);
 
+  template <typename I_other = I, typename FuncPost = decltype(func_noop)>
+  std::shared_ptr<I_other> DeepCopy(
+      const FuncPost& call_postprocess = func_noop);
+
   template <size_t i, typename Func = decltype(func_true)>
   const std::vector<std::shared_ptr<I>>& Refine(
       const std::vector<std::tuple_element_t<i, Ts>>& children_i,
@@ -122,7 +126,7 @@ class MultiNodeView
   explicit MultiNodeView(const TupleNodes& nodes,
                          const ArrayVectorImpls& parents)
       : nodes_(nodes), parents_(parents) {}
-  explicit MultiNodeView(TupleNodes&& nodes) : nodes_(nodes) {
+  explicit MultiNodeView(const TupleNodes& nodes) : nodes_(nodes) {
     assert(this->is_metaroot());
   }
 

### 2019-10-30 C++/space/triangulation_view.cpp

@@ -47,8 +47,15 @@ int main() {
   bfs = root_view->Bfs();
   for (auto nv : bfs) {
     std::cout << *nv << std::endl;
-    nv->Refine<0>();
   }
+
+  auto root_cpy = root_view->DeepCopy();
+  std::cout << " --- copied -- " << std::endl;
+  bfs = root_cpy->Bfs();
+  for (auto nv : bfs) {
+    std::cout << *nv << std::endl;
+  }
+
   //
   //  int x = child_view->level();
   //

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
fixed some comments

@@ -11,6 +11,8 @@
 
 namespace datastructures {
 
+// Below are two helper functions to create a compile-time-unrolled loop.
+// It is implemented using operator folding for the comma operator.
 template <typename Func, size_t... Is>
 constexpr void static_for_impl(Func&& f, std::index_sequence<Is...>) {
   (f(std::integral_constant<size_t, Is>{}), ...);
@@ -54,7 +56,7 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
         self().nodes());
   }
 
-  // Returns this multinode is the root of the multi-tree. That is, whether
+  // Returns if this multinode is the root of the multi-tree. That is, whether
   // all axis hold metaroots.
   bool is_root() const {
     return std::apply(

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
added template parameters

@@ -23,6 +23,10 @@ constexpr void static_for(Func&& f) {
   static_for_impl(std::forward<Func>(f), std::make_index_sequence<N>{});
 }
 
+// Template arguments are as follows:
+// I - The final implementation of this class, i.e. the derived class;
+// Ts - The tuple type that represents a node;
+// dim - The dimension of this multitree.
 template <typename I, typename Ts, size_t dim = std::tuple_size_v<Ts>>
 class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
  public:

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
added template parameters

@@ -172,6 +172,8 @@ template <typename T>
 class NodeView : public MultiNodeView<NodeView<T>, T> {
  public:
   using MultiNodeView<NodeView<T>, T>::MultiNodeView;
+
+  using MultiNodeView<NodeView<T>, T>::Refine<0>;
 };
 };  // namespace datastructures
 

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
revert

@@ -172,8 +172,6 @@ template <typename T>
 class NodeView : public MultiNodeView<NodeView<T>, T> {
  public:
   using MultiNodeView<NodeView<T>, T>::MultiNodeView;
-
-  using MultiNodeView<NodeView<T>, T>::Refine<0>;
 };
 };  // namespace datastructures
 

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
revert

@@ -88,7 +88,7 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
   std::shared_ptr<I_other> DeepCopy(
       const FuncPost& call_postprocess = func_noop);
 
-  template <size_t i, typename Func = decltype(func_true)>
+  template <size_t i = -1, typename Func = decltype(func_true)>
   const std::vector<std::shared_ptr<I>>& Refine(
       const std::vector<std::tuple_element_t<i, Ts>>& children_i,
       const Func& call_filter = func_true, bool make_conforming = false);

### 2019-10-30 C++/datastructures/multi_tree_view.hpp
added convenient overloads for nodeview

@@ -88,7 +88,7 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
   std::shared_ptr<I_other> DeepCopy(
       const FuncPost& call_postprocess = func_noop);
 
-  template <size_t i = -1, typename Func = decltype(func_true)>
+  template <size_t i, typename Func = decltype(func_true)>
   const std::vector<std::shared_ptr<I>>& Refine(
       const std::vector<std::tuple_element_t<i, Ts>>& children_i,
       const Func& call_filter = func_true, bool make_conforming = false);
@@ -99,6 +99,11 @@ class MultiNodeViewInterface : public std::enable_shared_from_this<I> {
                      bool make_conforming = false) {
     return Refine<i>(std::get<i>(self().nodes())->children(), call_filter,
                      make_conforming);
+  }
+  template <typename FuncFilt = decltype(func_true)>
+  void Refine(const FuncFilt& call_filter = func_true,
+              bool make_conforming = false) {
+    static_for<dim>([&](auto i) { Refine<i>(call_filter, make_conforming); });
   }
 
   // Some convenient debug function.
@@ -170,8 +175,15 @@ class MultiNodeView
 
 template <typename T>
 class NodeView : public MultiNodeView<NodeView<T>, T> {
+ private:
+  using Super = MultiNodeView<NodeView<T>, T>;
+
  public:
   using MultiNodeView<NodeView<T>, T>::MultiNodeView;
+
+  inline auto& children(size_t i = 0) { return Super::children(0); }
+  inline const auto& children(size_t i = 0) const { return Super::children(0); }
+  inline const auto& parents(size_t i = 0) const { return Super::parents(0); }
 };
 };  // namespace datastructures
 

### 2019-10-30 C++/space/triangulation_view.cpp

@@ -13,29 +13,29 @@ int main() {
   std::cout << root_view->level() << std::endl;
   std::cout << root_view->is_metaroot() << std::endl;
 
-  std::cout << root_view->children(0).size() << std::endl;
-  root_view->Refine<0>();
+  std::cout << root_view->children().size() << std::endl;
+  root_view->Refine();
 
-  std::cout << root_view->children(0).size() << std::endl;
+  std::cout << root_view->children().size() << std::endl;
 
   auto bfs = root_view->Bfs();
   for (auto nv : bfs) {
     std::cout << *nv << std::endl;
-    nv->Refine<0>();
+    nv->Refine();
   }
 
   std::cout << " ---- after uniform refinemenet --- " << std::endl;
   bfs = root_view->Bfs();
   for (auto nv : bfs) {
     std::cout << *nv << std::endl;
-    nv->Refine<0>();
+    nv->Refine();
   }
 
   std::cout << " ---- after uniform refinemenet --- " << std::endl;
   bfs = root_view->Bfs();
   for (auto nv : bfs) {
     std::cout << *nv << std::endl;
-    nv->Refine<0>();
+    nv->Refine();
   }
 
   auto bla_view =
@@ -60,6 +60,6 @@ int main() {
   //  int x = child_view->level();
   //
   //  std::cout << x << std::endl;
-  //  root_view.Refine<0>();
+  //  root_view.Refine();
   return 0;
 }

### 2020-05-15 Latex/cmam2020-westerdiep.tex
first pass latex

\documentclass[12pt]{amsart}
\usepackage[T1]{fontenc}
\usepackage{amsmath}

\title{Space-time adaptivity for parabolic evolution equations}
\author{Rob Stevenson \and Raymond van Veneti\"e \and Jan Westerdiep$^\star$}
\address{$^\star$ Korteweg-de Vries Institute for Mathematics, University of Amsterdam}
\email[Jan Westerdiep]{j.h.westerdiep@uva.nl}

\begin{document}

\maketitle

\section*{ABSTRACT}

Taking the well-posed mixed simultaneous space-time variational formulation
introduced in~\cite{And13}, we use methods previously developed in~\cite{SW20}
to construct an adaptive loop that produces space-time approximations as linear
combinations of tensor-products of wavelets in time and finite elements in space.

Using an efficient and reliable `hierarchical basis' error estimator, we apply bulk
chasing to show convergence of the iterands. Moreover, we provide an algorithm for
linear-complexity application of the system matrix and an optimal preconditioner.
Lastly, we include an extensive numerical study to show that this method is competitive
in terms of speed and moreover exhibits optimal convergence rate in the number of
degrees of freedom.

\begin{thebibliography}{99}

\bibitem[And13]{And13}
R.~Andreev.
\newblock Stability of sparse space-time finite element discretizations of
 linear parabolic evolution equations.
\newblock {\em IMA J. Numer. Anal.}, 33(1):242--260, 2013.

\bibitem[SW20]{SW20}
\newblock Stability of Galerkin discretizations of a mixed space-time variational
   formulation of parabolic evolution equations.
\newblock {\em IMA J. Numer. Anal.}, 2020.

\end{thebibliography}

\end{document}

### 2020-05-15 Latex/paper.tex
first pass latex

\documentclass[11pt,a4paper,oneside,english]{amsart}

\usepackage{amsaddr}
\usepackage{babel}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{commath}
\usepackage{colortbl}
\usepackage{dsfont}
\usepackage{enumerate}
\usepackage{esint}
\usepackage[left=0.09\paperwidth, right=0.15\paperwidth, bottom=0.12\paperheight]{geometry}
\usepackage{graphicx}
\usepackage[displaymath, mathlines]{lineno}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow,bigdelim}
\usepackage{url}
\usepackage{pdflscape}
\usepackage{rotating}
%\usepackage[outer]{showlabels}
\usepackage[color=white,linecolor=black, textwidth=0.12\paperwidth]{todonotes}
%\setlength{\parindent}{0pt}
\usepackage{wrapfig}

\numberwithin{equation}{section}
\newtheorem{theorem}{Theorem}
\numberwithin{theorem}{section}
\newtheorem{assum}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corol}[theorem]{Corollary}
\newtheorem{prop}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

\usepackage{color}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\providecommand\C{}
\renewcommand{\C}{\mathbb{C}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathbb{F}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\K}{\mathcal{K}}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\bbD}{\mathbb{D}}
\newcommand{\f}{\varphi}
\newcommand{\e}{\varepsilon}
\newcommand{\eps}{\varepsilon}
\renewcommand{\a}{\alpha}
\renewcommand{\b}{\beta}
\newcommand{\z}{\zeta}
\renewcommand{\O}{\Omega}
\renewcommand{\P}{\mathbb P}
\newcommand{\intRn}{\int_{\R^n}}
\newcommand{\sumi}{\sum_{i=1}^n}
\newcommand{\sumij}{\sum_{i,j=1}^n}
\renewcommand{\d}{\delta}
\newcommand{\p}{\partial}
\DeclareMathOperator{\osc}{osc}
\DeclareMathOperator{\supp}{supp}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\gen}{gen}
\DeclareMathOperator{\cat}{Cat}
\DeclareMathOperator{\Cat}{Cat}
\DeclareMathOperator{\Div}{div}
\DeclareMathOperator{\blockdiag}{blockdiag}
\renewcommand{\div}{\Div}
\providecommand{\spann}{\mathrm{span}}
\providecommand{\diam}{\mathrm{diam}}
\providecommand{\dist}{\mathrm{dist}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\usepackage{stmaryrd}
\newcommand{\jump}[1]{\left\llbracket #1 \right\rrbracket}
\newcommand{\rhu}{\rightharpoonup}
\newcommand{\T}{\mathcal{T}}
\newcommand{\V}{\mathcal{V}}
\DeclareMathOperator*{\dof}{\#\!DoF}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\Id}{Id}
\definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\providecommand\note{}
\newcommand{\new}[1]{{\color{ao(english)} #1}}
\renewcommand{\L}{\mathcal L}
\newcommand{\Y}{\mathcal Y}
\newcommand{\Lto}{{L_2(\Omega)}}
\newcommand{\bbT}{\mathbb{T}}
\newcommand{\cL}{\mathcal L}
\newcommand{\Lis}{\cL\mathrm{is}}
\newcommand{\udelta}{{\underline{\delta}}}
\newcommand{\jw}[1]{{\color{red}{JW: #1}}}
\newcommand{\rvv}[1]{{\color{teal}{RvV: #1}}}
\DeclareMathOperator*{\ran}{ran}

% Math symbol font matha
\DeclareFontFamily{U}{matha}{\hyphenchar\font45}
\DeclareFontShape{U}{matha}{m}{n}{
      <5> <6> <7> <8> <9> <10> gen * matha
      <10.95> matha10 <12> <14.4> <17.28> <20.74> <24.88> matha12
      }{}
\DeclareSymbolFont{matha}{U}{matha}{m}{n}
\DeclareFontSubstitution{U}{matha}{m}{n}

% Math symbol font mathb
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
      <5> <6> <7> <8> <9> <10>
      <10.95> <12> <14.4> <17.28> <20.74> <24.88>
      mathx10
      }{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareFontSubstitution{U}{mathx}{m}{n}

% Symbol definition
\DeclareMathDelimiter{\vvvert}{0}{matha}{"7E}{mathx}{"17}

\DeclareMathOperator{\ext}{ext}
\DeclareMathOperator{\eff}{eff}
\DeclareMathOperator{\loc}{loc}
\DeclareMathOperator{\irr}{int}
\DeclareMathOperator{\RT}{RT}
\DeclareMathOperator{\meas}{meas}
\newcommand{\bbone}{\mathds{1}}
\newcommand{\bbzero}{\mathds{0}}
\newcommand{\seminorm}[1]{\vvvert #1 \vvvert}
\newcommand{\ltwonorm}[1]{\lVert #1 \rVert}
\newcommand{\hta}{{\check \T_a}}
\newcommand{\hT}{{\check T}}
\newcommand{\he}{{\check e}}
\newcommand{\hE}{{\check \E}}
\newcommand{\dr}{{\tilde r}}
\renewcommand{\emptyset}{{\varnothing}}
\newcommand{\hprime}[1]{H^1_{*, #1}(\hta)}
\newcommand{\laplace}{\mathcal{4}}

\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
      <5> <6> <7> <8> <9> <10>
      <10.95> <12> <14.4> <17.28> <20.74> <24.88>
      mathx10
      }{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareFontSubstitution{U}{mathx}{m}{n}
\DeclareMathAccent{\widecheck}{0}{mathx}{"71}
\DeclareMathAccent{\wideparen}{0}{mathx}{"75}
%\linenumbers

\def\cs#1{\texttt{\char`\\#1}}

\title{Space-time adaptivity for parabolic evolution equations}
\author{Rob Stevenson \and Raymond van Veneti\"e \and Jan Westerdiep}
\address{Korteweg-de Vries Institute for Mathematics, University of Amsterdam, \\
P.O. Box 94248, 1090 GE Amsterdam, The Netherlands}
\date{\today}
\subjclass[2010]{}

\thanks{\emph{Funding:} The second and third authors are supported by the Netherlands Organisation for Scientific Research (NWO) under contract.~no.~613.001.652}

\begin{document}
\begin{abstract}
\end{abstract}

\maketitle

\section{Introduction}
\section{Problem statement and solution}
Let $V$ and $H$ be separable Hilbert spaces on a spatial domain $\Omega$ such that
$V \hookrightarrow H$ with dense and compact embedding. Identifying $H$ with its
dual, we obtain the Gelfand triple $V \hookrightarrow H \simeq H' \hookrightarrow V'$.

We use the notation $\la \cdot,\cdot \ra$ to denote both the scalar product
on $H \times H$, and its unique extension by continuity to the duality pairing on
$V' \times V$. Correspondingly, the norm on $H$ will be denoted by $\|\cdot\|$.

Let $T \in (0, \infty)$ be given, and set $I := (0, T)$. We make the following assumption.
\begin{assum}
  \label{assum:a}
  For a.e. $t$, let $a(t;\cdot,\cdot)$ denote a bilinear form on $V \times V$ that is
  \begin{alignat*}{3}
    t \mapsto a(t; \eta, \zeta) & ~~ \text{is measurable} \quad && (\eta, \zeta \in V) \\
    |a(t;\eta,\zeta)| & \lesssim  \|\eta\|_{V} \|\zeta\|_{V} &&(\eta,\zeta \in V) &&\text{(bounded)}, \\
     a(t;\eta,\eta)   & \gtrsim   \|\eta\|_{V}^2 &&(\eta \in {V}) &&\text{(coercive)}.
  \end{alignat*}
\end{assum}

With $A(t) \in \Lis({V},V')$ being defined by $ (A(t) \eta)(\zeta)=a(t;\eta,\zeta)$,
we are interested in solving the {\em parabolic initial value problem} of finding
$u$ such that
\begin{equation}
  \label{eqn:ibvp}
  \left\{
    \begin{array}{rl} 
      \frac{\dif u}{\dif t}(t) +A(t) u(t)&\!\!\!= g(t) \quad(t \in I),\\
      u(0) &\!\!\!= u_0.
    \end{array}
  \right.
\end{equation}

In a simultaneous space-time variational formulation, the parabolic PDE reads as
finding $u$ from a suitable space of functions of time and space such that
\begin{equation}
  \label{eqn:B-var-form}
  (Bw)(v):=\int_I \la \tfrac{\dif w}{\dif t}(t), v(t) \ra + a(t;w(t),v(t)) \dif t = \int_I \la g(t), v(t) \ra =:g(v)
\end{equation}
for all $v$ from another suitable space of functions of time and space.

\begin{theorem}
  \label{thm:varform}
  With $X:=L_2(I;{V}) \cap H^1(I;V')$, $Y:=L_2(I;{V})$, under assumption \ref{assum:a}, we have
  \begin{equation}
    \label{eqn:B-lis}
    \begin{bmatrix} B \\ \gamma_0\end{bmatrix}\in \Lis(X,Y' \times H),
  \end{equation}
  where for $t \in \bar I$, $\gamma_t\colon u \mapsto u(t,\cdot)$ denotes the trace map.
  In other words, given $(g, u_0) \in Y' \times H$, finding $u \in X$ such that 
  \begin{equation}
    \label{eqn:B-varform}
    (Bu)(v_1)+\la u(0,\cdot),v_2\ra=g(v_1)+\la u_0,v_2\ra\quad ((v_1,v_2) \in Y \times H),
  \end{equation}
  is a well-posed variational formulation of \eqref{eqn:ibvp}.
\end{theorem}

\subsection{Andreev formulation}
Defining $A, A_s \in \Lis(Y,Y')$, $A_a \in \cL(Y,Y')$, and $\partial_t \in \cL(X,Y')$ by 
\[
(Au)(v):=\int_I a(t;u(t),v(t)) \dif t, \quad A_s:=\tfrac12(A+A'), \quad A_a:=\tfrac12(A-A'), \quad \partial_t:=B-A,
\]
an equivalent self-adjoint saddle point formulation of \eqref{eqn:B-varform} was
introduced by Andreev in \cite{Andreev2013} as finding the solution $(\mu,u) \in Y\times X$ of
\[
  \begin{bmatrix} A_s & B\\ B' & -\gamma_0' \gamma_0 \end{bmatrix}
  \begin{bmatrix} \mu \\ u \end{bmatrix}=
  \begin{bmatrix} g \\ -\gamma_0' u_0 \end{bmatrix}.
\]
Thanks to \eqref{eqn:B-lis}, the Schur complement $S := B' A_s^{-1} B+\gamma_0'\gamma_0$ is in $\Lis(X, X')$, is self-adjoint, and coercive.

\begin{remark}
  For some quantitative estimates to be valid, we assume that $A=A_s$, but we do not expect this to be essential.
\end{remark}
We equip $Y$ and $X$ with the `energy' norms
$$
\|\cdot\|_Y^2:=(A_s\cdot)(\cdot),\quad \|\cdot\|_X^2:=\|\cdot\|_Y^2+\|\partial_t \cdot\|_{Y'}^2+\|\gamma_T \cdot\|^2,
$$
which are equivalent to the canonical norms on $Y$ and $X$.

\begin{lemma}
  \label{lem:energy-norm}
  It holds that $\|\cdot\|_X^2=(S\cdot)(\cdot)$.
\end{lemma}
\begin{proof} 
  We find \jw{dit ben ik nog niet nagegaan}
  \[
    \|w\|_X^2=\sup_{0 \neq v_1 \in Y} \frac{(B w)(v_1)}{\|v_1\|_Y^2}+\|\gamma_0 w\|^2
    =\sup_{0 \neq (v_1,v_2) \in Y \times H} \frac{((B w)(v_1)+\la \gamma_0 w,v_2\ra)^2}{\|v_1\|_Y^2+\|v_2\|^2} =(Sw)(w),
  \]
  where the first equality can be found in e.g.~\cite[Thm.~2.1]{Ern2017a}, and,
  when realising that $S$ is also the Schur complement of 
  $\begin{bsmallmatrix} A_s & 0 & B\\ 0 & \Id & \gamma_0\\ B' & \gamma_0' & 0\end{bsmallmatrix} \in \cL(Y\times H \times X, Y'\times H'\times X')$, the last one follows from e.g.~\cite[Lemma~2.2]{Kondratyuk2008}.
\end{proof}

Let $(X^\delta)_{\delta \in \Delta}$ be a collection of closed subspaces of $X$. 
We define a \emph{partial order} on $\Delta$ by $\delta \preceq \tilde{\delta}$
when $X^\delta \subseteq X^{\tilde{\delta}}$. For $\delta \in \Delta$, let
$u_\delta \in X^\delta$ denote the {\em Galerkin approximation to $u$}, i.e., the solution of
\begin{equation}
(Su_\delta)(v)=f(v) \quad (v \in X^\delta), 
  \label{eqn:galerkin}
\end{equation}
being the best approximation to $u$ from $X^\delta$ w.r.t.~$\|\cdot\|_X$.

\subsection{Discretized problem}
Applying $S$ requires the application of the nonlocal operator $A_s^{-1}$, making it
unsuitable for practical computation. To that end, let $(Y^\delta)_{\delta \in \Delta}$
be a collection of closed subspaces of $Y$ satisfying $X^\delta \subseteq Y^\delta$
that is \emph{uniformly inf-sup stable}, i.e.
\begin{equation}
  \gamma_\Delta:=\inf_{\delta \in \Delta}\inf_{0 \neq w \in X^\delta}\sup_{0\neq y^\delta \in Y^\delta} \frac{(\partial_t w)(v)}{\|\partial_t w\|_{Y'}\|v\|_Y} \in (0,1].
  \label{eqn:infinfsup}
\end{equation}
Uniform stability essentially establishes the norm equivalences
\[
  \gamma_\Delta ||\partial_t w||_{Y'} \leq ||\partial_t w||_{{Y^\delta}'} \leq ||\partial_t w||_{Y'} \quad (w \in Y, \delta \in \Delta).
\]
Note that $\gamma_\Delta$ can be made arbitrarily close to $1$ by selecting $Y^\delta$ sufficiently large.

For $\delta, \hat{\delta} \in \Delta$ with $Y^{\hat{\delta}} \supseteq Y^\delta$, and 
$E_Y^{\hat{\delta}}$, $E_X^\delta$ denoting the embeddings $Y^{\hat{\delta}} \rightarrow Y$,
$X^\delta \rightarrow X$, let $(\mu^{\hat{\delta} \delta},u^{\hat{\delta} \delta}) \in Y^{\hat{\delta}} \times X^\delta$ be the solution of 
\[
  \begin{bmatrix}{E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}& {E_Y^{\hat{\delta}}}' B E^\delta_X\\ {E^\delta_X}' B' E_Y^{\hat{\delta}}& -{E^\delta_X}' \gamma_0' \gamma_0 E^\delta_X \end{bmatrix}
  \begin{bmatrix} \mu^{\hat{\delta} \delta} \\ u^{\hat{\delta} \delta} \end{bmatrix}
  \begin{bmatrix} {E^{\hat{\delta}}_Y}' g \\ -{E^\delta_X}' \gamma_0' u_0 \end{bmatrix}
\]
or
\begin{equation}
  \label{eqn:discr-schur}
  \underbrace{
    {E^\delta_X}'(B' E^{\hat{\delta}}_Y({E^{\hat{\delta}}_Y}' A_s E^{\hat{\delta}}_Y)^{-1} {E^{\hat{\delta}}_Y}' B +\gamma_0'\gamma_0)E^\delta_X
  }_{S^{\hat{\delta} \delta}:=}u^{\hat{\delta} \delta}
  =
  \underbrace{
    {E^\delta_X}' (B' E^{\hat{\delta}}_Y({E^{\hat{\delta}}_Y}' A_s E^{\hat{\delta}}_Y)^{-1} {E^{\hat{\delta}}_Y}' g+\gamma_0' u_0)
  }_{f^{\hat{\delta} \delta}:=}.
\end{equation}\rvv{heavy notatie uitstellen?}
Unless $Y^{\hat{\delta}}=Y$, it holds that $S^{\hat{\delta} \delta} \neq {E_X^\delta}' S E_X^\delta$ and $f^{\hat{\delta} \delta} \neq {E_X^\delta}'f$, and so generally $u^{\hat{\delta} \delta} \neq u_\delta$.

Equipping $X^\delta$ with `energy' norm $\|w\|^2_{X^{\hat{\delta}\delta}}:=\|w\|^2_{Y}+ \|\partial_t w\|_{{Y^\delta}'}+\|\gamma_T w\|^2$,
from \cite[Lemma~3.3]{Stevenson2020a} it follows that
$\|\cdot\|_{X^{\hat{\delta}\delta}}^2=(S^{\hat{\delta}\delta}\cdot)(\cdot)$,
similar to Lemma~\ref{lem:energy-norm}. As shown in \cite[Thm.~3.6]{Stevenson2020a}, we have
\begin{equation}
  \label{eqn:galerkin-ortho}
  \|u-u^{\hat{\delta}\delta}\|_X \leq \gamma_\Delta^{-1} \|u-u_\delta\|_X,
\end{equation}
whereas by definition of $\gamma_\Delta$ it holds that
\begin{equation}
  \label{eqn:X-norm-equiv}
  \gamma_\Delta\|\cdot\|_{X} \leq \|\cdot\|_{X^{\hat{\delta}\delta}} \leq \|\cdot\|_{X} \quad\text{on } X^\delta.
\end{equation}

\section{Realization of uniform inf-sup stability \eqref{eqn:infinfsup}}
In our previous work, we realized the inf-inf-sup condition \eqref{eqn:infinfsup}
for families $(Y^\delta, X^\delta)_{\delta \in \Delta}$ of finite element spaces
wrt.~partitions of the spacetime domain into prismatic elements, where the partition
in time is independent of the spatial location, and the spatial mesh in each
\emph{time slab} is such that the corresponding $H$-orthogonal projection is
uniformly $V$-stable.

Let us first argue why uniform stability for general nonuniform prismatic partitions
is out of reach. \jw{follows from $\dim Y^\delta \lesssim \dim X^\delta$ en
de non-locality van $A_s^{-1}$: gegeven een $X^\delta$ die naar een spacetime-hoekje
gerefined is, moet $Y^\delta$ de gehele time-slab refinen. Ik ben het precieze argument
even vergeten maar met een plaatje wordt het snel duidelijk.}

\subsection{The spatial meshes}
The following result generalizes \cite[Lemma 6.2]{11} and \cite[Proposition 2.5]{258.4}
by allowing $W \neq \tilde W$.

\begin{prop} \label{prop:space-infsup}
  Let $W$ and $\tilde W$ be closed subspaces of $H$ such that there exists a
  (biorthogonal) projector $Q \in \cL(H,H)$ with $\ran Q=W$ and $\ran (\Id -Q)={\tilde W}^\perp$. 
  Then, when $W \subset V$, it holds that 
  \[
    \gamma:=\inf_{0 \neq \tilde w \in \tilde W}\sup_{0 \neq w \in W}\frac{\la \tilde w,w\ra}{\|\tilde w\|_{V'}\|w\|_V}>0
  \]
  if and only if $Q \in \cL(V,V)$, in which case $\gamma=\|Q\|_{\cL(V,V)}^{-1}$.
\end{prop}
\begin{proof} \jw{Dit ben ik nog niet nagegaan.} If $Q\in \cL(V,V)$, then for $\tilde w \in \tilde W$,
$$
\sup_{0 \neq w \in W}\frac{\la \tilde w,w\ra}{\|w\|_V}=\sup_{0 \neq v \in V}\frac{\la \tilde w, v\ra}{\|Qv\|_V}
\geq \|Q\|_{\cL(V,V)}^{-1} \|\tilde w\|_{V'},
$$
or $\gamma \geq \|Q\|_{\cL(V,V)}^{-1}$. If on the other hand $\gamma>0$, then for any $u \in H$,
$$
\gamma \|Q' u\|_{V'} \leq \sup_{0 \neq w \in W}\frac{\la Q'u,w\ra}{\|w\|_V} =
 \sup_{0 \neq w \in W}\frac{\la u,w\ra}{\|w\|_V} \leq \|u\|_{V'},
 $$
so that from $H$ being dense in $V'$, $\|Q\|_{\cL(V,V)}=\|Q'\|_{\cL(V',V')}\leq \gamma^{-1}$.
\end{proof}
\begin{remark}
  In the typical case where $H = L_2(\Omega)$ and $V = H_0^1(\Omega)$ on a bounded
  polytopal domain $\Omega \subset \R^d$, the uniform boundedness of $\|Q\|_{\cL(V,V)}$
  has been demonstrated for families of locally refined partitions, for $d=2$
  including those that are generated by the newest vertex bisection algorithm.
  We refer to \cite{Carstensen2001a,Gaspoz2016}. \jw{Misschien is het ``$H^1$-stability
  of $L^2$-projection" inmiddels wel uitgespeeld. ik vond ook een werk uit
  2014 van de groep van Dirk, en dit werkte voor elke dimensie, maar enkel op
  vreemde fin.elts. Ook een werk uit 2019 van Gaspoz, Heine, Siebert waar het
  2d-NVB geval wordt bekeken voor Lagrange-fin elts tot graad 4.}
\end{remark}

\subsection{Wavelets in time, finite elements in space}
Let ${\mathcal O}$ be a collection of pairs of closed subspaces $(W,\tilde{W})$
as in Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$.

Let $\Sigma=\{\sigma_\lambda : \lambda \in \vee_\Sigma\}$ be, properly scaled, a Riesz basis for $L_2(I)$ and $H^1(I)$.\rvv{Definitie maken ergens?}
Let $\Psi=\{\psi_\lambda \colon \lambda \in \vee_\Psi\}$ be a Riesz basis for $L_2(I)$, with dual basis denoted as $\tilde{\Psi}$.

Given
\[
\framebox{$X^\delta=\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes \tilde{W}_\lambda^\delta,$}
\]
where \framebox{$(W_\lambda^\delta,\tilde{W}_\lambda^\delta) \in {\mathcal O}$
or $W_\lambda^\delta=\tilde{W}_\lambda^\delta=\{0\}$}, let
\[
\framebox{$Y^\delta=\sum_{\mu \in \vee_\Psi} \psi_\mu \otimes V_\mu^\delta$}
\]
and $\tilde{Y}^\delta=\sum_{\mu \in \vee_\Psi} \tilde{\psi}_\mu \otimes \tilde{V}_\mu^\delta$ 
where \framebox{$(V^\delta_\mu,\tilde{V}^\delta_\mu) \in {\mathcal O}$ or $V_\lambda^\delta=\tilde{V}_\lambda^\delta=\{0\}$} is such that 
\begin{equation}
  \label{eqn:pre-doubletree}
  \la \sigma'_\lambda,\psi_\mu\ra_{L_2(I)} \neq 0 \implies \tilde{V}^\delta_\mu \supset \tilde{W}^\delta_\lambda.
\end{equation}
\jw{hoe werkt deze eis precies?}
Thanks to $\sigma'_\lambda=\sum_{\mu \in \vee_\Psi} \la \sigma'_\lambda,\psi_\mu\ra_{L_2(I)}\tilde{\psi}_\mu$, the previous implication ensures that
\begin{equation}
  \label{eqn:d_tX-in-Y}
 \framebox{$\partial_t X^\delta \subset \tilde{Y}^\delta.$}
\end{equation}
\rvv{deze uitlijning is slecht}
\begin{theorem}
In the previous notation, $(X^\delta,Y^\delta)_{\delta \in \Delta}$ satisfies
  \eqref{eqn:infinfsup} with uniform inf-sup constant $\gamma$.
\end{theorem}
\begin{proof}
  \jw{Ik ben dit nog niet nagegaan.}
Let $\tilde v=\sum_{\mu \in \vee_\Psi} \tilde{\psi}_\mu \otimes \tilde{v}_\mu \in \tilde{Y}^\delta$.
Then, for any constant $\eps\in (0,\gamma)$, there exists a $v=\sum_{\mu \in \vee_\Psi} \psi_\mu \otimes v_\mu \in Y^\delta$  with $\la \tilde v_\mu,v_\mu\ra \geq (\gamma-\eps) \|\tilde{v}_\mu\|_{V'} \|v_\mu\|_V$ and $\|\tilde{v}_\mu\|_{V'}= \|v_\mu\|_V$, and so
\[
  \la\tilde v, v\ra_{L_2(I)\otimes H}
  = \sum_{\mu \in \vee_\Psi} \la \tilde{v}_\mu,v_\mu\ra \geq (\gamma-\eps) 
  \sum_{\mu \in \vee_\Psi} \|\tilde{v}_\lambda\|_{V'}^2
  \eqsim \|\tilde v\|_{Y'} \|v\|_Y.
\]
  Together with \eqref{eqn:d_tX-in-Y}, the result follows.
\end{proof}

\subsubsection{Sufficient conditions for \eqref{eqn:pre-doubletree}}
\label{sec:spaces}
To ensure \eqref{eqn:pre-doubletree}, we will assume that that for $\ell \in \N_0$,
\[
\spann\{\sigma'_\lambda\colon |\lambda|\leq \ell\} \subset \spann\{\tilde{\psi}_\lambda\colon |\lambda|\leq \ell\}.
\]
 so that
\begin{equation}
  \label{eqn:deriv-orthogonal}
  |\mu| >|\lambda| \implies \la \sigma'_\lambda,\psi_\mu\ra_{L_2(I)}=0.
 \end{equation}
Now \eqref{eqn:pre-doubletree} is ensured under the condition that
\begin{equation}
\label{eqn:Y-space-refined}
  |\mu| \leq |\lambda| \wedge|\supp \psi_\mu \cap \supp \sigma_\lambda|>0 \implies \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda.
\end{equation}

We will assume that $\Sigma$ and $\Psi$ are locally supported, which in particular
means that $|\supp \sigma_\lambda| \lesssim 2^{-|\lambda|}$ and
$|\supp \psi_\lambda| \lesssim 2^{-|\lambda|}$. Furthermore, to ensure an
efficient stiffness matrix evaluation, we will shirtly impose a \emph{doubletree constraint}
on $X^\delta$ and $Y^\delta$, which will have the consequence that for any
$\lambda \in \vee_\Sigma$ with $\tilde{W}_\lambda^\delta \neq \{0\}$, and any
$\ell <|\lambda|$, there exists a $\mu \in \vee_\Sigma$ with $|\mu|=\ell$
($\mu$ being the `ancestor' of $\lambda$) with
$\dist(\supp \sigma_\mu,\supp \sigma_\lambda)\lesssim 2^{-|\mu|}$ and
$\tilde{W}_\mu^\delta \supset \tilde{W}_\lambda^\delta$.
We infer that \eqref{eqn:Y-space-refined} and thus \eqref{eqn:pre-doubletree} can
be ensured for $Y^\delta$ with $\dim Y^\delta \lesssim \dim X^\delta$.

In particular, for $\lambda \in \vee_\Psi$, let $S(\lambda)$ be the neighborhood
of $\supp \psi_\lambda$ with $\diam S(\lambda) \lesssim 2^{-|\lambda|}$ s.t.~$S(\lambda) \subset S(\mu)$
whenever $\lambda$ is a child of $\mu \in \vee_\Psi$. Then, assuming the doubletree
constraint on $X^\delta$, \eqref{eqn:Y-space-refined} is ensured under the condition
that
\[
  |\mu| = |\lambda| \wedge|\supp \psi_\mu \cap \supp S(\lambda)|>0 \implies \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda.
\]
\jw{Dit is min-of-meer copy-paste uit followup maar ik vind het nu nog wel cryptisch :-(}

\begin{example}
  Consider pairs $(W, \tilde W) \in \mathcal O$ where $W = \tilde W$.
  Let any $\sigma_\lambda$ be a continuous piecewise polynomial of degree $d-1$\jw{voor $d$ de dimensie van je space domain?! of gewoon *een* $d$? in het tweede geval: rename naar $p$?}
  w.r.t.~a uniform partition of $I$ into $2^{|\lambda|}$ subintervals, and let 
  $\Psi=\tilde{\Psi}$ be such that $\spann\{\tilde{\psi}_\lambda : |\mu| \leq |\lambda|\}$
  includes all piecewise polynomials of degree $d-2$ w.r.t.~a uniform partition
  of $I$ into $2^{|\lambda|}$ subintervals so that \eqref{eqn:deriv-orthogonal}
  is satisfied. If $\spann\{\tilde{\psi}_\lambda : |\mu| \leq |\lambda|\}$ even
  includes all piecewise polynomials of degree $d-1$ w.r.t.~a uniform partition
  of $I$ into $2^{|\lambda|}$ subintervals, then $Y^\delta$ as constructed above
  additionally satisfies
  \begin{equation}
    \label{eqn:inclusion}
      \framebox{$X^\delta \subset Y^\delta$}.
  \end{equation}
\end{example}

\begin{example}
  \label{ex:wavelets}
  \jw{dit moet misschien niet een voorbeeld zijn maar een eigen subsectie}
  Consider linear temporal wavelets and linear spatial finite element spaces.
  For $\Sigma$, we take the continuous 3-point from Figure~\ref{fig:3pt-wavelet}
  \jw{waarom is dit een Riesz basis? dit heeft een citatie nodig met wat meer info, bv \cite{Stevenson1998}?}.
  For this basis, $S(\lambda) = \supp \sigma_\lambda$.
  \begin{figure}
    \caption{3-point wavelet.}
    \label{fig:3pt-wavelet}
  \end{figure}

  For $\Psi$, we take the discontinuous orthogonal piecewise linear wavelet basis
  of Figure~\ref{fig:ortho-wavelet}\jw{again, needs citation; ik kon niks vinden zo 1-2-3;
  misschien omdat het niet echt een ``wavelet'' in de typische zin is?} For this
  basis, again $S(\lambda) = \supp \psi_\lambda)$.
  \begin{figure}
    \caption{orthogonal wavelet.}
    \label{fig:ortho-wavelet}
  \end{figure}

  In the spatial axis, we consider the collection $\mathbb T$ of conforming
  triangulations generated by NVB from some coarsest triangulation $\T_\perp$,
  and define $V_\T = V_\T$ the space of continuous piecewise linears w.r.t.~$\T$
  that vanish on $\partial \Omega$ such that $X^\delta := \sum_{\lambda \in \vee^\delta_\Sigma} \sigma_\lambda \otimes V_{\T_{\vee^\delta_\Sigma, \lambda}}$ satisfies the doubletree constraint.
  This means that when $\mu \in \vee^\delta_\Sigma$ is a parent of $\lambda \in \vee^\delta_\Sigma$,
      then $\T_{\vee^\delta_\Sigma, \mu}$ is a refinement of $\T_{\vee^\delta_\Sigma, \lambda}$, and that $\vee^\delta_\Sigma$ is a subtree of $\vee_\Sigma$, meaning
  \begin{itemize}
    \item it contains all wavelets on level 0, and
    \item if it contains a wavelet of level $\lambda \geq 1$, then it also contains
      its parents.
  \end{itemize}
\end{example}

\begin{remark}
When equipping the finite element spaces $V_{\T}$ with the hierarchical basis,
this doubletree constraint coincides with the constraint on the index set of
the basis for $X^\delta$; see e.g.~\cite{TODO}.
\end{remark}

We consider spaces $Y^\delta = \tilde Y^\delta$ of the form
$Y^\delta := \sum_{\lambda \in \vee^\delta_\Psi} \psi_\lambda \otimes V_{\T_{\vee^\delta_\Psi, \lambda}}$
for $\T_{\vee^\delta_\Psi, \lambda} \in \mathbb T$, again satisfying the doubletree
constraint. The choice
\[
  Y^\delta = \sum_{\lambda \in \vee_\Sigma^\delta} \sum_{\{\mu \in \vee_\Psi\colon |\mu|=|\lambda|,\,
|\supp \sigma_\lambda \cap \supp \psi_\mu|>0\}} \psi_\mu \otimes V_{\T_{\vee_\Sigma^\delta,\lambda}}
\]
is a doubletree by virtue of $X^\delta$ being one, and for this pair $(X^\delta, Y^\delta)$, \eqref{eqn:d_tX-in-Y} and \eqref{eqn:inclusion} are valid.

\section{Adaptive refinement}
We aim at solving a discretized version of the Schur complement equation
\begin{equation}
  \label{eqn:schur}
 (B' A_s^{-1} B + \gamma_0' \gamma_0)u =: Su = f := B' A_s^{-1}g+\gamma_0' u_0.
\end{equation}
Recalling the definition of the Galerkin approximation $u_\delta$ from \eqref{eqn:galerkin},
we introduce the following.

\begin{assum}[Saturation] \label{assum:saturation}
There exists a collection of subspaces $(F^\delta)_{\delta \in \Delta} \subset X'$, and
a constant $\rho<1$ such that for all $\delta \in \Delta$, when $f \in F^\delta$,
there exists a $\Delta \ni \udelta=\udelta(\delta) \succeq \delta$ with
\begin{equation}
  \label{eqn:reduction}
  \|u-u_\udelta\|_X \leq \rho \|u-u_\delta\|_X.
\end{equation}
\end{assum}
\begin{remark} \label{data-oscillation}
  Notice that above assumption cannot be valid without a restriction on the
  right-hand side $f =B' A_s^{-1}g+\gamma_0' u_0\in X'$. Indeed: given any
  $\{0\} \neq X^\delta \subsetneq X^{\udelta} \subsetneq X$, consider a non-zero
  $f \in X'$ that vanishes on $X^{\udelta}$. Then $u_\delta=u_\udelta=0 \neq u$,
  meaning that \eqref{eqn:reduction} does not hold. 
\end{remark}

For the time being we will operate under the restrictive assumption that whenever
we apply \eqref{eqn:reduction} we simply assume that $f \in F^\delta$. We will
later remove this assumption.

In view of obtaining an efficient implementation, in the definition of
$(\mu^{\hat{\delta} \delta},u^{\hat{\delta} \delta})$, $S^{\hat{\delta} \delta}$
and $f^{\hat{\delta} \delta}$ in \eqref{eqn:discr-schur}, we replace
${E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}$ by some \emph{preconditioner}
$A_s^{\hat{\delta}}={A_s^{\hat{\delta}}}' \in \Lis(Y^{\hat{\delta}},{Y^{\hat \delta}}')$,
whose inverse can be applied in linear computational complexity, and which satisfies for some $\eps_\Delta>0$,
\[
\|{E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}-A_s^{\hat{\delta}}\|_{\cL(Y^{\hat{\delta}},{Y^{\hat{\delta}}}')}\leq \eps_\Delta, \quad(\hat{\delta} \in \Delta).
\]
\jw{dit zou misschien een leuke numerieke test zijn, niet?}
Despite this modification, we keep using the old notations for $\mu^{\hat{\delta} \delta}$,
$u^{\hat{\delta} \delta}$, $S^{\hat{\delta} \delta}$,
$\|\cdot\|_{X^{\hat{\delta}\delta}}:=(S^{\hat{\delta}\delta}\cdot)(\cdot)^{\frac12}$,
and $f^{\hat{\delta} \delta}$.

From \eqref{eqn:galerkin-ortho} and \eqref{eqn:X-norm-equiv},
we infer that there exists a $C_\Delta=C_\Delta(\gamma_\Delta,\eps_{\Delta})\geq 1$ such that 
$\lim_{\gamma_\Delta \uparrow 1,\eps_\Delta \downarrow 0} C_\Delta = 1$, and for
$\delta,\hat{\delta} \in \Delta$ with $Y^{\hat{\delta}} \supseteq Y^\delta$, again we have
\[
  \|u-u^{\hat{\delta}\delta}\|_X \leq C_\Delta \|u-u_\delta\|_X, \quad \text{and} \quad
C_\Delta^{-1} \|\cdot\|_{X} \leq \|\cdot\|_{X^{\hat{\delta}\delta}} \leq C_\Delta \|\cdot\|_{X} \quad\text{on } X^\delta.
\]

\subsection{Towards convergence}
For $\delta \in \Delta$, we consider the modified discretized problem taking
$\hat{\delta}:=\udelta=\udelta(\delta)$ from Assumption~\ref{assum:saturation}.
We are going to construct a sequence $(\delta_i) \subset \Delta$ with $\delta_i \prec \delta_{i+1}$
such that $(u^{\udelta(\delta_i)\delta_i})_i$ converges to $u$.

In the next lemma it is shown that if one constructs from $w \in X^\delta$ a
$v \in X^\udelta$ that is closer to the best approximation $u_\udelta$ of $u$ from
$X^\udelta$, then, thanks to Assumption~\ref{assum:saturation}, $v$ is also closer to $u$.

\begin{lemma}
  Let $w \in X^\delta$, $v \in X^\udelta$ be such that for some $\beta$,
  \[
    \|u_\udelta-v\|_X \leq \beta \|u_\udelta-w\|_X.
  \]
  Then
  \[
    \|u-v\|_X \leq \sqrt{\rho^2+\beta^2(1-\rho^2)}\,\|u-w\|_X.
  \]
\end{lemma}
\begin{proof}
  Using $u-u_\udelta \perp_X X^\udelta$ twice, we obtain
  \begin{align*}
    \|u-v\|_X^2& =\|u-u_\udelta\|_X^2+\|u_\udelta-v\|_X^2\\
    &\leq \|u-u_\udelta\|_X^2+\beta^2 \|u_\udelta-w\|_X^2\\
    &=\|u-u_\udelta\|_X^2+\beta^2 (\|u-w\|_X^2-\|u-u_\udelta\|_X^2)\\
    &=(1-\beta^2)\|u-u_\udelta\|_X^2+\beta^2\|u-w\|_X^2\\
    &\leq (\rho^2(1-\beta^2)+\beta^2)\|u-w\|_X^2,
  \end{align*}
  where we used Assumption~\ref{assum:saturation} and $\|u-u_\delta\|_X \leq \|u-w\|_X$.
\end{proof}

Recall the abbreviation $\udelta=\udelta(\delta)$ for given $\delta \in \Delta$.
Recalling that $S^{\udelta \udelta} u^{\udelta \udelta}=f^{\udelta \udelta}$,
the function $u^{\udelta \delta}$ is the Galerkin approximation to
$u^{\udelta \udelta}$, i.e., it is its best approximation from $X^\delta$
w.r.t.~$\|\cdot\|_{X^{\udelta \udelta}}$.

In the next proposition it is shown that an improved Galerkin approximation to
$u^{\udelta \udelta}$ from a space $X^{\tilde{\delta}} \supset X^\delta$, i.e.,
the function $u^{\udelta \tilde{\delta}}$, is, for $c_\Delta$ sufficiently close
to $1$, also an improved approximation to $u$, and that this holds true for
$u^{\udelta(\tilde{\delta}) \tilde{\delta}}$ as well. The latter function will be
the successor of $u^{\udelta(\delta) \delta}$ in our converging sequence.

\jw{todo finish}

\subsection{A posteriori error estimation}
\begin{prop}
  \label{prop:residual-apost}
  Definitie van residual error estim
\end{prop}

\begin{prop}
  \label{prop:global-apost}
  Definitie van global. Let $\Xi^{\udelta}$ be a basis for $Y^{\udelta}$. Then,
  for any $v^\delta \in X^\delta$,
  \[
    \|u - v^\delta\|_X^2 \simeq \la (\mathbf A_s^{\udelta})^{-1}(g - B E_X^{\delta} v^\delta), (g - B E_X^{\delta} v^\delta)(\Xi^{\udelta}) \ra + \|u_0 - \gamma_0 v^\delta\|_{L_2(\Omega)}^2,
  \]
  where $\mathbf A_s^{\udelta} := (A_s E_Y^{\udelta} \Xi^{\udelta})(E_Y^{\udelta} \Xi^{\udelta})$.
\end{prop}

\section{Application in linear time}
XYZ inleiding. Here
\begin{align*}
(A_s v_1 \otimes w_1)(v_2 \otimes w_2)&=\int_0^1 v_1(t) v_2(t)\dif t \int_\Omega \nabla w_1({\bf x})\cdot  \nabla w_2({\bf x}) \dif{\bf x},\\
(B v_1 \otimes w_1)(v_2 \otimes w_2)&=\int_0^1 \frac{\dif v_1}{\dif t}(t) v_2(t)\dif t \int_\Omega w_1({\bf x}) w_2({\bf x}) \dif {\bf x}+(A_s v_1 \otimes w_1)(v_2 \otimes w_2),\\
(\gamma_0' \gamma_0 v_1 \otimes w_1)(v_2 \otimes w_2)&=
v_1(0) v_2(0)\int_\Omega w_1({\bf x}) w_2({\bf x}) \dif {\bf x}\\
(\gamma_0' u_0)(v_2 \otimes w_2)&=v_2(0) \int_\Omega u_0({\bf x}) w_2({\bf x}) \dif {\bf x}\\
g(v_2 \otimes w_2)&=\sum_i \int_0^1g_{1i}(t) v_2(t)\dif t \int_\Omega g_{2i}({\bf x}) w_2({\bf x}) \dif  {\bf x},
\end{align*}
where we assumed that $g$ is given or approximated in the form $g(t,{\bf x})=\sum_i g_{1i}(t) g_{2i}({\bf x})$.
\subsection{Efficient evaluation of operators in space and time}
TODO
\subsection{Efficient evaluation of tensor-product operators}
TODO
\subsection{Preconditioning the arising system}
\jw{introduce operator preconditioning}

\subsubsection{A preconditioner for bilinear forms on $Y \times Y$}
Let $\Psi$ be a Riesz basis for $L_2(I)$, any $y \in Y$ is of the form
$\sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$.
Using that $||w||_Y \eqsim \sum_{\lambda \in \vee_\Psi} ||w_\lambda||_V^2 < \infty$,
on $Y \times Y$, we define the bounded, symmetric, coercive bilinear form
\[
  (D_Y \sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Phi}} \psi_\mu \otimes v_\mu ) := \sum_{\lambda \in \vee_\Psi} \la w_\lambda, v_\lambda \ra_V.
\]
With $E_Y^\delta$ the trivial embedding $Y^\delta \to Y$,
we set $D_Y^\delta := {E_Y^\delta}' D_Y E_Y^\delta \in \Lis(Y^\delta, {Y^\delta}')$,
whose norm and norm of its inverse are uniformly bounded in $\delta \in \Delta$.
Equipping $Y^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Phi_\lambda^\delta$,
\jw{$\Phi_\lambda^\delta$ moet een beter teken voor}
the matrix representation of $D_Y^\delta$ reads
\[
  \mathbf D_Y^\delta = \blockdiag [\mathbf A_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_V.
\]
For given $\mathbf B_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta)^{-1}$, we have
\[
  (\mathbf D_Y^\delta)^{-1} \eqsim \blockdiag[\mathbf B_\lambda^\delta]_\lambda.
\]
In \S below, we will construct such $\mathbf B_\lambda^\delta$ in the form of a
\emph{multigrid preconditioner}, which can be applied in linear complexity. A
simpler though more computationally intensive approach would be to perform direct
solves in the spatial axis, i.e.~$\mathbf B_\lambda^\delta := (\mathbf A_\lambda^\delta)^{-1}$.

\subsubsection{A preconditioner for bilinear forms on $X \times X$}
Let $\Sigma$ be a Riesz basis for $L_2(I)$ such that
$\{2^{-|\lambda|}\sigma_\lambda : \lambda \in \vee_\Sigma\}$ is a Riesz basis for $H^1(I)$.
Using that any $w \in X$ is of the form $\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$,
on $X \times X$ we define the bilinear form
\[
  (D_X \sum_{\lambda \in \vee_{\Sigma}} \sigma_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Sigma}} \sigma_\mu \otimes v_\mu) := \sum_{\lambda \in \vee_\Sigma} \la w_\lambda, v_\lambda \ra_V + 4^{|\lambda|} \la w_\lambda, v_\lambda \ra_{V'}.
\]
\begin{lemma}
  The symmetric bilinear form $D_X$ is bounded and coercive.
\end{lemma}
\begin{proof}
  The result follows directly from the norm equivalence
  \[
    ||w||_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} ||w_\lambda||_V^2 + 4^{|\lambda|} ||w_\lambda||_{V'}^2.
  \]
  \jw{TODO prove; hier kwamen we vorige keer niet netjes uit.}
\end{proof}
Denoting with $E_X^\delta$ the trivial embedding $X^\delta \to X$, we again set
$D_X^\delta := {E_X^\delta}' D_X E_X^\delta \in \Lis(X^\delta, {X^\delta}')$, whose
norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$. Equipping
$X^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Phi_\lambda^\delta$,
the matrix representation of $D_X^\delta$ reads
\[
  \mathbf D_X^\delta = \blockdiag[\mathbf A_\lambda^\delta + 4^{|\lambda|} \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_V.
\]
\begin{theorem}
If the collection $\mathcal O := \{ \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$,
then
  \[
    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1} \quad \text{and} \quad \mathbf M_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra.
  \]
  In other words, constructing the $\mathbf C_\lambda^\delta$ leads to an optimal preconditioner on $X \times X$.
\end{theorem}
\begin{proof}

  With $\mathbf u$ the representation of $u$ in $\Phi_\lambda^\delta$, we have
  $||u||_{{Y^\delta}'} = ||(\mathbf A_\lambda^\delta)^{-1/2} \mathbf M_\lambda^\delta \mathbf u||$.
For $u \in \tilde W_\lambda^\delta$, we have $\gamma ||u||_{V'} \leq ||u||_{(\tilde W_\lambda^\delta)'} \leq ||u||_{V'}$, so that
  \[
    \gamma^2 \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'} \leq \mathbf M_\lambda^\delta (\mathbf A_\lambda^\delta)^{-1} \mathbf M_\lambda^\delta \leq \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}.
  \]
  From \cite[Thm.~4]{Pearson2012}, we find the equivalence
  \[
\tfrac12\big( {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda\big)
\leq ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda) ({\bf A}_\lambda^\delta)^{-1} ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda)
\leq {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda.
\]
Therefore, given $\mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$ uniformly in $\delta$ and $|\lambda|$, the result holds.
\end{proof}
In \S below we will construct such $\mathbf C_\lambda^\delta$ as a multigrid preconditioner. An alternative is to take $\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$ directly.

\section{Numerical results}
We study the numerical solutions to the heat equation, posed on a two-dimensional
polygonal spatial domain. Recall from \S\ref{sec:spaces} the three-point (continuous piecewise linear) 
wavelet basis $\Sigma$, the orthonormal (discontinuous piecewise linear)
wavelet basis $\Psi$, and the hierarchical (continuous piecewise linear) basis $\Upsilon$.
We equip the primal trial space $X^\delta$ with
$\Sigma \otimes \Upsilon$ and the auxiliary trial space $Y^\delta$ with $\Psi \otimes \Upsilon$.

The span of the members of $\Sigma$ up to level $L$ coincides with that of the
piecewise linears on a dyadic refinement of $I$ into $2^L$ subintervals, and the
span of the restriction of $\Upsilon$ to level $2L$ coincides with that of the
piecewise linears on a uniform refinement of $\T_\perp$ into $2^{2L}$ subelements.
Therefore, taking $X^\delta$ as the \emph{full grid} tensor product of $\Psi$ up to
level $L$ and $\Upsilon$ until level $2L$, we are able to make a direct comparison
with a typical mesh of $I \times \Omega$ into simplicial elements. \jw{meh slecht geformuleerd}
Further restriction of the full grid to those functions of level $\ell$ in time and
$k$ in space such that $2 \ell + k \leq 2L$ produces a \emph{sparse grid}.

\subsection{Condition numbers of preconditioner}
\subsection{Smooth solution}
We consider the unit-square domain $\Omega := [0,1]^2$ and prescribe the solution
\[
  u(t,x,y) := (1 + t^2) x (1-x) y (1-y)
\]
with derived data $u_0$ and $g$. For this smooth solution, selecting for $X^\delta$
the full grid of level $L$ results in $N \simeq 2^{3L}$ degrees of freedom, with
known error convergence rate of $N^{-1/3}$.\jw{citatie}
\begin{itemize}
  \item Adaptive vs full vs sparse grid \jw{ik kreeg de non-adapt sparse grid niet aan de praat}
  \item True error [als mogelijk] vs estimated error vs residual error?
  \item Robustness wrt hyperparameters $\theta$, PCG rtol, V-cycles?
\end{itemize}

Figure~\ref{fig:smooth} shows the error progressions for the smooth problem. We
plot the residual error estimator $\|\mathbf r^{\delta}\|$ from Proposition~\ref{prop:residual-apost},
the global error estimator from Proposition~\ref{prop:global-apost} with $v^\delta=u^{\udelta \delta}$,
and the time-slice error $\|u_0 - \gamma_0 u^{\udelta \delta}\|_{L_2(\Omega)}$ at $t=0$.
We see that the $X$-norm error progression for full grid refinement is indeed $\mathcal O(N^{-1/3})$,
whereas adaptive refinement yields the optimal\jw{cite} rate of $\mathcal O(N^{-1/2})$.
We moreover note the superconvergence at time $t=0$ for both refinement schemes.

Although the results are measured in a different norm, these results markedly improve
on the recent results by F\"uhrer and Karkulik, \cite[Ex.~5.2.1]{Fuhrer2019}, where an
adaptive rate of $\tfrac{1}{2}$ was observed for the time-slice error and $\tfrac{1}{3}$
for the primary norm\jw{kunnen we *iets* van een vergelijking maken met onze oplossing? We zouden natuurlijk $||u - u^{\udelta \delta}||_Y$ kunnen uitrekenen maar dat is nog niet zo 1-2-3 gedaan, ik denk via
\[
  ||u - u^{\udelta \delta}||_Y^2 = ||u||_Y - 2\la u, u^{\udelta \delta} \ra_Y + \mathbf u^\top \mathbf A_s^\delta \mathbf u
\]
moet het wel lukken, maar is niet super easy-peasy om die inproducten uit te rekenen. Of evt
\[
  \eta_{\text{FK19}} := \|\partial_t u^{\udelta \delta} - \div \nabla u^{\udelta \delta} - g\|_{I \times \Omega}^2 + \|u_0 - \gamma_0 u^{\udelta \delta}\|_{L_2(\Omega)}
\]}
\begin{figure}
  \includegraphics[width=0.5\linewidth]{smooth_adaptive_errors}
  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
  \label{fig:smooth}
\end{figure}

\subsection{First singular solution}
Selecting the L-shaped domain $\Omega := [-1,1]^2 \setminus [0,1]^2$ with data
$u_0 \equiv 0$ and $g(x,y) := \bbone_{{x^2 + y^2 \leq 1/4}}(x,y)$, the true
solution is unknown in closed form but known to be singular in the re-entrant
corner.\jw{hier wil ik een citatie of *iets* anders} Figure~\ref{fig:cylinder}
shows the error progression

\begin{itemize}
  \item Adaptive vs full vs sparse grid
  \item Misschien de spatial meshes voor wat verschillende $\lambda$, of sowieso
    tellen hoeveel $\lambda$'s er zijn in dit geval? lijkt veel op de numres van FK19
\end{itemize}

\subsection{Second singular solution}
Selecting again the unit-square domain $\Omega := [0,1]^2$ with data $u_0 := \bbone$
and $g \equiv 0$, we see BLA.

\begin{itemize}
  \item Adaptive vs full vs sparse grid
  \item Memory en time graphs
\end{itemize}

\bibliography{library} 
\bibliographystyle{siam}

\end{document}

### 2020-05-15 Latex/pint2020-westerdiep.tex
first pass latex

--- 2020-05-15 Latex/cmam2020-westerdiep.tex
@@
 introduced in~\cite{And13}, we use methods previously developed in~\cite{SW20}
-to construct an adaptive loop that produces space-time approximations as linear
-combinations of tensor-products of wavelets in time and finite elements in space.
+to construct an adaptive refinement loop that produces nested space-time
+approximations as linear combinations of tensor-products of wavelets in time and
+finite elements in space.
 
-Using an efficient and reliable `hierarchical basis' error estimator, we apply bulk
-chasing to show convergence of the iterands. Moreover, we provide an algorithm for
-linear-complexity application of the system matrix and an optimal preconditioner.
-Lastly, we include an extensive numerical study to show that this method is competitive
-in terms of speed and moreover exhibits optimal convergence rate in the number of
-degrees of freedom.
+The sequence of approximations produced by this loop exhibits optimal convergence
+rate in the number of degrees of freedom, at an optimal computational cost. The
+result is that for certain problems, large-scale parallel space-time solvers may
+be unnecessary and can be replaced by a method that adapts itself to the problem.
 

### 2020-05-15 C++/applications/uniform.cpp

@@ -1,0 +1,155 @@
+#include <stdlib.h>
+
+#include <boost/program_options.hpp>
+#include <chrono>
+#include <limits>
+
+#include "../space/initial_triangulation.hpp"
+#include "../time/basis.hpp"
+#include "../tools/util.hpp"
+#include "adaptive_heat_equation.hpp"
+#include "problems.hpp"
+
+using applications::AdaptiveHeatEquation;
+using datastructures::DoubleTreeView;
+using space::HierarchicalBasisFn;
+using Time::OrthonormalWaveletFn;
+using Time::ThreePointWaveletFn;
+
+using namespace applications;
+namespace po = boost::program_options;
+
+space::InitialTriangulation InitialTriangulation(std::string domain,
+                                                 size_t initial_refines) {
+  if (domain == "square" || domain == "unit-square")
+    return space::InitialTriangulation::UnitSquare(initial_refines);
+  else if (domain == "lshape" || domain == "l-shape")
+    return space::InitialTriangulation::LShape(initial_refines);
+  else {
+    std::cout << "domain not recognized :-(" << std::endl;
+    exit(1);
+  }
+}
+
+int main(int argc, char* argv[]) {
+  std::string problem, domain;
+  size_t initial_refines = 0;
+  size_t max_level = 0;
+  size_t max_dofs = 0;
+  bool estimate_global_error = true;
+  bool sparse_refine = true;
+  boost::program_options::options_description problem_optdesc(
+      "Problem options");
+  problem_optdesc.add_options()(
+      "problem", po::value<std::string>(&problem)->default_value("singular"))(
+      "domain", po::value<std::string>(&domain)->default_value("square"))(
+      "initial_refines", po::value<size_t>(&initial_refines))(
+      "max_level",
+      po::value<size_t>(&max_level)
+          ->default_value(std::numeric_limits<std::size_t>::max()))(
+      "max_dofs", po::value<size_t>(&max_dofs)->default_value(
+                      std::numeric_limits<std::size_t>::max()))(
+      "estimate_global_error", po::value<bool>(&estimate_global_error))(
+      "sparse_refine", po::value<bool>(&sparse_refine));
+  boost::program_options::options_description cmdline_options;
+  cmdline_options.add(problem_optdesc);
+
+  po::variables_map vm;
+  po::store(po::command_line_parser(argc, argv).options(cmdline_options).run(),
[…]

### 2020-05-15 Latex/paper.tex
paper

@@
 the global error estimator from Proposition~\ref{prop:global-apost} with $v^\delta=u^{\udelta \delta}$,
-and the time-slice error $\|u_0 - \gamma_0 u^{\udelta \delta}\|_{L_2(\Omega)}$ at $t=0$.
+and the time-slice error $\|u_0 - \gamma_0 u^{\udelta \delta}\|_{L_2(\Omega)}$ at $t=0$.\jw{misschien ook de $X$-norm meten?}
 We see that the $X$-norm error progression for full grid refinement is indeed $\mathcal O(N^{-1/3})$,
@@
 \end{itemize}
+\begin{figure}
+  \includegraphics[width=0.5\linewidth]{cylinder_adaptive_errors}
+  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
+  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
+  \label{fig:cylinder}
+\end{figure}
 
@@
 \end{itemize}
+\begin{figure}
+  \includegraphics[width=0.5\linewidth]{singular_adaptive_errors}
+  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
+  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
+  \label{fig:singular}
+\end{figure}
 

### 2020-05-18 Latex/paper.tex
testy testy

@@
   \includegraphics[width=0.5\linewidth]{smooth_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
+  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the smooth problem as
   a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
@@
 solution is unknown in closed form but known to be singular in the re-entrant
-corner.\jw{hier wil ik een citatie of *iets* anders} Figure~\ref{fig:cylinder}
+corner.\jw{is het wel singulier? of alleen minder regulier? hier wil ik een citatie of *iets* anders} Figure~\ref{fig:cylinder}
 shows the error progression
@@
   \includegraphics[width=0.5\linewidth]{cylinder_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
+  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the cylinder problem as
   a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
@@
   \includegraphics[width=0.5\linewidth]{singular_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ as
+  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the singular problem as
   a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}

### 2020-05-18 C++/applications/uniform.cpp

@@ -81,9 +81,9 @@ int main(int argc, char* argv[]) {
       std::cout << "problem not recognized :-(" << std::endl;
       return 1;
     }
-    B.ortho_tree.UniformRefine(level);
-    B.three_point_tree.UniformRefine(level);
-    T.hierarch_basis_tree.UniformRefine(2 * level);
+    B.ortho_tree.UniformRefine(level + 2);
+    B.three_point_tree.UniformRefine(level + 2);
+    T.hierarch_basis_tree.UniformRefine(2 * (level + 2));
 
     auto vec_Xd = std::make_shared<
         DoubleTreeVector<ThreePointWaveletFn, HierarchicalBasisFn>>(

### 2020-05-18 Latex/paper.tex
current state

@@
 solution is unknown in closed form but known to be singular in the re-entrant
-corner.\jw{is het wel singulier? of alleen minder regulier? hier wil ik een citatie of *iets* anders} Figure~\ref{fig:cylinder}
-shows the error progression
-
+corner.\jw{is het wel singulier? of alleen minder regulier? hier wil ik een citatie of *iets* anders}
+Figure~\ref{fig:cylinder} shows the error progression for this cylinder problem.
+We see that the full grid rate of $1/4$ is improved to $1/2$ for adaptive refinement,
+and again note the superconvergent rate for the time-slice error at $t=0$.
 \begin{itemize}

### 2020-05-18 C++/applications/uniform.cpp

@@ -66,6 +66,9 @@ int main(int argc, char* argv[]) {
 
   auto T = InitialTriangulation(domain, initial_refines);
   auto B = Time::Bases();
+  auto vec_Xd = std::make_shared<
+      DoubleTreeVector<ThreePointWaveletFn, HierarchicalBasisFn>>(
+      B.three_point_tree.meta_root(), T.hierarch_basis_tree.meta_root());
 
   for (int level = 1; level < max_level; level++) {
     std::pair<std::unique_ptr<LinearFormBase<Time::OrthonormalWaveletFn>>,
@@ -81,19 +84,26 @@ int main(int argc, char* argv[]) {
       std::cout << "problem not recognized :-(" << std::endl;
       return 1;
     }
-    B.ortho_tree.UniformRefine(level + 2);
-    B.three_point_tree.UniformRefine(level + 2);
-    T.hierarch_basis_tree.UniformRefine(2 * (level + 2));
-
-    auto vec_Xd = std::make_shared<
-        DoubleTreeVector<ThreePointWaveletFn, HierarchicalBasisFn>>(
-        B.three_point_tree.meta_root(), T.hierarch_basis_tree.meta_root());
+    B.ortho_tree.UniformRefine(level + 1);
+    B.three_point_tree.UniformRefine(level + 1);
+    T.hierarch_basis_tree.UniformRefine(2 * (level + 1));
     if (sparse_refine)
-      vec_Xd->SparseRefine(level, {2, 1});
+      vec_Xd->SparseRefine(2 * level, {2, 1});
     else
       vec_Xd->UniformRefine({level, 2 * level});
     size_t ndof = vec_Xd->Bfs().size();  // A slight overestimate.
-    std::cout << ndof << std::endl;
+    int max_node_time = 0, max_node_space = 0;
+    for (auto node : vec_Xd->Bfs()) {
+      max_node_time =
+          std::max(max_node_time, std::get<0>(node->nodes())->level());
+      max_node_space =
+          std::max(max_node_space, std::get<1>(node->nodes())->level());
+    }
+    int max_space_tree_lvl = 0;
+    for (auto node : T.hierarch_basis_tree.Bfs())
+      max_space_tree_lvl = std::max(max_space_tree_lvl, node->level());
+    std::cout << ndof << " " << max_node_time << " " << max_node_space << " "
+              << max_space_tree_lvl << std::endl;
     if (ndof == 0) continue;
     if (ndof > max_dofs) break;
     AdaptiveHeatEquationOptions adapt_opts;

### 2020-05-26 C++/applications/adaptive_heat_equation.hpp
works

@@ -83,7 +83,8 @@ class AdaptiveHeatEquation {
     return Solve(Eigen::VectorXd::Zero(vec_Xd_->container().size()));
   }
 
-  std::pair<double, std::pair<double, double>> EstimateGlobalError(const Eigen::VectorXd &u_dd_d);
+  std::pair<double, std::pair<double, double>> EstimateGlobalError(
+      const Eigen::VectorXd &u_dd_d);
   std::pair<TypeXVector *, double> Estimate(const Eigen::VectorXd &u_dd_d);
   std::vector<TypeXNode *> Mark(TypeXVector *residual);
 
@@ -93,6 +94,7 @@ class AdaptiveHeatEquation {
   std::shared_ptr<TypeXVector> vec_Xd() { return vec_Xd_; }
   std::shared_ptr<TypeXVector> vec_Xdd() { return vec_Xdd_; }
   std::shared_ptr<TypeYVector> vec_Ydd() { return vec_Ydd_; }
+  HeatEquation *heat_d_dd() { return heat_d_dd_.get(); }
 
  protected:
   Eigen::VectorXd RHS(HeatEquation &heat);

### 2020-05-26 C++/applications/uniform.cpp
merge conflicts

@@ -38,10 +38,7 @@ int main(int argc, char* argv[]) {
   size_t max_dofs = 0;
   bool estimate_global_error = true;
   bool sparse_refine = true;
-<<<<<<< HEAD
   bool calculate_condition_numbers = false;
-=======
->>>>>>> cpp/master
   boost::program_options::options_description problem_optdesc(
       "Problem options");
   problem_optdesc.add_options()(
@@ -54,14 +51,10 @@ int main(int argc, char* argv[]) {
       "max_dofs", po::value<size_t>(&max_dofs)->default_value(
                       std::numeric_limits<std::size_t>::max()))(
       "estimate_global_error", po::value<bool>(&estimate_global_error))(
-<<<<<<< HEAD
       "sparse_refine", po::value<bool>(&sparse_refine))(
       "calculate_condition_numbers",
       po::value<bool>(&calculate_condition_numbers));
 
-=======
-      "sparse_refine", po::value<bool>(&sparse_refine));
->>>>>>> cpp/master
   boost::program_options::options_description cmdline_options;
   cmdline_options.add(problem_optdesc);
 

### 2020-05-26 C++/applications/uniform.cpp
fix build

@@ -106,10 +106,10 @@ int main(int argc, char* argv[]) {
     int max_space_tree_lvl = 0;
     for (auto node : T.hierarch_basis_tree.Bfs())
       max_space_tree_lvl = std::max(max_space_tree_lvl, node->level());
-    std::cout << ndof << " " << max_node_time << " " << max_node_space << " "
+    std::cout << ndof_X << " " << max_node_time << " " << max_node_space << " "
               << max_space_tree_lvl << std::endl;
-    if (ndof == 0) continue;
-    if (ndof > max_dofs) break;
+    if (ndof_X == 0) continue;
+    if (ndof_X > max_dofs) break;
     AdaptiveHeatEquationOptions adapt_opts;
     adapt_opts.use_cache = false;
     AdaptiveHeatEquation heat_eq(vec_Xd, std::move(problem_data.first),

### 2020-05-26 Latex/paper.tex
current status

@@
 
+\usepackage{showlabels}
+\usepackage[boxed]{algorithm2e}
 \usepackage{amsaddr}
@@
 \renewcommand{\div}{\Div}
-\providecommand{\spann}{\mathrm{span}}
-\providecommand{\diam}{\mathrm{diam}}
-\providecommand{\dist}{\mathrm{dist}}
+\DeclareMathOperator{\spann}{span}
+\DeclareMathOperator{\diam}{diam}
+\DeclareMathOperator{\dist}{dist}
 \newcommand{\la}{\langle}
@@
 \DeclareMathOperator*{\ran}{ran}
+\SetKwRepeat{Do}{do}{while}%
 
@@
   \label{assum:a}
-  For a.e. $t$, let $a(t;\cdot,\cdot)$ denote a bilinear form on $V \times V$ that is
+  For a.e.~$t$, let $a(t;\cdot,\cdot)$ denote a bilinear form on $V \times V$ that is
   \begin{alignat*}{3}
-    t \mapsto a(t; \eta, \zeta) & ~~ \text{is measurable} \quad && (\eta, \zeta \in V) \\
-    |a(t;\eta,\zeta)| & \lesssim  \|\eta\|_{V} \|\zeta\|_{V} &&(\eta,\zeta \in V) &&\text{(bounded)}, \\
+    t \mapsto a(t; \eta, \zeta) & ~~ \text{is measurable} \quad && (\eta, \zeta \in V), \\
+    |a(t;\eta,\zeta)| & \lesssim  \|\eta\|_{V} \|\zeta\|_{V} &&(\eta,\zeta \in V) &\quad&\text{(bounded)}, \\
      a(t;\eta,\eta)   & \gtrsim   \|\eta\|_{V}^2 &&(\eta \in {V}) &&\text{(coercive)}.
@@
 \]
-Thanks to \eqref{eqn:B-lis}, the Schur complement $S := B' A_s^{-1} B+\gamma_0'\gamma_0$ is in $\Lis(X, X')$, is self-adjoint, and coercive.
+Thanks to \eqref{eqn:B-lis}, the Schur complement operator $S := B' A_s^{-1} B+\gamma_0'\gamma_0$ is self-adjoint, coercive, and in $\Lis(X, X')$.
 
@@
 \end{equation}
-being the best approximation to $u$ from $X^\delta$ w.r.t.~$\|\cdot\|_X$.
+being the best approximation to $u$ from $X^\delta$ wrt.~$\|\cdot\|_X$.
 
@@
 \begin{equation}
-  \gamma_\Delta:=\inf_{\delta \in \Delta}\inf_{0 \neq w \in X^\delta}\sup_{0\neq y^\delta \in Y^\delta} \frac{(\partial_t w)(v)}{\|\partial_t w\|_{Y'}\|v\|_Y} \in (0,1].
+  \gamma_\Delta:=\inf_{\delta \in \Delta}\inf_{0 \neq w \in X^\delta}\sup_{0\neq v \in Y^\delta} \frac{(\partial_t w)(v)}{\|\partial_t w\|_{Y'}\|v\|_Y} \in (0,1].
   \label{eqn:infinfsup}
@@
 \[
-  \gamma_\Delta ||\partial_t w||_{Y'} \leq ||\partial_t w||_{{Y^\delta}'} \leq ||\partial_t w||_{Y'} \quad (w \in Y, \delta \in \Delta).
+  \gamma_\Delta \|\partial_t w\|_{Y'} \leq \|\partial_t w\|_{{Y^\delta}'} \leq \|\partial_t w\|_{Y'} \quad (w \in Y, \delta \in \Delta).
 \]
@@
 
-For $\delta, \hat{\delta} \in \Delta$ with $Y^{\hat{\delta}} \supseteq Y^\delta$, and 
+For $\delta, \hat{\delta} \in \Delta$ with $Y^\delta \subseteq Y^{\hat{\delta}}$, and 
 $E_Y^{\hat{\delta}}$, $E_X^\delta$ denoting the embeddings $Y^{\hat{\delta}} \rightarrow Y$,
@@
 Equipping $X^\delta$ with `energy' norm $\|w\|^2_{X^{\hat{\delta}\delta}}:=\|w\|^2_{Y}+ \|\partial_t w\|_{{Y^\delta}'}+\|\gamma_T w\|^2$,
-from \cite[Lemma~3.3]{Stevenson2020a} it follows that
+from \cite[Lemma~3.4]{Stevenson2020a} it follows that
 $\|\cdot\|_{X^{\hat{\delta}\delta}}^2=(S^{\hat{\delta}\delta}\cdot)(\cdot)$,
-similar to Lemma~\ref{lem:energy-norm}. As shown in \cite[Thm.~3.6]{Stevenson2020a}, we have
+similar to Lemma~\ref{lem:energy-norm}. As shown in \cite[Thm.~3.7]{Stevenson2020a}, we have
 \begin{equation}
@@
 
+We will instead build our spaces as linear combinations ofwavelets in time with
+finite elements in space, thus preserving some of the usual methodology for solving
+PDEs numerically.
+
 \subsection{The spatial meshes}
-The following result generalizes \cite[Lemma 6.2]{11} and \cite[Proposition 2.5]{258.4}
-by allowing $W \neq \tilde W$.
+The following result generalizes~\cite[Lemma 6.2]{Andreev2013}
+and~\cite[Proposition 2.5]{Tantardini2016} by allowing $W \neq \tilde W$.
 
@@
 \subsection{Wavelets in time, finite elements in space}
+\jw{Willen we hier $\tilde Y^\delta$ houden ookal is hij gewoon $=Y^\delta$?}
 Let ${\mathcal O}$ be a collection of pairs of closed subspaces $(W,\tilde{W})$
@@
 $|\supp \psi_\lambda| \lesssim 2^{-|\lambda|}$. Furthermore, to ensure an
-efficient stiffness matrix evaluation, we will shirtly impose a \emph{doubletree constraint}
+efficient stiffness matrix evaluation, we will shortly impose a \emph{doubletree constraint}\jw{definitie maken?}
 on $X^\delta$ and $Y^\delta$, which will have the consequence that for any
@@
 
-In particular, for $\lambda \in \vee_\Psi$, let $S(\lambda)$ be the neighborhood
-of $\supp \psi_\lambda$ with $\diam S(\lambda) \lesssim 2^{-|\lambda|}$ s.t.~$S(\lambda) \subset S(\mu)$
-whenever $\lambda$ is a child of $\mu \in \vee_\Psi$. Then, assuming the doubletree
-constraint on $X^\delta$, \eqref{eqn:Y-space-refined} is ensured under the condition
-that
+In particular, under the doubletree constraint on $X^\delta$, \eqref{eqn:Y-space-refined}
+is ensured under the condition that\jw{ik heb dit niet gecheckt, bovendien heb ik $S(\lambda)$ eruit gesloopt}
 \[
-  |\mu| = |\lambda| \wedge|\supp \psi_\mu \cap \supp S(\lambda)|>0 \implies \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda.
+  |\mu| = |\lambda| \wedge|\supp \psi_\mu \cap \supp \sigma_\lambda|>0 \implies \tilde{V}^\delta_\mu \supset  \tilde{W}^\delta_\lambda \quad (\lambda \in \vee_\Psi).
 \]
@@
   Let any $\sigma_\lambda$ be a continuous piecewise polynomial of degree $d-1$\jw{voor $d$ de dimensie van je space domain?! of gewoon *een* $d$? in het tweede geval: rename naar $p$?}
-  w.r.t.~a uniform partition of $I$ into $2^{|\lambda|}$ subintervals, and let 
+  wrt.~a uniform partition of $I$ into $2^{|\lambda|}$ subintervals, and let 
   $\Psi=\tilde{\Psi}$ be such that $\spann\{\tilde{\psi}_\lambda : |\mu| \leq |\lambda|\}$
-  includes all piecewise polynomials of degree $d-2$ w.r.t.~a uniform partition
+  includes all piecewise polynomials of degree $d-2$ wrt.~a uniform partition
   of $I$ into $2^{|\lambda|}$ subintervals so that \eqref{eqn:deriv-orthogonal}
   is satisfied. If $\spann\{\tilde{\psi}_\lambda : |\mu| \leq |\lambda|\}$ even
-  includes all piecewise polynomials of degree $d-1$ w.r.t.~a uniform partition
+  includes all piecewise polynomials of degree $d-1$ wrt.~a uniform partition
   of $I$ into $2^{|\lambda|}$ subintervals, then $Y^\delta$ as constructed above
@@
 
-\begin{example}
-  \label{ex:wavelets}
-  \jw{dit moet misschien niet een voorbeeld zijn maar een eigen subsectie}
-  Consider linear temporal wavelets and linear spatial finite element spaces.
-  For $\Sigma$, we take the continuous 3-point from Figure~\ref{fig:3pt-wavelet}
-  \jw{waarom is dit een Riesz basis? dit heeft een citatie nodig met wat meer info, bv \cite{Stevenson1998}?}.
-  For this basis, $S(\lambda) = \supp \sigma_\lambda$.
+\subsubsection{Our choice of wavelets}
+\label{sec:wavelets}
+In the temporal axis, we consider piecewise linear wavelets. For $\Sigma$, we take the continuous three-point from \cite{Stevenson1998}; cf.~Figure~\ref{fig:3pt-wavelet}.
 \begin{figure}
-    \caption{3-point wavelet.}
+  \caption{TODO: Three-point wavelet.}
   \label{fig:3pt-wavelet}
@@
 of Figure~\ref{fig:ortho-wavelet}\jw{again, needs citation; ik kon niks vinden zo 1-2-3;
-  misschien omdat het niet echt een ``wavelet'' in de typische zin is?} For this
-  basis, again $S(\lambda) = \supp \psi_\lambda)$.
+misschien omdat het niet echt een ``wavelet'' in de typische zin is?}
 \begin{figure}
@@
 triangulations generated by NVB from some coarsest triangulation $\T_\perp$,
-  and define $V_\T = V_\T$ the space of continuous piecewise linears w.r.t.~$\T$
+and define $V_\T = \tilde V_\T$ the space of continuous piecewise linears wrt.~$\T$
 that vanish on $\partial \Omega$ such that $X^\delta := \sum_{\lambda \in \vee^\delta_\Sigma} \sigma_\lambda \otimes V_{\T_{\vee^\delta_\Sigma, \lambda}}$ satisfies the doubletree constraint.
@@
 \end{itemize}
-\end{example}
 
@@
   When equipping the finite element spaces $V_{\T}$ with the hierarchical basis,
-this doubletree constraint coincides with the constraint on the index set of
-the basis for $X^\delta$; see e.g.~\cite{TODO}.
+  this doubletree constraint coincides with the multi-index constraint on the index
+  set of the basis for $X^\delta$; see e.g.~\cite{TODO}.
 \end{remark}
@@
 constraint. The choice
-\[
+\begin{equation}
   Y^\delta = \sum_{\lambda \in \vee_\Sigma^\delta} \sum_{\{\mu \in \vee_\Psi\colon |\mu|=|\lambda|,\,
   |\supp \sigma_\lambda \cap \supp \psi_\mu|>0\}} \psi_\mu \otimes V_{\T_{\vee_\Sigma^\delta,\lambda}}
-\]
-is a doubletree by virtue of $X^\delta$ being one, and for this pair $(X^\delta, Y^\delta)$, \eqref{eqn:d_tX-in-Y} and \eqref{eqn:inclusion} are valid.
+  \label{eqn:generate-Ydelta}
+\end{equation}
+is a doubletree by virtue of $X^\delta$ being one, and for this pair $(X^\delta, Y^\delta)$,
+\eqref{eqn:d_tX-in-Y} and \eqref{eqn:inclusion} are valid.
 
@@
 There exists a collection of subspaces $(F^\delta)_{\delta \in \Delta} \subset X'$, and
-a constant $\rho<1$ such that for all $\delta \in \Delta$, when $f \in F^\delta$,
-there exists a $\Delta \ni \udelta=\udelta(\delta) \succeq \delta$ with
+a constant $\zeta<1$ such that for all $\delta \in \Delta$, when $f \in F^\delta$,
+  there exists a refinement $\udelta = \udelta(\delta) \in \Delta$ of $\delta$ with
 \begin{equation}
   \label{eqn:reduction}
-  \|u-u_\udelta\|_X \leq \rho \|u-u_\delta\|_X.
+  \|u-u_\udelta\|_X \leq \zeta \|u-u_\delta\|_X.
 \end{equation}
@@
 For the time being we will operate under the restrictive assumption that whenever
-we apply \eqref{eqn:reduction} we simply assume that $f \in F^\delta$. We will
-later remove this assumption.
+we apply \eqref{eqn:reduction}, we simply assume that $f \in F^\delta$. We will
+remove this assumption, as well as give a full specification of $F^\delta$ and $\udelta$,
+in \S TODO.
 
@@
 and $f^{\hat{\delta} \delta}$ in \eqref{eqn:discr-schur}, we replace
-${E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}$ by some \emph{preconditioner}
-$A_s^{\hat{\delta}}={A_s^{\hat{\delta}}}' \in \Lis(Y^{\hat{\delta}},{Y^{\hat \delta}}')$,
-whose inverse can be applied in linear computational complexity, and which satisfies for some $\eps_\Delta>0$,
-\[
-\|{E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}-A_s^{\hat{\delta}}\|_{\cL(Y^{\hat{\delta}},{Y^{\hat{\delta}}}')}\leq \eps_\Delta, \quad(\hat{\delta} \in \Delta).
-\]
-\jw{dit zou misschien een leuke numerieke test zijn, niet?}
+$A_s^{\hat \delta} := {E_Y^{\hat{\delta}}}' A_s E_Y^{\hat{\delta}}$ by some \emph{preconditioner}
+$\tilde A_s^{\hat{\delta}}=\tilde A_s^{\hat{\delta}}{}' \in \Lis(Y^{\hat{\delta}},{Y^{\hat \delta}}')$,
+whose inverse can be applied in linear computational complexity, and which is
+\emph{uniformly spectrally equivalent} to $A_s^{\hat \delta}$ in that it satisfies
+for some $\eps_\Delta>0$,
+\jw{dit is anders dan followup maar komt overeen met \cite[Remark 3.8]{Stevenson2020a} denk/hoop ik}
+\begin{equation}
+  \label{eqn:A_s-precond}
+  \kappa_2\big(A_s^{\hat\delta} (\tilde A_s^{\hat\delta})^{-1}\big) :=
+  \|A_s^{\hat \delta}(\tilde A_s^{\hat{\delta}})^{-1}\|_{\cL(Y^{\hat{\delta}},{Y^{\hat{\delta}}}')} \|\tilde A_s^{\hat{\delta}}(A_s^{\hat \delta})^{-1}\|_{\cL(Y^{\hat{\delta}},{Y^{\hat{\delta}}}')}\leq 1 + \eps_\Delta, \quad(\hat{\delta} \in \Delta).
+\end{equation}
 Despite this modification, we keep using the old notations for $\mu^{\hat{\delta} \delta}$,
@@
 $\lim_{\gamma_\Delta \uparrow 1,\eps_\Delta \downarrow 0} C_\Delta = 1$, and for
-$\delta,\hat{\delta} \in \Delta$ with $Y^{\hat{\delta}} \supseteq Y^\delta$, again we have
-\[
+$\delta,\hat{\delta} \in \Delta$ with $Y^{\hat{\delta}} \supseteq Y^\delta$,
+using~\cite[Rem.~3.8]{Stevenson2020a}, again we have
+\begin{equation}
   \|u-u^{\hat{\delta}\delta}\|_X \leq C_\Delta \|u-u_\delta\|_X, \quad \text{and} \quad
 C_\Delta^{-1} \|\cdot\|_{X} \leq \|\cdot\|_{X^{\hat{\delta}\delta}} \leq C_\Delta \|\cdot\|_{X} \quad\text{on } X^\delta.
-\]
+  \label{eqn:precond-norm-equiv}
+\end{equation}
 
@@
 \begin{lemma}
-  Let $w \in X^\delta$, $v \in X^\udelta$ be such that for some $\beta$,
+  Let $w \in X^\delta$, $v \in X^\udelta$ be such that for some $\rho$,
   \[
-    \|u_\udelta-v\|_X \leq \beta \|u_\udelta-w\|_X.
+    \|u_\udelta-v\|_X \leq \rho \|u_\udelta-w\|_X.
   \]
@@
   \[
-    \|u-v\|_X \leq \sqrt{\rho^2+\beta^2(1-\rho^2)}\,\|u-w\|_X.
+    \|u-v\|_X \leq \sqrt{\zeta^2+\rho^2(1-\zeta^2)}\,\|u-w\|_X.
   \]
+  \label{lem:closer-solution}
 \end{lemma}
@@
     \|u-v\|_X^2& =\|u-u_\udelta\|_X^2+\|u_\udelta-v\|_X^2\\
-    &\leq \|u-u_\udelta\|_X^2+\beta^2 \|u_\udelta-w\|_X^2\\
-    &=\|u-u_\udelta\|_X^2+\beta^2 (\|u-w\|_X^2-\|u-u_\udelta\|_X^2)\\
-    &=(1-\beta^2)\|u-u_\udelta\|_X^2+\beta^2\|u-w\|_X^2\\
-    &\leq (\rho^2(1-\beta^2)+\beta^2)\|u-w\|_X^2,
+    &\leq \|u-u_\udelta\|_X^2+\rho^2 \|u_\udelta-w\|_X^2\\
+    &=\|u-u_\udelta\|_X^2+\rho^2 (\|u-w\|_X^2-\|u-u_\udelta\|_X^2)\\
+    &=(1-\rho^2)\|u-u_\udelta\|_X^2+\rho^2\|u-w\|_X^2\\
+    &\leq (\zeta^2(1-\rho^2)+\rho^2)\|u-w\|_X^2,
   \end{align*}
-  where we used Assumption~\ref{assum:saturation} and $\|u-u_\delta\|_X \leq \|u-w\|_X$.
+  where we used Assumption~\ref{assum:saturation} and $\|u-u_\delta\|_X \leq \|u-w\|_X$.\jw{deze proof heb ik niet gecontroleerd}
 \end{proof}
@@
 $u^{\udelta \udelta}$, i.e., it is its best approximation from $X^\delta$
-w.r.t.~$\|\cdot\|_{X^{\udelta \udelta}}$.
+wrt.~$\|\cdot\|_{X^{\udelta \udelta}}$.
 
@@
 
-\jw{todo finish}
+\begin{prop}
+  \label{prop:construct-improved-solution}
+  Let $\delta \preceq \tilde{\delta} \preceq \udelta=\udelta(\delta)$ be such that
+  \begin{equation}
+    \|u^{\udelta \udelta}-u^{\udelta \tilde{\delta}}\|_{X^{\udelta \udelta}} \leq \rho\|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}.
+    \label{eqn:construct-improved-solution}
+  \end{equation}
+  Then for any $\check{\delta} \succeq \tilde{\delta}$, it holds that
+  \[
+    \|u-u^{\udelta(\check{\delta}) \check{\delta}}\|_X \leq
+    \underbrace{C_\Delta \sqrt{\zeta^2+\hat{\rho}^2(1-\zeta^2)}}_{\bar{\rho}:=}\,\|u-u^{\udelta(\delta) \delta}\|_X
+  \]
+  where $\hat{\rho}:=(1+\rho C_\Delta^2)\sqrt{C_\Delta^2-1}\sqrt{\frac{\zeta^2}{1-\zeta^2}}+\rho C_\Delta^2$.
+  Notice that $\hat{\rho}$ and $\bar{\rho}$ are $<1$ when $\rho<1$ and $C_\Delta$ is sufficiently close to $1$.\jw{oef alle diacritics duizelen me hier}
+\end{prop}
+\begin{proof}
+  Using that $u-u_\udelta \perp_X X^\udelta$, it follows from \eqref{eqn:precond-norm-equiv} that 
+  $\|u-u^{\udelta \delta}\|_X^2 \leq C_\Delta^2 \|u-u_\delta\|_X^2$ is equivalent to
+  $\|u_\udelta-u^{\udelta \delta}\|_X \leq \sqrt{C_\Delta^2 -1}\, \|u-u_{\delta}\|_X$.
+  Similarly, Assumption~\ref{assum:saturation} implies $\|u-u_\delta\|_X \leq \sqrt{\frac{\zeta^2}{1-\zeta^2}} \|u_\udelta-w\|_X$ for any $w \in X^\delta$.
+  We infer that
+  \begin{align*}
+    \|u_\udelta-u^{\udelta \tilde{\delta}}\|_X &\leq \|u_\udelta-u^{\udelta \udelta}\|_X+\|u^{\udelta \udelta}-u^{\udelta \tilde{\delta}}\|_X\\
+    &\leq \|u_\udelta-u^{\udelta \udelta}\|_X+C_\Delta\|u^{\udelta \udelta}-u^{\udelta \tilde{\delta}}\|_{X^{\udelta \udelta}}\\
+    &\leq \|u_\udelta-u^{\udelta \udelta}\|_X+\rho C_\Delta^2  \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X}\\
+    &\leq (1+\rho C_\Delta^2) \|u_\udelta-u^{\udelta \udelta}\|_X+\rho C_\Delta^2  \|u_\udelta-u^{\udelta \delta}\|_{X}\\
+    &\leq \Big[{\textstyle (1+\rho C_\Delta^2)\sqrt{C_\Delta^2-1}\,\sqrt{\frac{\zeta^2}{1-\zeta^2}}+\rho C_\Delta^2} \Big]\|u_\udelta-u^{\udelta \delta}\|_{X}
+  \end{align*}
+  From Lemma~\ref{lem:closer-solution} we conclude that
+  $\|u-u^{\udelta \tilde{\delta}}\|_X \leq \sqrt{\zeta^2+\hat{\rho}^2(1-\zeta^2)} \|u-u^{\udelta \delta}\|_{X}$.
+  Thanks to \eqref{eqn:precond-norm-equiv}, it holds that
+  \[
+    \|u-u^{\udelta(\check{\delta}) \check{\delta}}\|_X
+    \leq C_\Delta \|u-u_{\check{\delta}}\|_X
+    \leq C_\Delta \|u-u_{\tilde{\delta}}\|_X
+    \leq C_\Delta \|u-u^{\udelta \tilde{\delta}}\|_X,
+  \]
+  which completes the proof.\jw{deze proof heb ik niet gecontroleerd}
+\end{proof}
 
 \subsection{A posteriori error estimation}
+To realize \eqref{eqn:construct-improved-solution}, i.e., to construct an improved
+Galerkin approximation to $u^{\udelta \udelta}$, we apply the concept of bulk
+chasing, also known as D\"{o}rfler marking, on a collection of a posteriori error
+indicators that constitute an efficient and reliable error estimator. We will apply
+an estimator of `hierarchical basis' type.
+
+\begin{assum}[$X$-stable splitting of $X^{\udelta} \ominus X^\delta$]
+  \jw{is dit een juiste naam voor deze assumption?}
+  Assume we have $\Psi^\delta=\{\psi^\delta_\lambda\colon \lambda \in I_\delta^\udelta\} \subset X^\udelta$
+  such that $X^\delta + \spann \Psi^\delta=X^\udelta$ and, for some constants $0<m\leq M$,
+  for all $z \in X^\delta$ and ${\bf c} := (c_\lambda)_{\lambda \in I_\delta^\udelta} \in \R^{\# I_\delta^\udelta}$, we have
+  \begin{equation}
+    m^2 \|z+{\bf c}^\top \Psi^\delta\|_X^2 
+    \leq \|z\|_X^2+\|{\bf c}\|^2
+    \leq M^2 \|z+{\bf c}^\top \Psi^\delta\|_X^2.
+  \label{eqn:X-stable-splitting}
+  \end{equation}
+  \label{assum:X-stable-splitting}
+\end{assum}
+\begin{remark}
+  A suitable basis is not currently known to us. However, in \S TODO, we will construct
+  such $\Psi^\delta$ with constants $m$ and $M$ that deteriorate with increasing $|\lambda|$.
+  \jw{is die constructie iets dat we misschien in een appendix willen zetten?}
+\end{remark}
+
+\begin{prop}[Error reduction upon refinement]
+  \label{prop:residual-apost}
+  With the notation from Assumption~\ref{assum:X-stable-splitting}, define
+  \framebox{${\bf r}^{\udelta \delta}:=(f^{\udelta \udelta}-S^{\udelta \udelta} u^{\udelta \delta})(\Psi^\delta)$}.
+  Let $J \subseteq I_\delta^\udelta$ be such that for some constant $\theta \in (0,1]$,
+  \[
+    \|{\bf r}^{\udelta \delta}|_J\| \geq \theta \|{\bf r}^{\udelta \delta}\|,
+  \]
+  and, for some $\tilde{\delta} \preceq \udelta$, let $X^\delta +\spann \Psi^\delta|_J \subseteq X^{\tilde{\delta}}$.
+  Then with $\rho:=\sqrt{1-\big(\frac{m}{M} C_\Delta^{-4} \theta\big)^2}$,
+  \eqref{eqn:construct-improved-solution} is valid, i.e.,
+  \[
+    \|u^{\udelta \udelta}-u^{\udelta \tilde{\delta}}\|_{X^{\udelta \udelta}} \leq \rho \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}},
+  \]
+  and so, for any $\check{\delta} \succeq \tilde{\delta}$,
+  \[
+    \|u-u^{\udelta(\check{\delta}) \check{\delta}}\|_X \leq \bar{\rho} \|u-u^{\udelta(\delta) \delta}\|_X
+  \]
+  with $\bar{\rho}$ defined in Proposition~\ref{prop:construct-improved-solution}.
+\end{prop}
+\begin{proof}
+  As a consequence of \eqref{eqn:X-stable-splitting} and \eqref{eqn:precond-norm-equiv}, we have
+  \[
+    C_\Delta^{-4} m^2 \|z+{\bf c}^\top \Psi^\delta\|_{X^{\udelta \udelta}}^2 
+    \leq
+    \|z\|_{X^{\udelta \udelta}}^2+ \|{\bf c}\|^2
+    \leq 
+    C_\Delta^{4} M^2 \|z+{\bf c}^\top \Psi^\delta\|_{X^{\udelta \udelta}}^2 \quad (z \in X^d, {\bf c} \in \R^{\# I_\delta^\udelta}).
+  \]
+  We infer that 
+  \begin{align*}
+    \|u^{\udelta \tilde{\delta}}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}&=\sup_{0 \neq (z,{\bf c}) \in X^\delta \times \R^{\# I_\delta^\udelta}}
+    \frac{(S^{\udelta \udelta}(u^{\udelta \tilde{\delta}}-u^{\udelta \delta}))(z+{\bf c}^\top \Psi^\delta)}{\|z+{\bf c}^\top \Psi^\delta\|_{X^{\udelta \udelta}}}\\
+    & \geq m C_\Delta^{-2}  \sup_{0 \neq (z,{\bf c})}
+    \frac{(S^{\udelta \udelta}(u^{\udelta \tilde{\delta}}-u^{\udelta \delta}))({\bf c}^\top \Psi^\delta)}{\sqrt{\|z\|_{X^{\udelta \udelta}}^2+\|{\bf c}\|^2}}\\
+    & =m C_\Delta^{-2} \sup_{0 \neq {\bf c}}
+    \frac{\langle {\bf c},(f^\udelta -S^\udelta u_{\delta}^\udelta)(\Psi^\delta|_J)\rangle}{\|{\bf c}\|}\\
+    &=m C_\Delta^{-2} \|{\bf r}^{\udelta \delta}|_J\|  \geq m C_\Delta^{-2} \theta \|{\bf r}^{\udelta \delta}\|\\
+    & =m C_\Delta^{-2} \theta \sup_{0 \neq (z,{\bf c})}
+    \frac{(S^\udelta(u^{\udelta \udelta}-u^{\udelta \delta}))({\bf c}^\top \Psi^\delta)}{\sqrt{\|z\|_{X^{\udelta \udelta}}^2+\|{\bf c}\|^2}}\\
+    & \geq \frac{m}{M} C_\Delta^{-4} \theta
+    \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}
+  \end{align*}
+  so that
+  \[
+    \|u^{\udelta \udelta}-u^{\udelta \tilde{\delta}}\|_{X^{\udelta \udelta}}^2=
+    \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}^2-
+    \|u^{\udelta \tilde{\delta}}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}^2
+    \leq \Big(1-\big(\frac{m}{M} C_\Delta^{-4} \theta\big)^2\Big) \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}}^2,
+  \]
+  which completes the proof of the first inequality; the final inequality follows
+  from an application of Proposition~\ref{prop:construct-improved-solution}.
+  \jw{dit bewijs ben ik nog niet nagegaan}
+\end{proof}
+
+Additionally we have that $\|{\bf r}^{\udelta \delta}\|$ provides an efficient and reliable a posteriori estimator.
 \begin{prop}
-  \label{prop:residual-apost}
-  Definitie van residual error estim
+  \label{prop:residual-efficient-reliable}
+  Under Assumption~\ref{assum:X-stable-splitting}, let $\zeta C_\Delta<1$. Then
+  for $\delta \in \Delta$ and $\udelta=\udelta(\delta)$,
+  \[
+    \tfrac{m C_\Delta^{-3}}{1+\zeta \sqrt{C_\Delta^2-1}} \|{\bf r}^{\udelta \delta}\|
+    \leq \|u-u^{\udelta \delta}\|_X
+    \leq \tfrac{M C_\Delta^3}{\sqrt{1-\zeta^2}-\zeta\sqrt{C_\Delta^2-1}} \|{\bf r}^{\udelta \delta}\|.
+  \]
 \end{prop}
+\begin{proof}
+  Assumption~\ref{assum:saturation} gives $\|u-u_\udelta\|_X \leq \zeta \|u-u^{\udelta \delta}\|_X$,
+  which by $u-u_\udelta\perp_{X} X^\udelta$ yields
+  \begin{equation}
+   \|u_\udelta-u^{\udelta \delta}\|_{X} \leq \|u-u^{\udelta \delta}\|_{X} \leq \tfrac{1}{\sqrt{1-\zeta^2}} \|u_\udelta-u^{\udelta \delta}\|_{X}.
+    \label{eqn:galerkin-error-equiv}
+  \end{equation}
+  In the proof of Proposition~\ref{prop:construct-improved-solution} we already
+  saw that \eqref{eqn:precond-norm-equiv} is equivalent to
+  $\|u_\udelta-u^{\udelta \udelta}\|_X \leq \sqrt{C_\Delta^2 -1}\, \|u-u_{\udelta}\|_X$.
+  Together with Assumption~\ref{assum:saturation}, it gives
+  \[
+    \Big|\|u_\udelta-u^{\udelta \delta}\|_X-\|u^{\udelta \udelta}-u^{\udelta \delta}\|_X \Big| \leq \zeta \sqrt{C_\Delta^2-1}\,\|u-u^{\udelta\delta}\|_X
+  \]
+  which in combination with \eqref{eqn:galerkin-error-equiv} and $\zeta C_\Delta<1$ yields
+  \[
+    \tfrac{1}{1+\zeta \sqrt{C_\Delta^2-1}} \|u^{\udelta \udelta}-u^{\udelta \delta}\|_X
+    \leq \|u-u^{\udelta \delta}\|_X
+    \leq \tfrac{1}{\sqrt{1-\zeta^2}-\zeta\sqrt{C_\Delta^2-1}} \|u^{\udelta \udelta}-u^{\udelta \delta}\|_X.
+  \]
+  Under Assumption~\ref{assum:X-stable-splitting}, together with \eqref{eqn:precond-norm-equiv},
+  we have $m C^{-2}_\Delta \|{\bf r}^{\udelta \delta}\| \leq \|u^{\udelta \udelta}-u^{\udelta \delta}\|_{X^{\udelta \udelta}} \leq M C_\Delta^2 \|{\bf r}^{\udelta \delta}\|$
+  where the latter inequalities were shown in the proof of Prop.~\ref{prop:residual-apost}
+  when reading $(J,\theta,\tilde{\delta})$ as $(I_\delta^\udelta,1,\udelta)$. This completes the proof.\jw{dit heb ik nog niet gecheckt}
+\end{proof}
+
+This error estimator will drive our adaptive loop. Even without Assumption~\ref{assum:X-stable-splitting} however, we derive the following efficient and reliable \emph{global estimator}.
+\begin{prop}
+  \label{prop:global-apost}
+  Let $\Xi^{\udelta}$ be a basis for $Y^{\udelta}$. Then, modulo some data-oscillation
+  term we expect to be of higher order, for any $v^\delta \in X^\delta$,
+  \[
+    \|u - v^\delta\|_X^2 \eqsim \la (\mathbf A_s^{\udelta})^{-1}{\bf r}^\udelta_Y, {\bf r}^\udelta_Y \ra + \|u_0 - \gamma_0 v^\delta\|_H^2,
+  \] 
+  where $\mathbf A_s^{\udelta} := (A_s E_Y^{\udelta} \Xi^{\udelta})(E_Y^{\udelta} \Xi^{\udelta})$ and ${\bf r}^\udelta_Y := (g - B E_X^{\delta} v^\delta)(\Xi^\udelta)$.
+  \jw{zouden we in staat zijn om de constantes hidden in $\eqsim$ hierboven, precies te maken? Mss de data oscillation erin?}
+\end{prop}
+\begin{proof}
+  Define $e^\delta := u - E_X^\delta v^\delta$. From~\eqref{eqn:B-lis}, we derive
+  \[
+    \|e^\delta\|_X^2 \eqsim \|B e^\delta\|_{Y' \times H}^2
+    = \|B e^\delta\|_{Y'}^2 + \|\gamma_0 e^\delta\|_H^2 = \|B e^\delta\|_{Y'}^2 + \|u_0 - \gamma_0 v^\delta\|_H^2,
+  \]
+  and using~\cite[Lemma 3.4]{Stevenson2020a} together with~\eqref{eqn:infinfsup}, we find
+  \begin{align*}
+    \|B e^\delta\|_{Y'}^2 &= \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{Y'}^2 + \|\gamma_T e^\delta\|_H^2 - \|\gamma_0 e^\delta\|_H^2 \\
+    &\eqsim \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{{Y^\udelta}'}^2 + \|\gamma_T e^\delta\|_H^2 - \|\gamma_0 e^\delta\|_H^2 = \|B e^\delta\|_{{Y^\udelta}'}^2.
+  \end{align*}
+  The result then follows from
+  \begin{align*}
+    \|B e^\delta\|_{{Y^\udelta}'}^2 &= \|g - BE_X^\delta v^\delta\|_{{Y^\udelta}'}^2
+    = \sup_{\mathclap{0 \not= w \in Y^\udelta}} \frac{[(g - Bv^\delta)(w)]^2}{\|w\|_Y^2} = ~~\sup_{\mathclap{0 \not= {\bf w} \in \R^{\# Y^\udelta}}} ~~\frac{[(g - Bv^\delta)({\bf w}^\top \Xi^\udelta)]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} \\
+    &= \sup_{0 \not= {\bf w} \in \R^{\# Y^\udelta}} \frac{[{\bf w}^\top {\bf r}^\udelta_Y]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} = \sup_{0 \not= {\bf x} \in \R^{\# Y^\udelta}} \frac{[(({\bf A}_s^\udelta)^{-1/2}{\bf x})^\top {\bf r}^\udelta_Y]^2}{{\bf x}^\top {\bf x}} = \la ({\bf A}_s^\udelta)^{-1} {\bf r}_Y^\udelta, {\bf r}_Y^\udelta \ra.
+\qedhere
+  \end{align*}
+\end{proof}
+\begin{remark}
+  \jw{deze afschatting geldt voro elke $v^\delta \in X^\delta$ en niet alleen voor de oplossing van ons systeem; ik zie twee mogelijk-leuke toepassingen hiervan, namelijk
+  \begin{itemize}
+    \item het stelt ons in staat om de $X$-norm fout van PCG iterands te plotten
+    \item het stelt ons in staat om de $X$-norm fout van de FK19-oplossing te plotten
+  \end{itemize}
+  is daar iets nuttig aan?}
+\end{remark}
+
+\subsection{Data oscillation}
+In view of the discussion following Assumption~\ref{assum:saturation}, notice
+that the results that were obtained so far, in particular Propositions~\ref{prop:residual-apost}
+and~\ref{prop:residual-efficient-reliable}, are valid under the condition that
+$f \in F^\delta$.
+
+Let us now consider the situation that the expressions $u$, $u^{\udelta(\delta) \delta}$,
+${\bf r}^{\udelta(\delta) \delta}$, and $u^{\udelta(\tilde{\delta}) \tilde{\delta}}$
+considered in these propositions refer to the solution, Galerkin approximation,
+residual vector, and `refined' Galerkin approximation with the true right-hand side
+$f \in X'$ being \emph{replaced} by an approximation $f^\delta \in F^\delta$.
+In the following we (temporarily) denote these four and similar expressions with
+an \emph{additional left superscript} $\delta$, or more generally $\tilde\delta$
+when a right-hand side $f^{\tilde\delta} \in F^{\tilde \delta}$ has been used for
+their computation.
 
 \begin{prop}
-  \label{prop:global-apost}
-  Definitie van global. Let $\Xi^{\udelta}$ be a basis for $Y^{\udelta}$. Then,
-  for any $v^\delta \in X^\delta$,
+  \label{prop:contraction}
+  Assume Assumption~\ref{assum:X-stable-splitting}. Let $\theta \in (0,1]$ be a constant. Then for $1-\gamma_\Delta$ and $\eps_\Delta$
+  sufficiently small, and $\omega>0$ a sufficiently small constant, there exists
+  a constant $\check{\rho}<1$ such that for 
+  \begin{itemize}
+    \item $J \subseteq I_\delta^\udelta$ with $\|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}|_J\| \geq \theta \|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}\|$, and
+    \item $X^\delta+\spann \Psi^\delta|_{J} \subseteq X^{\check{\delta}}$, and
+    \item $\max(\|f-f^\delta\|_{X'}, \|f-f^{\check{\delta}}\|_{X'}) \leq \omega \|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}\|$,
+  \end{itemize}
+  it holds that
   \[
-    \|u - v^\delta\|_X^2 \simeq \la (\mathbf A_s^{\udelta})^{-1}(g - B E_X^{\delta} v^\delta), (g - B E_X^{\delta} v^\delta)(\Xi^{\udelta}) \ra + \|u_0 - \gamma_0 v^\delta\|_{L_2(\Omega)}^2,
+    \|u-{}^{\check{\delta}\!}u^{\udelta(\check{\delta}) \check{\delta}}\|_X \leq \check{\rho} \|u-{}^{\delta\!}u^{\udelta(\delta)\delta}\|_X.
   \]
-  where $\mathbf A_s^{\udelta} := (A_s E_Y^{\udelta} \Xi^{\udelta})(E_Y^{\udelta} \Xi^{\udelta})$.
 \end{prop}
+\begin{proof}
+  In the newly introduced notations, the statements of Propositions~\ref{prop:residual-apost}
+  and \ref{prop:residual-efficient-reliable} read as
+  \[
+    \|{}^{\delta\!}u^\delta-{}^{\delta\!}u^{\udelta(\check{\delta}) \check{\delta}}\|_X \leq \bar{\rho} \|{}^{\delta\!}u-{}^{\delta\!}u^{\udelta(\delta)\delta}\|_X \text{ and }
+    \|{}^{\delta\!}u-{}^{\delta\!}u^{\udelta(\delta)\delta}\|_X \eqsim \|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}\|.
+  \]
+  For $\omega$ sufficiently small, the proof is easily completed by
+  \[
+    \|u-{}^{\delta\!}u\|_X \lesssim \|f-f^\delta\|_{X'} \leq \omega \|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}\| \eqsim \omega \|{}^{\delta\!}u-{}^{\delta\!}u^{\udelta(\delta)\delta}\|_X,
+  \]
+  and
+  \[
+    \|{}^{\check{\delta}\!}u^{\udelta(\check{\delta}) \check{\delta}}-{}^{\delta\!}u^{\udelta(\check{\delta}) \check{\delta}}\|_X \lesssim
+    \|f^{\check{\delta}}-f^\delta\|_{X'}
+    \leq 2 \omega \|{}^{\delta\!}{\bf r}^{\udelta(\delta) \delta}\|\eqsim \omega \|{}^{\delta\!}u-{}^{\delta\!}u^{\udelta(\delta)\delta}\|_X.\qedhere
+  \]
+\end{proof}
+
+\begin{assum}
+  \label{assum:rhs}
+  We assume to have maps of the following type available:
+  \begin{align*}
+    &\Delta \rightarrow X' \colon \delta \mapsto f^\delta \in F^\delta,\\
+    &\eta\colon \Delta \rightarrow \R \text{ such that } \|f-f^\delta\|_{X'} \leq \eta(\delta), \text{ and } \eta(\tilde \delta)\leq \eta(\delta) \text{ when } \tilde{\delta} \succeq \delta, \\
+    &\R_{>0} \rightarrow \Delta\colon \eps \mapsto \delta(\eps) \text{ such that } \eta(\delta(\eps)) \leq \eps.
+  \end{align*}
+\end{assum}
+This in particular means that for any $\eps>0$ we are able to find an $f^\delta \in F^\delta$
+with $\|f-f^\delta\|_{X'} \leq \eps$.
+
+Given a $\delta \in \Delta$, and thinking of $f^\delta$ being a quasi-best
+approximation to $f$ from $F^\delta$, the difference $f-f^\delta$ is often referred
+to as \emph{data-oscillation}.
+
+\subsection{A convergent algorithm}
+In view of Proposition~\ref{prop:contraction}, we will use the short-hand notations
+\[
+  u^\delta := {}^{\delta\!}u^{\udelta(\delta)\delta}, \quad {\bf r}^\delta := {}^{\delta\!}{\bf r}^{\udelta(\delta) \delta},
+\]
+i.e., $u^\delta$ is the solution of~\eqref{eqn:discr-schur} where ${E_Y^\udelta} A_s E_Y^\udelta$
+is replaced by a preconditioner (cf.~\eqref{eqn:A_s-precond}, the right-hand side
+$f$ should be read as $f^\delta$, and ${\bf r}^\delta$ is its subsequent residual
+vector. The simple Algorithm~\ref{alg:adapt} shows the resulting adaptive loop.
+
+\begin{algorithm}[t]
+  $\delta := \delta_{\text{init}} \in \Delta$; \\
+  \Do{$\|{\bf r}^{\delta}\| > \eps$}{
+    compute $u^{\delta}$ and ${\bf r}^{\delta}$; \\
+    \eIf{$\eta(\delta) > \omega \|{\bf r}^\delta\|$}{
+      refine $\delta$ to a smallest $\tilde \delta$ s.t.~$\eta(\tilde \delta) \leq \eta(\delta)/2$
+    }{
+      mark a smallest $J \subset I_\delta^\udelta$ with $\|{\bf r}^{\udelta \delta}|_J\| \geq \theta \|{\bf r}^{\udelta \delta}\|$; \\
+      refine $\delta$ to a smallest $\tilde \delta$ s.t.~$X^\delta + \spann \Psi^\delta|_J \subseteq X^{\tilde \delta}$; \\
+    }
+    set $\delta := \tilde \delta$.
+  }
+  \caption{A convergent algorithm. \jw{Kon geen bijschrift verzinnen}}
+  \label{alg:adapt}
+\end{algorithm}
+
+\begin{theorem}
+  Assume Assumptions~\ref{assum:saturation} and~\ref{assum:X-stable-splitting}.
+  Under the conditions of Proposition~\ref{prop:contraction},
+  there exist constants $\theta \in (0,1]$ and $\overline \rho < 1$ such that
+  in each iteration of Algorithm~\ref{alg:adapt}, $\theta\|u - u^{\udelta \delta}\|_X + \eta(\delta)$
+  decreases with a factor $\overline \rho$. The algorithm terminates, and at termination,
+  we have $\|u - u^{\udelta \delta}\|_X + \eta(\delta) \lesssim \eps$.
+\end{theorem}
+\begin{proof}
+  \jw{dit bewijs heb ik nog niet nagelopen.}
+  In the case that $\eta(\delta)>\omega \|{\bf r}^\delta\|$, then
+  \[
+    \|u-u^\delta\|_X \leq \|{}^{\delta\!}u-u^\delta\|_X+\|u-{}^{\delta\!}u\|_X \eqsim \|{\bf r}^\delta\|+\|u-{}^{\delta\!}u\|_X \lesssim \|{\bf r}^\delta\|+\eta(\delta) \lesssim \eta(\delta),
+  \]
+  so that
+  \begin{align*}
+    \|u-u^{\tilde{\delta}}\|_X &\leq \|{}^{\tilde{\delta}\!}u-u^{\tilde{\delta}}\|_X+\|u-{}^{\tilde{\delta}\!}u\|_X\\
+    &\leq \gamma_\Delta^{-1}\|{}^{\tilde{\delta}\!}u-{}^{\tilde{\delta}\!}u_{\tilde{\delta}}\|_X+\|u-{}^{\tilde{\delta}\!}u\|_X\\
+    &\leq \gamma_\Delta^{-1}\|{}^{\tilde{\delta}\!}u-u^\delta\|_X+\|u-{}^{\tilde{\delta}\!}u\|_X\\
+    &\leq
+    \gamma_\Delta^{-1}\|u-u^\delta\|_X+(1+\gamma_\Delta^{-1})\|u-{}^{\tilde{\delta}\!}u\|_X\\
+    & \lesssim \eta(\delta)+\eta(\tilde \delta) \lesssim \eta(\delta).
+  \end{align*}
+  Since $\eta(\tilde{\delta})\leq \eta(\delta)/2$, we conclude that there exists
+  constants $\vartheta>0$ and $\breve{\rho}<1$ such that
+  \begin{equation}
+    \label{eqn:contraction}
+    \theta \|u-u^{\tilde{\delta}}\|_X +\eta(\tilde{\delta}) \leq \breve{\rho}\big(\vartheta \|u-u^{\delta}\|_X +\eta(\delta)\big).
+  \end{equation}
+
+  In the case that $\eta(\delta)<\omega \|{\bf r}^\delta\|$, from
+  $\|{\bf r}^\delta\| \eqsim \|{}^{\delta\!}u-u^\delta\|_X \lesssim \|u-u^\delta\|_X+\eta(\delta)$,
+  we infer that when $\omega$ is sufficiently small, it follows that $\eta(\delta) \lesssim \|u-u^\delta\|_X$.
+  Since in this case $\eta(\tilde{\delta})\leq \eta(\delta)$ and 
+  $\|u-u^{\tilde{\delta}}\|_X \leq \check{\rho} \|u-u^\delta\|_X$ by Proposition~\ref{prop:contraction},
+  we conclude \eqref{eqn:contraction} possibly by taking $\breve{\rho}$ closer to $1$.
+
+  Since $\|{\bf r}^\delta\| \eqsim \|{}^{\delta\!}u-u^\delta\|_X \lesssim \|u-u^\delta\|_X+\eta(\delta)$,
+  we conclude that the \texttt{do while} loop terminates after finitely many iterations.
+  At termination, we have that
+  $\|u-u^\delta\|_X \lesssim \|{}^{\delta\!}u-u^\delta\|_X+\eta(\delta) \eqsim \|{\bf r}^\delta\|+\eta(\delta)<\eps$, which completes the proof.
+\end{proof}
 
 \section{Application in linear time}
-XYZ inleiding. Here
+\jw{XYZ inleiding.} Here
 \begin{align*}
@@
 $\sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$.
-Using that $||w||_Y \eqsim \sum_{\lambda \in \vee_\Psi} ||w_\lambda||_V^2 < \infty$,
+Using that $\|w\|_Y \eqsim \sum_{\lambda \in \vee_\Psi} \|w_\lambda\|_V^2 < \infty$,
 on $Y \times Y$, we define the bounded, symmetric, coercive bilinear form
@@
 whose norm and norm of its inverse are uniformly bounded in $\delta \in \Delta$.
-Equipping $Y^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Phi_\lambda^\delta$,
-\jw{$\Phi_\lambda^\delta$ moet een beter teken voor}
+Equipping $Y^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Xi_\lambda^\delta$,
 the matrix representation of $D_Y^\delta$ reads
 \[
-  \mathbf D_Y^\delta = \blockdiag [\mathbf A_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_V.
+  \mathbf D_Y^\delta = \blockdiag [\mathbf A_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Xi_\lambda^\delta, \Xi_\lambda^\delta \ra_V.
 \]
@@
 \]
-In \S below, we will construct such $\mathbf B_\lambda^\delta$ in the form of a
+In \S\ref{sec:mgprecond} below, we construct such $\mathbf B_\lambda^\delta$ in the form of a
 \emph{multigrid preconditioner}, which can be applied in linear complexity. A
-simpler though more computationally intensive approach would be to perform direct
+simpler approach, although at superlinear complexity, would be to perform direct
 solves in the spatial axis, i.e.~$\mathbf B_\lambda^\delta := (\mathbf A_\lambda^\delta)^{-1}$.
@@
   \[
-    ||w||_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} ||w_\lambda||_V^2 + 4^{|\lambda|} ||w_\lambda||_{V'}^2.
+    \|w\|_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} \|w_\lambda\|_V^2 + 4^{|\lambda|} \|w_\lambda\|_{V'}^2.
   \]
@@
 norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$. Equipping
-$X^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Phi_\lambda^\delta$,
+$X^\delta$ with a basis of the form $\cup_\lambda \sigma_\lambda \otimes \Phi_\lambda^\delta$,
 the matrix representation of $D_X^\delta$ reads
@@
 \begin{theorem}
-If the collection $\mathcal O := \{ \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
+If the collection $\mathcal O := \{W_\lambda^\delta = \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
 satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$,
@@
   With $\mathbf u$ the representation of $u$ in $\Phi_\lambda^\delta$, we have
-  $||u||_{{Y^\delta}'} = ||(\mathbf A_\lambda^\delta)^{-1/2} \mathbf M_\lambda^\delta \mathbf u||$.
-For $u \in \tilde W_\lambda^\delta$, we have $\gamma ||u||_{V'} \leq ||u||_{(\tilde W_\lambda^\delta)'} \leq ||u||_{V'}$, so that
+  $\|u\|_{{Y^\delta}'} = \|(\mathbf A_\lambda^\delta)^{-1/2} \mathbf M_\lambda^\delta \mathbf u\|$.
+For $u \in W_\lambda^\delta$, we have $\gamma \|u\|_{V'} \leq \|u\|_{(W_\lambda^\delta)'} \leq \|u\|_{V'}$, so that
   \[
@@
 \end{proof}
-In \S below we will construct such $\mathbf C_\lambda^\delta$ as a multigrid preconditioner.
-An alternative is to take $\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$ directly.
+We will momentarily construct such $\mathbf C_\lambda^\delta$ as a multigrid
+preconditioner. A (suboptimal) alternative is to take
+$\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$
+directly.
+
+\subsubsection{A multigrid preconditioner in space}
+\label{sec:mgprecond}
+\jw{ray: kan jij hier iets over schrijven?}
 
@@
 wavelet basis $\Psi$, and the hierarchical (continuous piecewise linear) basis $\Upsilon$.
-We equip the primal trial space $X^\delta$ with
-$\Sigma \otimes \Upsilon$ and the auxiliary trial space $Y^\delta$ with $\Psi \otimes \Upsilon$.
+We equip the primal trial space $X^\delta$ with $\Sigma \otimes \Upsilon$ and
+the auxiliary trial space $Y^\delta$ with $\Psi \otimes \Upsilon$. We construct
+$X^\udelta$ by means of \eqref{eqn:todo} and then $Y^\udelta$ as in~\eqref{eqn:generate-Ydelta}.
 
@@
 level $L$ and $\Upsilon$ until level $2L$, we are able to make a direct comparison
-with a typical mesh of $I \times \Omega$ into simplicial elements. \jw{meh slecht geformuleerd}
+with a typical tensor product mesh into prismatic elements. \jw{meh slecht geformuleerd}
 Further restriction of the full grid to those functions of level $\ell$ in time and
@@
 \begin{itemize}
-  \item Adaptive vs full vs sparse grid \jw{ik kreeg de non-adapt sparse grid niet aan de praat}
   \item True error [als mogelijk] vs estimated error vs residual error?
+  \item Iets over het aantal PCG-iters?
   \item Robustness wrt hyperparameters $\theta$, PCG rtol, V-cycles?
@@
 We moreover note the superconvergence at time $t=0$ for both refinement schemes.
+We lastly note that for this smooth solution, adaptive refinement does not improve
+on classical sparse grid refinement.
 
-Although the results are measured in a different norm, these results markedly improve
-on the recent results by F\"uhrer and Karkulik, \cite[Ex.~5.2.1]{Fuhrer2019}, where an
-adaptive rate of $\tfrac{1}{2}$ was observed for the time-slice error and $\tfrac{1}{3}$
-for the primary norm\jw{kunnen we *iets* van een vergelijking maken met onze oplossing?
-We zouden natuurlijk $||u - u^{\udelta \delta}||_Y$ kunnen uitrekenen maar dat is nog niet zo 1-2-3 gedaan, ik denk via
-\[
-  ||u - u^{\udelta \delta}||_Y^2 = ||u||_Y - 2\la u, u^{\udelta \delta} \ra_Y + \mathbf u^\top \mathbf A_s^\delta \mathbf u
-\]
-moet het wel lukken, maar is niet super easy-peasy om die inproducten uit te rekenen. Of evt
-\[
-  \eta_{\text{FK19}} := \|\partial_t u^{\udelta \delta} - \div \nabla u^{\udelta \delta} - g\|_{I \times \Omega}^2 + \|u_0 - \gamma_0 u^{\udelta \delta}\|_{L_2(\Omega)}
-\]}
 \begin{figure}
   \includegraphics[width=0.5\linewidth]{smooth_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the smooth problem as
-  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
+  \caption{Error progressions for the smooth problem of residual norm (solid line),
+  estimated $X$-norm error (dashed), and $t=0$ time-slice error (dotted) as a function
+  of $\dim X^\delta$ for adaptive (black), sparse grid (red), and full grid refinement (orange).}
   \label{fig:smooth}
@@
 
-\subsection{First singular solution}
+\subsection{Cylinder solution}
 Selecting the L-shaped domain $\Omega := [-1,1]^2 \setminus [0,1]^2$ with data
@@
 Figure~\ref{fig:cylinder} shows the error progression for this cylinder problem.
-We see that the full grid rate of $1/4$ is improved to $1/2$ for adaptive refinement,
-and again note the superconvergent rate for the time-slice error at $t=0$.
+We see that the full grid rate of $\mathcal O(N^{-1/4})$ is improved to $\mathcal O(N^{-1/2}$
+for adaptive refinement, recovering the rate of the smooth case.
+
 \begin{itemize}
-  \item Adaptive vs full vs sparse grid
   \item Misschien de spatial meshes voor wat verschillende $\lambda$, of sowieso
@@
   \includegraphics[width=0.5\linewidth]{cylinder_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the cylinder problem as
-  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
+  \caption{Error progressions for the cylinder problem of residual norm (solid line),
+  estimated $X$-norm error (dashed), and $t=0$ time-slice error (dotted) as a function
+  of $\dim X^\delta$ for adaptive (black), sparse grid (red), and full grid refinement (orange).}
   \label{fig:cylinder}
@@
 \subsection{Second singular solution}
-Selecting again the unit-square domain $\Omega := [0,1]^2$ with data $u_0 := \bbone$
-and $g \equiv 0$, we see BLA.
+Selecting the unit-square domain $\Omega := [0,1]^2$ with data $u_0 := \bbone$
+and $g \equiv 0$, we see the error progressions plotted in Figure~\ref{fig:singular-error}.
+For the uniform refinement, we see the abhorrent convergence rate of $\O(N^{-1/11})$.
+Luckily, the adaptive rate is much better at $\O(N^{-1/3})$. It is unknown
+if this rate is optimal.
 
-\begin{itemize}
-  \item Adaptive vs full vs sparse grid
-  \item Memory en time graphs
-\end{itemize}
+Figure~\ref{fig:singular-time-memory} shows the peak memory consumption in kilobytes
+over $\# \dim X^\delta$ and computation time of a single application of $S^{\udelta \delta}$.
+We see that the memory consumption is asymptotically linear in the number of
+unknowns, whereas for the time per apply, we observe a rate of $1.04$.\jw{dit is natuurlijk groter dan 1. Is het wel acceptabel?}
+
 \begin{figure}
@@
   a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
-  \label{fig:singular}
+  \label{fig:singular-error}
+\end{figure}
+\begin{figure}
+  \includegraphics[width=\linewidth]{singular_mem_time.pdf}
+  \caption{Time and memory progression for the singular problem for adaptive refinement.\jw{excuse the terrible graph}}
+  \label{fig:singular-time-memory}
 \end{figure}

### 2020-05-26 Latex/paper.tex
minor edits in the preconditioning section

@@
 \]
-In \S\ref{sec:mgprecond} below, we construct such $\mathbf B_\lambda^\delta$ in the form of a
-\emph{multigrid preconditioner}, which can be applied in linear complexity. A
-simpler approach, although at superlinear complexity, would be to perform direct
+Suitable $\mathbf B_\lambda^\delta$, that can also be applied in linear complexity, are provided by \emph{multi-level methods},
+e.g.~one can use a V-cycle method based on the stable subspace decomposition introduced in~\cite{Wu2017}.
+ A simpler approach, although at superlinear complexity, would be to perform direct
 solves in the spatial axis, i.e.~$\mathbf B_\lambda^\delta := (\mathbf A_\lambda^\delta)^{-1}$.
@@
 If the collection $\mathcal O := \{W_\lambda^\delta = \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
-satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$,
-then
-  \[
-    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1} \quad \text{and} \quad \mathbf M_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra.
+satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$, and one has $\mathbf C_\lambda^\delta$ such that
+\begin{equation}\label{eqn:robustmg}
+    \mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1} \quad \text{uniformly in $\delta$ and $|\lambda|\in \N_0$},
+\end{equation}
+where $M_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra$, then
+\[
+    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda.
 \]
@@
 \end{proof}
-We will momentarily construct such $\mathbf C_\lambda^\delta$ as a multigrid
-preconditioner. A (suboptimal) alternative is to take
+In~\cite{olshanskii2000convergence, kolev2016multilevel} it was shown that a V-cycle multi-grid method yields a ${\bf C}^\delta_{\lambda}$ that satisfies~\eqref{eqn:robustmg}, and that moreover  can be applied in linear complexity.
+These results were shown for uniformly refined meshes, but similarly as for $4^{-|\lambda|} = 0$, we expect them to hold for locally refined meshes as well.
+\rvv{Hier had Rob iets snuggers over te zeggen.}
+
+ A (suboptimal) alternative is to take
 $\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$
 directly.
-
-\subsubsection{A multigrid preconditioner in space}
-\label{sec:mgprecond}
-\jw{ray: kan jij hier iets over schrijven?}
 

### 2020-05-26 Latex/paper.tex
try try

--- 2020-05-26 Latex/paper.tex
@@
     \item het stelt ons in staat om de $X$-norm fout van PCG iterands te plotten
-    \item het stelt ons in staat om de $X$-norm fout van de FK19-oplossing te plotten
+    \item het stelt ons in staat om de $X$-norm fout van (1 component van) de FK19-oplossing te plotten
   \end{itemize}
@@
 \section{Application in linear time}
-\jw{XYZ inleiding.} Here
+\jw{XYZ inleiding.}
+
+Here
 \begin{align*}
@@
 where we assumed that $g$ is given or approximated in the form $g(t,{\bf x})=\sum_i g_{1i}(t) g_{2i}({\bf x})$.
+
 \subsection{Efficient evaluation of operators in space and time}
-TODO
+\jw{willen we hier de sectie uit followup copy-pasten of is een referentie naar nicos' phd of dat paper met kestler voldoende?}
 \subsection{Efficient evaluation of tensor-product operators}
-TODO
+\jw{willen we hier de sectie uit followup copy-pasten of is een referentie naar nicos' phd of dat paper met kestler voldoende?}
+
 \subsection{Preconditioning the arising system}
@@
 $k$ in space such that $2 \ell + k \leq 2L$ produces a \emph{sparse grid}.
+
+\jw{ik mis nog een paar rake zinnen om te vergelijken met FK19 hieronder. De moeilijkheid
+zit 'm in het verschil in norm + dat zelfs hun uniform-refinement solutions tov
+een simplicial mesh zijn, en wij dus een tensor mesh hebben}
 
@@
 
-\subsection{Second singular solution}
+\subsection{Singular solution}
 Selecting the unit-square domain $\Omega := [0,1]^2$ with data $u_0 := \bbone$
@@
 We see that the memory consumption is asymptotically linear in the number of
-unknowns, whereas for the time per apply, we observe a rate of $1.04$.\jw{dit is natuurlijk groter dan 1. Is het wel acceptabel?}
+unknowns, whereas for the time per apply, we observe a rate of $1.03$.\jw{dit is natuurlijk groter dan 1. Is het wel acceptabel?}
+The memory consumption is relatively high because $\dim Y^\udelta \approx 16 \times \dim X^\delta$,
+resulting in very large systems.\jw{oef, ja wel goed om te vertellen maar 16 is een hoge constante zeg}
 
@@
   \includegraphics[width=0.5\linewidth]{singular_adaptive_errors}
-  \caption{Error progression of $e^\delta := u - u^{\underline{\delta}\delta}$ for the singular problem as
-  a function of $\dim X^\delta$ for adaptive and full grid refinement.\jw{excuse the terrible graph}}
+  \caption{Error progressions for the singular problem of residual norm (solid line),
+  estimated $X$-norm error (dashed), and $t=0$ time-slice error (dotted) as a function
+  of $\dim X^\delta$ for adaptive (black), sparse grid (red), and full grid refinement (orange).}
   \label{fig:singular-error}

### 2020-05-29 Latex/paper.tex
new version

@@
 
+\usepackage{booktabs}
 \usepackage{showlabels}
@@
 \DeclareMathOperator{\vol}{vol}
+\DeclareMathOperator{\children}{children}
 \DeclareMathOperator{\gen}{gen}
@@
 \section{Introduction}
+Een paar open vragen:
+\begin{itemize}
+  \item Willen we de algoritmes van \S 5.1, \S 5.2 hierin zetten of citen?
+  \item Willen we de details van onze implementatie hier bespreken of willen we die in een (korte?) followup doen?
+  \item Wat doen we met secties \S 5.8 5.9, 5.10 uit followup, wat een bewijs geeft van $X$-stability voor een basis die we niet meer gebruiken?
+
+\end{itemize}
+\subsection{Notation}
+TODO
 \section{Problem statement and solution}
@@
 \begin{proof}
-  We find \jw{dit ben ik nog niet nagegaan}
+  We find
   \[
@@
 Let us first argue why uniform stability for general nonuniform prismatic partitions
-is out of reach. \jw{follows from $\dim Y^\delta \lesssim \dim X^\delta$ en
+is out of reach for our current formulation. \jw{follows from $\dim Y^\delta \lesssim \dim X^\delta$ en
 de non-locality van $A_s^{-1}$: gegeven een $X^\delta$ die naar een spacetime-hoekje
-gerefined is, moet $Y^\delta$ de gehele time-slab refinen. Ik ben het precieze argument
-even vergeten maar met een plaatje wordt het snel duidelijk.}
+gerefined is, moet $Y^\delta$ de gehele time-slab refinen om de inf-sup constante
+goed te krijgen. Ik ben het precieze argument
+even vergeten, dat staat in een schrift dat op KdV ligt, maar misschien weet Rob dit nog?}
 
@@
 \end{prop}
-\begin{proof} \jw{Dit ben ik nog niet nagegaan.} If $Q\in \cL(V,V)$, then for $\tilde w \in \tilde W$,
+\begin{proof} If $Q\in \cL(V,V)$, then for $\tilde w \in \tilde W$,
 $$
 \sup_{0 \neq w \in W}\frac{\la \tilde w,w\ra}{\|w\|_V}=\sup_{0 \neq v \in V}\frac{\la \tilde w, v\ra}{\|Qv\|_V}
-\geq \|Q\|_{\cL(V,V)}^{-1} \|\tilde w\|_{V'},
+  \geq \sup_{0 \neq v \in V} \frac{\la \tilde w, v \ra}{\|Q\|_{\cL(V, V)} \|v\|_V} = \|Q\|_{\cL(V,V)}^{-1} \|\tilde w\|_{V'},
 $$
@@
 and define $V_\T = \tilde V_\T$ the space of continuous piecewise linears wrt.~$\T$
-that vanish on $\partial \Omega$ such that $X^\delta := \sum_{\lambda \in \vee^\delta_\Sigma} \sigma_\lambda \otimes V_{\T_{\vee^\delta_\Sigma, \lambda}}$ satisfies the doubletree constraint.
+that vanish on $\partial \Omega$ such that $X^\delta := \sum_{\lambda \in \vee^\delta_\Sigma} \sigma_\lambda \otimes V_{\T_{\delta, \lambda}}$ satisfies the doubletree constraint.
 This means that when $\mu \in \vee^\delta_\Sigma$ is a parent of $\lambda \in \vee^\delta_\Sigma$,
-    then $\T_{\vee^\delta_\Sigma, \mu}$ is a refinement of $\T_{\vee^\delta_\Sigma, \lambda}$, and that $\vee^\delta_\Sigma$ is a subtree of $\vee_\Sigma$, meaning
+    then $\T_{\delta, \mu}$ is a refinement of $\T_{\delta, \lambda}$, and that $\vee^\delta_\Sigma$ is a subtree of $\vee_\Sigma$, meaning
 \begin{itemize}
@@
 We consider spaces $Y^\delta = \tilde Y^\delta$ of the form
-$Y^\delta := \sum_{\lambda \in \vee^\delta_\Psi} \psi_\lambda \otimes V_{\T_{\vee^\delta_\Psi, \lambda}}$
+$Y^\delta := \sum_{\lambda \in \vee^\delta_\Psi} \psi_\lambda \otimes V_{\T_{\delta, \lambda}}$
 for $\T_{\vee^\delta_\Psi, \lambda} \in \mathbb T$, again satisfying the doubletree
@@
   Y^\delta = \sum_{\lambda \in \vee_\Sigma^\delta} \sum_{\{\mu \in \vee_\Psi\colon |\mu|=|\lambda|,\,
-  |\supp \sigma_\lambda \cap \supp \psi_\mu|>0\}} \psi_\mu \otimes V_{\T_{\vee_\Sigma^\delta,\lambda}}
+  |\supp \sigma_\lambda \cap \supp \psi_\mu|>0\}} \psi_\mu \otimes V_{\T_{\delta,\lambda}}
   \label{eqn:generate-Ydelta}
@@
 \end{assum}
+Given $X^\delta = \sum_{\lambda \in \vee_\Sigma^\delta} \sigma_\lambda \otimes V_{\T_{\delta, \lambda}}$, we define
+\begin{equation}
+  \label{eqn:X-udelta}
+  X^\udelta := \sum_{\lambda \in \vee_\Sigma^\delta} \left[\sum_{\mu \in \children(\lambda)} \sigma_\mu \otimes V_{\T_{\delta, \lambda}} +  \sigma_\lambda \otimes V_{\T^{++}_{\delta, \lambda}} \right]
+\end{equation}
+where $\T^{++}$ is found from $\T$ by two uniform bisections.
+This definition of $X^\udelta$ defines some nodes twice.\jw{slechte zin maar die "+" maar het dus niet-uniek; sommige nodes komen van "een pijltje omhoog" en "een pijltje naar rechts"}
 \begin{remark}
-  A suitable basis is not currently known to us. However, in \S TODO, we will construct
+  When $X^\delta$ is a sparse grid until level $2L$ \jw{de definitie hiervan wordt pas in \S 6 gegeven}, then $X^\udelta$ is a sparse grid until level $2(L+1)$.
+\end{remark}
+\begin{remark}
+  A suitable basis is not currently known to us. However, in Appendix~\ref{sec:PsiDelta}, we will construct
   such $\Psi^\delta$ with constants $m$ and $M$ that deteriorate with increasing $|\lambda|$.
@@
   \[
-    \|u - v^\delta\|_X^2 \eqsim \la (\mathbf A_s^{\udelta})^{-1}{\bf r}^\udelta_Y, {\bf r}^\udelta_Y \ra + \|u_0 - \gamma_0 v^\delta\|_H^2,
+    \|u - v^\delta\|_X^2 \eqsim \la (\mathbf A_s^{\udelta})^{-1}{\bf r}^\udelta_Y, {\bf r}^\udelta_Y \ra + \|u_0 - \gamma_0 v^\delta\|^2,
   \]
@@
     \|e^\delta\|_X^2 \eqsim \|B e^\delta\|_{Y' \times H}^2
-    = \|B e^\delta\|_{Y'}^2 + \|\gamma_0 e^\delta\|_H^2 = \|B e^\delta\|_{Y'}^2 + \|u_0 - \gamma_0 v^\delta\|_H^2,
+    = \|B e^\delta\|_{Y'}^2 + \|\gamma_0 e^\delta\|^2 = \|B e^\delta\|_{Y'}^2 + \|u_0 - \gamma_0 v^\delta\|^2,
   \]
@@
   \begin{align*}
-    \|B e^\delta\|_{Y'}^2 &= \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{Y'}^2 + \|\gamma_T e^\delta\|_H^2 - \|\gamma_0 e^\delta\|_H^2 \\
-    &\eqsim \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{{Y^\udelta}'}^2 + \|\gamma_T e^\delta\|_H^2 - \|\gamma_0 e^\delta\|_H^2 = \|B e^\delta\|_{{Y^\udelta}'}^2.
+    \|B e^\delta\|_{Y'}^2 &= \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{Y'}^2 + \|\gamma_T e^\delta\|^2 - \|\gamma_0 e^\delta\|^2 \\
+    &\eqsim \|e^\delta\|_Y^2 + \|\partial_t e^\delta\|_{{Y^\udelta}'}^2 + \|\gamma_T e^\delta\|^2 - \|\gamma_0 e^\delta\|^2 = \|B e^\delta\|_{{Y^\udelta}'}^2.
   \end{align*}
@@
     \|B e^\delta\|_{{Y^\udelta}'}^2 &= \|g - BE_X^\delta v^\delta\|_{{Y^\udelta}'}^2
-    = \sup_{\mathclap{0 \not= w \in Y^\udelta}} \frac{[(g - Bv^\delta)(w)]^2}{\|w\|_Y^2} = ~~\sup_{\mathclap{0 \not= {\bf w} \in \R^{\# Y^\udelta}}} ~~\frac{[(g - Bv^\delta)({\bf w}^\top \Xi^\udelta)]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} \\
-    &= \sup_{0 \not= {\bf w} \in \R^{\# Y^\udelta}} \frac{[{\bf w}^\top {\bf r}^\udelta_Y]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} = \sup_{0 \not= {\bf x} \in \R^{\# Y^\udelta}} \frac{[(({\bf A}_s^\udelta)^{-1/2}{\bf x})^\top {\bf r}^\udelta_Y]^2}{{\bf x}^\top {\bf x}} = \la ({\bf A}_s^\udelta)^{-1} {\bf r}_Y^\udelta, {\bf r}_Y^\udelta \ra.
+    = \sup_{\mathclap{0 \not= w \in Y^\udelta}} \frac{[(g - Bv^\delta)(w)]^2}{\|w\|_Y^2} = ~~\sup_{\mathclap{0 \not= {\bf w} \in \R^{\dim Y^\udelta}}} ~~\frac{[(g - Bv^\delta)({\bf w}^\top \Xi^\udelta)]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} \\
+    &= \sup_{0 \not= {\bf w} \in \R^{\dim Y^\udelta}} \frac{[{\bf w}^\top {\bf r}^\udelta_Y]^2}{{\bf w}^\top {\bf A}_s^\udelta {\bf w}} = \sup_{0 \not= {\bf x} \in \R^{\dim Y^\udelta}} \frac{[(({\bf A}_s^\udelta)^{-1/2}{\bf x})^\top {\bf r}^\udelta_Y]^2}{{\bf x}^\top {\bf x}} = \la ({\bf A}_s^\udelta)^{-1} {\bf r}_Y^\udelta, {\bf r}_Y^\udelta \ra.
 \qedhere
@@
 to as \emph{data-oscillation}.
+\jw{hier iets over de specificatie van $F^\delta$ en $f^\delta$ zoals \S 5.8 uit followup}
 
@@
 \subsection{Preconditioning the arising system}
-\jw{introduce operator preconditioning}
+\jw{introduce operator preconditioning}. \jw{onze norm-inducing bilinear forms lijken wat op die van [arXiv:1212.6037]. Maar onze $D_X$-operator is mooier, door dat inzicht van [PW12]}
 
 \subsubsection{A preconditioner for bilinear forms on $Y \times Y$}
-Let $\Psi$ be a Riesz basis for $L_2(I)$, any $y \in Y$ is of the form
+Let $\Psi$ be a Riesz basis for $L_2(I)$,\jw{moet deze basis niet ook orthogonaal zijn zodat ${\bf D}_Y^\delta$ inderdaad $\Id$ is in time?} any $y \in Y$ is of the form
 $\sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$.
@@
 For given $\mathbf B_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta)^{-1}$, we have
-\[
-  (\mathbf D_Y^\delta)^{-1} \eqsim \blockdiag[\mathbf B_\lambda^\delta]_\lambda.
-\]
+\begin{equation}
+  (\mathbf D_Y^\delta)^{-1} \eqsim (\tilde {\bf D}_Y^{\delta})^{-1} := \blockdiag[\mathbf B_\lambda^\delta]_\lambda.
+  \label{eqn:DY-precond}
+\end{equation}
 Suitable $\mathbf B_\lambda^\delta$, that can also be applied in linear complexity, are provided by \emph{multi-level methods},
@@
 \begin{theorem}
+  \label{thm:DX-precond}
   If the collection $\mathcal O := \{W_\lambda^\delta = \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
@@
   \[
-    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda.
+    (\mathbf D_X^\delta)^{-1} \eqsim (\tilde {\bf D}_X^{\delta})^{-1} := \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda.
   \]
@@
 the auxiliary trial space $Y^\delta$ with $\Psi \otimes \Upsilon$. We construct
-$X^\udelta$ by means of \eqref{eqn:todo} and then $Y^\udelta$ as in~\eqref{eqn:generate-Ydelta}.
+$X^\udelta$ by means of \eqref{eqn:X-udelta} and then $Y^\udelta$ as in~\eqref{eqn:generate-Ydelta}.
+\jw{zou het interessant zijn om niet tov $Y^\udelta$ maar tov $Y^\delta$ te solven? Dan kunnen we een stevige factor verder, naar alle waarschijnlijkheid. en stuk sneller}
 
@@
 $k$ in space such that $2 \ell + k \leq 2L$ produces a \emph{sparse grid}.
+Sparse grids are a useful tool in overcoming the \emph{curse of dimensionality},
+allowing one to reduce the number of degrees of freedom from $N$ to $\mathcal O\big(N^{\tfrac{d-1}{d}}\big)$
+at virtually no loss of approximation power. We refer to the seminal work of Bungartz
+and Griebel \cite{Bungartz2004} for more details.
 
@@
 \subsection{Condition numbers of preconditioner}
+We consider the singular problem discussed below for the calibration of our preconditioners.
+
+Replacing the nonlocal operator $({E_Y^\udelta}' A_s E_Y^\udelta)^{-1}$ in the
+forward application of $S^{\udelta \delta}$ with the block-diagonal preconditioner
+of~\eqref{eqn:DY-precond} is only permitted if it is \emph{uniformly spectrally equivalent}
+as in~\eqref{eqn:A_s-precond}. In Table~\ref{table:DY-precond}, we investigate
+this equivalence by computing the spectral condition numbers of the resulting
+matrix representations using Lanczos iteration. Here, we choose the multigrid-in-space
+operators ${\bf B}_\lambda^\udelta$ with $n$ V-cycles.\jw{slechte bewoording}
+We see that for each refinement scheme $\Delta$ and $n$, the condition numbers
+converge to some value $1 + \eps_\Delta$ for some modest $\eps_\Delta$. This defends
+taking $n=3$ in the sequel.
+
+\begin{table}
+  \begin{tabular}{rrrrrrrrrrr}\toprule
+    $\dim X^\delta$ & $n=1$ & $n=2$ & $n=3$ & $n=4$ & $n=5$  & $n=6$\\\midrule
+    \emph{adaptive} \quad
+    %22     & 1.20065  & 1.02872  & 1.00469  & 1.00078  & 1.00013 \\
+    %67     & 1.29296  & 1.05413  & 1.01177  & 1.00252  & 1.00055 \\
+    %221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 \\
+    %777    & 1.35620  & 1.07408  & 1.01836  & 1.00457  & 1.00110 \\
+    %2865   & 1.35994  & 1.07516  & 1.01841  & 1.00470  & 1.00119 \\
+    %10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 \\
+    %42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 \\
+    %167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 \\
+    %662273 & 1.36598  & 1.07707  & 1.01929  & 1.00494  & 1.00124 \\
+    \\\midrule
+    \emph{sparse grid} \quad
+    %22     & 1.20065  & 1.02872  & 1.00469  & 1.00078  & 1.00013 & 1.00002 \\
+    %67     & 1.29296  & 1.05413  & 1.01177  & 1.00252  & 1.00055 & 1.00011 \\
+    221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 & 1.00024 \\
+    %777    & 1.35620  & 1.07408  & 1.01836  & 1.00457  & 1.00110 & 1.00026 \\
+    %2865   & 1.35994  & 1.07516  & 1.01841  & 1.00470  & 1.00119 & 1.00027 \\
+    10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 & 1.00028 \\
+    %42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 & 1.00028 \\
+    %167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 & 1.00028 \\
+    662273 & 1.36598  & 1.07707  & 1.01929  & 1.00494  & 1.00125 & 1.00028 \\
+    \midrule
+    \emph{full grid} \quad
+    %27      & 1.20065  & 1.02871  & 1.00469  & 1.00078  & 1.00013  & 1.00002 \\
+    125     & 1.29332  & 1.05427  & 1.01182  & 1.00249  & 1.00054  & 1.00011 \\
+    %729     & 1.34276  & 1.06965  & 1.01647  & 1.00409  & 1.00103  & 1.00025 \\
+    4913    & 1.35600    & 1.07395  & 1.01815  & 1.00462  & 1.00111  & 1.00027 \\
+    %35937   & 1.3598   & 1.07516  & 1.01871  & 1.00476  & 1.00116  & 1.00028 \\
+    274625  & 1.36216  & 1.07600    & 1.01902  & 1.00485  & 1.00123  & 1.00028
+    \\\bottomrule
+  \end{tabular}
+  \caption{Spectral condition numbers $\kappa_2\big((\tilde {\bf D}_Y^\udelta)^{-1} {\bf A}_s^{\udelta}\big)$ with the ${\bf B}_\lambda^\udelta$ multigrid-in-space preconditioners with $n$ V-cycles.\jw{wording; ik wacht nog op numres}}
+  \label{table:DY-precond}
+\end{table}
+
+Fixing $n=3$ for the forward application of $S^{\udelta \delta}$, we want to precondition
+the resulting matrix ${\bf S}^{\udelta \delta}$ as well. Following Theorem~\ref{thm:DX-precond},
+we build a block diagonal preconditioner
+Table~\ref{table:DX-precond} then shows the condition numbers of the preconditioned
+system. We compare a preconditioner using a direct solve in the spatial axis with
+multigrid-in-space with $m$ V-cycles. We again see fast convergence of the multigrid
+preconditioners to the direct solve, and moreover notice stabilization of the condition
+numbers regardless of the choice of preconditioner, thus defending $m=3$ in the sequel.
+
+\begin{table}
+  \begin{tabular}{rrrrrrrrrrr}\toprule
+    $\dim X^\delta$ & $m=1$ & $m=2$ & $m=3$ & $m=4$ & $m=5$ & $m=6$ & D.S.\\\midrule
+    \emph{adaptive} \quad
+    \\\midrule
+    \emph{sparse grid} \quad
+    %22       & 2.73105  & 2.74699& 2.74942& 2.74982& 2.74989&2.74990 & 2.74990\\
+    %67       & 4.61358  & 3.60135& 3.55394& 3.54991& 3.54938&3.54929 & 3.54928\\
+    221      & 6.73666  & 4.65982& 4.65159& 4.65347& 4.65404&4.65414 & 4.65416\\
+    %777      & 7.54550  & 4.89895& 4.85749& 4.86621& 4.86912&4.86971 & 4.86985\\
+    %2865     & 7.79973  & 5.21031& 5.21667& 5.22923& 5.23239&5.23302 & 5.23317\\
+    10913    & 8.43870  & 5.79534& 5.77102& 5.77038& 5.77060&5.77065 & 5.77066\\
+    %42433    & 9.37352  & 6.40080& 6.37194& 6.37121& 6.37136&6.37140 & 6.37141\\
+    %167041   & 10.0961  & 7.19988& 7.19469& 7.19370& 7.19357&7.19356 & 7.19356\\
+    662273   & 10.6490  & 7.82146& 7.81987& 7.81913& 7.81910&7.81909 & 7.81910\\
+    \midrule
+    \emph{full grid} \quad
+    %27      & 3.50861  & 3.50338  & 3.50784  & 3.50848  & 3.50859  & 3.5086   & 3.50861 \\
+    125     & 4.60722  & 6.05995  & 4.64478  & 4.60486  & 4.60643  & 4.60706  & 4.60719 \\
+    %729     & 5.46211  & 7.86126  & 5.49222  & 5.45141  & 5.45939  & 5.46158  & 5.46201 \\
+    4913    & 5.89239  & 9.19603  & 5.91405  & 5.87484  & 5.88790   & 5.89189  & 5.89269 \\
+    %35937   & 6.05852  & 9.71829  & 6.104    & 6.04536  & 6.05475  & 6.05884  & 6.05828 \\
+    274625  & 6.41674  & 9.99095  & 6.43572  & 6.40801  & 6.41185  & 6.41532  & 6.41641
+    \\\bottomrule
+  \end{tabular}
+  \caption{Spectral condition numbers $\kappa_2\big((\tilde {\bf D}_X^\delta)^{-1} {\bf S}^{\udelta \delta}\big)$ with the ${\bf C}_\lambda^\delta$ multigrid-in-space preconditioners with $m$ V-cycles, versus a direct solve (D.S.).\jw{wording; ik wacht nog op numres}}
+  \label{table:DX-precond}
+\end{table}
+
 \subsection{Smooth solution}
@@
 We see that the $X$-norm error progression for full grid refinement is indeed $\mathcal O(N^{-1/3})$,
-whereas adaptive refinement yields the optimal\jw{cite} rate of $\mathcal O(N^{-1/2})$.
+whereas sparse grid refinement yields the expected improved\jw{cite} rate $O(N^{-1/2})$.
+It is obvious that for this smooth solution, adaptive refinement yields no advantage over sparse grid refinement.
 We moreover note the superconvergence at time $t=0$ for both refinement schemes.
@@
 and $g \equiv 0$, we see the error progressions plotted in Figure~\ref{fig:singular-error}.
-For the uniform refinement, we see the abhorrent convergence rate of $\O(N^{-1/11})$.
-Luckily, the adaptive rate is much better at $\O(N^{-1/3})$. It is unknown
+For the uniform refinement, we see the abhorrent convergence rate of $\mathcal O(N^{-1/11})$.
+Luckily, the adaptive rate is much better at $\mathcal O(N^{-1/3})$. It is unknown
 if this rate is optimal.
@@
 Figure~\ref{fig:singular-time-memory} shows the peak memory consumption in kilobytes
-over $\# \dim X^\delta$ and computation time of a single application of $S^{\udelta \delta}$.
+over $\dim X^\delta$ and computation time of a single application of $S^{\udelta \delta}$.
 We see that the memory consumption is asymptotically linear in the number of
@@
 
+\appendix
+\section{Construction of basis $\Psi^\delta$}
+\label{sec:PsiDelta}
+TODO
+
 \bibliography{library}

### 2020-06-17 Latex/paper.tex
draft

@@
   \item Willen we de details van onze implementatie hier bespreken of willen we die in een (korte?) followup doen?
+    \begin{itemize}
+      \item We zouden ook de implementatie + die algo's uit \S 5.1 en \S 5.2 samen in een nieuw papertje kunnen zetten, dat klinkt wel als een logische scheidingslijn
+    \end{itemize}
   \item Wat doen we met secties \S 5.8 5.9, 5.10 uit followup, wat een bewijs geeft van $X$-stability voor een basis die we niet meer gebruiken?
@@
 \begin{equation}
-(Su_\delta)(v)=f(v) \quad (v \in X^\delta),
+  (Su_\delta)(v)=f(v) =: (B' A^{-1} g + \gamma_0' u_0)(v) \quad (v \in X^\delta),
   \label{eqn:galerkin}
@@
 \end{equation}
+  \jw{in zekere zin gebruiken we niet dat *er zo'n refinement is*, maar dat *onze* refinement hieraan voldoet, toch?}
 \end{assum}
@@
     $\dim X^\delta$ & $n=1$ & $n=2$ & $n=3$ & $n=4$ & $n=5$  & $n=6$\\\midrule
-    \emph{adaptive} \quad
-    %22     & 1.20065  & 1.02872  & 1.00469  & 1.00078  & 1.00013 \\
-    %67     & 1.29296  & 1.05413  & 1.01177  & 1.00252  & 1.00055 \\
-    %221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 \\
-    %777    & 1.35620  & 1.07408  & 1.01836  & 1.00457  & 1.00110 \\
-    %2865   & 1.35994  & 1.07516  & 1.01841  & 1.00470  & 1.00119 \\
-    %10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 \\
-    %42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 \\
-    %167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 \\
-    %662273 & 1.36598  & 1.07707  & 1.01929  & 1.00494  & 1.00124 \\
-    \\\midrule
+    %\emph{adaptive} \quad
+    %%22     & 1.20065  & 1.02872  & 1.00469  & 1.00078  & 1.00013 \\
+    %%67     & 1.29296  & 1.05413  & 1.01177  & 1.00252  & 1.00055 \\
+    %%221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 \\
+    %%777    & 1.35620  & 1.07408  & 1.01836  & 1.00457  & 1.00110 \\
+    %%2865   & 1.35994  & 1.07516  & 1.01841  & 1.00470  & 1.00119 \\
+    %%10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 \\
+    %%42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 \\
+    %%167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 \\
+    %%662273 & 1.36598  & 1.07707  & 1.01929  & 1.00494  & 1.00124 \\
+    %\\\midrule
     \emph{sparse grid} \quad
@@
     %67     & 1.29296  & 1.05413  & 1.01177  & 1.00252  & 1.00055 & 1.00011 \\
-    221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 & 1.00024 \\
+    %221    & 1.34285  & 1.06806  & 1.01631  & 1.00405  & 1.00101 & 1.00024 \\
     %777    & 1.35620  & 1.07408  & 1.01836  & 1.00457  & 1.00110 & 1.00026 \\
     %2865   & 1.35994  & 1.07516  & 1.01841  & 1.00470  & 1.00119 & 1.00027 \\
-    10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 & 1.00028 \\
-    %42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 & 1.00028 \\
-    %167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 & 1.00028 \\
+    %10913  & 1.36209  & 1.07597  & 1.01901  & 1.00485  & 1.00123 & 1.00028 \\
+    42433  & 1.36533  & 1.07672  & 1.01920  & 1.00491  & 1.00124 & 1.00028 \\
+    167041 & 1.36589  & 1.07716  & 1.01924  & 1.00492  & 1.00124 & 1.00028 \\
     662273 & 1.36598  & 1.07707  & 1.01929  & 1.00494  & 1.00125 & 1.00028 \\
+    2636289 & 1.36602  & 1.07716  & 1.01934  & 1.00494  & 1.00126 & 1.00029 \\
     \midrule
@@
     %27      & 1.20065  & 1.02871  & 1.00469  & 1.00078  & 1.00013  & 1.00002 \\
-    125     & 1.29332  & 1.05427  & 1.01182  & 1.00249  & 1.00054  & 1.00011 \\
+    %125     & 1.29332  & 1.05427  & 1.01182  & 1.00249  & 1.00054  & 1.00011 \\
     %729     & 1.34276  & 1.06965  & 1.01647  & 1.00409  & 1.00103  & 1.00025 \\
     4913    & 1.35600    & 1.07395  & 1.01815  & 1.00462  & 1.00111  & 1.00027 \\
-    %35937   & 1.3598   & 1.07516  & 1.01871  & 1.00476  & 1.00116  & 1.00028 \\
-    274625  & 1.36216  & 1.07600    & 1.01902  & 1.00485  & 1.00123  & 1.00028
+    35937   & 1.3598   & 1.07516  & 1.01871  & 1.00476  & 1.00116  & 1.00028 \\
+    274625  & 1.36216  & 1.07600    & 1.01902  & 1.00485  & 1.00123  & 1.00028 \\
+    2146689  & 1.36514  & 1.07673    & 1.01919  & 1.0049  & 1.00124  & 1.00028
     \\\bottomrule
@@
     $\dim X^\delta$ & $m=1$ & $m=2$ & $m=3$ & $m=4$ & $m=5$ & $m=6$ & D.S.\\\midrule
-    \emph{adaptive} \quad
-    \\\midrule
+    %\emph{adaptive} \quad
+    %\\\midrule
     \emph{sparse grid} \quad
@@
     %67       & 4.61358  & 3.60135& 3.55394& 3.54991& 3.54938&3.54929 & 3.54928\\
-    221      & 6.73666  & 4.65982& 4.65159& 4.65347& 4.65404&4.65414 & 4.65416\\
+    %221      & 6.73666  & 4.65982& 4.65159& 4.65347& 4.65404&4.65414 & 4.65416\\
     %777      & 7.54550  & 4.89895& 4.85749& 4.86621& 4.86912&4.86971 & 4.86985\\
     %2865     & 7.79973  & 5.21031& 5.21667& 5.22923& 5.23239&5.23302 & 5.23317\\
-    10913    & 8.43870  & 5.79534& 5.77102& 5.77038& 5.77060&5.77065 & 5.77066\\
-    %42433    & 9.37352  & 6.40080& 6.37194& 6.37121& 6.37136&6.37140 & 6.37141\\
-    %167041   & 10.0961  & 7.19988& 7.19469& 7.19370& 7.19357&7.19356 & 7.19356\\
+    %10913    & 8.43870  & 5.79534& 5.77102& 5.77038& 5.77060&5.77065 & 5.77066\\
+    42433    & 9.37352  & 6.40080& 6.37194& 6.37121& 6.37136&6.37140 & 6.37141\\
+    167041   & 10.0961  & 7.19988& 7.19469& 7.19370& 7.19357&7.19356 & 7.19356\\
     662273   & 10.6490  & 7.82146& 7.81987& 7.81913& 7.81910&7.81909 & 7.81910\\
+    2636289  & 11.1022  & 8.19769& 8.20189& 8.20195& 8.20281& 8.20239& 8.20236 \\
     \midrule
@@
     %27      & 3.50861  & 3.50338  & 3.50784  & 3.50848  & 3.50859  & 3.5086   & 3.50861 \\
-    125     & 4.60722  & 6.05995  & 4.64478  & 4.60486  & 4.60643  & 4.60706  & 4.60719 \\
+    %125     & 4.60722  & 6.05995  & 4.64478  & 4.60486  & 4.60643  & 4.60706  & 4.60719 \\
     %729     & 5.46211  & 7.86126  & 5.49222  & 5.45141  & 5.45939  & 5.46158  & 5.46201 \\
     4913    & 5.89239  & 9.19603  & 5.91405  & 5.87484  & 5.88790   & 5.89189  & 5.89269 \\
-    %35937   & 6.05852  & 9.71829  & 6.104    & 6.04536  & 6.05475  & 6.05884  & 6.05828 \\
-    274625  & 6.41674  & 9.99095  & 6.43572  & 6.40801  & 6.41185  & 6.41532  & 6.41641
+    35937   & 6.05852  & 9.71829  & 6.104    & 6.04536  & 6.05475  & 6.05884  & 6.05828 \\
+    274625  & 6.41674  & 9.99095  & 6.43572  & 6.40801  & 6.41185  & 6.41532  & 6.41641 \\
+    2146689  & 10.11500  & 7.07961& 7.08715& 7.08764& 7.08767& 7.08767 & 7.08767
     \\\bottomrule
@@
   estimated $X$-norm error (dashed), and $t=0$ time-slice error (dotted) as a function
-  of $\dim X^\delta$ for adaptive (black), sparse grid (red), and full grid refinement (orange).}
+  of $\dim X^\delta$ for adaptive (black), sparse grid (red), and full grid refinement (orange).\jw{dit plaatje kan opnieuw want die rates zijn eigenlijk net anders}}
   \label{fig:cylinder}
